### 2023-10-26 paper/main.qmd
Add quarto paper

---
title: "Hierarchical Bayesian regression modelling predicts biochemical rection energies and facilitates quantitative modelling of cell metabolism"
toc: true
abstract: >
  This paper presents Maud, a command-line application that implements
  Bayesian statistical inference for kinetic models of biochemical metabolic
  reaction networks. Maud takes into account quantitative information from omics
  experiments and background knowledge, as well as structural information about
  kinetic mechanisms, regulatory interactions and enzyme knockouts. Below, we
  review the existing options in this area, explain how Maud improves on the state
  of the art, describe the intended modelling workflow and illustrate its use with
  an example application.
author: 
  - name: "Teddy Groves"
    orcid: 0000-0002-7109-3270
    email: tedgro@biosustain.dtu.dk
    affiliation: DTU Biosustain
  - name: "Nicholas Luke Cowie"
    affiliation: DTU Biosustain
  - name: "Lars Keld Nielsen"
    affiliation: DTU Biosustain
bibliography: bibliography.bib
format:
  acs-pdf:
    keep-tex: true
  html:
   include-in-header: mathjax.html
  docx: default
---

# Introduction

A kinetic model of cellular metabolism aims to express what is known
about a cellular process in the form of an *in silico* representation
of the underlying network of chemical reactions. Kinetic models can be
used to improve production of target molecules, determine regulatory
networks [@christodoulou_reserve_2018] and identify potential drug targets
[@deberardinis_fundamentals_2016;@Liberti2017]. However, the use of kinetic
models in practice is hindered by their dependence on noisy and uncertain
information sources. Quantitative in vivo measurements of chemical abundances,
and in vitro measurements relating to kinetic parameters, both contain vital
information but are notoriously inaccurate (REFERENCES). Practically useful
kinetic modelling therefore requires a principled statistical approach that
encompasses multiple possible model parameterisations.

Bayesian statistical inference can combine the structural information implicit
in kinetic models with knowledge about metabolic parameters and information from
omics measurements [@saa_construction_2016; @gopalakrishnan_k-fit_2020]. However,
kinetic models pose serious computational challenges for Bayesian inference
[@gutenkunst_2007; @raue_identifiability_2010].

The scope of a kinetic model is defined by a stoichiometric matrix, $S$, in
which rows represent metabolites, columns represent reactions, and matrix
elements $s_{ij}$ represent the stoichiometric coefficient of metabolite $i$ in
reaction $j$. The change in metabolite concentrations is

\begin{equation}\label{eq-1}
\frac{dC}{dt} = S\cdot v - \mu\cdot C 
\end{equation}

In equation \eqref{eq-1}, C represents a vector of metabolite concentrations,
$v$ is a vector of reaction rates, and $\mu$ is the growth rate. The second term
represents the dilution due to cell growth.

In a kinetic model, the rates, $v$, are expressed as a function of the enzyme
concentrations, $E$, the metabolite concentrations, $C$, and a set of parameters,
$\theta$ as shown in equation \eqref{eq-2}

\begin{equation}\label{eq-2}
v = f(C, E, \theta)
\end{equation}

The parameters must include sufficient boundary concentrations and fluxes to
solve \eqref{eq-1}.

It is common to assume pseudo-steady state for metabolites, i.e., the rate
of fluxes towards any metabolite is much greater than the rate of change in
concentration, $𝑣 \gg \frac{𝑑𝐶}{𝑑𝑡}$. Moreover, the dilution effect is assumed
minimal, $\mu\cdot C \ll \vec{v}$ (true unless the concentration is very high).
Finally, the enzyme concentration is assumed constant for the period considered
and hence part of the parameters.

Given these assumptions and a set of values for \theta, a set of steady state
metabolite concentrations and fluxes can be found by solving for $C$ the
algebraic equation:

\begin{equation}\label{eq-3}
S\cdot f(C;\theta) = 0
\end{equation}

In a fermentation context, \eqref{eq-3} captures the rapid kinetics inside
the cell, while another set of ODEs would be used to describe the external
substrate and product concentrations, which could act as boundary parameters
to \eqref{eq-3}.

In the context of kinetic modelling, Bayesian inference is appealing because
it allows uncertainty to be represented appropriately without sacrificing
mechanistic accuracy. Measurement uncertainty can naturally be represented in a
Bayesian measurement model, whereas the prior model can represent quantitative
uncertainty about kinetic parameters. Finally, kinetic rate laws can be
represented in Bayesian data generation models with arbitrarily high fidelity.
See (REFERENCES) for more about Bayesian inference and (REFERENCE) for a
discussion of practical Bayesian workflow.

Another advantage is that Bayesian inference problems are well-posed even when
not all parameters are strongly identified. Sloppy models in which measurable
quantities are sensitive to combinations of parameters but not to individual
marginal parameter values are ubiquitous in models of biological systems
[gutenkunst_2007; @white_2016]. The parameter correlation structure represents
the set of potential models that describe the observed data. As we demonstrate
in our case study below, capturing this correlation structure is difficult
outside a Bayesian context.

Previous Bayesian kinetic models have either sacrificed mechanistic accuracy
or have attempted to fit realistic kinetic models using obsolete or unreliable
computational methods.

The most popular algorithm for fitting Bayesian statistical models is
Markov Chain Monte Carlo (MCMC). Modern MCMC algorithms allow exploration of
high- dimensional posterior distributions, have robust failure diagnostics
[@vehtariRankNormalizationFoldingLocalization2021] and can incorporate fast
numerical solvers, thereby making inference feasible for Bayesian kinetic
models. Nonetheless, the kinetic modelling literature reports an aversion
to MCMC, rooted mainly in concerns about sampling time and the presumed
difficulty of implementing the required statistical model [@raue_joining_2013;
@saa_construction_2016]. We are only aware of two recent attempts to implement
a Bayesian kinetic modelling approach using MCMC. @stapor_pesto_2018 fitted
detailed kinetic models using relatively inefficient MCMC algorithms that
do not scale well to high dimensional parameter spaces limiting the scope
of modelling. Conversely, @st.johnBayesianInferenceMetabolic2018 utilises an
efficient sampling algorithm but uses approximate kinetics, namely lin-log
kinetics @visser_dynamic_2003, limiting the scope of interpreting parameters and
inferring cellular behaviour in experimental conditions outside the reference
dataset.

There have also been efforts to implement Bayesian inference for kinetic models
without the use of MCMC. Examples of alternative inference methods include
variational inference as in @st.johnBayesianInferenceMetabolic2018, rejection
sampling and approximate Bayesian computation @saa_construction_2016 and Laplace
approximation, in which the Fisher information matrix is used to calculate a
normal approximation around the maximum a posteriori parameter configuration
[@liebermeister_model_2021; @gopalakrishnan_k-fit_2020; @stapor_pesto_2018;
@raue_joining_2013]. Non-MCMC-based Bayesian kinetic models have limited
utility because they lack reliable diagnostic tools for verifying that their
results approximate the target posterior distribution. This is a problem because
realistic kinetic models tend to induce highly correlated, non-Gaussian, joint
probability distributions [@gutenkunst_2007; @stapor_pesto_2018].

Our application Maud is the first Bayesian kinetic model to combine biologically
realistic mechanistic accuracy---including accurate rate laws, post-
translational modification and thermodynamics---with fast, robust MCMC sampling
using adaptive Hamiltonian Monte Carlo. Further, Maud is a general-purpose
application that can be used to fit a wide range of Bayesian kinetic models.

# Results and Discussion

To demonstrate our application’s capabilities, we used Maud to analyse an
artificial dataset based on the human methionine cycle. We first used Maud
to generate simulated training and validation measurements based on plausible
parameter values, then performed posterior sampling. Next, we used Maud to
predict the validation data.

To show that Maud is robust to missing measurements we compared the results
of fitting the full dataset with an intentionally incomplete dataset. To
demonstrate why a full Bayesian approach is preferable to an approach based on a
Laplace approximation of the posterior distribution, we also fit our methionine
cycle model using the latter method and compared the results with MCMC sampling.

Finally, we investigated our results to find out what our model learned about
the contributions of different regulatory factors to the flux through GNMT, an
important reaction. This analysis illustrates how Maud can be used to generate
actionable insights about metabolism without the need for further statistical
analysis.

The methionine cycle, illustrated in @fig-methionine-reactions, is a fundamental
pathway in human metabolism, whose intermediate metabolites participate in a
variety of mechanisms which must compete for the same resources. Due to this
competition, as well as the fact that all the functions occur simultaneously,
the methionine cycle is highly regulated, with 6 known allosteric effectors
(REFERENCE). This complex regulation means that quantitative modelling of the
methionine cycle requires a detailed kinetic model: this is why we chose it as a
case study for Maud.

![CAPTION - FILL THIS IN!](./figures/methionine-reactions.png){#fig-methionine-reactions}

## Dataset and model specification

The simulated dataset and underlying kinetic model that we used for our analysis
can be found in [#Supplementary Info].

We constructed a kinetic model of the methionine cycle in Maud’s format using
the description in @korendyaseva_allosteric_2008. The ordinary differential
equation system describing this model is shown in Equation \eqref{eq-meth-ode}.

After specifying the qualitative aspects of the kinetic model, we chose
parameter values to use as ground truth by Monte Carlo sampling using a previous
model of the methionine cycle as a starting point (see @saa_general_2015 for
this model).

We used these parameters to simulate steady states in a range of plausible
experimental conditions, again using @saa_general_2015 as a starting point.
These steady states were then used to generate simulated measurements using the
measurement model described below.

\begin{equation}
a = b \label{eq-meth-ode}
\end{equation}

### Measurement model

Generating an artificial dataset required a specification of the true
measurement error; we specified a standard deviation of 0.1 on logarithmic
scale for enzyme and metabolite concentration measurements, corresponding to
approximately 10% measurement error and a standard deviation for each reaction
measurement approximately 10% of the simulated value.

These measurement error specifications are somewhat optimistic considering the
many sources of variation and uncertainty affecting quantitative proteomics,
metabolomics and fluxomics analyses, but are a reasonable first approximation to
a realistic set of measurements.

For our main model run, we assumed that all metabolite and enzyme concentrations
were measured, and that there was a reaction measurement for each of the
network’s elementary flux modes.

### Training/validation split

The training testing split was selected to achieve a large difference between
the fluxes of the training and testing dataset. The split was determined as
we are interested in showing how our model can fit to varied conditions, and
conditions closer to the training set are likely to be predicted well without
necessarily learning the system.

### Additional dataset with missing measurements

To gain insight into our model’s robustness to missing measurements, we
also performed a model run with the same 6 experimental datasets, but with
measurements of the metabolite S-Adenosyl-L-homocysteine, or “ahcys” removed.
Since ahcys regulates three enzymes in the methionine cycle, including one
enzyme which is also thermodynamically regulated, we expected the removal of
these measurements to yield interesting results.

### Maud input specification

We constructed inputs in Maud’s format for each of the analysed datasets, based on the scenario that the true kinetic model was known except for parameter values, which needed to be inferred from the training data and priors. These inputs can be found at XXXX.

The prior distributions and corresponding true parameter values used in our case
study are shown in [Supplementary]. The first two columns show the 1% and 99%
quantiles of each marginal prior distribution. True parameter value are shown
in column three, and the last column shows the z-score on log scale of the true
parameter value according the marginal prior distribution. As seen in table
2 [REFERENCE] there are 7 parameters for which the true value is outside the
1%-99% range. This is desirable, making the case study more realistic, because
extreme deviance from the prior distribution is likely to occur in practice due
to in vivo to in vitro measurement differences.

### Computation

We conducted adaptive Hamiltonian Monte Carlo sampling for the full and missing-
data datasets, obtaining XXX post-warmup samples after XXX of total computation
time. For the Laplace approximation comparison, we used Maud’s Laplace mode.
Full details as well as instructions for reproducing our analysis can be found
at XXX.

## Findings

### Posterior inference

Running standard diagnostic checks indicated that the samples we generated
were from the target posterior distribution. The improved $\hat{R}$ statistic
[@vehtariRankNormalizationFoldingLocalization2021] for every sampled variable
was within 1% of 1, indicating appropriate mixing within and between Markov
chains. Additionally, the number of effective samples was high, indicating
that we generated enough posterior samples to support inferences about the
bulks of the distributions of the sampled parameters. Furthermore, we observed
no post-warmup divergent transitions, indicating that the sampler was able to
transform the log-posterior distribution, avoiding any regions with excessive
curvature that might inhibit exploration via HMC.

Posterior predictive checking indicated that our model achieved a good fit
to the simulated reaction and metabolite concentration measurements, as shown
by the graphs in the top row of figure 3. Note that the fit was good for both
training and validation measurements.

Analysis of the posterior distributions for kinetic parameters indicated that
these are highly correlated. The marginal posterior distributions for most
kinetic parameters did not shrink significantly compared with the corresponding
marginal prior distributions, even though these parameters’ joint posterior
distribution contained enough information to make accurate out of sample
predictions. In some cases, there were two-dimensional correlations such as the
one shown in the bottom left of figure 3; in this case the marginal distribution
of the two parameters is roughly banana shaped. More commonly, however, two-
dimensional pair plots were insufficient to reveal the underlying correlation
structure, as seen in the bottom-right plot in figure 3. This does not mean that
the parameters were uncorrelated, but rather that the correlations involve more
than two parameters.

Overall, our results show that Maud can fit a realistic pathway-sized dataset.
This was achieved without fixing the marginal values of kinetic parameters: the
information required to make good predictions was contained in the correlation
structure of the joint posterior distribution. This finding is consistent with
previous analyses of biological systems that found they are “sloppy”, that is,
sensitive to parameter combinations rather than marginal parameter values, with
important combinations, scales and regions of sensitivity being difficult to
ascertain in advance @gutenkunst_2007, @poirier_revising_1998.

The question naturally arises whether the crucial high-dimensional parameter
correlations are linear or non-linear. This is relevant to the question of
model performance, as linear correlations are easier to correct for. A linearly
correlated posterior space would also be easier to summarise. We address this
question below in section XXX.

This case study illustrates the type of kinetic model and dataset that Maud
can fit. The model we analysed has XXX reactions, XXX state variables and XXX
parameters. Posterior sampling with adaptive Hamiltonian Monte Carlo generated
XXX ESS/hour. Generalising from this result, we conclude that it is feasible
to use this method to fit models in the same order of magnitude, but not, for
example, genome-scale kinetic models. To fit larger models, faster steady state
solving methods or alternative inference algorithms will be required. Section
XXX addresses whether Laplace approximation is a suitable candidate.

### Comparison with Laplace approximation

We found that the Laplace method was not able to produce an accurate posterior
approximation for our model and dataset.

The Laplace approximation yielded XXX samples in XXX time. The diagnostics
indicated that our algorithm was able to find the maximum a posteriori parameter
configuration, approximate the Hessian  and use these quantities to generate
approximate posterior samples. The results can be found at XXX.

@fig-laplace summarises the results of comparing the Laplace approximation of our
case study’s posterior distribution with the true posterior. As can be seen
from the top left plot, the Laplace method does not provide a good approximation
to the true posterior distribution, as the marginal distribution of the
total log probability density is clearly different. This was confirmed using
the Kolmogorov-Smirnov test, which is a test to differentiate two empirical
univariate distributions. The two distributions were significantly different
with a p-value indistinguishable from 0.

The difference between the Laplace approximation output and the true posterior
distribution manifests itself not only in the parameter space, but also in
the measurement space. Figure 5 frame B compares the 5%-95% interval for
flux measurement log likelihoods in the true posterior with the Laplace
approximation; lower log likelihood values indicate that the modelled and
measured values are further away. The graph shows that the Laplace approximation
yielded significantly worse predictions than the true posterior, even for the
training data.

To further explore why this is the case we compared samples from the true
posterior and the Laplace approximation for the pairwise marginal distributions
of two Michaelis-Menten constants [LATEX CONSTANT]- and [LATEX CONSTANT]: see
the bottom right cell of figure 5. This comparison demonstrates that the Laplace
method is not able to capture the correct relationships between parameters’
distributions.

This result shows that MCMC, while slower than Laplace approximation, is
unfortunately preferable for posterior inference in this case. We expect that
this is typical of kinetic models of realistic metabolic networks in general, so
we recommend that Maud users use MCMC sampling if possible.

Our results here also provide circumstantial evidence that the parameter
correlations in Bayesian kinetic model posteriors tend to be non-linear, as
a posterior with only linear correlations would likely be more germane to
Laplace approximation.A conclusion that we drew from this analysis was that
the results of fitting our model cannot be summarised simply, for example by
fitting a multivariate normal distribution to the posterior draws. We therefore
recommend that Maud users store the full set of MCMC draws rather than using
such an approximation. This does not preclude the possibility that there is
an alternative, more compact, way to summarise the results of Bayesian kinetic
model inference; we leave research into this topic to future work.

![CAPTION: FILL THIS IN!](./figures/laplace.png){#fig-laplace}

### Effect of missing metabolite concentration measurements

Comparing model runs with and without the ahcys measurements showed that Maud
can produce sensible results even from incomplete metabolomics data.

@fig-missing shows that, as might be expected, the model with missing
measurements did not correctly infer the missing ahcys concentrations.
Nonetheless, the remaining measured metabolites were still well predicted,
suggesting that information about the network is still preserved despite the
missing measurements. Comparison of flux measurements in both models also
indicated that removing the ahcys measurement did not result in catastrophic
model failure.

The missing measurements did affect Maud’s ability to infer parameter values
correctly. As we can see in the [#lower left plot of @fig-missing], the model
with full metabolomics learned the true value for the displayed dissociation
constant, despite this value being far from the mean of the corresponding
marginal prior distribution, whereas the model with the missing measurements
stayed in the neighbourhood of the prior.

This result is reassuring because not having access to all measurements is a
common situation in multi-omics studies. For instance, measuring all metabolites
in a pathway can be infeasible because of limitations of mass spectrometers,
availability of standards, column effects, and compartmentalisation. However,
provided that sufficient information is available from other sources, our
approach can produce sensible results from incomplete metabolomics data.

![CAPTION: FILL THIS IN!](./figures/missing.png){#fig-missing}

### Application to regulatory understanding

To investigate how the flux of GNMT between conditions [X] and [Y] changes
we compared the regulatory contributions in the posterior distribution.
Presented in [@fig-decomposition top left] is the regulatory description of
GNMT, an irreversible enzyme that is homotropically activated by its substrate,
competitively inhibited by its product and heterotropically inhibited. The
diverse regulation makes it the ideal test case to elucidate regulatory changes.

Here we found that saturation and allosteric effects were the main drivers
of regulation between the two conditions. @fig-decomposition shows the distributions
of the regulatory components as the log-ratio of each component with [X] in
the numerator and [Y] in the denominator, positive values suggest that this
component was increased in experiment [X] relative to [Y], with 0 indicating no
difference. Naturally, this presents the ideal statistical test when evaluating
the regulation, with the probability of regulation given by the relative
proportion of samples above and below the 0 point.

We demonstrate that Maud can accurately reproduce the regulatory effect between
the two conditions by comparing it to the ground truth. As the true values are
well within our probability mass of the posterior we can say that Maud is able
to accurately reproduce the regulatory behaviour of metabolic networks and also
provides a tool for statistical inference of the regulatory behaviour.

Here we have demonstrated a key feature in Maud, that is, direct inference on
any parameter that arises from the data generating process. More succinctly,
we show the benefit of error propagation that Bayesian inference provides as
researchers can directly use quantiles for hypothesis testing.

Our tool allows inference directly on the joint posterior distribution rather
than the marginal parameter. The implication is that inference will be better
than that of the marginal distributions due to correlated parameter values, if
the data generating process can be directly modelled.

![CAPTION: FILL THIS IN!](./figures/decomposition.png){#fig-decomposition}

# Methods

Maud is a command line application implementing Bayesian inference for
a wide range of realistic kinetic models. Maud is written in Python
[@vanrossumPythonReferenceManual2009], designed for use on Windows, Macos
and Linux, registered on the Python Package Index as `maud-metabolic-models`,
documented at <https://maud- metabolic-models.readthedocs.io> and actively
developed and maintained at <https://github.com/biosustain/Maud/>.

To use Maud, a user must first collate appropriate input information, represent
it in files with Maud’s required formats (see section (REFERENCE)below).  Maud’s
command line interface provides commands for inference, simulation and making
out-of-sample predictions. Results are stored in files, using a structured,
interoperable format.

## Input format

Maud inputs are structured directories, somewhat inspired by the PEtab
format @SchmiesterSch2021. A Maud input directory must contain a toml
[@preston-wernertomandgedampradyunTOMLSpecification0rc2020] file called
`config.toml` which gives the input a name, configures how Maud will be run
and tells Maud where to find the other files, allowing these to have custom
names. It must also include a file containing a kinetic model definition,
a file specifying information about parameters and a file with information
experiments. The required structure of these files is documented at <https://
maud-metabolic-models.readthedocs.io/en/latest/inputting.html>. The input is
validated against a Pydantic [@pydanticdevelopersPydantic2022] data model which can be
inspected at <https://github.com/biosustain/Maud/tree/main/maud/data_model>.

We chose to implement a custom input format despite the existence of standard
formats in similar areas, including SBML [@keatingSBMLLevelExtensible2020]
and PEtab [@SchmiesterSch2021]. This choice was partly motivated by the need
to ensure flexibility as Maud was developed, but there are also features of
SBML and PEtab that make them structurally unsuitable in this context. Our
requirements for an input format included that it be mathematics-free, so that
all mathematical details are encapsulated in source code, and that it has a
detailed, verifiable structure. These requirements made toml more attractive
than SBML: toml is easier for humans to read and edit and can straightforwardly
be validated using tools like pydantic. Further, an SBML representation of our
desired input would not contain differential equations.  It would therefore not
be interoperable with most SBML-targeting software, which typically assumes that
differential equations are available and does not know about Maud’s structure.

## Kinetic model

Maud’s kinetic model decomposes into factors as shown in equation
\eqref{eq-decomposition}.

\begin{equation}
F(C;\theta) = Enzyme\cdot k_{cat}\cdot Reversibility \cdot Saturation \cdot Allostery \label{eq-decomposition}
\end{equation}

 Each of the terms on the right-hand side of (REFER TO EQUATION) is a function
of  $C$ and $\theta$. This idea is taken from @noor_note_2013. The terms
usefully gather physically meaningful and conceptually distinct factors
contributing to reaction fluxes. $Enzyme$ captures the effect of enzyme
concentration, $k_{cat}$ that of enzyme efficiency, $Reversibility$ quantifies
thermodynamic effects, $Saturation$ the effect of enzyme availability and
$Allostery$ the effect of post translational modifications.

We used the model of enzyme saturation from @liebermeister_modular_2010
and the generalised Monod-Wyman-Changeux model of Allosteric
regulation introduced in [@monod_nature_1965; @changeux_2013;
@popova_generalization_1975; @popova_description_1979] and used more recently
in @matosGRASPComputationalPlatform2022. To capture the effect on enzyme
activity of coupled phosphorylation and dephosphorylation processes we developed
a new mathematical model inspired by the generalised MWC model of allosteric
regulation. Full details of all mathematical aspects of Maud’s kinetic model can
be found in supplementary material section (REFERENCE).

## Statistical model

Maud represents information from measurements using generalised linear
regression models that probabilistically connect realised measurements with
true values of the measurable quantities. Information from other sources
is represented using a prior distribution over a set of latent parameters.
The parameters and the measurable quantities are connected by a generative
model encompassing Maud’s kinetic model as well as the steady state equation
\eqref{eq-3}. Together, the measurement model, prior model and generative model
determine a joint probability function that assigns a probability density to any
possible combination of measurements and parameters.

Below we describe the prior and measurement models, as the generative model has
already been discussed above.

### Prior model

Maud’s prior model includes unknown parameters corresponding to quantities
in the kinetic model that are assumed to be unknown, other than steady state
metabolite concentrations and fluxes, which are derived from the values of
other parameters by solving the steady state problem. See Table 1 (REFERENCE)
for a description of all these parameters and their dimensions. Note that some
quantities in Maud’s kinetic model are not treated as parameters: for example,
temperatures, compartment volumes and the formation energy of water. Maud
treats these quantities as if they were known precisely: they can be configured
by the user or default values can be used. Although in practice there can be
considerable uncertainty regarding these quantities, we chose to disregard this
uncertainty in the interest of simplicity.

Except for metabolites’ standard condition Gibbs energy changes of formation,
Maud uses independent normal prior distributions for parameters that can in
principle be both negative and positive. For parameters that are constrained to
be positive, Maud uses independent log-normal distributions. Formation energy
parameters have a multivariate normal prior distribution. Location, scale and
covariance parameters for all these prior distributions can be chosen freely by
the user.

### Measurement model

Maud’s measurement model considers three types of measurement: metabolite
concentration measurements, enzyme concentration measurements and flux
measurements, represented by vectors $𝑦_{𝑐𝑜𝑛𝑐}$ $𝑦_{𝑒𝑛𝑧}$ and $𝑦_{𝑓𝑙𝑢𝑥}$
respectively.

All measurements are specific to an experimental condition - that is, a case
where the true state of the network, including knockouts, boundary conditions
and state variables as well as kinetic and thermodynamic parameters, can safely
be assumed to be the same. Maud’s statistical model allows for arbitrarily many
experimental conditions, and for any measurable quantity to be measured any
number of times in any condition.

Metabolite and enzyme measurements are intended to represent the results of
quantitative metabolomics and proteomics experiments. The likelihood functions
for such measurements are shown in equations Equation 6 and Equation 7.

EQUATIONS

Both equations are log-normal generalised linear models with a standard
link function (the natural logarithm $\ln$) and known standard deviation
$ \sigma_{𝑐𝑜𝑛𝑐}$. The use of this measurement model is motivated by the
consideration that concentrations are constrained to be non-negative, so the
measurement model should avoid assigning positive probability mass to negative
metabolite concentration values. In addition, we expect the precision of most
metabolomics and proteomics experiments to be roughly proportional to the value
of the true measured quantity, which supports a measurement model with constant
coefficient of variation. The measurement standard deviations $\sigma_{𝑐𝑜𝑛𝑐}$
and $\sigma_{𝑒𝑛𝑧}$ are assumed to be known exactly for simplicity; plausible
values can be elicited by considering the likely coefficient of variation of the
measuring apparatus.

Flux measurements, representing the results of quantitative fluxomics analyses,
are modelled using a likelihood function from a standard linear regression
model, as shown in equation Equation 8 (REFERENCE). Flux measurements can be
obtained from the results of isotope labelling experiments using metabolic
flux analysis, for example as described in (Young 2014). When entering flux
measurements, it is important only to specify measurements for a network’s free
fluxes, as the values of some steady state fluxes in a metabolic network are
constrained by others, with the result that dependent fluxes cannot typically be
measured separately. If measurements of multiple dependent fluxes are entered,
information will inappropriately be double counted.

EQUATION

Our measurement model improves on analyses of metabolomics and proteomics
data that assume a regression model with normally distributed errors, whether
explicitly using a standard linear model or implicitly using ordinary least
squares fitting. This assumption is undesirable because it implies that the
measured quantity could in principle be negative, and assumes an additive
underlying random process, whereas multiplicative processes tend to better
describe real concentration data.

The use of independent measurement models for metabolite, enzyme and flux
measurements carries an implicit assumption that there are no systematic
correlations in the measurement errors. This choice was motivated by simplicity
- it would be better to use a model with potentially correlated measurements.
Similarly, it would be preferable to include measurement errors as model
parameters, thereby avoiding possible bias due to incorrect assessments of
measurement accuracy. However, we chose to use a simpler measurement model
to avoid the complexity and potential fitting issues that these changes would
entail.

Finally, the reader may wonder why we chose a linear regression model for
reaction flux measurements even though this creates the potential for erroneous
double counting and requires non-trivial upstream modelling, as intracellular
fluxes typically cannot be measured directly, but must be inferred from isotope
labelling experiments using metabolic flux analysis (see REFERENCE). Ideally
Maud’s measurement model for fluxes would extend from fluxes to the results of
potential labelling experiments, thereby removing the need for upstream analysis
and avoiding any double counting. This option has not yet been pursued, again
for the sake of simplicity.

## Implementation

Maud uses the Python library click [@clickdevelopersClickPythonComposable2022]
to implement a command line interface. The command line interface loads
input files as Python dictionaries, which are parsed using the Python library
`toml` [@pearsonTomlPythonLibrary2020] and then validated and converted into
structured `MaudInput` objects using Pydantic [@pydanticdevelopersPydantic2022].
Maud’s statistical model is implemented in the probabilistic programming
language Stan [@carpenterStanProbabilisticProgramming2017] and accessed using
the interface cmdstanpy [@standevelopmentteamCmdStanPy2022]. For posterior
sampling, Maud uses the `MaudInput` to create a Stan input json file and
obtain configuration information for cmdstanpy, which it uses to (if necessary)
generate a model executable file and then trigger posterior sampling using
adaptive Hamiltonian Monte Carlo. When sampling is complete, Maud converts to
the output into the standard format InferenceData using the Python library arviz
[@kumarArviZUnifiedLibrary2019] and saves it as a json file, along with some
information for debugging.

Two details of Maud’s implementation are important to highlight: the method of
posterior sampling and how Maud solves the steady state problem \eqref{eq-3}.

### Posterior sampling

Although integrals of the joint probability model for kinetic models
are typically analytically intractable, they can be approximated
numerically using Markov Chain Monte Carlo (MCMC) and other methods.
Maud uses MCMC primarily because there exist many methods for verifying
that MCMC samples really do approximate the target probability
distribution: see@vehtariRankNormalizationFoldingLocalization2021 and
@taltsValidatingBayesianInference2018 for discussion of this point. In
addition, there are several examples of successful Bayesian kinetic modelling
projects using MCMC including @st.johnBayesianInferenceMetabolic2018 and
@xingModelingKineticsLargescale2010.

### Solving the steady state problem

In order to implement adaptive Hamiltonian Monte Carlo, Maud must
repeatedly solve the steady state problem \eqref{eq-3} and find its
gradients with respect to all parameters. To achieve this, Maud uses a
hybrid method involving two numerical solvers from the SUNDIALS suite
[@serbanCVODESSensitivityEnabledODE2005]: CVODES and IDAS. These solvers are
accessed via their interface from Stan. The hybrid method follows that propsed
by @margossianComputingSteadyStates2018, and involves numerically evolving
the ODE system for a short period of time, then using the difference between
the evolved and starting concentrations as the target for a numerical algebra
solver.

# Acknowledgements

# References

::: {#refs}
:::

### 2023-10-26 paper/main.tex
Add quarto paper

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\KOMAoption{captions}{tableheading}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Hierarchical Bayesian regression modelling predicts biochemical rection energies and facilitates quantitative modelling of cell metabolism},
  pdfauthor={Teddy Groves; Nicholas Luke Cowie; Lars Keld Nielsen},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Hierarchical Bayesian regression modelling predicts biochemical
rection energies and facilitates quantitative modelling of cell
metabolism}
\author{Teddy Groves \and Nicholas Luke Cowie \and Lars Keld Nielsen}
\date{}

\begin{document}
\maketitle
\begin{abstract}
This paper presents Maud, a command-line application that implements
Bayesian statistical inference for kinetic models of biochemical
metabolic reaction networks. Maud takes into account quantitative
information from omics experiments and background knowledge, as well as
structural information about kinetic mechanisms, regulatory interactions
and enzyme knockouts. Below, we review the existing options in this
area, explain how Maud improves on the state of the art, describe the
intended modelling workflow and illustrate its use with an example
application.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, breakable, sharp corners, interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, frame hidden]}{\end{tcolorbox}}\fi

\renewcommand*\contentsname{Table of contents}
{
\hypersetup{linkcolor=}
\setcounter{tocdepth}{3}
\tableofcontents
}
\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

A kinetic model of cellular metabolism aims to express what is known
about a cellular process in the form of an \emph{in silico}
representation of the underlying network of chemical reactions. Kinetic
models can be used to improve production of target molecules, determine
regulatory networks (Christodoulou et al. 2018) and identify potential
drug targets (DeBerardinis and Chandel 2016; Liberti et al. 2017).
However, the use of kinetic models in practice is hindered by their
dependence on noisy and uncertain information sources. Quantitative in
vivo measurements of chemical abundances, and in vitro measurements
relating to kinetic parameters, both contain vital information but are
notoriously inaccurate (REFERENCES). Practically useful kinetic
modelling therefore requires a principled statistical approach that
encompasses multiple possible model parameterisations.

Bayesian statistical inference can combine the structural information
implicit in kinetic models with knowledge about metabolic parameters and
information from omics measurements (P. A. Saa and Nielsen 2016;
Gopalakrishnan, Dash, and Maranas 2020). However, kinetic models pose
serious computational challenges for Bayesian inference (Gutenkunst et
al. 2007; A. Raue et al. 2010).

The scope of a kinetic model is defined by a stoichiometric matrix,
\(S\), in which rows represent metabolites, columns represent reactions,
and matrix elements \(s_{ij}\) represent the stoichiometric coefficient
of metabolite \(i\) in reaction \(j\). The change in metabolite
concentrations is

\begin{equation}\label{eq-1}
\frac{dC}{dt} = S\cdot v - \mu\cdot C 
\end{equation}

In equation \eqref{eq-1}, C represents a vector of metabolite
concentrations, \(v\) is a vector of reaction rates, and \(\mu\) is the
growth rate. The second term represents the dilution due to cell growth.

In a kinetic model, the rates, \(v\), are expressed as a function of the
enzyme concentrations, \(E\), the metabolite concentrations, \(C\), and
a set of parameters, \(\theta\) as shown in equation \eqref{eq-2}

\begin{equation}\label{eq-2}
v = f(C, E, \theta)
\end{equation}

The parameters must include sufficient boundary concentrations and
fluxes to solve \eqref{eq-1}.

It is common to assume pseudo-steady state for metabolites, i.e., the
rate of fluxes towards any metabolite is much greater than the rate of
change in concentration, \(𝑣 \gg \frac{𝑑𝐶}{𝑑𝑡}\). Moreover, the dilution
effect is assumed minimal, \(\mu\cdot C \ll \vec{v}\) (true unless the
concentration is very high). Finally, the enzyme concentration is
assumed constant for the period considered and hence part of the
parameters.

Given these assumptions and a set of values for \theta, a set of steady
state metabolite concentrations and fluxes can be found by solving for
\(C\) the algebraic equation:

\begin{equation}\label{eq-3}
S\cdot f(C;\theta) = 0
\end{equation}

In a fermentation context, \eqref{eq-3} captures the rapid kinetics
inside the cell, while another set of ODEs would be used to describe the
external substrate and product concentrations, which could act as
boundary parameters to \eqref{eq-3}.

In the context of kinetic modelling, Bayesian inference is appealing
because it allows uncertainty to be represented appropriately without
sacrificing mechanistic accuracy. Measurement uncertainty can naturally
be represented in a Bayesian measurement model, whereas the prior model
can represent quantitative uncertainty about kinetic parameters.
Finally, kinetic rate laws can be represented in Bayesian data
generation models with arbitrarily high fidelity. See (REFERENCES) for
more about Bayesian inference and (REFERENCE) for a discussion of
practical Bayesian workflow.

Another advantage is that Bayesian inference problems are well-posed
even when not all parameters are strongly identified. Sloppy models in
which measurable quantities are sensitive to combinations of parameters
but not to individual marginal parameter values are ubiquitous in models
of biological systems {[}gutenkunst\_2007; White et al. (2016){]}. The
parameter correlation structure represents the set of potential models
that describe the observed data. As we demonstrate in our case study
below, capturing this correlation structure is difficult outside a
Bayesian context.

Previous Bayesian kinetic models have either sacrificed mechanistic
accuracy or have attempted to fit realistic kinetic models using
obsolete or unreliable computational methods.

The most popular algorithm for fitting Bayesian statistical models is
Markov Chain Monte Carlo (MCMC). Modern MCMC algorithms allow
exploration of high- dimensional posterior distributions, have robust
failure diagnostics (Vehtari et al. 2021) and can incorporate fast
numerical solvers, thereby making inference feasible for Bayesian
kinetic models. Nonetheless, the kinetic modelling literature reports an
aversion to MCMC, rooted mainly in concerns about sampling time and the
presumed difficulty of implementing the required statistical model
(Andreas Raue et al. 2013; P. A. Saa and Nielsen 2016). We are only
aware of two recent attempts to implement a Bayesian kinetic modelling
approach using MCMC. Stapor et al. (2018) fitted detailed kinetic models
using relatively inefficient MCMC algorithms that do not scale well to
high dimensional parameter spaces limiting the scope of modelling.
Conversely, St. John et al. (2018) utilises an efficient sampling
algorithm but uses approximate kinetics, namely lin-log kinetics Visser
and Heijnen (2003), limiting the scope of interpreting parameters and
inferring cellular behaviour in experimental conditions outside the
reference dataset.

There have also been efforts to implement Bayesian inference for kinetic
models without the use of MCMC. Examples of alternative inference
methods include variational inference as in St. John et al. (2018),
rejection sampling and approximate Bayesian computation P. A. Saa and
Nielsen (2016) and Laplace approximation, in which the Fisher
information matrix is used to calculate a normal approximation around
the maximum a posteriori parameter configuration (Liebermeister and Noor
2021; Gopalakrishnan, Dash, and Maranas 2020; Stapor et al. 2018;
Andreas Raue et al. 2013). Non-MCMC-based Bayesian kinetic models have
limited utility because they lack reliable diagnostic tools for
verifying that their results approximate the target posterior
distribution. This is a problem because realistic kinetic models tend to
induce highly correlated, non-Gaussian, joint probability distributions
(Gutenkunst et al. 2007; Stapor et al. 2018).

Our application Maud is the first Bayesian kinetic model to combine
biologically realistic mechanistic accuracy---including accurate rate
laws, post- translational modification and thermodynamics---with fast,
robust MCMC sampling using adaptive Hamiltonian Monte Carlo. Further,
Maud is a general-purpose application that can be used to fit a wide
range of Bayesian kinetic models.

\hypertarget{results-and-discussion}{%
\section{Results and Discussion}\label{results-and-discussion}}

To demonstrate our application's capabilities, we used Maud to analyse
an artificial dataset based on the human methionine cycle. We first used
Maud to generate simulated training and validation measurements based on
plausible parameter values, then performed posterior sampling. Next, we
used Maud to predict the validation data.

To show that Maud is robust to missing measurements we compared the
results of fitting the full dataset with an intentionally incomplete
dataset. To demonstrate why a full Bayesian approach is preferable to an
approach based on a Laplace approximation of the posterior distribution,
we also fit our methionine cycle model using the latter method and
compared the results with MCMC sampling.

Finally, we investigated our results to find out what our model learned
about the contributions of different regulatory factors to the flux
through GNMT, an important reaction. This analysis illustrates how Maud
can be used to generate actionable insights about metabolism without the
need for further statistical analysis.

The methionine cycle, illustrated in
Figure~\ref{fig-methionine-reactions}, is a fundamental pathway in human
metabolism, whose intermediate metabolites participate in a variety of
mechanisms which must compete for the same resources. Due to this
competition, as well as the fact that all the functions occur
simultaneously, the methionine cycle is highly regulated, with 6 known
allosteric effectors (REFERENCE). This complex regulation means that
quantitative modelling of the methionine cycle requires a detailed
kinetic model: this is why we chose it as a case study for Maud.

\begin{figure}

{\centering \includegraphics{./figures/methionine-reactions.png}

}

\caption{\label{fig-methionine-reactions}CAPTION - FILL THIS IN!}

\end{figure}

\hypertarget{dataset-and-model-specification}{%
\subsection{Dataset and model
specification}\label{dataset-and-model-specification}}

The simulated dataset and underlying kinetic model that we used for our
analysis can be found in {[}\#Supplementary Info{]}.

We constructed a kinetic model of the methionine cycle in Maud's format
using the description in Korendyaseva et al. (2008). The ordinary
differential equation system describing this model is shown in
Equation\,\eqref{eq-meth-ode}.

After specifying the qualitative aspects of the kinetic model, we chose
parameter values to use as ground truth by Monte Carlo sampling using a
previous model of the methionine cycle as a starting point (see P. Saa
and Nielsen (2015) for this model).

We used these parameters to simulate steady states in a range of
plausible experimental conditions, again using P. Saa and Nielsen (2015)
as a starting point. These steady states were then used to generate
simulated measurements using the measurement model described below.

\begin{equation}
a = b \label{eq-meth-ode}
\end{equation}

\hypertarget{measurement-model}{%
\subsubsection{Measurement model}\label{measurement-model}}

Generating an artificial dataset required a specification of the true
measurement error; we specified a standard deviation of 0.1 on
logarithmic scale for enzyme and metabolite concentration measurements,
corresponding to approximately 10\% measurement error and a standard
deviation for each reaction measurement approximately 10\% of the
simulated value.

These measurement error specifications are somewhat optimistic
considering the many sources of variation and uncertainty affecting
quantitative proteomics, metabolomics and fluxomics analyses, but are a
reasonable first approximation to a realistic set of measurements.

For our main model run, we assumed that all metabolite and enzyme
concentrations were measured, and that there was a reaction measurement
for each of the network's elementary flux modes.

\hypertarget{trainingvalidation-split}{%
\subsubsection{Training/validation
split}\label{trainingvalidation-split}}

The training testing split was selected to achieve a large difference
between the fluxes of the training and testing dataset. The split was
determined as we are interested in showing how our model can fit to
varied conditions, and conditions closer to the training set are likely
to be predicted well without necessarily learning the system.

\hypertarget{additional-dataset-with-missing-measurements}{%
\subsubsection{Additional dataset with missing
measurements}\label{additional-dataset-with-missing-measurements}}

To gain insight into our model's robustness to missing measurements, we
also performed a model run with the same 6 experimental datasets, but
with measurements of the metabolite S-Adenosyl-L-homocysteine, or
``ahcys'' removed. Since ahcys regulates three enzymes in the methionine
cycle, including one enzyme which is also thermodynamically regulated,
we expected the removal of these measurements to yield interesting
results.

\hypertarget{maud-input-specification}{%
\subsubsection{Maud input
specification}\label{maud-input-specification}}

We constructed inputs in Maud's format for each of the analysed
datasets, based on the scenario that the true kinetic model was known
except for parameter values, which needed to be inferred from the
training data and priors. These inputs can be found at XXXX.

The prior distributions and corresponding true parameter values used in
our case study are shown in {[}Supplementary{]}. The first two columns
show the 1\% and 99\% quantiles of each marginal prior distribution.
True parameter value are shown in column three, and the last column
shows the z-score on log scale of the true parameter value according the
marginal prior distribution. As seen in table 2 {[}REFERENCE{]} there
are 7 parameters for which the true value is outside the 1\%-99\% range.
This is desirable, making the case study more realistic, because extreme
deviance from the prior distribution is likely to occur in practice due
to in vivo to in vitro measurement differences.

\hypertarget{computation}{%
\subsubsection{Computation}\label{computation}}

We conducted adaptive Hamiltonian Monte Carlo sampling for the full and
missing- data datasets, obtaining XXX post-warmup samples after XXX of
total computation time. For the Laplace approximation comparison, we
used Maud's Laplace mode. Full details as well as instructions for
reproducing our analysis can be found at XXX.

\hypertarget{findings}{%
\subsection{Findings}\label{findings}}

\hypertarget{posterior-inference}{%
\subsubsection{Posterior inference}\label{posterior-inference}}

Running standard diagnostic checks indicated that the samples we
generated were from the target posterior distribution. The improved
\(\hat{R}\) statistic (Vehtari et al. 2021) for every sampled variable
was within 1\% of 1, indicating appropriate mixing within and between
Markov chains. Additionally, the number of effective samples was high,
indicating that we generated enough posterior samples to support
inferences about the bulks of the distributions of the sampled
parameters. Furthermore, we observed no post-warmup divergent
transitions, indicating that the sampler was able to transform the
log-posterior distribution, avoiding any regions with excessive
curvature that might inhibit exploration via HMC.

Posterior predictive checking indicated that our model achieved a good
fit to the simulated reaction and metabolite concentration measurements,
as shown by the graphs in the top row of figure 3. Note that the fit was
good for both training and validation measurements.

Analysis of the posterior distributions for kinetic parameters indicated
that these are highly correlated. The marginal posterior distributions
for most kinetic parameters did not shrink significantly compared with
the corresponding marginal prior distributions, even though these
parameters' joint posterior distribution contained enough information to
make accurate out of sample predictions. In some cases, there were
two-dimensional correlations such as the one shown in the bottom left of
figure 3; in this case the marginal distribution of the two parameters
is roughly banana shaped. More commonly, however, two- dimensional pair
plots were insufficient to reveal the underlying correlation structure,
as seen in the bottom-right plot in figure 3. This does not mean that
the parameters were uncorrelated, but rather that the correlations
involve more than two parameters.

Overall, our results show that Maud can fit a realistic pathway-sized
dataset. This was achieved without fixing the marginal values of kinetic
parameters: the information required to make good predictions was
contained in the correlation structure of the joint posterior
distribution. This finding is consistent with previous analyses of
biological systems that found they are ``sloppy'', that is, sensitive to
parameter combinations rather than marginal parameter values, with
important combinations, scales and regions of sensitivity being
difficult to ascertain in advance Gutenkunst et al. (2007), Poirier
(1998).

The question naturally arises whether the crucial high-dimensional
parameter correlations are linear or non-linear. This is relevant to the
question of model performance, as linear correlations are easier to
correct for. A linearly correlated posterior space would also be easier
to summarise. We address this question below in section XXX.

This case study illustrates the type of kinetic model and dataset that
Maud can fit. The model we analysed has XXX reactions, XXX state
variables and XXX parameters. Posterior sampling with adaptive
Hamiltonian Monte Carlo generated XXX ESS/hour. Generalising from this
result, we conclude that it is feasible to use this method to fit models
in the same order of magnitude, but not, for example, genome-scale
kinetic models. To fit larger models, faster steady state solving
methods or alternative inference algorithms will be required. Section
XXX addresses whether Laplace approximation is a suitable candidate.

\hypertarget{comparison-with-laplace-approximation}{%
\subsubsection{Comparison with Laplace
approximation}\label{comparison-with-laplace-approximation}}

We found that the Laplace method was not able to produce an accurate
posterior approximation for our model and dataset.

The Laplace approximation yielded XXX samples in XXX time. The
diagnostics indicated that our algorithm was able to find the maximum a
posteriori parameter configuration, approximate the Hessian and use
these quantities to generate approximate posterior samples. The results
can be found at XXX.

Figure~\ref{fig-laplace} summarises the results of comparing the Laplace
approximation of our case study's posterior distribution with the true
posterior. As can be seen from the top left plot, the Laplace method
does not provide a good approximation to the true posterior
distribution, as the marginal distribution of the total log probability
density is clearly different. This was confirmed using the
Kolmogorov-Smirnov test, which is a test to differentiate two empirical
univariate distributions. The two distributions were significantly
different with a p-value indistinguishable from 0.

The difference between the Laplace approximation output and the true
posterior distribution manifests itself not only in the parameter space,
but also in the measurement space. Figure 5 frame B compares the
5\%-95\% interval for flux measurement log likelihoods in the true
posterior with the Laplace approximation; lower log likelihood values
indicate that the modelled and measured values are further away. The
graph shows that the Laplace approximation yielded significantly worse
predictions than the true posterior, even for the training data.

To further explore why this is the case we compared samples from the
true posterior and the Laplace approximation for the pairwise marginal
distributions of two Michaelis-Menten constants {[}LATEX CONSTANT{]}-
and {[}LATEX CONSTANT{]}: see the bottom right cell of figure 5. This
comparison demonstrates that the Laplace method is not able to capture
the correct relationships between parameters' distributions.

This result shows that MCMC, while slower than Laplace approximation, is
unfortunately preferable for posterior inference in this case. We expect
that this is typical of kinetic models of realistic metabolic networks
in general, so we recommend that Maud users use MCMC sampling if
possible.

Our results here also provide circumstantial evidence that the parameter
correlations in Bayesian kinetic model posteriors tend to be non-linear,
as a posterior with only linear correlations would likely be more
germane to Laplace approximation.A conclusion that we drew from this
analysis was that the results of fitting our model cannot be summarised
simply, for example by fitting a multivariate normal distribution to the
posterior draws. We therefore recommend that Maud users store the full
set of MCMC draws rather than using such an approximation. This does not
preclude the possibility that there is an alternative, more compact, way
to summarise the results of Bayesian kinetic model inference; we leave
research into this topic to future work.

\begin{figure}

{\centering \includegraphics{./figures/laplace.png}

}

\caption{\label{fig-laplace}CAPTION: FILL THIS IN!}

\end{figure}

\hypertarget{effect-of-missing-metabolite-concentration-measurements}{%
\subsubsection{Effect of missing metabolite concentration
measurements}\label{effect-of-missing-metabolite-concentration-measurements}}

Comparing model runs with and without the ahcys measurements showed that
Maud can produce sensible results even from incomplete metabolomics
data.

Figure~\ref{fig-missing} shows that, as might be expected, the model
with missing measurements did not correctly infer the missing ahcys
concentrations. Nonetheless, the remaining measured metabolites were
still well predicted, suggesting that information about the network is
still preserved despite the missing measurements. Comparison of flux
measurements in both models also indicated that removing the ahcys
measurement did not result in catastrophic model failure.

The missing measurements did affect Maud's ability to infer parameter
values correctly. As we can see in the \#lower left plot
of~\ref{fig-missing}, the model with full metabolomics learned the true
value for the displayed dissociation constant, despite this value being
far from the mean of the corresponding marginal prior distribution,
whereas the model with the missing measurements stayed in the
neighbourhood of the prior.

This result is reassuring because not having access to all measurements
is a common situation in multi-omics studies. For instance, measuring
all metabolites in a pathway can be infeasible because of limitations of
mass spectrometers, availability of standards, column effects, and
compartmentalisation. However, provided that sufficient information is
available from other sources, our approach can produce sensible results
from incomplete metabolomics data.

\begin{figure}

{\centering \includegraphics{./figures/missing.png}

}

\caption{\label{fig-missing}CAPTION: FILL THIS IN!}

\end{figure}

\hypertarget{application-to-regulatory-understanding}{%
\subsubsection{Application to regulatory
understanding}\label{application-to-regulatory-understanding}}

To investigate how the flux of GNMT between conditions {[}X{]} and
{[}Y{]} changes we compared the regulatory contributions in the
posterior distribution. Presented in Figure~\ref{fig-decomposition} is
the regulatory description of GNMT, an irreversible enzyme that is
homotropically activated by its substrate, competitively inhibited by
its product and heterotropically inhibited. The diverse regulation makes
it the ideal test case to elucidate regulatory changes.

Here we found that saturation and allosteric effects were the main
drivers of regulation between the two conditions.
Figure~\ref{fig-decomposition} shows the distributions of the regulatory
components as the log-ratio of each component with {[}X{]} in the
numerator and {[}Y{]} in the denominator, positive values suggest that
this component was increased in experiment {[}X{]} relative to {[}Y{]},
with 0 indicating no difference. Naturally, this presents the ideal
statistical test when evaluating the regulation, with the probability of
regulation given by the relative proportion of samples above and below
the 0 point.

We demonstrate that Maud can accurately reproduce the regulatory effect
between the two conditions by comparing it to the ground truth. As the
true values are well within our probability mass of the posterior we can
say that Maud is able to accurately reproduce the regulatory behaviour
of metabolic networks and also provides a tool for statistical inference
of the regulatory behaviour.

Here we have demonstrated a key feature in Maud, that is, direct
inference on any parameter that arises from the data generating process.
More succinctly, we show the benefit of error propagation that Bayesian
inference provides as researchers can directly use quantiles for
hypothesis testing.

Our tool allows inference directly on the joint posterior distribution
rather than the marginal parameter. The implication is that inference
will be better than that of the marginal distributions due to correlated
parameter values, if the data generating process can be directly
modelled.

\begin{figure}

{\centering \includegraphics{./figures/decomposition.png}

}

\caption{\label{fig-decomposition}CAPTION: FILL THIS IN!}

\end{figure}

\hypertarget{methods}{%
\section{Methods}\label{methods}}

Maud is a command line application implementing Bayesian inference for a
wide range of realistic kinetic models. Maud is written in Python (Van
Rossum and Drake 2009), designed for use on Windows, Macos and Linux,
registered on the Python Package Index as
\texttt{maud-metabolic-models}, documented at \textless https://maud-
metabolic-models.readthedocs.io\textgreater{} and actively developed and
maintained at \url{https://github.com/biosustain/Maud/}.

To use Maud, a user must first collate appropriate input information,
represent it in files with Maud's required formats (see section
(REFERENCE)below). Maud's command line interface provides commands for
inference, simulation and making out-of-sample predictions. Results are
stored in files, using a structured, interoperable format.

\hypertarget{input-format}{%
\subsection{Input format}\label{input-format}}

Maud inputs are structured directories, somewhat inspired by the PEtab
format Schmiester et al. (2021). A Maud input directory must contain a
toml (Preston-Werner, Tom and Gedam, Pradyun 2020) file called
\texttt{config.toml} which gives the input a name, configures how Maud
will be run and tells Maud where to find the other files, allowing these
to have custom names. It must also include a file containing a kinetic
model definition, a file specifying information about parameters and a
file with information experiments. The required structure of these files
is documented at \textless https://
maud-metabolic-models.readthedocs.io/en/latest/inputting.html\textgreater.
The input is validated against a Pydantic (Pydantic developers 2022)
data model which can be inspected at
\url{https://github.com/biosustain/Maud/tree/main/maud/data_model}.

We chose to implement a custom input format despite the existence of
standard formats in similar areas, including SBML (Keating et al. 2020)
and PEtab (Schmiester et al. 2021). This choice was partly motivated by
the need to ensure flexibility as Maud was developed, but there are also
features of SBML and PEtab that make them structurally unsuitable in
this context. Our requirements for an input format included that it be
mathematics-free, so that all mathematical details are encapsulated in
source code, and that it has a detailed, verifiable structure. These
requirements made toml more attractive than SBML: toml is easier for
humans to read and edit and can straightforwardly be validated using
tools like pydantic. Further, an SBML representation of our desired
input would not contain differential equations. It would therefore not
be interoperable with most SBML-targeting software, which typically
assumes that differential equations are available and does not know
about Maud's structure.

\hypertarget{kinetic-model}{%
\subsection{Kinetic model}\label{kinetic-model}}

Maud's kinetic model decomposes into factors as shown in equation
\eqref{eq-decomposition}.

\begin{equation}
F(C;\theta) = Enzyme\cdot k_{cat}\cdot Reversibility \cdot Saturation \cdot Allostery \label{eq-decomposition}
\end{equation}

Each of the terms on the right-hand side of (REFER TO EQUATION) is a
function of \(C\) and \(\theta\). This idea is taken from Noor et al.
(2013). The terms usefully gather physically meaningful and conceptually
distinct factors contributing to reaction fluxes. \(Enzyme\) captures
the effect of enzyme concentration, \(k_{cat}\) that of enzyme
efficiency, \(Reversibility\) quantifies thermodynamic effects,
\(Saturation\) the effect of enzyme availability and \(Allostery\) the
effect of post translational modifications.

We used the model of enzyme saturation from Liebermeister, Uhlendorf,
and Klipp (2010) and the generalised Monod-Wyman-Changeux model of
Allosteric regulation introduced in (Monod, Wyman, and Changeux 1965;
Changeux 2013; Popova and Sel'kov 1975, 1979) and used more recently in
Matos et al. (2022). To capture the effect on enzyme activity of coupled
phosphorylation and dephosphorylation processes we developed a new
mathematical model inspired by the generalised MWC model of allosteric
regulation. Full details of all mathematical aspects of Maud's kinetic
model can be found in supplementary material section (REFERENCE).

\hypertarget{statistical-model}{%
\subsection{Statistical model}\label{statistical-model}}

Maud represents information from measurements using generalised linear
regression models that probabilistically connect realised measurements
with true values of the measurable quantities. Information from other
sources is represented using a prior distribution over a set of latent
parameters. The parameters and the measurable quantities are connected
by a generative model encompassing Maud's kinetic model as well as the
steady state equation \eqref{eq-3}. Together, the measurement model,
prior model and generative model determine a joint probability function
that assigns a probability density to any possible combination of
measurements and parameters.

Below we describe the prior and measurement models, as the generative
model has already been discussed above.

\hypertarget{prior-model}{%
\subsubsection{Prior model}\label{prior-model}}

Maud's prior model includes unknown parameters corresponding to
quantities in the kinetic model that are assumed to be unknown, other
than steady state metabolite concentrations and fluxes, which are
derived from the values of other parameters by solving the steady state
problem. See Table 1 (REFERENCE) for a description of all these
parameters and their dimensions. Note that some quantities in Maud's
kinetic model are not treated as parameters: for example, temperatures,
compartment volumes and the formation energy of water. Maud treats these
quantities as if they were known precisely: they can be configured by
the user or default values can be used. Although in practice there can
be considerable uncertainty regarding these quantities, we chose to
disregard this uncertainty in the interest of simplicity.

Except for metabolites' standard condition Gibbs energy changes of
formation, Maud uses independent normal prior distributions for
parameters that can in principle be both negative and positive. For
parameters that are constrained to be positive, Maud uses independent
log-normal distributions. Formation energy parameters have a
multivariate normal prior distribution. Location, scale and covariance
parameters for all these prior distributions can be chosen freely by the
user.

\hypertarget{measurement-model-1}{%
\subsubsection{Measurement model}\label{measurement-model-1}}

Maud's measurement model considers three types of measurement:
metabolite concentration measurements, enzyme concentration measurements
and flux measurements, represented by vectors \(𝑦_{𝑐𝑜𝑛𝑐}\) \(𝑦_{𝑒𝑛𝑧}\)
and \(𝑦_{𝑓𝑙𝑢𝑥}\) respectively.

All measurements are specific to an experimental condition - that is, a
case where the true state of the network, including knockouts, boundary
conditions and state variables as well as kinetic and thermodynamic
parameters, can safely be assumed to be the same. Maud's statistical
model allows for arbitrarily many experimental conditions, and for any
measurable quantity to be measured any number of times in any condition.

Metabolite and enzyme measurements are intended to represent the results
of quantitative metabolomics and proteomics experiments. The likelihood
functions for such measurements are shown in equations Equation\,6 and
Equation\,7.

EQUATIONS

Both equations are log-normal generalised linear models with a standard
link function (the natural logarithm \(\ln\)) and known standard
deviation \$ \sigma\_\{𝑐𝑜𝑛𝑐\}\$. The use of this measurement model is
motivated by the consideration that concentrations are constrained to be
non-negative, so the measurement model should avoid assigning positive
probability mass to negative metabolite concentration values. In
addition, we expect the precision of most metabolomics and proteomics
experiments to be roughly proportional to the value of the true measured
quantity, which supports a measurement model with constant coefficient
of variation. The measurement standard deviations \(\sigma_{𝑐𝑜𝑛𝑐}\) and
\(\sigma_{𝑒𝑛𝑧}\) are assumed to be known exactly for simplicity;
plausible values can be elicited by considering the likely coefficient
of variation of the measuring apparatus.

Flux measurements, representing the results of quantitative fluxomics
analyses, are modelled using a likelihood function from a standard
linear regression model, as shown in equation Equation\,8 (REFERENCE).
Flux measurements can be obtained from the results of isotope labelling
experiments using metabolic flux analysis, for example as described in
(Young 2014). When entering flux measurements, it is important only to
specify measurements for a network's free fluxes, as the values of some
steady state fluxes in a metabolic network are constrained by others,
with the result that dependent fluxes cannot typically be measured
separately. If measurements of multiple dependent fluxes are entered,
information will inappropriately be double counted.

EQUATION

Our measurement model improves on analyses of metabolomics and
proteomics data that assume a regression model with normally distributed
errors, whether explicitly using a standard linear model or implicitly
using ordinary least squares fitting. This assumption is undesirable
because it implies that the measured quantity could in principle be
negative, and assumes an additive underlying random process, whereas
multiplicative processes tend to better describe real concentration
data.

The use of independent measurement models for metabolite, enzyme and
flux measurements carries an implicit assumption that there are no
systematic correlations in the measurement errors. This choice was
motivated by simplicity - it would be better to use a model with
potentially correlated measurements. Similarly, it would be preferable
to include measurement errors as model parameters, thereby avoiding
possible bias due to incorrect assessments of measurement accuracy.
However, we chose to use a simpler measurement model to avoid the
complexity and potential fitting issues that these changes would entail.

Finally, the reader may wonder why we chose a linear regression model
for reaction flux measurements even though this creates the potential
for erroneous double counting and requires non-trivial upstream
modelling, as intracellular fluxes typically cannot be measured
directly, but must be inferred from isotope labelling experiments using
metabolic flux analysis (see REFERENCE). Ideally Maud's measurement
model for fluxes would extend from fluxes to the results of potential
labelling experiments, thereby removing the need for upstream analysis
and avoiding any double counting. This option has not yet been pursued,
again for the sake of simplicity.

\hypertarget{implementation}{%
\subsection{Implementation}\label{implementation}}

Maud uses the Python library click (Click Developers 2022) to implement
a command line interface. The command line interface loads input files
as Python dictionaries, which are parsed using the Python library
\texttt{toml} (Pearson 2020) and then validated and converted into
structured \texttt{MaudInput} objects using Pydantic (Pydantic
developers 2022). Maud's statistical model is implemented in the
probabilistic programming language Stan (Carpenter et al. 2017) and
accessed using the interface cmdstanpy (Stan Development Team 2022). For
posterior sampling, Maud uses the \texttt{MaudInput} to create a Stan
input json file and obtain configuration information for cmdstanpy,
which it uses to (if necessary) generate a model executable file and
then trigger posterior sampling using adaptive Hamiltonian Monte Carlo.
When sampling is complete, Maud converts to the output into the standard
format InferenceData using the Python library arviz (Kumar et al. 2019)
and saves it as a json file, along with some information for debugging.

Two details of Maud's implementation are important to highlight: the
method of posterior sampling and how Maud solves the steady state
problem \eqref{eq-3}.

\hypertarget{posterior-sampling}{%
\subsubsection{Posterior sampling}\label{posterior-sampling}}

Although integrals of the joint probability model for kinetic models are
typically analytically intractable, they can be approximated numerically
using Markov Chain Monte Carlo (MCMC) and other methods. Maud uses MCMC
primarily because there exist many methods for verifying that MCMC
samples really do approximate the target probability distribution:
see@vehtariRankNormalizationFoldingLocalization2021 and Talts et al.
(2018) for discussion of this point. In addition, there are several
examples of successful Bayesian kinetic modelling projects using MCMC
including St. John et al. (2018) and Xing et al. (2010 Jan-Feb).

\hypertarget{solving-the-steady-state-problem}{%
\subsubsection{Solving the steady state
problem}\label{solving-the-steady-state-problem}}

In order to implement adaptive Hamiltonian Monte Carlo, Maud must
repeatedly solve the steady state problem \eqref{eq-3} and find its
gradients with respect to all parameters. To achieve this, Maud uses a
hybrid method involving two numerical solvers from the SUNDIALS suite
(Serban and Hindmarsh 2005): CVODES and IDAS. These solvers are accessed
via their interface from Stan. The hybrid method follows that propsed by
Margossian (2018), and involves numerically evolving the ODE system for
a short period of time, then using the difference between the evolved
and starting concentrations as the target for a numerical algebra
solver.

\hypertarget{acknowledgements}{%
\section{Acknowledgements}\label{acknowledgements}}

\hypertarget{references}{%
\section{References}\label{references}}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-carpenterStanProbabilisticProgramming2017}{}}%
Carpenter, Bob, Andrew Gelman, Matthew D. Hoffman, Daniel Lee, Ben
Goodrich, Michael Betancourt, Marcus Brubaker, Jiqiang Guo, Peter Li,
and Allen Riddell. 2017. {``Stan: {A Probabilistic Programming
Language}.''} \emph{Journal of Statistical Software} 76 (1): 1--32.
\url{https://doi.org/10.18637/jss.v076.i01}.

\leavevmode\vadjust pre{\hypertarget{ref-changeux_2013}{}}%
Changeux, Jean-Pierre. 2013. {``50 Years of Allosteric Interactions: The
Twists and Turns of the Models.''} \emph{Nature Reviews. Molecular Cell
Biology} 14 (12): 819--29. \url{https://doi.org/10.1038/nrm3695}.

\leavevmode\vadjust pre{\hypertarget{ref-christodoulou_reserve_2018}{}}%
Christodoulou, Dimitris, Hannes Link, Tobias Fuhrer, Karl Kochanowski,
Luca Gerosa, and Uwe Sauer. 2018. {``Reserve {Flux} {Capacity} in the
{Pentose} {Phosphate} {Pathway} {Enables} {Escherichia} Coli's {Rapid}
{Response} to {Oxidative} {Stress}.''} \emph{Cell Systems} 6 (5):
569--578.e7. \url{https://doi.org/10.1016/j.cels.2018.04.009}.

\leavevmode\vadjust pre{\hypertarget{ref-clickdevelopersClickPythonComposable2022}{}}%
Click Developers. 2022. {``Click: {Python} Composable Command Line
Interface Toolkit.''} Pallets. \url{https://pypi.org/project/click/}.

\leavevmode\vadjust pre{\hypertarget{ref-deberardinis_fundamentals_2016}{}}%
DeBerardinis, Ralph J, and Navdeep S Chandel. 2016. {``Fundamentals of
Cancer Metabolism.''} \emph{Science Advances} 2 (5): e1600200.
\url{https://doi.org/10.1126/sciadv.1600200}.

\leavevmode\vadjust pre{\hypertarget{ref-gopalakrishnan_k-fit_2020}{}}%
Gopalakrishnan, Saratram, Satyakam Dash, and Costas Maranas. 2020.
{``K-{FIT}: {An} Accelerated Kinetic Parameterization Algorithm Using
Steady-State Fluxomic Data.''} \emph{Metabolic Engineering}, March.
\url{https://doi.org/10.1016/j.ymben.2020.03.001}.

\leavevmode\vadjust pre{\hypertarget{ref-gutenkunst_2007}{}}%
Gutenkunst, Ryan N, Joshua J Waterfall, Fergal P Casey, Kevin S Brown,
Christopher R Myers, and James P Sethna. 2007. {``Universally Sloppy
Parameter Sensitivities in Systems Biology Models.''} \emph{PLoS
Computational Biology} 3 (10): 1871--78.
\url{https://doi.org/10.1371/journal.pcbi.0030189}.

\leavevmode\vadjust pre{\hypertarget{ref-keatingSBMLLevelExtensible2020}{}}%
Keating, Sarah M, Dagmar Waltemath, Matthias König, Fengkai Zhang,
Andreas Dräger, Claudine Chaouiya, Frank T Bergmann, et al. 2020.
{``{\textsc{SBML}} {Level} 3: An Extensible Format for the Exchange and
Reuse of Biological Models.''} \emph{Molecular Systems Biology} 16 (8):
e9110. \url{https://doi.org/10.15252/msb.20199110}.

\leavevmode\vadjust pre{\hypertarget{ref-korendyaseva_allosteric_2008}{}}%
Korendyaseva, Tatyana K., Denis N. Kuvatov, Vladimir A. Volkov, Michael
V. Martinov, Victor M. Vitvitsky, Ruma Banerjee, and Fazoil I.
Ataullakhanov. 2008. {``An {Allosteric} {Mechanism} for {Switching}
Between {Parallel} {Tracks} in {Mammalian} {Sulfur} {Metabolism}.''}
Edited by Daniel Segre. \emph{PLoS Computational Biology} 4 (5):
e1000076. \url{https://doi.org/10.1371/journal.pcbi.1000076}.

\leavevmode\vadjust pre{\hypertarget{ref-kumarArviZUnifiedLibrary2019}{}}%
Kumar, Ravin, Colin Carroll, Ari Hartikainen, and Osvaldo Martin. 2019.
{``{ArviZ} a Unified Library for Exploratory Analysis of {Bayesian}
Models in {Python}.''} \emph{Journal of Open Source Software} 4 (33):
1143. \url{https://doi.org/10.21105/joss.01143}.

\leavevmode\vadjust pre{\hypertarget{ref-Liberti2017}{}}%
Liberti, Maria V., Ziwei Dai, Suzanne E. Wardell, Joshua A. Baccile,
Xiaojing Liu, Xia Gao, Robert Baldi, et al. 2017. {``A {Predictive}
{Model} for {Selective} {Targeting} of the {Warburg} {Effect} Through
{GAPDH} {Inhibition} with a {Natural} {Product}.''} \emph{Cell
Metabolism} 26 (4): 648--659.e8.
\url{https://doi.org/10.1016/j.cmet.2017.08.017}.

\leavevmode\vadjust pre{\hypertarget{ref-liebermeister_model_2021}{}}%
Liebermeister, Wolfram, and Elad Noor. 2021. {``Model {Balancing}: {A}
{Search} for {In}-{Vivo} {Kinetic} {Constants} and {Consistent}
{Metabolic} {States}.''} \emph{Metabolites} 11 (11): 749.
\url{https://doi.org/10.3390/metabo11110749}.

\leavevmode\vadjust pre{\hypertarget{ref-liebermeister_modular_2010}{}}%
Liebermeister, Wolfram, Jannis Uhlendorf, and Edda Klipp. 2010.
{``Modular Rate Laws for Enzymatic Reactions: Thermodynamics,
Elasticities and Implementation.''} \emph{Bioinformatics} 26 (12):
1528--34. \url{https://doi.org/10.1093/bioinformatics/btq141}.

\leavevmode\vadjust pre{\hypertarget{ref-margossianComputingSteadyStates2018}{}}%
Margossian, Charles. 2018. {``Computing Steady States with {Stan}'s
Nonlinear Algebraic Solver.''} In \emph{Stan {Conference} 2018}.
\url{https://doi.org/10.5281/zenodo.1284375}.

\leavevmode\vadjust pre{\hypertarget{ref-matosGRASPComputationalPlatform2022}{}}%
Matos, Marta R A, Pedro A Saa, Nicholas Cowie, Svetlana Volkova, Marina
de Leeuw, and Lars K Nielsen. 2022. {``{GRASP}: A Computational Platform
for Building Kinetic Models of Cellular Metabolism.''}
\emph{Bioinformatics Advances} 2 (1): vbac066.
\url{https://doi.org/10.1093/bioadv/vbac066}.

\leavevmode\vadjust pre{\hypertarget{ref-monod_nature_1965}{}}%
Monod, J, J Wyman, and J P Changeux. 1965. {``On the Nature of
Allosteric Transitions: A Plausible Model.''} \emph{Journal of Molecular
Biology} 12 (May): 88--118.
\url{https://doi.org/10.1016/S0022-2836(65)80285-6}.

\leavevmode\vadjust pre{\hypertarget{ref-noor_note_2013}{}}%
Noor, Elad, Avi Flamholz, Wolfram Liebermeister, Arren Bar-Even, and Ron
Milo. 2013. {``A Note on the Kinetics of Enzyme Action: {A}
Decomposition That Highlights Thermodynamic Effects.''} \emph{FEBS
Letters}, A century of {Michaelis} - {Menten} kinetics, 587 (17):
2772--77. \url{https://doi.org/10.1016/j.febslet.2013.07.028}.

\leavevmode\vadjust pre{\hypertarget{ref-pearsonTomlPythonLibrary2020}{}}%
Pearson, William. 2020. {``Toml: {Python Library} for {Tom}'s {Obvious},
{Minimal Language}.''} \url{https://pypi.org/project/toml/}.

\leavevmode\vadjust pre{\hypertarget{ref-poirier_revising_1998}{}}%
Poirier, Dale J. 1998. {``{REVISING} {BELIEFS} {IN} {NONIDENTIFIED}
{MODELS}.''} \emph{Econometric Theory} 14 (4): 483--509.
\url{https://doi.org/10.1017/S0266466698144043}.

\leavevmode\vadjust pre{\hypertarget{ref-popova_generalization_1975}{}}%
Popova, S V, and E E Sel'kov. 1975. {``Generalization of the Model by
{Monod}, {Wyman} and {Changeux} for the Case of a Reversible
Monosubtrate Reaction {SR},{TP}.''} \emph{FEBS Letters} 53 (3): 269--73.
\url{https://doi.org/10.1016/0014-5793(75)80034-2}.

\leavevmode\vadjust pre{\hypertarget{ref-popova_description_1979}{}}%
---------. 1979. {``{[}{Description} of the Kinetics of the Two
Substrate Reactions {S1}+{S2} Goes to and Comes from {S3}+{S4} by a
Generalized {Monod}, {Wyman}, {Changeux} Model{]}.''}
\emph{Molekuliarnaia Biologiia} 13 (1): 129--39.
\url{https://www.ncbi.nlm.nih.gov/pubmed/156878}.

\leavevmode\vadjust pre{\hypertarget{ref-preston-wernertomandgedampradyunTOMLSpecification0rc2020}{}}%
Preston-Werner, Tom and Gedam, Pradyun. 2020. {``{TOML Specification}
1.0.0-Rc.1.''} \url{https://toml.io/en/v1.0.0-rc.1/}.

\leavevmode\vadjust pre{\hypertarget{ref-pydanticdevelopersPydantic2022}{}}%
Pydantic developers. 2022. {``Pydantic.''}
\url{https://pypi.org/project/pydantic/}.

\leavevmode\vadjust pre{\hypertarget{ref-raue_identifiability_2010}{}}%
Raue, A., V. Becker, U. Klingmüller, and J. Timmer. 2010.
{``Identifiability and Observability Analysis for Experimental Design in
Nonlinear Dynamical Models.''} \emph{Chaos: An Interdisciplinary Journal
of Nonlinear Science} 20 (4): 045105.
\url{https://doi.org/10.1063/1.3528102}.

\leavevmode\vadjust pre{\hypertarget{ref-raue_joining_2013}{}}%
Raue, Andreas, Clemens Kreutz, Fabian Joachim Theis, and Jens Timmer.
2013. {``Joining Forces of {Bayesian} and Frequentist Methodology: A
Study for Inference in the Presence of Non-Identifiability.''}
\emph{Philosophical Transactions of the Royal Society A: Mathematical,
Physical and Engineering Sciences} 371 (1984): 20110544.
\url{https://doi.org/10.1098/rsta.2011.0544}.

\leavevmode\vadjust pre{\hypertarget{ref-saa_construction_2016}{}}%
Saa, Pedro A, and Lars K Nielsen. 2016. {``Construction of Feasible and
Accurate Kinetic Models of Metabolism: {A} {Bayesian} Approach.''}
\emph{Scientific Reports} 6 (July): 29635.
\url{https://doi.org/10.1038/srep29635}.

\leavevmode\vadjust pre{\hypertarget{ref-saa_general_2015}{}}%
Saa, Pedro, and Lars K Nielsen. 2015. {``A General Framework for
Thermodynamically Consistent Parameterization and Efficient Sampling of
Enzymatic Reactions.''} \emph{PLoS Computational Biology} 11 (4):
e1004195. \url{https://doi.org/10.1371/journal.pcbi.1004195}.

\leavevmode\vadjust pre{\hypertarget{ref-SchmiesterSch2021}{}}%
Schmiester, Leonard, Yannik Schälte, Frank T. Bergmann, Tacio Camba,
Erika Dudkin, Janine Egert, Fabian Fröhlich, et al. 2021.
{``{PEtab}\textemdash{{Interoperable}} Specification of Parameter
Estimation Problems in Systems Biology.''} \emph{PLOS Computational
Biology} 17 (1): 1--10.
\url{https://doi.org/10.1371/journal.pcbi.1008646}.

\leavevmode\vadjust pre{\hypertarget{ref-serbanCVODESSensitivityEnabledODE2005}{}}%
Serban, Radu, and Alan C. Hindmarsh. 2005. {``{CVODES}: {The
Sensitivity-Enabled ODE Solver} in {SUNDIALS}.''} In \emph{Volume 6: 5th
{International Conference} on {Multibody Systems}, {Nonlinear Dynamics},
and {Control}, {Parts A}, {B}, and {C}}, 257--69. {Long Beach,
California, USA}: {ASMEDC}.
\url{https://doi.org/10.1115/DETC2005-85597}.

\leavevmode\vadjust pre{\hypertarget{ref-st.johnBayesianInferenceMetabolic2018}{}}%
St. John, Peter, Jonathan Strutz, Linda J Broadbelt, Keith E J Tyo, and
Yannick J Bomble. 2018. {``Bayesian Inference of Metabolic Kinetics from
Genome-Scale Multiomics Data.''} \emph{bioRxiv}, October.
\url{https://doi.org/10.1101/450163}.

\leavevmode\vadjust pre{\hypertarget{ref-standevelopmentteamCmdStanPy2022}{}}%
Stan Development Team. 2022. {``{CmdStanPy}.''}
\url{https://github.com/stan-dev/cmdstanpy}.

\leavevmode\vadjust pre{\hypertarget{ref-stapor_pesto_2018}{}}%
Stapor, Paul, Daniel Weindl, Benjamin Ballnus, Sabine Hug, Carolin Loos,
Anna Fiedler, Sabrina Krause, et al. 2018. {``{PESTO}: Parameter
Estimation Toolbox.''} \emph{Bioinformatics} 34 (4): 705--7.
\url{https://doi.org/10.1093/bioinformatics/btx676}.

\leavevmode\vadjust pre{\hypertarget{ref-taltsValidatingBayesianInference2018}{}}%
Talts, Sean, Michael Betancourt, Daniel Simpson, Aki Vehtari, and Andrew
Gelman. 2018. {``Validating {Bayesian Inference Algorithms} with
{Simulation-Based Calibration}.''} \emph{arXiv:1804.06788 {[}Stat{]}},
April. \url{http://arxiv.org/abs/1804.06788}.

\leavevmode\vadjust pre{\hypertarget{ref-vanrossumPythonReferenceManual2009}{}}%
Van Rossum, Guido, and Fred L. Drake. 2009. \emph{Python 3 {Reference
Manual}}. {Scotts Valley, CA}: {CreateSpace}.

\leavevmode\vadjust pre{\hypertarget{ref-vehtariRankNormalizationFoldingLocalization2021}{}}%
Vehtari, Aki, Andrew Gelman, Daniel Simpson, Bob Carpenter, and
Paul-Christian Bürkner. 2021. {``Rank-{Normalization}, {Folding}, and
{Localization}: {An Improved R\^{}} for {Assessing Convergence} of
{MCMC} (with {Discussion}).''} \emph{Bayesian Analysis} 16 (2):
667--718. \url{https://doi.org/10.1214/20-BA1221}.

\leavevmode\vadjust pre{\hypertarget{ref-visser_dynamic_2003}{}}%
Visser, Diana, and Joseph J Heijnen. 2003. {``Dynamic Simulation and
Metabolic Re-Design of a Branched Pathway Using Linlog Kinetics.''}
\emph{Metabolic Engineering} 5 (3): 164--76.
\url{https://doi.org/10.1016/s1096-7176(03)00025-9}.

\leavevmode\vadjust pre{\hypertarget{ref-white_2016}{}}%
White, Andrew, Malachi Tolman, Howard D Thames, Hubert Rodney Withers,
Kathy A Mason, and Mark K Transtrum. 2016. {``The Limitations of
Model-Based Experimental Design and Parameter Estimation in Sloppy
Systems.''} \emph{PLoS Computational Biology} 12 (12): e1005227.
\url{https://doi.org/10.1371/journal.pcbi.1005227}.

\leavevmode\vadjust pre{\hypertarget{ref-xingModelingKineticsLargescale2010}{}}%
Xing, Zizhuo, Nikki Bishop, Kirk Leister, and Zheng Jian Li. 2010
Jan-Feb. {``Modeling Kinetics of a Large-Scale Fed-Batch {CHO} Cell
Culture by {Markov} Chain {Monte Carlo} Method.''} \emph{Biotechnology
Progress} 26 (1): 208--19. \url{https://doi.org/10.1002/btpr.284}.

\end{CSLReferences}

\end{document}

### 2023-10-27 paper/main.qmd
Update paper

@@
 elements $s_{ij}$ represent the stoichiometric coefficient of metabolite $i$ in
-reaction $j$. The change in metabolite concentrations is
+reaction $j$. The change in metabolite concentrations is:
 
@@
 marginal parameter values are ubiquitous in models of biological systems
-[gutenkunst_2007; @white_2016]. The parameter correlation structure represents
+[@gutenkunst_2007; @white_2016]. The parameter correlation structure represents
 the set of potential models that describe the observed data. As we demonstrate
@@
 
-After specifying the qualitative aspects of the kinetic model, we chose
+After specifying the qualitative aspects of the kinetic model, we selected
 parameter values to use as ground truth by Monte Carlo sampling using a previous
@@
 We conducted adaptive Hamiltonian Monte Carlo sampling for the full and missing-
-data datasets, obtaining XXX post-warmup samples after XXX of total computation
+-data datasets, obtaining XXX post-warmup samples after XXX of total computation
 time. For the Laplace approximation comparison, we used Maud’s Laplace mode.
@@
 bulks of the distributions of the sampled parameters. Furthermore, we observed
-no post-warmup divergent transitions, indicating that the sampler was able to
+no post warm-up divergent transitions, indicating that the sampler was able to
 transform the log-posterior distribution, avoiding any regions with excessive
@@
 distribution contained enough information to make accurate out of sample
-predictions. In some cases, there were two-dimensional correlations such as the
-one shown in the bottom left of figure 3; in this case the marginal distribution
-of the two parameters is roughly banana shaped. More commonly, however, two-
-dimensional pair plots were insufficient to reveal the underlying correlation
-structure, as seen in the bottom-right plot in figure 3. This does not mean that
-the parameters were uncorrelated, but rather that the correlations involve more
-than two parameters.
+predictions. In some cases, there were two-dimensional correlations such
+as the one shown in the bottom left of figure 3; in this case the marginal
+distribution of the two parameters is roughly banana shaped. More commonly,
+however, two-dimensional pair plots were insufficient to reveal the underlying
+correlation structure, as seen in the bottom-right plot in figure 3. This does
+not mean that the parameters were uncorrelated, but rather that the correlations
+involve more than two parameters.
 
@@
 a posterior with only linear correlations would likely be more germane to
-Laplace approximation.A conclusion that we drew from this analysis was that
+Laplace approximation. A conclusion that we drew from this analysis was that
 the results of fitting our model cannot be summarised simply, for example by
@@
 a wide range of realistic kinetic models. Maud is written in Python
-[@vanrossumPythonReferenceManual2009], designed for use on Windows, Macos
+[@vanrossumPythonReferenceManual2009], designed for use on Windows, macOS
 and Linux, registered on the Python Package Index as `maud-metabolic-models`,
@@
 than SBML: toml is easier for humans to read and edit and can straightforwardly
-be validated using tools like pydantic. Further, an SBML representation of our
+be validated using tools like Pydantic. Further, an SBML representation of our
 desired input would not contain differential equations.  It would therefore not
-be interoperable with most SBML-targeting software, which typically assumes that
+be interoperable with most SBML targeting software, which typically assumes that
 differential equations are available and does not know about Maud’s structure.
@@
 parameters have a multivariate normal prior distribution. Location, scale and
-covariance parameters for all these prior distributions can be chosen freely by
-the user.
+covariance parameters for all these prior distributions can be selected freely
+by the user.
 
@@
 
-All measurements are specific to an experimental condition - that is, a case
+All measurements are specific to an experimental condition; that is, a case
 where the true state of the network, including knockouts, boundary conditions
@@
 are modelled using a likelihood function from a standard linear regression
-model, as shown in equation Equation 8 (REFERENCE). Flux measurements can be
-obtained from the results of isotope labelling experiments using metabolic
-flux analysis, for example as described in (Young 2014). When entering flux
-measurements, it is important only to specify measurements for a network’s free
-fluxes, as the values of some steady state fluxes in a metabolic network are
-constrained by others, with the result that dependent fluxes cannot typically be
-measured separately. If measurements of multiple dependent fluxes are entered,
+model, as shown in equation 8 (REFERENCE). Flux measurements can be obtained
+from the results of isotope labelling experiments using metabolic flux analysis,
+for example as described in (Young 2014). When entering flux measurements,
+it is important only to specify measurements for a network’s free fluxes, as
+the values of some steady state fluxes in a metabolic network are constrained
+by others, with the result that dependent fluxes cannot typically be measured
+separately. If measurements of multiple dependent fluxes are entered,
 information will inappropriately be double counted.
@@
 
-Finally, the reader may wonder why we chose a linear regression model for
+Finally, the reader may wonder why Maud uses a linear regression model for
 reaction flux measurements even though this creates the potential for erroneous
@@
 Maud uses the Python library click [@clickdevelopersClickPythonComposable2022]
-to implement a command line interface. The command line interface loads
-input files as Python dictionaries, which are parsed using the Python library
-`toml` [@pearsonTomlPythonLibrary2020] and then validated and converted into
-structured `MaudInput` objects using Pydantic [@pydanticdevelopersPydantic2022].
-Maud’s statistical model is implemented in the probabilistic programming
-language Stan [@carpenterStanProbabilisticProgramming2017] and accessed using
-the interface cmdstanpy [@standevelopmentteamCmdStanPy2022]. For posterior
-sampling, Maud uses the `MaudInput` to create a Stan input json file and
-obtain configuration information for cmdstanpy, which it uses to (if necessary)
-generate a model executable file and then trigger posterior sampling using
-adaptive Hamiltonian Monte Carlo. When sampling is complete, Maud converts to
-the output into the standard format InferenceData using the Python library arviz
+to implement a command line interface. The command line interface loads input
+files as Python dictionaries, which are parsed using the Python library `toml`
+[@pearsonTomlPythonLibrary2020] and then validated and converted into structured
+`MaudInput` objects using Pydantic [@pydanticdevelopersPydantic2022]. Maud’s
+statistical model is implemented in the probabilistic programming language Stan
+[@carpenterStanProbabilisticProgramming2017] and accessed using the interface
+cmdstanpy [@standevelopmentteamCmdStanPy2022]. For posterior sampling, Maud
+uses the `MaudInput` to create a Stan input json file and obtain configuration
+information for cmdstanpy, which it uses to (if necessary) generate a
+model executable file and then trigger posterior sampling using adaptive
+Hamiltonian Monte Carlo. When sampling is complete, Maud converts to the
+output into the standard format `InferenceData` using the Python library arviz
 [@kumarArviZUnifiedLibrary2019] and saves it as a json file, along with some
@@
 [@serbanCVODESSensitivityEnabledODE2005]: CVODES and IDAS. These solvers are
-accessed via their interface from Stan. The hybrid method follows that propsed
+accessed via their interface from Stan. The hybrid method follows that proposed
 by @margossianComputingSteadyStates2018, and involves numerically evolving

### 2023-10-27 paper/main.tex
Update paper

@@
 %
-\documentclass[
-  letterpaper,
-  DIV=11,
-  numbers=noendperiod]{scrartcl}
+\documentclass[journal=,manuscript=]{achemso}
+\usepackage[version=3]{mhchem}
+\newcommand*\mycommand[1]{\texttt{\emph{#1}}}
 
@@
 \makeatother
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
-  \fi
-  % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-
-\KOMAoption{captions}{tableheading}
+
 \makeatletter
@@
 
-\title{Hierarchical Bayesian regression modelling predicts biochemical
+\author{Teddy Groves}
+\affiliation{ DTU Biosustain,  }
+
+\email{tedgro@biosustain.dtu.dk}
+\author{Nicholas Luke Cowie}
+\affiliation{ DTU Biosustain,  }
+
+\author{Lars Keld Nielsen}
+\affiliation{ DTU Biosustain,  }
+
+\title[]{Hierarchical Bayesian regression modelling predicts biochemical
 rection energies and facilitates quantitative modelling of cell
 metabolism}
-\author{Teddy Groves \and Nicholas Luke Cowie \and Lars Keld Nielsen}
-\date{}
-
+\makeatletter
 \begin{document}
@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, breakable, sharp corners, interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, enhanced, frame hidden, breakable, boxrule=0pt, sharp corners, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
 
@@
 models can be used to improve production of target molecules, determine
-regulatory networks (Christodoulou et al. 2018) and identify potential
-drug targets (DeBerardinis and Chandel 2016; Liberti et al. 2017).
-However, the use of kinetic models in practice is hindered by their
-dependence on noisy and uncertain information sources. Quantitative in
-vivo measurements of chemical abundances, and in vitro measurements
-relating to kinetic parameters, both contain vital information but are
-notoriously inaccurate (REFERENCES). Practically useful kinetic
-modelling therefore requires a principled statistical approach that
-encompasses multiple possible model parameterisations.
+regulatory networks \citep{christodoulou_reserve_2018} and identify
+potential drug targets
+\citep{deberardinis_fundamentals_2016, Liberti2017}. However, the use of
+kinetic models in practice is hindered by their dependence on noisy and
+uncertain information sources. Quantitative in vivo measurements of
+chemical abundances, and in vitro measurements relating to kinetic
+parameters, both contain vital information but are notoriously
+inaccurate (REFERENCES). Practically useful kinetic modelling therefore
+requires a principled statistical approach that encompasses multiple
+possible model parameterisations.
 
@@
 implicit in kinetic models with knowledge about metabolic parameters and
-information from omics measurements (P. A. Saa and Nielsen 2016;
-Gopalakrishnan, Dash, and Maranas 2020). However, kinetic models pose
-serious computational challenges for Bayesian inference (Gutenkunst et
-al. 2007; A. Raue et al. 2010).
+information from omics measurements
+\citep{saa_construction_2016, gopalakrishnan_k-fit_2020}. However,
+kinetic models pose serious computational challenges for Bayesian
+inference \citep{gutenkunst_2007, raue_identifiability_2010}.
 
@@
 of metabolite \(i\) in reaction \(j\). The change in metabolite
-concentrations is
+concentrations is:
 
@@
 but not to individual marginal parameter values are ubiquitous in models
-of biological systems {[}gutenkunst\_2007; White et al. (2016){]}. The
-parameter correlation structure represents the set of potential models
-that describe the observed data. As we demonstrate in our case study
-below, capturing this correlation structure is difficult outside a
-Bayesian context.
+of biological systems \citep{gutenkunst_2007, white_2016}. The parameter
+correlation structure represents the set of potential models that
+describe the observed data. As we demonstrate in our case study below,
+capturing this correlation structure is difficult outside a Bayesian
+context.
 
@@
 exploration of high- dimensional posterior distributions, have robust
-failure diagnostics (Vehtari et al. 2021) and can incorporate fast
-numerical solvers, thereby making inference feasible for Bayesian
-kinetic models. Nonetheless, the kinetic modelling literature reports an
-aversion to MCMC, rooted mainly in concerns about sampling time and the
-presumed difficulty of implementing the required statistical model
-(Andreas Raue et al. 2013; P. A. Saa and Nielsen 2016). We are only
-aware of two recent attempts to implement a Bayesian kinetic modelling
-approach using MCMC. Stapor et al. (2018) fitted detailed kinetic models
-using relatively inefficient MCMC algorithms that do not scale well to
-high dimensional parameter spaces limiting the scope of modelling.
-Conversely, St. John et al. (2018) utilises an efficient sampling
-algorithm but uses approximate kinetics, namely lin-log kinetics Visser
-and Heijnen (2003), limiting the scope of interpreting parameters and
-inferring cellular behaviour in experimental conditions outside the
-reference dataset.
+failure diagnostics
+\citep{vehtariRankNormalizationFoldingLocalization2021} and can
+incorporate fast numerical solvers, thereby making inference feasible
+for Bayesian kinetic models. Nonetheless, the kinetic modelling
+literature reports an aversion to MCMC, rooted mainly in concerns about
+sampling time and the presumed difficulty of implementing the required
+statistical model \citep{raue_joining_2013, saa_construction_2016}. We
+are only aware of two recent attempts to implement a Bayesian kinetic
+modelling approach using MCMC. \citet{stapor_pesto_2018} fitted detailed
+kinetic models using relatively inefficient MCMC algorithms that do not
+scale well to high dimensional parameter spaces limiting the scope of
+modelling. Conversely, \citet{st.johnBayesianInferenceMetabolic2018}
+utilises an efficient sampling algorithm but uses approximate kinetics,
+namely lin-log kinetics \citet{visser_dynamic_2003}, limiting the scope
+of interpreting parameters and inferring cellular behaviour in
+experimental conditions outside the reference dataset.
 
@@
 models without the use of MCMC. Examples of alternative inference
-methods include variational inference as in St. John et al. (2018),
-rejection sampling and approximate Bayesian computation P. A. Saa and
-Nielsen (2016) and Laplace approximation, in which the Fisher
-information matrix is used to calculate a normal approximation around
-the maximum a posteriori parameter configuration (Liebermeister and Noor
-2021; Gopalakrishnan, Dash, and Maranas 2020; Stapor et al. 2018;
-Andreas Raue et al. 2013). Non-MCMC-based Bayesian kinetic models have
-limited utility because they lack reliable diagnostic tools for
-verifying that their results approximate the target posterior
-distribution. This is a problem because realistic kinetic models tend to
-induce highly correlated, non-Gaussian, joint probability distributions
-(Gutenkunst et al. 2007; Stapor et al. 2018).
+methods include variational inference as in
+\citet{st.johnBayesianInferenceMetabolic2018}, rejection sampling and
+approximate Bayesian computation \citet{saa_construction_2016} and
+Laplace approximation, in which the Fisher information matrix is used to
+calculate a normal approximation around the maximum a posteriori
+parameter configuration
+\citep{liebermeister_model_2021, gopalakrishnan_k-fit_2020, stapor_pesto_2018, raue_joining_2013}.
+Non-MCMC-based Bayesian kinetic models have limited utility because they
+lack reliable diagnostic tools for verifying that their results
+approximate the target posterior distribution. This is a problem because
+realistic kinetic models tend to induce highly correlated, non-Gaussian,
+joint probability distributions
+\citep{gutenkunst_2007, stapor_pesto_2018}.
 
@@
 We constructed a kinetic model of the methionine cycle in Maud's format
-using the description in Korendyaseva et al. (2008). The ordinary
-differential equation system describing this model is shown in
+using the description in \citet{korendyaseva_allosteric_2008}. The
+ordinary differential equation system describing this model is shown in
 Equation\,\eqref{eq-meth-ode}.
 
-After specifying the qualitative aspects of the kinetic model, we chose
-parameter values to use as ground truth by Monte Carlo sampling using a
-previous model of the methionine cycle as a starting point (see P. Saa
-and Nielsen (2015) for this model).
+After specifying the qualitative aspects of the kinetic model, we
+selected parameter values to use as ground truth by Monte Carlo sampling
+using a previous model of the methionine cycle as a starting point (see
+\citet{saa_general_2015} for this model).
 
 We used these parameters to simulate steady states in a range of
-plausible experimental conditions, again using P. Saa and Nielsen (2015)
+plausible experimental conditions, again using \citet{saa_general_2015}
 as a starting point. These steady states were then used to generate
@@
 We conducted adaptive Hamiltonian Monte Carlo sampling for the full and
-missing- data datasets, obtaining XXX post-warmup samples after XXX of
+missing- -data datasets, obtaining XXX post-warmup samples after XXX of
 total computation time. For the Laplace approximation comparison, we
@@
 generated were from the target posterior distribution. The improved
-\(\hat{R}\) statistic (Vehtari et al. 2021) for every sampled variable
-was within 1\% of 1, indicating appropriate mixing within and between
-Markov chains. Additionally, the number of effective samples was high,
-indicating that we generated enough posterior samples to support
-inferences about the bulks of the distributions of the sampled
-parameters. Furthermore, we observed no post-warmup divergent
+\(\hat{R}\) statistic
+\citep{vehtariRankNormalizationFoldingLocalization2021} for every
+sampled variable was within 1\% of 1, indicating appropriate mixing
+within and between Markov chains. Additionally, the number of effective
+samples was high, indicating that we generated enough posterior samples
+to support inferences about the bulks of the distributions of the
+sampled parameters. Furthermore, we observed no post warm-up divergent
 transitions, indicating that the sampler was able to transform the
@@
 figure 3; in this case the marginal distribution of the two parameters
-is roughly banana shaped. More commonly, however, two- dimensional pair
+is roughly banana shaped. More commonly, however, two-dimensional pair
 plots were insufficient to reveal the underlying correlation structure,
@@
 important combinations, scales and regions of sensitivity being
-difficult to ascertain in advance Gutenkunst et al. (2007), Poirier
-(1998).
+difficult to ascertain in advance \citet{gutenkunst_2007},
+\citet{poirier_revising_1998}.
 
@@
 as a posterior with only linear correlations would likely be more
-germane to Laplace approximation.A conclusion that we drew from this
+germane to Laplace approximation. A conclusion that we drew from this
 analysis was that the results of fitting our model cannot be summarised
@@
 Maud is a command line application implementing Bayesian inference for a
-wide range of realistic kinetic models. Maud is written in Python (Van
-Rossum and Drake 2009), designed for use on Windows, Macos and Linux,
-registered on the Python Package Index as
+wide range of realistic kinetic models. Maud is written in Python
+\citep{vanrossumPythonReferenceManual2009}, designed for use on Windows,
+macOS and Linux, registered on the Python Package Index as
 \texttt{maud-metabolic-models}, documented at \textless https://maud-
@@
 Maud inputs are structured directories, somewhat inspired by the PEtab
-format Schmiester et al. (2021). A Maud input directory must contain a
-toml (Preston-Werner, Tom and Gedam, Pradyun 2020) file called
-\texttt{config.toml} which gives the input a name, configures how Maud
-will be run and tells Maud where to find the other files, allowing these
-to have custom names. It must also include a file containing a kinetic
-model definition, a file specifying information about parameters and a
-file with information experiments. The required structure of these files
-is documented at \textless https://
+format \citet{SchmiesterSch2021}. A Maud input directory must contain a
+toml \citep{preston-wernertomandgedampradyunTOMLSpecification0rc2020}
+file called \texttt{config.toml} which gives the input a name,
+configures how Maud will be run and tells Maud where to find the other
+files, allowing these to have custom names. It must also include a file
+containing a kinetic model definition, a file specifying information
+about parameters and a file with information experiments. The required
+structure of these files is documented at \textless https://
 maud-metabolic-models.readthedocs.io/en/latest/inputting.html\textgreater.
-The input is validated against a Pydantic (Pydantic developers 2022)
-data model which can be inspected at
-\url{https://github.com/biosustain/Maud/tree/main/maud/data_model}.
+The input is validated against a Pydantic
+\citep{pydanticdevelopersPydantic2022} data model which can be inspected
+at \url{https://github.com/biosustain/Maud/tree/main/maud/data_model}.
 
 We chose to implement a custom input format despite the existence of
-standard formats in similar areas, including SBML (Keating et al. 2020)
-and PEtab (Schmiester et al. 2021). This choice was partly motivated by
-the need to ensure flexibility as Maud was developed, but there are also
-features of SBML and PEtab that make them structurally unsuitable in
-this context. Our requirements for an input format included that it be
+standard formats in similar areas, including SBML
+\citep{keatingSBMLLevelExtensible2020} and PEtab
+\citep{SchmiesterSch2021}. This choice was partly motivated by the need
+to ensure flexibility as Maud was developed, but there are also features
+of SBML and PEtab that make them structurally unsuitable in this
+context. Our requirements for an input format included that it be
 mathematics-free, so that all mathematical details are encapsulated in
@@
 humans to read and edit and can straightforwardly be validated using
-tools like pydantic. Further, an SBML representation of our desired
+tools like Pydantic. Further, an SBML representation of our desired
 input would not contain differential equations. It would therefore not
-be interoperable with most SBML-targeting software, which typically
+be interoperable with most SBML targeting software, which typically
 assumes that differential equations are available and does not know
@@
 Each of the terms on the right-hand side of (REFER TO EQUATION) is a
-function of \(C\) and \(\theta\). This idea is taken from Noor et al.
-(2013). The terms usefully gather physically meaningful and conceptually
-distinct factors contributing to reaction fluxes. \(Enzyme\) captures
-the effect of enzyme concentration, \(k_{cat}\) that of enzyme
-efficiency, \(Reversibility\) quantifies thermodynamic effects,
-\(Saturation\) the effect of enzyme availability and \(Allostery\) the
-effect of post translational modifications.
-
-We used the model of enzyme saturation from Liebermeister, Uhlendorf,
-and Klipp (2010) and the generalised Monod-Wyman-Changeux model of
-Allosteric regulation introduced in (Monod, Wyman, and Changeux 1965;
-Changeux 2013; Popova and Sel'kov 1975, 1979) and used more recently in
-Matos et al. (2022). To capture the effect on enzyme activity of coupled
-phosphorylation and dephosphorylation processes we developed a new
-mathematical model inspired by the generalised MWC model of allosteric
-regulation. Full details of all mathematical aspects of Maud's kinetic
-model can be found in supplementary material section (REFERENCE).
+function of \(C\) and \(\theta\). This idea is taken from
+\citet{noor_note_2013}. The terms usefully gather physically meaningful
+and conceptually distinct factors contributing to reaction fluxes.
+\(Enzyme\) captures the effect of enzyme concentration, \(k_{cat}\) that
+of enzyme efficiency, \(Reversibility\) quantifies thermodynamic
+effects, \(Saturation\) the effect of enzyme availability and
+\(Allostery\) the effect of post translational modifications.
+
+We used the model of enzyme saturation from
+\citet{liebermeister_modular_2010} and the generalised
+Monod-Wyman-Changeux model of Allosteric regulation introduced in
+\citep{monod_nature_1965, changeux_2013, popova_generalization_1975, popova_description_1979}
+and used more recently in \citet{matosGRASPComputationalPlatform2022}.
+To capture the effect on enzyme activity of coupled phosphorylation and
+dephosphorylation processes we developed a new mathematical model
+inspired by the generalised MWC model of allosteric regulation. Full
+details of all mathematical aspects of Maud's kinetic model can be found
+in supplementary material section (REFERENCE).
 
@@
 multivariate normal prior distribution. Location, scale and covariance
-parameters for all these prior distributions can be chosen freely by the
-user.
+parameters for all these prior distributions can be selected freely by
+the user.
 
@@
 
-All measurements are specific to an experimental condition - that is, a
+All measurements are specific to an experimental condition; that is, a
 case where the true state of the network, including knockouts, boundary
@@
 analyses, are modelled using a likelihood function from a standard
-linear regression model, as shown in equation Equation\,8 (REFERENCE).
-Flux measurements can be obtained from the results of isotope labelling
+linear regression model, as shown in equation\,8 (REFERENCE). Flux
+measurements can be obtained from the results of isotope labelling
 experiments using metabolic flux analysis, for example as described in
@@
 
-Finally, the reader may wonder why we chose a linear regression model
+Finally, the reader may wonder why Maud uses a linear regression model
 for reaction flux measurements even though this creates the potential
@@
 
-Maud uses the Python library click (Click Developers 2022) to implement
-a command line interface. The command line interface loads input files
-as Python dictionaries, which are parsed using the Python library
-\texttt{toml} (Pearson 2020) and then validated and converted into
-structured \texttt{MaudInput} objects using Pydantic (Pydantic
-developers 2022). Maud's statistical model is implemented in the
-probabilistic programming language Stan (Carpenter et al. 2017) and
-accessed using the interface cmdstanpy (Stan Development Team 2022). For
+Maud uses the Python library click
+\citep{clickdevelopersClickPythonComposable2022} to implement a command
+line interface. The command line interface loads input files as Python
+dictionaries, which are parsed using the Python library \texttt{toml}
+\citep{pearsonTomlPythonLibrary2020} and then validated and converted
+into structured \texttt{MaudInput} objects using Pydantic
+\citep{pydanticdevelopersPydantic2022}. Maud's statistical model is
+implemented in the probabilistic programming language Stan
+\citep{carpenterStanProbabilisticProgramming2017} and accessed using the
+interface cmdstanpy \citep{standevelopmentteamCmdStanPy2022}. For
 posterior sampling, Maud uses the \texttt{MaudInput} to create a Stan
@@
 When sampling is complete, Maud converts to the output into the standard
-format InferenceData using the Python library arviz (Kumar et al. 2019)
-and saves it as a json file, along with some information for debugging.
+format \texttt{InferenceData} using the Python library arviz
+\citep{kumarArviZUnifiedLibrary2019} and saves it as a json file, along
+with some information for debugging.
 
@@
 samples really do approximate the target probability distribution:
-see@vehtariRankNormalizationFoldingLocalization2021 and Talts et al.
-(2018) for discussion of this point. In addition, there are several
-examples of successful Bayesian kinetic modelling projects using MCMC
-including St. John et al. (2018) and Xing et al. (2010 Jan-Feb).
+see@vehtariRankNormalizationFoldingLocalization2021 and
+\citet{taltsValidatingBayesianInference2018} for discussion of this
+point. In addition, there are several examples of successful Bayesian
+kinetic modelling projects using MCMC including
+\citet{st.johnBayesianInferenceMetabolic2018} and
+\citet{xingModelingKineticsLargescale2010}.
 
@@
 hybrid method involving two numerical solvers from the SUNDIALS suite
-(Serban and Hindmarsh 2005): CVODES and IDAS. These solvers are accessed
-via their interface from Stan. The hybrid method follows that propsed by
-Margossian (2018), and involves numerically evolving the ODE system for
-a short period of time, then using the difference between the evolved
-and starting concentrations as the target for a numerical algebra
-solver.
+\citep{serbanCVODESSensitivityEnabledODE2005}: CVODES and IDAS. These
+solvers are accessed via their interface from Stan. The hybrid method
+follows that proposed by \citet{margossianComputingSteadyStates2018},
+and involves numerically evolving the ODE system for a short period of
+time, then using the difference between the evolved and starting
+concentrations as the target for a numerical algebra solver.
 
@@
 
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-carpenterStanProbabilisticProgramming2017}{}}%
-Carpenter, Bob, Andrew Gelman, Matthew D. Hoffman, Daniel Lee, Ben
-Goodrich, Michael Betancourt, Marcus Brubaker, Jiqiang Guo, Peter Li,
-and Allen Riddell. 2017. {``Stan: {A Probabilistic Programming
-Language}.''} \emph{Journal of Statistical Software} 76 (1): 1--32.
-\url{https://doi.org/10.18637/jss.v076.i01}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-changeux_2013}{}}%
-Changeux, Jean-Pierre. 2013. {``50 Years of Allosteric Interactions: The
-Twists and Turns of the Models.''} \emph{Nature Reviews. Molecular Cell
-Biology} 14 (12): 819--29. \url{https://doi.org/10.1038/nrm3695}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-christodoulou_reserve_2018}{}}%
-Christodoulou, Dimitris, Hannes Link, Tobias Fuhrer, Karl Kochanowski,
-Luca Gerosa, and Uwe Sauer. 2018. {``Reserve {Flux} {Capacity} in the
-{Pentose} {Phosphate} {Pathway} {Enables} {Escherichia} Coli's {Rapid}
-{Response} to {Oxidative} {Stress}.''} \emph{Cell Systems} 6 (5):
-569--578.e7. \url{https://doi.org/10.1016/j.cels.2018.04.009}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-clickdevelopersClickPythonComposable2022}{}}%
-Click Developers. 2022. {``Click: {Python} Composable Command Line
-Interface Toolkit.''} Pallets. \url{https://pypi.org/project/click/}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-deberardinis_fundamentals_2016}{}}%
-DeBerardinis, Ralph J, and Navdeep S Chandel. 2016. {``Fundamentals of
-Cancer Metabolism.''} \emph{Science Advances} 2 (5): e1600200.
-\url{https://doi.org/10.1126/sciadv.1600200}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-gopalakrishnan_k-fit_2020}{}}%
-Gopalakrishnan, Saratram, Satyakam Dash, and Costas Maranas. 2020.
-{``K-{FIT}: {An} Accelerated Kinetic Parameterization Algorithm Using
-Steady-State Fluxomic Data.''} \emph{Metabolic Engineering}, March.
-\url{https://doi.org/10.1016/j.ymben.2020.03.001}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-gutenkunst_2007}{}}%
-Gutenkunst, Ryan N, Joshua J Waterfall, Fergal P Casey, Kevin S Brown,
-Christopher R Myers, and James P Sethna. 2007. {``Universally Sloppy
-Parameter Sensitivities in Systems Biology Models.''} \emph{PLoS
-Computational Biology} 3 (10): 1871--78.
-\url{https://doi.org/10.1371/journal.pcbi.0030189}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-keatingSBMLLevelExtensible2020}{}}%
-Keating, Sarah M, Dagmar Waltemath, Matthias König, Fengkai Zhang,
-Andreas Dräger, Claudine Chaouiya, Frank T Bergmann, et al. 2020.
-{``{\textsc{SBML}} {Level} 3: An Extensible Format for the Exchange and
-Reuse of Biological Models.''} \emph{Molecular Systems Biology} 16 (8):
-e9110. \url{https://doi.org/10.15252/msb.20199110}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-korendyaseva_allosteric_2008}{}}%
-Korendyaseva, Tatyana K., Denis N. Kuvatov, Vladimir A. Volkov, Michael
-V. Martinov, Victor M. Vitvitsky, Ruma Banerjee, and Fazoil I.
-Ataullakhanov. 2008. {``An {Allosteric} {Mechanism} for {Switching}
-Between {Parallel} {Tracks} in {Mammalian} {Sulfur} {Metabolism}.''}
-Edited by Daniel Segre. \emph{PLoS Computational Biology} 4 (5):
-e1000076. \url{https://doi.org/10.1371/journal.pcbi.1000076}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-kumarArviZUnifiedLibrary2019}{}}%
-Kumar, Ravin, Colin Carroll, Ari Hartikainen, and Osvaldo Martin. 2019.
-{``{ArviZ} a Unified Library for Exploratory Analysis of {Bayesian}
-Models in {Python}.''} \emph{Journal of Open Source Software} 4 (33):
-1143. \url{https://doi.org/10.21105/joss.01143}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Liberti2017}{}}%
-Liberti, Maria V., Ziwei Dai, Suzanne E. Wardell, Joshua A. Baccile,
-Xiaojing Liu, Xia Gao, Robert Baldi, et al. 2017. {``A {Predictive}
-{Model} for {Selective} {Targeting} of the {Warburg} {Effect} Through
-{GAPDH} {Inhibition} with a {Natural} {Product}.''} \emph{Cell
-Metabolism} 26 (4): 648--659.e8.
-\url{https://doi.org/10.1016/j.cmet.2017.08.017}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-liebermeister_model_2021}{}}%
-Liebermeister, Wolfram, and Elad Noor. 2021. {``Model {Balancing}: {A}
-{Search} for {In}-{Vivo} {Kinetic} {Constants} and {Consistent}
-{Metabolic} {States}.''} \emph{Metabolites} 11 (11): 749.
-\url{https://doi.org/10.3390/metabo11110749}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-liebermeister_modular_2010}{}}%
-Liebermeister, Wolfram, Jannis Uhlendorf, and Edda Klipp. 2010.
-{``Modular Rate Laws for Enzymatic Reactions: Thermodynamics,
-Elasticities and Implementation.''} \emph{Bioinformatics} 26 (12):
-1528--34. \url{https://doi.org/10.1093/bioinformatics/btq141}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-margossianComputingSteadyStates2018}{}}%
-Margossian, Charles. 2018. {``Computing Steady States with {Stan}'s
-Nonlinear Algebraic Solver.''} In \emph{Stan {Conference} 2018}.
-\url{https://doi.org/10.5281/zenodo.1284375}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-matosGRASPComputationalPlatform2022}{}}%
-Matos, Marta R A, Pedro A Saa, Nicholas Cowie, Svetlana Volkova, Marina
-de Leeuw, and Lars K Nielsen. 2022. {``{GRASP}: A Computational Platform
-for Building Kinetic Models of Cellular Metabolism.''}
-\emph{Bioinformatics Advances} 2 (1): vbac066.
-\url{https://doi.org/10.1093/bioadv/vbac066}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-monod_nature_1965}{}}%
-Monod, J, J Wyman, and J P Changeux. 1965. {``On the Nature of
-Allosteric Transitions: A Plausible Model.''} \emph{Journal of Molecular
-Biology} 12 (May): 88--118.
-\url{https://doi.org/10.1016/S0022-2836(65)80285-6}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-noor_note_2013}{}}%
-Noor, Elad, Avi Flamholz, Wolfram Liebermeister, Arren Bar-Even, and Ron
-Milo. 2013. {``A Note on the Kinetics of Enzyme Action: {A}
-Decomposition That Highlights Thermodynamic Effects.''} \emph{FEBS
-Letters}, A century of {Michaelis} - {Menten} kinetics, 587 (17):
-2772--77. \url{https://doi.org/10.1016/j.febslet.2013.07.028}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-pearsonTomlPythonLibrary2020}{}}%
-Pearson, William. 2020. {``Toml: {Python Library} for {Tom}'s {Obvious},
-{Minimal Language}.''} \url{https://pypi.org/project/toml/}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-poirier_revising_1998}{}}%
-Poirier, Dale J. 1998. {``{REVISING} {BELIEFS} {IN} {NONIDENTIFIED}
-{MODELS}.''} \emph{Econometric Theory} 14 (4): 483--509.
-\url{https://doi.org/10.1017/S0266466698144043}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-popova_generalization_1975}{}}%
-Popova, S V, and E E Sel'kov. 1975. {``Generalization of the Model by
-{Monod}, {Wyman} and {Changeux} for the Case of a Reversible
-Monosubtrate Reaction {SR},{TP}.''} \emph{FEBS Letters} 53 (3): 269--73.
-\url{https://doi.org/10.1016/0014-5793(75)80034-2}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-popova_description_1979}{}}%
----------. 1979. {``{[}{Description} of the Kinetics of the Two
-Substrate Reactions {S1}+{S2} Goes to and Comes from {S3}+{S4} by a
-Generalized {Monod}, {Wyman}, {Changeux} Model{]}.''}
-\emph{Molekuliarnaia Biologiia} 13 (1): 129--39.
-\url{https://www.ncbi.nlm.nih.gov/pubmed/156878}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-preston-wernertomandgedampradyunTOMLSpecification0rc2020}{}}%
-Preston-Werner, Tom and Gedam, Pradyun. 2020. {``{TOML Specification}
-1.0.0-Rc.1.''} \url{https://toml.io/en/v1.0.0-rc.1/}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-pydanticdevelopersPydantic2022}{}}%
-Pydantic developers. 2022. {``Pydantic.''}
-\url{https://pypi.org/project/pydantic/}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-raue_identifiability_2010}{}}%
-Raue, A., V. Becker, U. Klingmüller, and J. Timmer. 2010.
-{``Identifiability and Observability Analysis for Experimental Design in
-Nonlinear Dynamical Models.''} \emph{Chaos: An Interdisciplinary Journal
-of Nonlinear Science} 20 (4): 045105.
-\url{https://doi.org/10.1063/1.3528102}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-raue_joining_2013}{}}%
-Raue, Andreas, Clemens Kreutz, Fabian Joachim Theis, and Jens Timmer.
-2013. {``Joining Forces of {Bayesian} and Frequentist Methodology: A
-Study for Inference in the Presence of Non-Identifiability.''}
-\emph{Philosophical Transactions of the Royal Society A: Mathematical,
-Physical and Engineering Sciences} 371 (1984): 20110544.
-\url{https://doi.org/10.1098/rsta.2011.0544}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-saa_construction_2016}{}}%
-Saa, Pedro A, and Lars K Nielsen. 2016. {``Construction of Feasible and
-Accurate Kinetic Models of Metabolism: {A} {Bayesian} Approach.''}
-\emph{Scientific Reports} 6 (July): 29635.
-\url{https://doi.org/10.1038/srep29635}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-saa_general_2015}{}}%
-Saa, Pedro, and Lars K Nielsen. 2015. {``A General Framework for
-Thermodynamically Consistent Parameterization and Efficient Sampling of
-Enzymatic Reactions.''} \emph{PLoS Computational Biology} 11 (4):
-e1004195. \url{https://doi.org/10.1371/journal.pcbi.1004195}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-SchmiesterSch2021}{}}%
-Schmiester, Leonard, Yannik Schälte, Frank T. Bergmann, Tacio Camba,
-Erika Dudkin, Janine Egert, Fabian Fröhlich, et al. 2021.
-{``{PEtab}\textemdash{{Interoperable}} Specification of Parameter
-Estimation Problems in Systems Biology.''} \emph{PLOS Computational
-Biology} 17 (1): 1--10.
-\url{https://doi.org/10.1371/journal.pcbi.1008646}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-serbanCVODESSensitivityEnabledODE2005}{}}%
-Serban, Radu, and Alan C. Hindmarsh. 2005. {``{CVODES}: {The
-Sensitivity-Enabled ODE Solver} in {SUNDIALS}.''} In \emph{Volume 6: 5th
-{International Conference} on {Multibody Systems}, {Nonlinear Dynamics},
-and {Control}, {Parts A}, {B}, and {C}}, 257--69. {Long Beach,
-California, USA}: {ASMEDC}.
-\url{https://doi.org/10.1115/DETC2005-85597}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-st.johnBayesianInferenceMetabolic2018}{}}%
-St. John, Peter, Jonathan Strutz, Linda J Broadbelt, Keith E J Tyo, and
-Yannick J Bomble. 2018. {``Bayesian Inference of Metabolic Kinetics from
-Genome-Scale Multiomics Data.''} \emph{bioRxiv}, October.
-\url{https://doi.org/10.1101/450163}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-standevelopmentteamCmdStanPy2022}{}}%
-Stan Development Team. 2022. {``{CmdStanPy}.''}
-\url{https://github.com/stan-dev/cmdstanpy}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-stapor_pesto_2018}{}}%
-Stapor, Paul, Daniel Weindl, Benjamin Ballnus, Sabine Hug, Carolin Loos,
-Anna Fiedler, Sabrina Krause, et al. 2018. {``{PESTO}: Parameter
-Estimation Toolbox.''} \emph{Bioinformatics} 34 (4): 705--7.
-\url{https://doi.org/10.1093/bioinformatics/btx676}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-taltsValidatingBayesianInference2018}{}}%
-Talts, Sean, Michael Betancourt, Daniel Simpson, Aki Vehtari, and Andrew
-Gelman. 2018. {``Validating {Bayesian Inference Algorithms} with
-{Simulation-Based Calibration}.''} \emph{arXiv:1804.06788 {[}Stat{]}},
-April. \url{http://arxiv.org/abs/1804.06788}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-vanrossumPythonReferenceManual2009}{}}%
-Van Rossum, Guido, and Fred L. Drake. 2009. \emph{Python 3 {Reference
-Manual}}. {Scotts Valley, CA}: {CreateSpace}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-vehtariRankNormalizationFoldingLocalization2021}{}}%
-Vehtari, Aki, Andrew Gelman, Daniel Simpson, Bob Carpenter, and
-Paul-Christian Bürkner. 2021. {``Rank-{Normalization}, {Folding}, and
-{Localization}: {An Improved R\^{}} for {Assessing Convergence} of
-{MCMC} (with {Discussion}).''} \emph{Bayesian Analysis} 16 (2):
-667--718. \url{https://doi.org/10.1214/20-BA1221}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-visser_dynamic_2003}{}}%
-Visser, Diana, and Joseph J Heijnen. 2003. {``Dynamic Simulation and
-Metabolic Re-Design of a Branched Pathway Using Linlog Kinetics.''}
-\emph{Metabolic Engineering} 5 (3): 164--76.
-\url{https://doi.org/10.1016/s1096-7176(03)00025-9}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-white_2016}{}}%
-White, Andrew, Malachi Tolman, Howard D Thames, Hubert Rodney Withers,
-Kathy A Mason, and Mark K Transtrum. 2016. {``The Limitations of
-Model-Based Experimental Design and Parameter Estimation in Sloppy
-Systems.''} \emph{PLoS Computational Biology} 12 (12): e1005227.
-\url{https://doi.org/10.1371/journal.pcbi.1005227}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-xingModelingKineticsLargescale2010}{}}%
-Xing, Zizhuo, Nikki Bishop, Kirk Leister, and Zheng Jian Li. 2010
-Jan-Feb. {``Modeling Kinetics of a Large-Scale Fed-Batch {CHO} Cell
-Culture by {Markov} Chain {Monte Carlo} Method.''} \emph{Biotechnology
-Progress} 26 (1): 208--19. \url{https://doi.org/10.1002/btpr.284}.
-
-\end{CSLReferences}
+\renewcommand{\bibsection}{}
+\bibliography{bibliography.bib}
 

### 2023-10-27 paper/main.qmd
Update paper

@@
 concentration measurements, enzyme concentration measurements and flux
-measurements, represented by vectors $𝑦_{𝑐𝑜𝑛𝑐}$ $𝑦_{𝑒𝑛𝑧}$ and $𝑦_{𝑓𝑙𝑢𝑥}$
+measurements, represented by vectors $𝑦^{𝑐𝑜𝑛𝑐}$ $𝑦^{𝑒𝑛𝑧}$ and $𝑦^{𝑓𝑙𝑢𝑥}$
 respectively.
@@
 quantitative metabolomics and proteomics experiments. The likelihood functions
-for such measurements are shown in equations Equation 6 and Equation 7.
+for such measurements are shown in equations Equation \eqref{eq-yconc} and
+Equation \eqref{eq-yenz}.
 
-EQUATIONS
+\begin{equation}
+y_i^{conc} \sim LN(\ln{\hat{y}_i^{conc}}, \sigma_i^{conc})\label{eq-yconc}
+\end{equation}
+
+\begin{equation}
+y_i^{enz} \sim LN(\ln{\hat{y}_i^{enz}}, \sigma_i^{enz})\label{eq-yenz}
+\end{equation}
 
@@
 are modelled using a likelihood function from a standard linear regression
-model, as shown in equation 8 (REFERENCE). Flux measurements can be obtained
+model, as shown in equation \eqref{eq-yflux}. Flux measurements can be obtained
 from the results of isotope labelling experiments using metabolic flux analysis,
@@
 
-EQUATION
+\begin{equation}
+y_i^{flux} \sim LN(\ln{\hat{y}_i^{flux}}, \sigma_i^{flux})\label{eq-yflux}
+\end{equation}
 

### 2023-10-27 paper/main.tex
Update paper

@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, enhanced, frame hidden, breakable, boxrule=0pt, sharp corners, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable, frame hidden, interior hidden]}{\end{tcolorbox}}\fi
 
@@
 metabolite concentration measurements, enzyme concentration measurements
-and flux measurements, represented by vectors \(𝑦_{𝑐𝑜𝑛𝑐}\) \(𝑦_{𝑒𝑛𝑧}\)
-and \(𝑦_{𝑓𝑙𝑢𝑥}\) respectively.
+and flux measurements, represented by vectors \(𝑦^{𝑐𝑜𝑛𝑐}\) \(𝑦^{𝑒𝑛𝑧}\)
+and \(𝑦^{𝑓𝑙𝑢𝑥}\) respectively.
 
@@
 of quantitative metabolomics and proteomics experiments. The likelihood
-functions for such measurements are shown in equations Equation\,6 and
-Equation\,7.
+functions for such measurements are shown in equations
+Equation\,\eqref{eq-yconc} and Equation\,\eqref{eq-yenz}.
 
-EQUATIONS
+\begin{equation}
+y_i^{conc} \sim LN(\ln{\hat{y}_i^{conc}}, \sigma_i^{conc})\label{eq-yconc}
+\end{equation}
+
+\begin{equation}
+y_i^{enz} \sim LN(\ln{\hat{y}_i^{enz}}, \sigma_i^{enz})\label{eq-yenz}
+\end{equation}
 
@@
 analyses, are modelled using a likelihood function from a standard
-linear regression model, as shown in equation\,8 (REFERENCE). Flux
+linear regression model, as shown in equation\,\eqref{eq-yflux}. Flux
 measurements can be obtained from the results of isotope labelling

### 2023-10-27 paper/main.qmd
Update paper

@@
 represented in Bayesian data generation models with arbitrarily high fidelity.
-See (REFERENCES) for more about Bayesian inference and (REFERENCE) for a
-discussion of practical Bayesian workflow.
+See @gelmanBayesianDataAnalysis2020a for more about Bayesian inference and
+@gelmanBayesianWorkflow2020 for a discussion of practical Bayesian workflow.
 
@@
 The simulated dataset and underlying kinetic model that we used for our analysis
-can be found in [#Supplementary Info].
+can be found at <https://github.com/biosustain/Methionine_model/tree/main/data/methionine>
+and is described in supplementary information section [REFERENCE].
 
@@
 equation system describing this model is shown in Equation \eqref{eq-meth-ode}.
+
+\begin{align}
+\frac{d[met]}{dt} &= v_{Influx} - v_{PROT} - v_{MAT} +v_{MS} + v_{BHMT} \label{eq-meth-ode} \\
+\frac{d[amet]}{dt} &= v_{MAT} - v_{GNMT} - v_{METH} \nonumber \\
+\frac{d[ahyc]}{dt} &= v_{GNMT} + v_{METH} - v_{AHC} \nonumber \\
+\frac{d[hyc-L]}{dt} &= v_{AHC} - v_{CBS} - v_{MS} - v_{BHMT} \nonumber \\
+\frac{d[5mthf]}{dt} &= v_{MTHFR} - v_{MS} \nonumber 
+\end{align}
 
@@
 These steady states were then used to generate simulated measurements using the
-measurement model described below.
+measurement model described below in @sec-methionine-measurement-model.
 
-\begin{equation}
-a = b \label{eq-meth-ode}
-\end{equation}
-
-### Measurement model
+### Measurement model {#sec-methionine-measurement-model}
 
@@
 
-We constructed inputs in Maud’s format for each of the analysed datasets, based on the scenario that the true kinetic model was known except for parameter values, which needed to be inferred from the training data and priors.
-These inputs can be found at XXXX.
+We constructed inputs in Maud’s format for each of the analysed datasets, based
+on the scenario that the true kinetic model was known except for parameter
+values, which needed to be inferred from the training data and priors. These
+inputs can be found at <https://github.com/biosustain/Methionine_model/tree/main/data>.
 
-The prior distributions and corresponding true parameter values used in our case
-study are shown in [Supplementary]. The first two columns show the 1% and 99%
-quantiles of each marginal prior distribution. True parameter value are shown
-in column three, and the last column shows the z-score on log scale of the true
-parameter value according the marginal prior distribution. As seen in table
-2 [REFERENCE] there are 7 parameters for which the true value is outside the
-1%-99% range. This is desirable, making the case study more realistic, because
-extreme deviance from the prior distribution is likely to occur in practice due
-to in vivo to in vitro measurement differences.
+The prior distributions and corresponding true parameter values used in our
+case study are shown in supplementary materials section [REFERENCE]. The first
+two columns show the 1% and 99% quantiles of each marginal prior distribution.
+True parameter value are shown in column three, and the last column shows the
+z-score on log scale of the true parameter value according the marginal prior
+distribution. As can be seen from the table, there are 7 parameters for which
+the true value is outside the 1%-99% range. This is desirable, making the case
+study more realistic, because extreme deviance from the prior distribution is
+likely to occur in practice due to in vivo to in vitro measurement differences.
 
@@
 important combinations, scales and regions of sensitivity being difficult to
-ascertain in advance @gutenkunst_2007, @poirier_revising_1998.
+ascertain in advance [@gutenkunst_2007; @poirier_revising_1998].
 
@@
 correlated posterior space would also be easier to summarise. We address this
-question below in section XXX.
+question below in @sec-laplace.
 
@@
 XXX ESS/hour. Generalising from this result, we conclude that it is feasible
-to use this method to fit models in the same order of magnitude, but not, for
-example, genome-scale kinetic models. To fit larger models, faster steady state
-solving methods or alternative inference algorithms will be required. Section
-XXX addresses whether Laplace approximation is a suitable candidate.
+to use this method to fit models in the same order of magnitude, but not,
+for example, genome-scale kinetic models. To fit larger models, faster steady
+state solving methods or alternative inference algorithms will be required.
+@sec-laplace addresses whether Laplace approximation is a suitable candidate.
 
-### Comparison with Laplace approximation
+### Comparison with Laplace approximation {#sec-laplace}
 
@@
 configuration, approximate the Hessian and use these quantities to generate
-approximate posterior samples. The results can be found at XXX.
+approximate posterior samples. The results can be found at <https://github.com/biosustain/Methionine_model/tree/main/results>.
 
@@
 and Linux, registered on the Python Package Index as `maud-metabolic-models`,
-documented at <https://maud- metabolic-models.readthedocs.io> and actively
+documented at <https://maud-metabolic-models.readthedocs.io> and actively
 developed and maintained at <https://github.com/biosustain/Maud/>.
@@
 Maud inputs are structured directories, somewhat inspired by the PEtab
-format @SchmiesterSch2021. A Maud input directory must contain a toml
+format [@SchmiesterSch2021]. A Maud input directory must contain a toml
 [@preston-wernertomandgedampradyunTOMLSpecification0rc2020] file called
@@
 
-Each of the terms on the right-hand side of (REFER TO EQUATION) is a function
-of $C$ and $\theta$. This idea is taken from @noor_note_2013. The terms
-usefully gather physically meaningful and conceptually distinct factors
+Each of the terms on the right-hand side of \eqref{eq-decomposition} is a
+function of $C$ and $\theta$. This idea is taken from @noor_note_2013. The
+terms usefully gather physically meaningful and conceptually distinct factors
 contributing to reaction fluxes. $Enzyme$ captures the effect of enzyme
@@
 metabolite concentrations and fluxes, which are derived from the values of
-other parameters by solving the steady state problem. See Table 1 (REFERENCE)
-for a description of all these parameters and their dimensions. Note that some
-quantities in Maud’s kinetic model are not treated as parameters: for example,
-temperatures, compartment volumes and the formation energy of water. Maud
-treats these quantities as if they were known precisely: they can be configured
-by the user or default values can be used. Although in practice there can be
-considerable uncertainty regarding these quantities, we chose to disregard this
-uncertainty in the interest of simplicity.
+other parameters by solving the steady state problem. See table (REFERENCE) in
+this paper's supplementary materials for a description of all these parameters
+and their dimensions. Note that some quantities in Maud’s kinetic model are
+not treated as parameters: for example, temperatures, compartment volumes and
+the formation energy of water. Maud treats these quantities as if they were
+known precisely: they can be configured by the user or default values can be
+used. Although in practice there can be considerable uncertainty regarding
+these quantities, we chose to disregard this uncertainty in the interest of
+simplicity.
 
@@
 
-\begin{equation}
-y_i^{conc} \sim LN(\ln{\hat{y}_i^{conc}}, \sigma_i^{conc})\label{eq-yconc}
-\end{equation}
-
-\begin{equation}
-y_i^{enz} \sim LN(\ln{\hat{y}_i^{enz}}, \sigma_i^{enz})\label{eq-yenz}
-\end{equation}
+\begin{align}
+y_i^{conc} &\sim LN(\ln{\hat{y}_i^{conc}}, \sigma_i^{conc})\label{eq-yconc} \\
+y_i^{enz} &\sim LN(\ln{\hat{y}_i^{enz}}, \sigma_i^{enz})\label{eq-yenz}
+\end{align}
 
@@
 link function (the natural logarithm $\ln$) and known standard deviation
-$ \sigma_{𝑐𝑜𝑛𝑐}$. The use of this measurement model is motivated by the
+$\sigma_{𝑐𝑜𝑛𝑐}$. The use of this measurement model is motivated by the
 consideration that concentrations are constrained to be non-negative, so the
@@
 double counting and requires non-trivial upstream modelling, as intracellular
-fluxes typically cannot be measured directly, but must be inferred from isotope
-labelling experiments using metabolic flux analysis (see REFERENCE). Ideally
+fluxes typically cannot be measured directly. Instead, fluxes are typically
+inferred from isotope labelling experiments using metabolic flux analysis:
+see @daiUnderstandingMetabolismFlux2017 for more about this method. Ideally
 Maud’s measurement model for fluxes would extend from fluxes to the results of
@@
 that MCMC samples really do approximate the target probability
-distribution: see@vehtariRankNormalizationFoldingLocalization2021 and
+distribution: see @vehtariRankNormalizationFoldingLocalization2021 and
 @taltsValidatingBayesianInference2018 for discussion of this point. In

### 2023-10-27 paper/main.tex
Update paper

@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable, frame hidden, interior hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, enhanced]}{\end{tcolorbox}}\fi
 
@@
 Finally, kinetic rate laws can be represented in Bayesian data
-generation models with arbitrarily high fidelity. See (REFERENCES) for
-more about Bayesian inference and (REFERENCE) for a discussion of
+generation models with arbitrarily high fidelity. See
+\citet{gelmanBayesianDataAnalysis2020a} for more about Bayesian
+inference and \citet{gelmanBayesianWorkflow2020} for a discussion of
 practical Bayesian workflow.
@@
 The simulated dataset and underlying kinetic model that we used for our
-analysis can be found in {[}\#Supplementary Info{]}.
+analysis can be found at
+\url{https://github.com/biosustain/Methionine_model/tree/main/data/methionine}
+and is described in supplementary information section {[}REFERENCE{]}.
 
@@
 Equation\,\eqref{eq-meth-ode}.
+
+\begin{align}
+\frac{d[met]}{dt} &= v_{Influx} - v_{PROT} - v_{MAT} +v_{MS} + v_{BHMT} \label{eq-meth-ode} \\
+\frac{d[amet]}{dt} &= v_{MAT} - v_{GNMT} - v_{METH} \nonumber \\
+\frac{d[ahyc]}{dt} &= v_{GNMT} + v_{METH} - v_{AHC} \nonumber \\
+\frac{d[hyc-L]}{dt} &= v_{AHC} - v_{CBS} - v_{MS} - v_{BHMT} \nonumber \\
+\frac{d[5mthf]}{dt} &= v_{MTHFR} - v_{MS} \nonumber 
+\end{align}
 
@@
 as a starting point. These steady states were then used to generate
-simulated measurements using the measurement model described below.
+simulated measurements using the measurement model described below in
+Section~\ref{sec-methionine-measurement-model}.
 
-\begin{equation}
-a = b \label{eq-meth-ode}
-\end{equation}
-
-\hypertarget{measurement-model}{%
-\subsubsection{Measurement model}\label{measurement-model}}
+\hypertarget{sec-methionine-measurement-model}{%
+\subsubsection{Measurement
+model}\label{sec-methionine-measurement-model}}
 
@@
 except for parameter values, which needed to be inferred from the
-training data and priors. These inputs can be found at XXXX.
+training data and priors. These inputs can be found at
+\url{https://github.com/biosustain/Methionine_model/tree/main/data}.
 
 The prior distributions and corresponding true parameter values used in
-our case study are shown in {[}Supplementary{]}. The first two columns
-show the 1\% and 99\% quantiles of each marginal prior distribution.
-True parameter value are shown in column three, and the last column
-shows the z-score on log scale of the true parameter value according the
-marginal prior distribution. As seen in table 2 {[}REFERENCE{]} there
-are 7 parameters for which the true value is outside the 1\%-99\% range.
-This is desirable, making the case study more realistic, because extreme
-deviance from the prior distribution is likely to occur in practice due
-to in vivo to in vitro measurement differences.
+our case study are shown in supplementary materials section
+{[}REFERENCE{]}. The first two columns show the 1\% and 99\% quantiles
+of each marginal prior distribution. True parameter value are shown in
+column three, and the last column shows the z-score on log scale of the
+true parameter value according the marginal prior distribution. As can
+be seen from the table, there are 7 parameters for which the true value
+is outside the 1\%-99\% range. This is desirable, making the case study
+more realistic, because extreme deviance from the prior distribution is
+likely to occur in practice due to in vivo to in vitro measurement
+differences.
 
@@
 important combinations, scales and regions of sensitivity being
-difficult to ascertain in advance \citet{gutenkunst_2007},
-\citet{poirier_revising_1998}.
+difficult to ascertain in advance
+\citep{gutenkunst_2007, poirier_revising_1998}.
 
@@
 correct for. A linearly correlated posterior space would also be easier
-to summarise. We address this question below in section XXX.
+to summarise. We address this question below in
+Section~\ref{sec-laplace}.
 
@@
 kinetic models. To fit larger models, faster steady state solving
-methods or alternative inference algorithms will be required. Section
-XXX addresses whether Laplace approximation is a suitable candidate.
+methods or alternative inference algorithms will be required.
+Section~\ref{sec-laplace} addresses whether Laplace approximation is a
+suitable candidate.
 
-\hypertarget{comparison-with-laplace-approximation}{%
+\hypertarget{sec-laplace}{%
 \subsubsection{Comparison with Laplace
-approximation}\label{comparison-with-laplace-approximation}}
+approximation}\label{sec-laplace}}
 
@@
 these quantities to generate approximate posterior samples. The results
-can be found at XXX.
+can be found at
+\url{https://github.com/biosustain/Methionine_model/tree/main/results}.
 
@@
 macOS and Linux, registered on the Python Package Index as
-\texttt{maud-metabolic-models}, documented at \textless https://maud-
-metabolic-models.readthedocs.io\textgreater{} and actively developed and
-maintained at \url{https://github.com/biosustain/Maud/}.
+\texttt{maud-metabolic-models}, documented at
+\url{https://maud-metabolic-models.readthedocs.io} and actively
+developed and maintained at \url{https://github.com/biosustain/Maud/}.
 
@@
 Maud inputs are structured directories, somewhat inspired by the PEtab
-format \citet{SchmiesterSch2021}. A Maud input directory must contain a
+format \citep{SchmiesterSch2021}. A Maud input directory must contain a
 toml \citep{preston-wernertomandgedampradyunTOMLSpecification0rc2020}
@@
 
-Each of the terms on the right-hand side of (REFER TO EQUATION) is a
-function of \(C\) and \(\theta\). This idea is taken from
+Each of the terms on the right-hand side of \eqref{eq-decomposition} is
+a function of \(C\) and \(\theta\). This idea is taken from
 \citet{noor_note_2013}. The terms usefully gather physically meaningful
@@
 derived from the values of other parameters by solving the steady state
-problem. See Table 1 (REFERENCE) for a description of all these
-parameters and their dimensions. Note that some quantities in Maud's
-kinetic model are not treated as parameters: for example, temperatures,
-compartment volumes and the formation energy of water. Maud treats these
-quantities as if they were known precisely: they can be configured by
-the user or default values can be used. Although in practice there can
-be considerable uncertainty regarding these quantities, we chose to
-disregard this uncertainty in the interest of simplicity.
+problem. See table (REFERENCE) in this paper's supplementary materials
+for a description of all these parameters and their dimensions. Note
+that some quantities in Maud's kinetic model are not treated as
+parameters: for example, temperatures, compartment volumes and the
+formation energy of water. Maud treats these quantities as if they were
+known precisely: they can be configured by the user or default values
+can be used. Although in practice there can be considerable uncertainty
+regarding these quantities, we chose to disregard this uncertainty in
+the interest of simplicity.
 
@@
 
-\hypertarget{measurement-model-1}{%
-\subsubsection{Measurement model}\label{measurement-model-1}}
+\hypertarget{measurement-model}{%
+\subsubsection{Measurement model}\label{measurement-model}}
 
@@
 link function (the natural logarithm \(\ln\)) and known standard
-deviation \$ \sigma\_\{𝑐𝑜𝑛𝑐\}\$. The use of this measurement model is
+deviation \(\sigma_{𝑐𝑜𝑛𝑐}\). The use of this measurement model is
 motivated by the consideration that concentrations are constrained to be
@@
 modelling, as intracellular fluxes typically cannot be measured
-directly, but must be inferred from isotope labelling experiments using
-metabolic flux analysis (see REFERENCE). Ideally Maud's measurement
-model for fluxes would extend from fluxes to the results of potential
-labelling experiments, thereby removing the need for upstream analysis
-and avoiding any double counting. This option has not yet been pursued,
-again for the sake of simplicity.
+directly. Instead, fluxes are typically inferred from isotope labelling
+experiments using metabolic flux analysis: see
+\citet{daiUnderstandingMetabolismFlux2017} for more about this method.
+Ideally Maud's measurement model for fluxes would extend from fluxes to
+the results of potential labelling experiments, thereby removing the
+need for upstream analysis and avoiding any double counting. This option
+has not yet been pursued, again for the sake of simplicity.
 
@@
 primarily because there exist many methods for verifying that MCMC
-samples really do approximate the target probability distribution:
-see@vehtariRankNormalizationFoldingLocalization2021 and
+samples really do approximate the target probability distribution: see
+\citet{vehtariRankNormalizationFoldingLocalization2021} and
 \citet{taltsValidatingBayesianInference2018} for discussion of this

### 2023-10-27 paper/main.qmd
Add captions

@@
 
-![CAPTION - FILL THIS IN!](./figures/methionine-reactions.png){#fig-methionine-reactions}
+::: {#fig-methionine-reactions layout-ncol=1}
+
+![](./figures/methionine-reactions.png)
+
+The methionine cycle as modelled, with the solid black lines representing the
+reactions, the green lines representing allosteric interaction, and the red
+lines representing allosteric inhibition. The bold fonts are the reaction names
+and the regular font represents the metabolites.
+
+:::
 
@@
 
-![CAPTION: FILL THIS IN!](./figures/laplace.png){#fig-laplace}
+::: {#fig-laplace layout-ncol=1}
+
+![](./figures/laplace.png)
+
+CAPTION: FILL THIS IN!
+
+:::
 
@@
 
-![CAPTION: FILL THIS IN!](./figures/missing.png){#fig-missing}
+::: {#fig-missing layout-ncol=1}
+
+![](./figures/missing.png)
+
+Results of removing concentration measurements for the metabolite $ahcys_c$
+from our case study dataset.  (**Top**) A comparison of metabolite concentration
+residuals between the full measurement dataset (blue-green) and the missing-data
+dataset (grey), displayed on natural logarithmic scale. The missing-data
+model was unable to predict the withheld $ahcys_c$ concentrations.  (**Bottom
+Left**) The marginal posterior distribution for the Michaelis constant
+$K_m^{AHC1,ahcys_c}$ in each model, alongside the true parameter value used to
+generate both datasets. The true value is recovered by the complete-data model
+but not by the missing-data model.  (**Bottom Right**) The distribution of total
+log-likelihood for out-of-sample flux measurements in both models. There is a
+significant overlap between the two distributions, suggesting that removing the
+$ahcys_c$ measurement did not cause catastrophic prediction failure. However,
+overall the complete-data model tended to make better predictions.
+
+:::
 
@@
 
-![CAPTION: FILL THIS IN!](./figures/decomposition.png){#fig-decomposition}
+::: {#fig-decomposition layout-ncol=1}
+
+![](./figures/decomposition.png)
+
+Illustration of how analysing a system with Maud can yield actionable insights
+about the underlying metabolic network.  (**Top Left**) A schematic of the
+regulatory interactions associated with the enzyme $GNMT1$. Dashed green
+lines represents allosteric activation, dashed red lines indicate allosteric
+inhibition and solid red lines represent competitive inhibition. (**Bottom
+Left**) Comparison of marginal posterior distributions for $GNMT$ flux in
+conditions 1 and 2. (**Right**) Log-scale ratios of the regulatory elements
+defined in equation \eqref{eq-decomposition}. Note that the $Reversibility$
+and $k_cat$ components are excluded: this is because this reaction was modelled
+as irreversible, and $k_{cat}$ was modelled as constant across conditions.
+These plots identify why flux in condition 2 is higher than in condition 1: the
+flux increase is due to allostery and saturation with no control from enzyme
+concentration changes.
+
+:::
 

### 2023-10-27 paper/main.tex
Add captions

@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, breakable, enhanced, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 
-{\centering \includegraphics{./figures/methionine-reactions.png}
+\begin{minipage}[t]{\linewidth}
+
+{\centering
+
+\raisebox{-\height}{
+
+\includegraphics{./figures/methionine-reactions.png}
 
@@
 
-\caption{\label{fig-methionine-reactions}CAPTION - FILL THIS IN!}
+}
+
+\end{minipage}%
+
+\caption{\label{fig-methionine-reactions}The methionine cycle as
+modelled, with the solid black lines representing the reactions, the
+green lines representing allosteric interaction, and the red lines
+representing allosteric inhibition. The bold fonts are the reaction
+names and the regular font represents the metabolites.}
 
@@
 
-{\centering \includegraphics{./figures/laplace.png}
+\begin{minipage}[t]{\linewidth}
+
+{\centering
+
+\raisebox{-\height}{
+
+\includegraphics{./figures/laplace.png}
 
 }
+
+}
+
+\end{minipage}%
 
@@
 
-{\centering \includegraphics{./figures/missing.png}
+\begin{minipage}[t]{\linewidth}
+
+{\centering
+
+\raisebox{-\height}{
+
+\includegraphics{./figures/missing.png}
 
@@
 
-\caption{\label{fig-missing}CAPTION: FILL THIS IN!}
+}
+
+\end{minipage}%
+
+\caption{\label{fig-missing}Results of removing concentration
+measurements for the metabolite \(ahcys_c\) from our case study dataset.
+(\textbf{Top}) A comparison of metabolite concentration residuals
+between the full measurement dataset (blue-green) and the missing-data
+dataset (grey), displayed on natural logarithmic scale. The missing-data
+model was unable to predict the withheld \(ahcys_c\) concentrations.
+(\textbf{Bottom Left}) The marginal posterior distribution for the
+Michaelis constant \(K_m^{AHC1,ahcys_c}\) in each model, alongside the
+true parameter value used to generate both datasets. The true value is
+recovered by the complete-data model but not by the missing-data model.
+(\textbf{Bottom Right}) The distribution of total log-likelihood for
+out-of-sample flux measurements in both models. There is a significant
+overlap between the two distributions, suggesting that removing the
+\(ahcys_c\) measurement did not cause catastrophic prediction failure.
+However, overall the complete-data model tended to make better
+predictions.}
 
@@
 
-{\centering \includegraphics{./figures/decomposition.png}
+\begin{minipage}[t]{\linewidth}
+
+{\centering
+
+\raisebox{-\height}{
+
+\includegraphics{./figures/decomposition.png}
 
@@
 
-\caption{\label{fig-decomposition}CAPTION: FILL THIS IN!}
+}
+
+\end{minipage}%
+
+\caption{\label{fig-decomposition}Illustration of how analysing a system
+with Maud can yield actionable insights about the underlying metabolic
+network. (\textbf{Top Left}) A schematic of the regulatory interactions
+associated with the enzyme \(GNMT1\). Dashed green lines represents
+allosteric activation, dashed red lines indicate allosteric inhibition
+and solid red lines represent competitive inhibition. (\textbf{Bottom
+Left}) Comparison of marginal posterior distributions for \(GNMT\) flux
+in conditions 1 and 2. (\textbf{Right}) Log-scale ratios of the
+regulatory elements defined in equation \eqref{eq-decomposition}. Note
+that the \(Reversibility\) and \(k_cat\) components are excluded: this
+is because this reaction was modelled as irreversible, and \(k_{cat}\)
+was modelled as constant across conditions. These plots identify why
+flux in condition 2 is higher than in condition 1: the flux increase is
+due to allostery and saturation with no control from enzyme
+concentration changes.}
 

### 2023-10-27 paper/main.qmd
Small paper fixes.

@@
 univariate distributions. The two distributions were significantly different
-with a p-value indistinguishable from 0.
+with a p-value indistinguishable from zero.
 
@@
 The missing measurements did affect Maud’s ability to infer parameter values
-correctly. As we can see in the [#lower left plot of @fig-missing], the model
-with full metabolomics learned the true value for the displayed dissociation
-constant, despite this value being far from the mean of the corresponding
-marginal prior distribution, whereas the model with the missing measurements
-stayed in the neighbourhood of the prior.
+correctly. The lower left plot of @fig-missing shows that the model with full
+metabolomics learned the true value for the displayed dissociation constant,
+despite this value being far from the mean of the corresponding marginal prior
+distribution. In contrast, the model with missing measurements stayed in the
+neighbourhood of the prior.
 

### 2023-10-27 paper/main.tex
Small paper fixes.

@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, breakable, enhanced, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, sharp corners, boxrule=0pt, breakable, interior hidden, enhanced]}{\end{tcolorbox}}\fi
 
@@
 univariate distributions. The two distributions were significantly
-different with a p-value indistinguishable from 0.
+different with a p-value indistinguishable from zero.
 
@@
 The missing measurements did affect Maud's ability to infer parameter
-values correctly. As we can see in the \#lower left plot
-of~\ref{fig-missing}, the model with full metabolomics learned the true
-value for the displayed dissociation constant, despite this value being
-far from the mean of the corresponding marginal prior distribution,
-whereas the model with the missing measurements stayed in the
-neighbourhood of the prior.
+values correctly. The lower left plot of Figure~\ref{fig-missing} shows
+that the model with full metabolomics learned the true value for the
+displayed dissociation constant, despite this value being far from the
+mean of the corresponding marginal prior distribution. In contrast, the
+model with missing measurements stayed in the neighbourhood of the
+prior.
 

### 2023-10-27 paper/main.qmd
Update paper

@@
 Generating an artificial dataset required a specification of the true
-measurement error; we specified a standard deviation of 0.1 on logarithmic
-scale for enzyme and metabolite concentration measurements, corresponding to
-approximately 10% measurement error and a standard deviation for each reaction
-measurement approximately 10% of the simulated value.
+measurement error. For enzyme and metabolite concentration measurements
+we specified a standard deviation of 0.1 on natural logarithmic scale,
+corresponding to approximately 10% measurement error. For each reaction
+measurement a measurement standard deviation of approximately 10% of the
+simulated value.
 
@@
 
-To investigate how the flux of GNMT between conditions [X] and [Y] changes
-we compared the regulatory contributions in the posterior distribution.
-Presented in [@fig-decomposition top left] is the regulatory description of
-GNMT, an irreversible enzyme that is homotropically activated by its substrate,
-competitively inhibited by its product and heterotropically inhibited. The
-diverse regulation makes it the ideal test case to elucidate regulatory changes.
+To demonstrate how Maud's output can be used to yield useful metabolic insights
+we used the results of our case study to explain why the flux of the enzyme
+$GNMT$ is higher in condition 1 than in condition 2. GNMT is an irreversible
+enzyme that is homotropically activated by its substrate, competitively
+inhibited by its product and heterotropically inhibited. The diverse regulation
+makes it the ideal test case to elucidate regulatory changes.
 
-Here we found that saturation and allosteric effects were the main drivers
-of regulation between the two conditions. @fig-decomposition shows the distributions
-of the regulatory components as the log-ratio of each component with [X] in
-the numerator and [Y] in the denominator, positive values suggest that this
-component was increased in experiment [X] relative to [Y], with 0 indicating no
-difference. Naturally, this presents the ideal statistical test when evaluating
-the regulation, with the probability of regulation given by the relative
-proportion of samples above and below the 0 point.
+@fig-decomposition shows the regulatory description of GNMT, according to the
+results our main case study analysis. Each curve shows the marginal posterior
+distribution of the ratio of the corresponding regulatory component in condition
+1 compared with condition 2. A positive value indicates that the component
+was increased in condition 1 relative to condition 2, with 0 indicating no
+difference. The probability, according to our model, of the component acting in
+each direction is given by the relative area under the curve on each side of the
+0 point.
 
-We demonstrate that Maud can accurately reproduce the regulatory effect between
-the two conditions by comparing it to the ground truth. As the true values are
-well within our probability mass of the posterior we can say that Maud is able
-to accurately reproduce the regulatory behaviour of metabolic networks and also
-provides a tool for statistical inference of the regulatory behaviour.
+Our model correctly inferred that saturation and allosteric effects were the
+main drivers of regulation between the two conditions in this case, with the
+curves for each component aligning with the ground truth shown in red.
 
-Here we have demonstrated a key feature in Maud, that is, direct inference on
-any parameter that arises from the data generating process. More succinctly,
-we show the benefit of error propagation that Bayesian inference provides as
-researchers can directly use quantiles for hypothesis testing.
-
-Our tool allows inference directly on the joint posterior distribution rather
-than the marginal parameter. The implication is that inference will be better
-than that of the marginal distributions due to correlated parameter values, if
-the data generating process can be directly modelled.
+Importantly, this form of analysis takes into account all modelled sources of
+uncertainty, including uncertainty about the true values of the flux in each
+condition, and propagates this uncertainty into the final conclusion. Our result
+shows that Maud could be used in this realistic case not only to provide a user
+with an explanation for an observed difference in fluxes, but also a reasonable
+judgement as to the explanation's robustness.
 

### 2023-10-27 paper/main.tex
Rerender paper

@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, sharp corners, boxrule=0pt, breakable, interior hidden, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, breakable, sharp corners, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, enhanced]}{\end{tcolorbox}}\fi
 
@@
 Generating an artificial dataset required a specification of the true
-measurement error; we specified a standard deviation of 0.1 on
-logarithmic scale for enzyme and metabolite concentration measurements,
-corresponding to approximately 10\% measurement error and a standard
-deviation for each reaction measurement approximately 10\% of the
-simulated value.
+measurement error. For enzyme and metabolite concentration measurements
+we specified a standard deviation of 0.1 on natural logarithmic scale,
+corresponding to approximately 10\% measurement error. For each reaction
+measurement a measurement standard deviation of approximately 10\% of
+the simulated value.
 
@@
 
-To investigate how the flux of GNMT between conditions {[}X{]} and
-{[}Y{]} changes we compared the regulatory contributions in the
-posterior distribution. Presented in Figure~\ref{fig-decomposition} is
-the regulatory description of GNMT, an irreversible enzyme that is
-homotropically activated by its substrate, competitively inhibited by
-its product and heterotropically inhibited. The diverse regulation makes
-it the ideal test case to elucidate regulatory changes.
-
-Here we found that saturation and allosteric effects were the main
-drivers of regulation between the two conditions.
-Figure~\ref{fig-decomposition} shows the distributions of the regulatory
-components as the log-ratio of each component with {[}X{]} in the
-numerator and {[}Y{]} in the denominator, positive values suggest that
-this component was increased in experiment {[}X{]} relative to {[}Y{]},
-with 0 indicating no difference. Naturally, this presents the ideal
-statistical test when evaluating the regulation, with the probability of
-regulation given by the relative proportion of samples above and below
+To demonstrate how Maud's output can be used to yield useful metabolic
+insights we used the results of our case study to explain why the flux
+of the enzyme \(GNMT\) is higher in condition 1 than in condition 2.
+GNMT is an irreversible enzyme that is homotropically activated by its
+substrate, competitively inhibited by its product and heterotropically
+inhibited. The diverse regulation makes it the ideal test case to
+elucidate regulatory changes.
+
+Figure~\ref{fig-decomposition} shows the regulatory description of GNMT,
+according to the results our main case study analysis. Each curve shows
+the marginal posterior distribution of the ratio of the corresponding
+regulatory component in condition 1 compared with condition 2. A
+positive value indicates that the component was increased in condition 1
+relative to condition 2, with 0 indicating no difference. The
+probability, according to our model, of the component acting in each
+direction is given by the relative area under the curve on each side of
 the 0 point.
 
-We demonstrate that Maud can accurately reproduce the regulatory effect
-between the two conditions by comparing it to the ground truth. As the
-true values are well within our probability mass of the posterior we can
-say that Maud is able to accurately reproduce the regulatory behaviour
-of metabolic networks and also provides a tool for statistical inference
-of the regulatory behaviour.
-
-Here we have demonstrated a key feature in Maud, that is, direct
-inference on any parameter that arises from the data generating process.
-More succinctly, we show the benefit of error propagation that Bayesian
-inference provides as researchers can directly use quantiles for
-hypothesis testing.
-
-Our tool allows inference directly on the joint posterior distribution
-rather than the marginal parameter. The implication is that inference
-will be better than that of the marginal distributions due to correlated
-parameter values, if the data generating process can be directly
-modelled.
+Our model correctly inferred that saturation and allosteric effects were
+the main drivers of regulation between the two conditions in this case,
+with the curves for each component aligning with the ground truth shown
+in red.
+
+Importantly, this form of analysis takes into account all modelled
+sources of uncertainty, including uncertainty about the true values of
+the flux in each condition, and propagates this uncertainty into the
+final conclusion. Our result shows that Maud could be used in this
+realistic case not only to provide a user with an explanation for an
+observed difference in fluxes, but also a reasonable judgement as to the
+explanation's robustness.
 

### 2023-10-27 paper/main.qmd
Update paper and supplementary information

@@
 ---
-title: "Hierarchical Bayesian regression modelling predicts biochemical rection energies and facilitates quantitative modelling of cell metabolism"
-toc: true
+title: "Bayesian regression modelling facilitates quantitative modelling of cell metabolism"
+keywords: Bayesian inference, kinetic models of cell metabolism 
 abstract: >
@@
     email: tedgro@biosustain.dtu.dk
-    affiliation: DTU Biosustain
+    affiliations: 
+    - id: biosustain
+      name: DTU Biosustain
   - name: "Nicholas Luke Cowie"
-    affiliation: DTU Biosustain
+    affiliations: 
+    - ref: biosustain
   - name: "Lars Keld Nielsen"
-    affiliation: DTU Biosustain
+    affiliations: 
+    - ref: biosustain
 bibliography: bibliography.bib
@@
     keep-tex: true
-  html:
+    journal:
+      id: asbcd6  #This is an identifier for the target journal: from https://pubs.acs.org/, the identifier is the part of the URL after https://pubs.acs.org/journal/, e.g.
+for J.
+Org.
+Chem.
+it would be joceah.
+      type: article  #article, note, communication, review, letter, or perspectivee
+      layout: traditional #traditional or twocolumn, twocolumn currently broken for longtables
+      hide-email: false # if enabled, hides email addresses (and fax/phone)
+      abbreviations: IR,NMR,UV
+  acs-html:
    include-in-header: mathjax.html
+   embed-resources: true
   docx: default

### 2023-10-27 paper/main.tex
Update paper and supplementary information

@@
 %
-\documentclass[journal=,manuscript=]{achemso}
+\documentclass[journal=asbcd6,manuscript=article,layout=traditional]{achemso}
 \usepackage[version=3]{mhchem}
@@
 \hypersetup{
-  pdftitle={Hierarchical Bayesian regression modelling predicts biochemical rection energies and facilitates quantitative modelling of cell metabolism},
+  pdftitle={Bayesian regression modelling facilitates quantitative modelling of cell metabolism},
   pdfauthor={Teddy Groves; Nicholas Luke Cowie; Lars Keld Nielsen},
+  pdfkeywords={Bayesian inference, kinetic models of cell metabolism},
   colorlinks=true,
@@
 
-\title[]{Hierarchical Bayesian regression modelling predicts biochemical
-rection energies and facilitates quantitative modelling of cell
-metabolism}
+\keywords{Bayesian inference, kinetic models of cell metabolism}
+
+\title[]{Bayesian regression modelling facilitates quantitative
+modelling of cell metabolism}
 \makeatletter
@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, breakable, sharp corners, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, enhanced]}{\end{tcolorbox}}\fi
-
-\renewcommand*\contentsname{Table of contents}
-{
-\hypersetup{linkcolor=}
-\setcounter{tocdepth}{3}
-\tableofcontents
-}
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, interior hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, sharp corners, frame hidden]}{\end{tcolorbox}}\fi
+
 \hypertarget{introduction}{%

### 2023-10-27 paper/supplementary.qmd
Update paper and supplementary information

---
title: Supporting information
bibliography: bibliography.bib
format:
  html:
    embed-resources: true
    include-in-header: mathjax.html
  pdf: default
  docx: default
---

# Maud's kinetic model

## Parameters

@tbl-params shows all of Maud's unknown parameters along with their dimensions

Note that Maud’s metabolic model includes some quantities that are not treated
as parameters in its statistical model, including temperatures, compartment
volumes and the formation energy of water. Maud treats these quantities as
if they were known precisely: they can be configured by the user or default
values can be used. Although in practice there can be considerable uncertainty
regarding these quantities, we chose to disregard this uncertainty in the
interest of simplicity.

| Parameter | Modelled quantity | Dimensions | 
| --------- | ----------------- | ---------- |
| $\Delta_fG$| Formation energy | metabolites |
| $k_M$ | Michaelis Menten constants | Substrates of all enzyme/reactions and products of reversible enzyme/reactions |
| $k_I$ | Inhibition constants | Inhibiting metabolite/compartments of enzyme/reactions exhibiting competitive inhibition |
| $k_{cat}$ | Rate constants | Enzyme/reactions |
| $L_0$ | Transfer constants | Allosteric interactions |
| $e_T$ | T dissociation constants | Modifying metabolites of allosteric inhibitions |
| $e_R$ | R dissociation constants | Modifying metabolites of allosteric activations |
| $k_{cat\, pme}$ | Rate constants of phosphorylation modifying enzymes | Phosphorylation modifying enzymes |
| $v_{drain}$ | Drain fluxes | Drains, experiments |
| $Enzyme$ | Enzyme concentrations | Enzymes, experiments |
| $C_{unbalanced}$ | Unbalanced metabolite/compartment concentrations | Unbalanced metabolite/compartments, experiments |
| $C_{pme}$ | Phosphorylation modifying enzyme concentrations | Phosphorylation modifying enzymes, experiments|
| $\psi$ | Membrane potentials | Experiments |

: Parameters of Maud's statistical model {#tbl-params}

Solving the steady state problem for a given set of parameters in an experiment
yields a vector $C_{balanced}$ of balanced metabolite concentrations. These
are combined with the balanced metabolite concentrations $C_{unbalanced}$ to
produce a vector $C_{mic}$ with a concentration for each metabolite/compartment
combination.

$\Delta_fG$ parameters can optionally be fixed; this can be useful for
computational purposes, as for example to avoid estimating the formation energy
of a metabolite about which there is no available information due to it only
participating in irreversible reactions.

## Rate equations

As discussed in the main text, Maud’s kinetic model decomposes into factors
contributing to the flux in a metabolic network in an experiment as shown in
equation \eqref{eq-decomposition}. For succinctness, and since Maud's model
assumes that there are no interactions between experiments, we omit any notation
referring to experiments below. We also omit any reference to the network's
drain reactions: these are modelled as being exactly determined by the values of
the parameter vector $v_{drain}$.

\begin{equation}
F(C;\theta) = Enzyme\cdot k_{cat}\cdot Reversibility \cdot Saturation \cdot Allostery \label{eq-decomposition}
\end{equation}

The term $Enzyme$ in equation \eqref{eq-decomposition} is a vector of
non-negative real numbers representing the concentration of the enzyme
catalysing each reaction.

The term $k_{cat}$ in equation \eqref{eq-decomposition} is a vector of
non-negative real numbers representing the  amount of flux carried per unit of
saturated enzyme.

The term $Reversibility$ in equation \eqref{eq-decomposition} is a vector of
real numbers capturing the impact of thermodynamic effects on the reaction's
flux, as shown in equation \eqref{eq-reversibility}.

\begin{align} 
  Reversibility &= 1 - \exp(\frac{\Delta_{r}G + RT \cdot S^T \ln(C_{mic})}{RT}) \label{eq-reversibility} \\ 
  \Delta_{r}G &= S^{T}\Delta_{f}G + n F \psi \nonumber 
\end{align}

The terms in \eqref{eq-reversibility} have the following meanings:

- $T$ is the temperature in Kelvin (a number), 
- $R$ is the gas constant (a number), 
- $\Delta_rG$ is a vector representing the Gibbs free energy change of each reaction in standard conditions, 
- $\Delta_fG$ is a vector representing the standard condition Gibbs free energy change of each metabolite's formation reaction, or in other words each metabolite's 'formation energy'. 
- $n$ is a vector representing the number of charges transported by each reaction. 
- $F$ is the Faraday constant (a number) 
- $\psi$ is a vector representing each reaction's membrane potential (these numbers only matter for reactions that transport non-zero charge)

Note that, for reactions with zero transported charge, the thermodynamic effect
on each reaction is derived from metabolite formation energies. This formulation
is helpful because, provided that all reactions' rates are calculated from the
same formation energies, they are guaranteed to be thermodynamically consistent.

The term $n$ accounts for both the charge and the directionality. For instance,
a reaction that exports 2 protons to the extracellular space in the forward
direction would have -2 charge. If a negatively charged molecule like acetate is
exported in the forward direction, $n$ would be 1.

Note that this way of modelling the effect of transported charge does not take
into account that the concentration gradient used by the transport is that of
the dissociated molecules. Thus, this expression is only correct for ions whose
concentration can be expressed in the model only in the charged form; e.g.,
protons, $K^+$, $Na^+$, $Cl^-$, etc.

The term $Saturation$ in equation \eqref{eq-decomposition} is a vector of
non-negative real numbers representing, for each reaction, the fraction of
enzyme that is saturated, i.e. bound to one of the reaction's substrates.
To describe saturation we use equation \eqref{eq-saturation}, which is taken
from @liebermeister_modular_2010. Additionally, this term captures competitive
inhibition: as competitive inhibitor concentration increases, the saturation
denominator increases, effectively decreasing the saturation of the substrate on
the total enzyme pool. Conversely, as the substrate concentration increases this
term approaches 1.

\begin{align} 
Saturation_r &= a \cdot \text{free enzyme ratio}\label{eq-saturation} \\ 
a &= \prod_{\text{s substrate}}\frac{C_{mic}^s}{k_{M}^{rs}} \nonumber \\ 
\text{free enzyme ratio} &= \begin{cases} 
  \prod_{\text{s sustrate}} (1 + \frac{C_{mic}^s}{k_{M}^{rs}})^{S_sr} 
  + \sum_{\text{c inhibitor}}\frac{C_{mic}^c}{k_I^{rc}} & r\text{ irreversible} \\ 
  -1 
  + \prod_{\text{s sustrate}} (1 + \frac{C_{mic}^s}{k_{M}^{rs}})^{S_sr} 
     + \sum_{\text{c inhibitor}}\frac{C_{mic}^c}{k_I^{rc}} 
     + \prod_{\text{p product}} (1 + \frac{C_{mic}^p}{k_{M}^{rp}})^{S_pr}  & r\text{ reversible} 
    \end{cases} \nonumber
\end{align}

The term $Allostery$ in equation \eqref{eq-decomposition} is a vector of
non-negative numbers describing the effect of allosteric regulation on each
reaction. Allosteric regulation happens when binding to a certain molecule
changes an enzyme's shape in a way that changes its catalytic behaviour. We
use equation \eqref{eq-allostery} to describe this phenomenon, following the
generalised MWC approach described in @monod_nature_1965, @changeux_2013,
@popova_generalization_1975 and @popova_description_1979.

\begin{align} 
  Allostery_r &= \frac{1}{1 + L_0^r \cdot (\text{free enzyme ratio}_r \cdot \frac{Qtense}{Qrelaxed})^{subunits}}\label{eq-allostery} \\ 
       Qtense &= 1 + \sum_{\text{i inhibitor}} \frac{C_{mic}^i}{e_T^{ri}}\nonumber \\ 
     Qrelaxed &= 1 + \sum_{\text{a activator}} \frac{C_{mic}^a}{e_R^{ra}}\nonumber
\end{align}

The parameter $L_0$ in equation \eqref{eq-decomposition} is called the transfer
constant, and the parameter vectors $e_T$ and $e_R$ are called tense and relaxed
dissociation constants respectively.

Finally, the term $Phosphorylation$ in equation \eqref{eq-decomposition}
captures the important effect whereby enzyme activity is altered due to a
coupled process of phosphorylation and dephosphorylation. This description
achieves a similar behaviour to the MWC formalism for describing allosteric
regulation, but using the rates of phosphorylation and dephosphorylation rather
than concentrations of metabolites.

\begin{align}
Phosphorylation_r &= (\frac{\alpha}{\alpha + \beta})^{subunits} \label{eq-phosphorylation}\\
\alpha &= \sum_{\text{p phosphoylator}} k_{cat\,pme}^{p} \cdot C_{pme}^p \nonumber \\
\beta &= \sum_{\text{d dephosphoylator}} k_{cat\,pme}^{d} \cdot C_{pme}^d \nonumber 
\end{align}

# Methionine case study {#sec-methionine-case-study}

## Dataset generation

## Prior distributions compared with true parameter values

@tbl-case-study-params shows the prior distributions we used for independent
parameters.

| parameter name                      | 1% prior quantile | 99% prior quantile | true value   | prior Z-score of true value |
| ----------------------------------- | ----------------- | ------------------ | ------------ | --------------------------- |
| $e_{𝑅}^{𝐶𝐵𝑆1,𝑎𝑚𝑒𝑡𝑐}$              | 3.430e-06      | 0.002480           | 9.3e-05 | 0.004                       |
| $e_{𝑅}^{𝐺𝑁𝑀𝑇1,𝑎𝑚𝑒𝑡𝑐}$            | 3.000e-05      | 0.002000           | 2.000e-05 | -2.787                     |
| $e_{𝑅}^{𝑀𝐴𝑇3,𝑎𝑚𝑒𝑡𝑐}$              | 1.000e-04      | 0.001000           | 3.170e-04 | 0.003                       |
| $e_{𝑅}^{𝑀𝐴𝑇3,𝑚𝑒𝑡−𝐿𝑐}$             | 4.500e-04      | 0.000800           | 6.000e-04 | 0.000                       |
| $e_{𝑅}^{𝑀𝑇𝐻𝐹𝑅1,𝑎ℎ𝑐𝑦𝑠𝑐}$         | 1.120e-07      | 0.000081           | 2.000e-06 | -0.101                     |
| $e_{𝑇}^{𝐺𝑁𝑀𝑇1,𝑚𝑙𝑡ℎ𝑓𝑐}$           | 1.120e-05      | 0.008050           | 2.290e-04 | -0.136                     |
| $e_{𝑇}^{𝑀𝑇𝐻𝐹𝑅1,𝑎𝑚𝑒𝑡𝑐}$          | 1.120e-07      | 0.000081           | 1.500e-05 | 0.549306                    |
| $k_{𝑐𝑎𝑡}^{𝐴𝐻𝐶1}$                     | 1.200e+02      | 400.000000         | 2.340e+02 | 0.179861                    |
| $k_{𝑐𝑎𝑡}^{𝐵𝐻𝑀𝑇1}$                   | 6.000e+00      | 35.000000          | 1.380e+01 | -0.135                     |
| $k_{𝑐𝑎𝑡}^{𝐶𝐵𝑆1}$                     | 1.000e+01      | 188.000000         | 7.020e+00 | -2.887                     |
| $k_{𝑐𝑎𝑡}^{𝐺𝑁𝑀𝑇1}$                   | 7.000e-01      | 60.000000          | 1.050e+01 | 0.352083                    |
| $k_{𝑐𝑎𝑡}^{𝑀𝐴𝑇1}$                     | 8.200e-02      | 59.100000          | 7.900e+00 | 0.44375                     |
| $k_{𝑐𝑎𝑡}^{𝑀𝐴𝑇3}$             | 5.890e-01      | 424.000000         | 1.990e+01 | 0.080556                    |
| $k_{𝑐𝑎𝑡}^{𝑀𝐸𝑇𝐻−𝐺𝑒𝑛}$ | 4.840e-01      | 349.000000         | 1.160e+00 | -1.209                     |
| $k_{𝑐𝑎𝑡}^{𝑀𝑆1kcatMS1}$                | 1.000e+00      | 3.300000           | 1.770e+00 | -0.091                     |
| $k_{𝑐𝑎𝑡}^{𝑀𝑇𝐻𝐹𝑅1}$       | 1.300e+00      | 4.200000           | 3.170e+00 | 0.183333                    |
| $k_{𝑐𝑎𝑡}^{𝑃𝑅𝑂𝑇1}$                   | 1.590e-01      | 0.222000           | 2.650e-01 | 0.41875                     |
| $k_{𝐼}^{𝐺𝑁𝑀𝑇1,𝑎ℎ𝑐𝑦𝑠𝑐}$           | 2.000e-06      | 0.001400           | 5.300e-05 | 0.010                       |
| $k_{𝐼}^{𝑀𝐴𝑇1,𝑎𝑚𝑒𝑡𝑐}$              | 3.000e-04      | 0.000400           | 3.470e-04 | 0.014                       |
| $k_{𝐼}^{𝑀𝐸𝑇𝐻−𝐺𝑒𝑛,𝑎ℎ𝑐𝑦𝑠𝑐}$     | 1.000e-06      | 0.000030           | 6.000e-06 | 0.021                       |
| $k_{𝑀}^{𝐴𝐻𝐶1,𝑎ℎ𝑐𝑦𝑠𝑐}$             | 5.220e-05      | 0.037600           | 2.320e-05 | -2.050                     |
| $k_{𝑀}^{𝐴𝐻𝐶1,𝑎𝑑𝑛𝑐}$                | 1.670e-07      | 0.000120           | 5.660e-06 | 0.081944                    |
| $k_{𝑀}^{𝐴𝐻𝐶1,ℎ𝑐𝑦𝑠−𝐿𝑐}$            | 1.580e-07      | 0.000114           | 1.060e-05 | 0.318056                    |
| $k_{𝑀}^{𝐵𝐻𝑀𝑇1,ℎ𝑐𝑦𝑠−𝐿𝑐}$          | 1.200e-05      | 0.000032           | 1.980e-05 | 0.049                       |
| $k_{𝑀}^{𝐵𝐻𝑀𝑇1,𝑔𝑙𝑦𝑏𝑐}$            | 4.720e-05      | 0.034000           | 8.460e-03 | 0.659028                    |
| $k_{𝑀}^{𝐶𝐵𝑆1,ℎ𝑐𝑦𝑠−𝐿𝑐}$            | 1.000e-06      | 0.000025           | 4.240e-05 | 3.090                       |
| $k_{𝑀}^{𝐶𝐵𝑆1,𝑠𝑒𝑟−𝐿𝑐}$             | 2.000e-06      | 0.000004           | 2.830e-06 | 0.004                       |
| $k_{𝑀}^{𝐺𝑁𝑀𝑇1,𝑎𝑚𝑒𝑡𝑐}$            | 1.300e-05      | 0.009400           | 5.200e-04 | 0.1375                      |
| $k_{𝑀}^{𝐺𝑁𝑀𝑇1,𝑎ℎ𝑐𝑦𝑠𝑐}$           | 4.100e-07      | 0.000295           | 1.100e-05 | 0.000                       |
| $k_{𝑀}^{𝐺𝑁𝑀𝑇1,𝑔𝑙𝑦𝑐}$              | 5.480e-05      | 0.039500           | 2.540e-03 | 0.189583                    |
| $k_{𝑀}^{𝐺𝑁𝑀𝑇1,𝑠𝑎𝑟𝑐𝑠𝑐}$          | 3.730e-09      | 0.000003           | 1.000e-07 | 0.000                       |
| $k_{𝑀}^{𝑀𝐴𝑇1,𝑚𝑒𝑡−𝐿𝑐}$             | 1.400e-05      | 0.000720           | 1.070e-04 | 0.074                       |
| $k_{𝑀}^{𝑀𝐴𝑇1,𝑎𝑡𝑝𝑐}$                | 5.270e-05      | 0.038000           | 2.030e-03 | 0.125694                    |
| $k_{𝑀}^{𝑀𝐴𝑇3,𝑚𝑒𝑡−𝐿𝑐}$             | 4.470e-05      | 0.032200           | 1.130e-03 | -0.029                     |
| $k_{𝑀}^{𝑀𝐴𝑇3,𝑎𝑡𝑝𝑐}$                | 5.270e-05      | 0.038000           | 2.370e-03 | 0.179167                    |
| $k_{𝑀}^{𝑀𝐸𝑇𝐻−𝐺𝑒𝑛,𝑎𝑚𝑒𝑡𝑐}$      | 7.000e-06      | 0.000013           | 9.370e-06 | -0.135                     |
| $k_{𝑀}^{𝑀𝑆1,5𝑚𝑡ℎ𝑓𝑐}$                | 3.320e-06      | 0.002390           | 6.940e-05 | -0.124                     |
| $k_{𝑀}^{𝑀𝑆1,ℎ𝑐𝑦𝑠−𝐿𝑐}$              | 1.000e-06      | 0.000003           | 1.710e-06 | -0.054                     |
| $k_{𝑀}^{𝑀𝑇𝐻𝐹𝑅1,𝑚𝑙𝑡ℎ𝑓𝑐}$         | 7.500e-05      | 0.000088           | 8.080e-05 | -0.158                     |
| $k_{𝑀}^{𝑀𝑇𝐻𝐹𝑅1,𝑛𝑎𝑑𝑝ℎ𝑐}$         | 1.600e-05      | 0.000028           | 2.090e-05 | -0.105                     |
| $k_{𝑀}^{𝑃𝑅𝑂𝑇1,𝑚𝑒𝑡−𝐿𝑐}$           | 4.500e-05      | 0.000085           | 4.390e-05 | -2.507                     |
| $𝐿_{0}^{𝐶𝐵𝑆1}$                          | 3.730e-02      | 26.800000          | 1.030e+00 | 0.017                       |
| $𝐿_{0}^{𝐺𝑁𝑀𝑇1}$                        | 3.730e-02      | 26.800000          | 1.310e+02 | 0.3875                      |
| $𝐿_{0}^{𝑀𝐴𝑇3}$                          | 3.730e-03      | 2.680000           | 1.080e-01 | 0.037                       |
| $𝐿_{0}^{𝑀𝑇𝐻𝐹𝑅1}$                      | 1.120e-01      | 80.500000          | 3.920e-01 | -1.018                     |

: Parameter specification, marginal prior distributions and true parameter values used in our case study. {#tbl-case-study-params}

$\Delta_fG$ parameters for most metabolites were fixed; those that were
modelled as unknown had a multivariate normal prior distribution derived from
eQuilibrator [@beberEQuilibratorPlatformEstimation2021].

The values for $\Delta_fG$ parameters, as well as all other model parameters,
can be found by inspecting the file `priors.toml` which is online at <https://github.com/biosustain/Methionine_model/blob/main/data/methionine/priors.toml>.

The results of all reported Maud runs can be found at <https://github.com/biosustain/Methionine_model/blob/main/results>.

# References

::: {#refs}
:::

### 2023-10-28 paper/main.qmd
Update paper and supplementary, add main .gitignore

@@
 
-The Laplace approximation yielded XXX samples in XXX time. The diagnostics
-indicated that our algorithm was able to find the maximum a posteriori parameter
-configuration, approximate the Hessian and use these quantities to generate
-approximate posterior samples. The results can be found at <https://github.com/biosustain/Methionine_model/tree/main/results>.
+Our overall strategy was to compare samples generated using MCMC, which we
+were confident could be treated as draws from the true posterior distribution,
+with approximate posterior samples generated using the Laplace method. We were
+unable to generate approximate posterior samples for our main methionine cycle
+case study using Laplace approximation, as the algorithm could not recover from
+solver failures caused by unrealistic parameter configurations. We therefore
+made a comparison for a simpler model. This input can be found at <https://github.com/biosustain/Methionine_model/tree/main/data/example_ode>.
 
-@fig-laplace summarises the results of comparing the Laplace approximation of our
-case study’s posterior distribution with the true posterior. As can be seen
-from the top left plot, the Laplace method does not provide a good approximation
-to the true posterior distribution, as the marginal distribution of the
-total log probability density is clearly different. This was confirmed using
-the Kolmogorov-Smirnov test, which is a test to differentiate two empirical
-univariate distributions. The two distributions were significantly different
-with a p-value indistinguishable from zero.
+MCMC sampling for our comparison input yielded XXX samples in XXX time; Laplace
+sampling yielded XXX samples in XXX time. The diagnostics indicated that our
+algorithm was able to find the maximum a posteriori parameter configuration,
+approximate the Hessian and use these quantities to generate approximate
+posterior samples. The results can be found at <https://github.com/biosustain/Methionine_model/tree/main/results>.
+
+@fig-laplace summarises the results of comparing the samples generated using
+each method. As can be seen from the top left plot, the Laplace method does not
+provide a good approximation to the true posterior distribution in this case,
+as the marginal distribution of the total log probability density is clearly
+different. This impression was confirmed using the Kolmogorov-Smirnov test,
+which is a test to differentiate two empirical univariate distributions. The
+two distributions were significantly different with a p-value indistinguishable
+from zero.
 
 The difference between the Laplace approximation output and the true posterior
-distribution manifests itself not only in the parameter space, but also in
-the measurement space. Figure 5 frame B compares the 5%-95% interval for
-flux measurement log likelihoods in the true posterior with the Laplace
+distribution manifests not only in the parameter space, but also in the
+measurement space. Figure @fig-laplace frame B compares the 5%-95% interval
+for flux measurement log likelihoods in the true posterior with the Laplace
 approximation; lower log likelihood values indicate that the modelled and
@@
 posterior and the Laplace approximation for the pairwise marginal distributions
-of two Michaelis-Menten constants [LATEX CONSTANT]- and [LATEX CONSTANT]: see
-the bottom right cell of figure 5. This comparison demonstrates that the Laplace
-method is not able to capture the correct relationships between parameters’
-distributions.
+of two Michaelis-Menten constants [XXX]- and [XXX]: see the bottom right cell of
+figure @fig-laplace. This comparison demonstrates that the Laplace method is not
+able to capture the correct relationships between parameters’ distributions.
 
@@
 unfortunately preferable for posterior inference in this case. We expect that
-this is typical of kinetic models of realistic metabolic networks in general, so
-we recommend that Maud users use MCMC sampling if possible.
+the Laplace method will produce worse approximations the more complex the target
+model. Since the approximation is already unacceptable for our simple test
+model, we recommend that Maud users use MCMC sampling in preference to Laplace
+approximation if possible when fitting realistic Bayesian kinetic models.
 

### 2023-10-28 paper/main.tex
Update paper and supplementary, add main .gitignore

@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, interior hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, sharp corners, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, breakable]}{\end{tcolorbox}}\fi
 
@@
 
-The Laplace approximation yielded XXX samples in XXX time. The
-diagnostics indicated that our algorithm was able to find the maximum a
-posteriori parameter configuration, approximate the Hessian and use
-these quantities to generate approximate posterior samples. The results
-can be found at
+Our overall strategy was to compare samples generated using MCMC, which
+we were confident could be treated as draws from the true posterior
+distribution, with approximate posterior samples generated using the
+Laplace method. We were unable to generate approximate posterior samples
+for our main methionine cycle case study using Laplace approximation, as
+the algorithm could not recover from solver failures caused by
+unrealistic parameter configurations. We therefore made a comparison for
+a simpler model. This input can be found at
+\url{https://github.com/biosustain/Methionine_model/tree/main/data/example_ode}.
+
+MCMC sampling for our comparison input yielded XXX samples in XXX time;
+Laplace sampling yielded XXX samples in XXX time. The diagnostics
+indicated that our algorithm was able to find the maximum a posteriori
+parameter configuration, approximate the Hessian and use these
+quantities to generate approximate posterior samples. The results can be
+found at
 \url{https://github.com/biosustain/Methionine_model/tree/main/results}.
 
-Figure~\ref{fig-laplace} summarises the results of comparing the Laplace
-approximation of our case study's posterior distribution with the true
-posterior. As can be seen from the top left plot, the Laplace method
-does not provide a good approximation to the true posterior
-distribution, as the marginal distribution of the total log probability
-density is clearly different. This was confirmed using the
-Kolmogorov-Smirnov test, which is a test to differentiate two empirical
-univariate distributions. The two distributions were significantly
-different with a p-value indistinguishable from zero.
+Figure~\ref{fig-laplace} summarises the results of comparing the samples
+generated using each method. As can be seen from the top left plot, the
+Laplace method does not provide a good approximation to the true
+posterior distribution in this case, as the marginal distribution of the
+total log probability density is clearly different. This impression was
+confirmed using the Kolmogorov-Smirnov test, which is a test to
+differentiate two empirical univariate distributions. The two
+distributions were significantly different with a p-value
+indistinguishable from zero.
 
 The difference between the Laplace approximation output and the true
-posterior distribution manifests itself not only in the parameter space,
-but also in the measurement space. Figure 5 frame B compares the
-5\%-95\% interval for flux measurement log likelihoods in the true
-posterior with the Laplace approximation; lower log likelihood values
-indicate that the modelled and measured values are further away. The
-graph shows that the Laplace approximation yielded significantly worse
-predictions than the true posterior, even for the training data.
+posterior distribution manifests not only in the parameter space, but
+also in the measurement space. Figure Figure~\ref{fig-laplace} frame B
+compares the 5\%-95\% interval for flux measurement log likelihoods in
+the true posterior with the Laplace approximation; lower log likelihood
+values indicate that the modelled and measured values are further away.
+The graph shows that the Laplace approximation yielded significantly
+worse predictions than the true posterior, even for the training data.
 
@@
 true posterior and the Laplace approximation for the pairwise marginal
-distributions of two Michaelis-Menten constants {[}LATEX CONSTANT{]}-
-and {[}LATEX CONSTANT{]}: see the bottom right cell of figure 5. This
-comparison demonstrates that the Laplace method is not able to capture
-the correct relationships between parameters' distributions.
+distributions of two Michaelis-Menten constants {[}XXX{]}- and
+{[}XXX{]}: see the bottom right cell of figure Figure~\ref{fig-laplace}.
+This comparison demonstrates that the Laplace method is not able to
+capture the correct relationships between parameters' distributions.
 
@@
 unfortunately preferable for posterior inference in this case. We expect
-that this is typical of kinetic models of realistic metabolic networks
-in general, so we recommend that Maud users use MCMC sampling if
-possible.
+that the Laplace method will produce worse approximations the more
+complex the target model. Since the approximation is already
+unacceptable for our simple test model, we recommend that Maud users use
+MCMC sampling in preference to Laplace approximation if possible when
+fitting realistic Bayesian kinetic models.
 

### 2023-10-29 paper/main.qmd
Update paper

@@
 We conducted adaptive Hamiltonian Monte Carlo sampling for the full and missing-
--data datasets, obtaining XXX post-warmup samples after XXX of total computation
-time. For the Laplace approximation comparison, we used Maud’s Laplace mode.
-Full details as well as instructions for reproducing our analysis can be found
-at XXX.
+-data datasets. For the full dataset we obtained 1000 post-warmup samples each
+from 4 independent Markov chains after 1000 warm-up samples and "hot-starting"
+with a mass metric output by a previous model run.
+
+For the missing-data dataset XXX.
+
+For the Laplace approximation comparison, we used Maud’s Laplace mode. Full
+details as well as instructions for reproducing our analysis can be found at
+XXX.
 
@@
 This case study illustrates the type of kinetic model and dataset that Maud
-can fit. The model we analysed has XXX reactions, XXX state variables and XXX
-parameters. Posterior sampling with adaptive Hamiltonian Monte Carlo generated
-XXX ESS/hour. Generalising from this result, we conclude that it is feasible
-to use this method to fit models in the same order of magnitude, but not,
-for example, genome-scale kinetic models. To fit larger models, faster steady
-state solving methods or alternative inference algorithms will be required.
-@sec-laplace addresses whether Laplace approximation is a suitable candidate.
+can fit. The model we analysed has 10 reactions, 5 state variables and 212
+parameters. Generalising from our ability to fit this model in a reasonable time
+using Maud, we expect that Maud can be used to fit realistic Bayesian models
+of approximately the same size, but not, for example, genome-scale kinetic
+models. To fit larger models, faster steady state solving methods or alternative
+inference algorithms will be required. @sec-laplace addresses whether Laplace
+approximation is a suitable candidate.
 

### 2023-10-29 paper/main.tex
Update paper

@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, breakable, enhanced, borderline west={3pt}{0pt}{shadecolor}, sharp corners, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 We conducted adaptive Hamiltonian Monte Carlo sampling for the full and
-missing- -data datasets, obtaining XXX post-warmup samples after XXX of
-total computation time. For the Laplace approximation comparison, we
-used Maud's Laplace mode. Full details as well as instructions for
-reproducing our analysis can be found at XXX.
+missing- -data datasets. For the full dataset we obtained 1000
+post-warmup samples each from 4 independent Markov chains after 1000
+warm-up samples and ``hot-starting'' with a mass metric output by a
+previous model run.
+
+For the missing-data dataset XXX.
+
+For the Laplace approximation comparison, we used Maud's Laplace mode.
+Full details as well as instructions for reproducing our analysis can be
+found at XXX.
 
@@
 This case study illustrates the type of kinetic model and dataset that
-Maud can fit. The model we analysed has XXX reactions, XXX state
-variables and XXX parameters. Posterior sampling with adaptive
-Hamiltonian Monte Carlo generated XXX ESS/hour. Generalising from this
-result, we conclude that it is feasible to use this method to fit models
-in the same order of magnitude, but not, for example, genome-scale
-kinetic models. To fit larger models, faster steady state solving
-methods or alternative inference algorithms will be required.
-Section~\ref{sec-laplace} addresses whether Laplace approximation is a
-suitable candidate.
+Maud can fit. The model we analysed has 10 reactions, 5 state variables
+and 212 parameters. Generalising from our ability to fit this model in a
+reasonable time using Maud, we expect that Maud can be used to fit
+realistic Bayesian models of approximately the same size, but not, for
+example, genome-scale kinetic models. To fit larger models, faster
+steady state solving methods or alternative inference algorithms will be
+required. Section~\ref{sec-laplace} addresses whether Laplace
+approximation is a suitable candidate.
 

### 2023-10-30 paper/main.qmd
Add acknowledgements, edit supplementary

@@
     - id: biosustain
-      name: DTU Biosustain
+      name: DTU
+      department: DTU Biosustain
+      city: Kongens Lyngby
+      country: Denmark
   - name: "Nicholas Luke Cowie"
@@
     - ref: biosustain
+    - id: UQ
+    - name: The University of Queensland
+    - department: Australian Institute for Bioengineering and Nanotechnology (AIBN)
+    - city: St Lucia
+    - country: Australia
 bibliography: bibliography.bib
@@
 parameter values to use as ground truth by Monte Carlo sampling using a previous
-model of the methionine cycle as a starting point (see @saa_general_2015 for
+model of the methionine cycle as a starting point (see @saa_construction_2016 for
 this model).
@@
 We used these parameters to simulate steady states in a range of plausible
-experimental conditions, again using @saa_general_2015 as a starting point.
+experimental conditions, again using @saa_construction_2016 as a starting point.
 These steady states were then used to generate simulated measurements using the
@@
 were from the target posterior distribution. The improved $\hat{R}$ statistic
-[@vehtariRankNormalizationFoldingLocalization2021] for every sampled variable
-was within 1% of 1, indicating appropriate mixing within and between Markov
-chains. Additionally, the number of effective samples was high, indicating
-that we generated enough posterior samples to support inferences about the
-bulks of the distributions of the sampled parameters. Furthermore, we observed
-no post warm-up divergent transitions, indicating that the sampler was able to
-transform the log-posterior distribution, avoiding any regions with excessive
-curvature that might inhibit exploration via HMC.
+[@vehtariRankNormalizationFoldingLocalization2021] for every variable of
+interest was within 2% of 1, indicating appropriate mixing within and between
+Markov chains. Additionally, the number of effective samples was high,
+indicating that we generated enough posterior samples to support inferences
+about the bulks of the distributions of the sampled parameters. Furthermore,
+we observed no post warm-up divergent transitions, indicating that the sampler
+was able to transform the log-posterior distribution, avoiding any regions with
+excessive curvature that might inhibit exploration via HMC.
 
@@
 
+# Supporting information
+
+The document `supplementary.pdf` contains the following information:
+
+- Description of Maud's kinetic model including parameter dimensions and rate
+equations.
+- Description of the procedure used to generate the artificial case study
+datasets.
+- Specification of the prior distribution used for the case studies
+
 # Acknowledgements
+
+This research was funded by the Novo Nordisk foundation using grant number
+NNF14OC0009473.
 

### 2023-10-30 paper/main.tex

@@
 \author{Teddy Groves}
-\affiliation{ DTU Biosustain,  }
+\affiliation{DTU Biosustain, DTU, Kongens Lyngby, Denmark}
 
@@
 \author{Nicholas Luke Cowie}
-\affiliation{ DTU Biosustain,  }
+\affiliation{DTU Biosustain, DTU, Kongens Lyngby, Denmark}
 
 \author{Lars Keld Nielsen}
-\affiliation{ DTU Biosustain,  }
+\affiliation{DTU Biosustain, DTU, Kongens Lyngby, Denmark}
+\affiliation{   }
 
@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, breakable, enhanced, borderline west={3pt}{0pt}{shadecolor}, sharp corners, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, borderline west={3pt}{0pt}{shadecolor}, frame hidden, enhanced, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 using a previous model of the methionine cycle as a starting point (see
-\citet{saa_general_2015} for this model).
+\citet{saa_construction_2016} for this model).
 
 We used these parameters to simulate steady states in a range of
-plausible experimental conditions, again using \citet{saa_general_2015}
-as a starting point. These steady states were then used to generate
-simulated measurements using the measurement model described below in
-Section~\ref{sec-methionine-measurement-model}.
+plausible experimental conditions, again using
+\citet{saa_construction_2016} as a starting point. These steady states
+were then used to generate simulated measurements using the measurement
+model described below in Section~\ref{sec-methionine-measurement-model}.
 
@@
 \citep{vehtariRankNormalizationFoldingLocalization2021} for every
-sampled variable was within 1\% of 1, indicating appropriate mixing
+variable of interest was within 2\% of 1, indicating appropriate mixing
 within and between Markov chains. Additionally, the number of effective
@@
 
+\hypertarget{supporting-information}{%
+\section{Supporting information}\label{supporting-information}}
+
+The document \texttt{supplementary.pdf} contains the following
+information:
+
+\begin{itemize}
+\tightlist
+\item
+  Description of Maud's kinetic model including parameter dimensions and
+  rate equations.
+\item
+  Description of the procedure used to generate the artificial case
+  study datasets.
+\item
+  Specification of the prior distribution used for the case studies
+\end{itemize}
+
 \hypertarget{acknowledgements}{%
@@
 
+This research was funded by the Novo Nordisk foundation using grant
+number NNF14OC0009473.
+
 \hypertarget{references}{%

### 2023-10-30 paper/supplementary.qmd

@@
 
-## Prior distributions compared with true parameter values
+Starting with the model in @saa_construction_2016, we extracted values for
+enzyme concentrations, boundary conditions and fluxes. We used these values
+to generate MCMC samples using Maud using the priors specified in section
+@sec-methionine-priors. When this was finished, we selected one sample with
+relatively high log probability to use as a ground truth in our case study.
+These parameter values are shown below in table @tbl-case-study-params. We
+manually inspected the parameter values to screen for any obviously implausible
+values; we did not find any of these.
+
+## Prior distributions compared with true parameter values {#sec-methionine-priors}
 

### 2023-10-30 plot_figure_4.py
Code to make plot

@@ -1,0 +1,282 @@
+# Copyright (C) 2020 Novo Nordisk Foundation Center for Biosustainability,
+# Technical University of Denmark.
+#
+# This program is free software: you can redistribute it and/or modify
+# it under the terms of the GNU General Public License as published by
+# the Free Software Foundation, either version 3 of the License, or
+# (at your option) any later version.
+#
+# This program is distributed in the hope that it will be useful,
+# but WITHOUT ANY WARRANTY; without even the implied warranty of
+# MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+# GNU General Public License for more details.
+#
+# You should have received a copy of the GNU General Public License
+# along with this program.  If not, see <https://www.gnu.org/licenses/>.
+
+"""Code for plotting posterior distribution."""
+import os
+import argparse
+from pathlib import Path
+import tempfile
+import shutil
+import seaborn as sns
+
+import arviz as az
+import numpy as np
+import pandas as pd
+import scipy
+import seaborn as sns
+import matplotlib.pyplot as plt
+import matplotlib.patches as mpatches
+import matplotlib.image as mpimg
+import matplotlib.transforms as transforms
+
+from maud.loading_maud_inputs import load_maud_input
+from maud.getting_idatas import get_idata
+
+
+plt.style.use('ipynb')
+
+
+HELP_MSG = """
+This script plots boxplots of enzyme concentrations,
+metabolite concentrations, and fluxes of out-of-sample
+predictions. Measurements are also included in these
+plots.
+"""
+
+MAUD_OUTPUT_TRAINING = Path.cwd() / "results" / "example_ode"
+MAUD_OUTPUT_VALIDATION = MAUD_OUTPUT_TRAINING / "validation"
+MAUD_OUTPUT_LAPLACE_TRAINING = Path.cwd() / "results" / "example_ode_laplace"
+MAUD_OUTPUT_VALIDATION = MAUD_OUTPUT_LAPLACE_TRAINING / "validation"
+PLOT_OUTPUT = Path.cwd() / "figures" / "figure_4.png"
+
+colour_scheme = {
+    "measurement": "0.2",
+    "laplace": "#005F73",
+    "posterior": "#001219"
+}
+
+OFFSET = 0.05
+cm = 1/2.54  # centimeters in inches
+
+def plot_lp_testing(axes, lp):
+    color_map = {
+        "posterior": "#001219",
[…]

### 2023-10-30 plot_figure_2.py
results files example_ode example_ode_laplace and respective validation

@@ -1,0 +1,197 @@
+import builtins
+from pathlib import Path
+import tempfile
+import shutil
+import os
+
+import seaborn as sns
+import matplotlib.pyplot as plt
+import matplotlib.patches as mpatches
+import matplotlib.image as mpimg
+import numpy as np
+import pandas as pd
+import arviz as az
+
+from maud.loading_maud_inputs import load_maud_input
+from maud.getting_idatas import get_idata
+
+from plotting_functions import get_measurements_of_type, get_csvs
+
+MAUD_OUTPUT = Path.cwd() / "methionine_model"/ "results" / "methionine"
+PLOT_OUTPUT = Path.cwd() / "methionine_model"/ "plots" / "figure_2"
+
+plt.style.use('ipynb')
+
+colour_scheme = {
+    "measurement": "0.6",
+    "sample": "#8F2727"
+}
+
+def plot_flux_forest(mi, infd, axes):
+    """Return forest plot axes of fluxes."""
+    flux_measurements = get_measurements_of_type("flux", "yflux", mi.measurements)
+    flux_measurements["id"] = flux_measurements["experiments"] + '-' + flux_measurements["flux"]
+    flux_samples = infd.posterior["flux"].to_dataframe().reset_index()
+    flux_samples["id"] = flux_samples["experiments"] + '-' + flux_samples["reactions"]
+    flux_samples = flux_samples.groupby("id") \
+                               .quantile([0.05, 0.95]) \
+                               .reset_index() \
+                               .pivot(index="id", columns="level_1", values="flux")
+    sorted_fluxes = list(flux_measurements.sort_values("measurement")["id"])
+    for id, sample in flux_samples.reindex(sorted_fluxes).iterrows():
+        axes[0,0].vlines(
+            x=id,
+            ymin=sample[0.05],
+            ymax=sample[0.95],
+            linewidth=2.5,
+            color=colour_scheme["sample"],
+        )
+    for _, meas in flux_measurements.sort_values("measurement").iterrows():
+        axes[0,0].vlines(
+            x=meas["id"],
+            ymin=meas["measurement"]-1.65*meas["error_scale"],
+            ymax=meas["measurement"]+1.65*meas["error_scale"],
+            linewidth=2,
[…]

### 2023-10-30 plot_figure_3.py

@@ -1,0 +1,239 @@
+# Copyright (C) 2020 Novo Nordisk Foundation Center for Biosustainability,
+# Technical University of Denmark.
+#
+# This program is free software: you can redistribute it and/or modify
+# it under the terms of the GNU General Public License as published by
+# the Free Software Foundation, either version 3 of the License, or
+# (at your option) any later version.
+#
+# This program is distributed in the hope that it will be useful,
+# but WITHOUT ANY WARRANTY; without even the implied warranty of
+# MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+# GNU General Public License for more details.
+#
+# You should have received a copy of the GNU General Public License
+# along with this program.  If not, see <https://www.gnu.org/licenses/>.
+
+"""Code for plotting posterior distribution."""
+from pathlib import Path
+import tempfile
+
+import arviz as az
+import numpy as np
+import pandas as pd
+import seaborn as sns
+import matplotlib.pyplot as plt
+import matplotlib.patches as mpatches
+import matplotlib.image as mpimg
+import matplotlib.transforms as transforms
+from matplotlib.gridspec import GridSpec
+from matplotlib.lines import Line2D
+
+from maud import io
+
+plt.style.use('ipynb')
+
+
+HELP_MSG = """
+This script plots boxplots of enzyme concentrations,
+metabolite concentrations, and fluxes of out-of-sample
+predictions. Measurements are also included in these
+plots.
+"""
+
+MAUD_OUTPUT_MISSING_AHCYS = Path.cwd() / "results" / "methionine_cycle_6_missing_ahcys_validation"
+MAUD_OUTPUT = Path.cwd() / "results" / "methionine_cycle_6_validation"
+PLOT_OUTPUT = Path.cwd() / "plots" / "figure_4"
+
+colour_scheme = {
+    "complete_measurement_set": "#001219",
+    "missing_ahcys_measurement": "#005F73"
+}
+
+OFFSET = 0.05
+cm = 1/2.54  # centimeters in inches
+
+def plot_residual_state_variables(residual_df, ax):
+
+    offset = lambda p: transforms.ScaledTranslation(p/72.,0, plt.gcf().dpi_scale_trans)
+    for id, sample in residual_df.loc[residual_df["model"]=="missing ahcys_c measurement"].reset_index().iterrows():
+        ax.vlines(
+            x=id,
+            ymin=sample[0.05],
[…]

### 2023-10-30 plot_figure_5.py

@@ -1,0 +1,166 @@
+# Copyright (C) 2020 Novo Nordisk Foundation Center for Biosustainability,
+# Technical University of Denmark.
+#
+# This program is free software: you can redistribute it and/or modify
+# it under the terms of the GNU General Public License as published by
+# the Free Software Foundation, either version 3 of the License, or
+# (at your option) any later version.
+#
+# This program is distributed in the hope that it will be useful,
+# but WITHOUT ANY WARRANTY; without even the implied warranty of
+# MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+# GNU General Public License for more details.
+#
+# You should have received a copy of the GNU General Public License
+# along with this program.  If not, see <https://www.gnu.org/licenses/>.
+
+"""Code for plotting DOE and answering question about Allostery"""
+import os
+import argparse
+from pathlib import Path
+import tempfile
+import shutil
+
+import arviz as az
+import numpy as np
+import pandas as pd
+import plotnine as p9
+from plotnine import aes, geom_point, ggplot, labs
+import scipy
+import seaborn as sns
+import matplotlib.pyplot as plt
+import matplotlib.patches as mpatches
+import matplotlib.image as mpimg
+from matplotlib.gridspec import GridSpec
+
+from plotting_functions import get_measurements_of_type
+
+from maud import io
+
+plt.style.use('ipynb')
+
+met_13_path = Path.cwd() / "results" / "methionine_cycle_13"
+met_13_true_values = Path.cwd() / "simulate"
+PLOT_OUTPUT = Path.cwd() / "plots" / "figure_5"
+
+def plot_regulation(regulation_df, true_df, ax):
+    sns.violinplot(data=regulation_df, x="value", y="regulation", color="#005F73",
+              linewidth=1, orient="h", ax=ax, inner="quart", marker="")
+    for line in ax.lines:
+        line.set_marker(None)
+    ax.set_xlabel(r"ln $\frac{Regulatory\ Components_{2}}{Regulatory\ Components_{1}}$")
+    ax.axvline(x=0, color="k")
+    for i, value in enumerate(true_df):
+        ax.plot(value, i, "r")
+    return ax
+
+def plot_flux(flux_df, ax):
[…]

### 2023-10-30 paper/main.qmd
Update paper

@@
   an example application.
+number-sections: true
 author: 
@@
       abbreviations: IR,NMR,UV
-  acs-html:
-   include-in-header: mathjax.html
-   embed-resources: true
-  docx: default
+#  acs-html:
+#   include-in-header: mathjax.html
+#   embed-resources: true
+#  docx: default
 ---
@@
 
-### Computation
-
-We conducted adaptive Hamiltonian Monte Carlo sampling for the full and missing-
--data datasets. For the full dataset we obtained 1000 post-warmup samples each
-from 4 independent Markov chains after 1000 warm-up samples and "hot-starting"
-with a mass metric output by a previous model run.
-
-For the missing-data dataset XXX.
-
-For the Laplace approximation comparison, we used Maud’s Laplace mode. Full
-details as well as instructions for reproducing our analysis can be found at
-XXX.
-
 ## Findings
@@
 
-MCMC sampling for our comparison input yielded XXX samples in XXX time; Laplace
-sampling yielded XXX samples in XXX time. The diagnostics indicated that our
-algorithm was able to find the maximum a posteriori parameter configuration,
-approximate the Hessian and use these quantities to generate approximate
-posterior samples. The results can be found at <https://github.com/biosustain/Methionine_model/tree/main/results>.
+MCMC sampling for our comparison input yielded 200 samples in 625 minutes;
+Laplace sampling yielded the same number of samples in only one minute.
+The diagnostics indicated that our algorithm was able to find the maximum
+a posteriori parameter configuration, approximate the Hessian and use these
+quantities to generate approximate posterior samples. The results can be found
+at <https://github.com/biosustain/Methionine_model/tree/main/results>.
 
@@
 posterior and the Laplace approximation for the pairwise marginal distributions
-of two Michaelis-Menten constants [XXX]- and [XXX]: see the bottom right cell of
-figure @fig-laplace. This comparison demonstrates that the Laplace method is not
-able to capture the correct relationships between parameters’ distributions.
+of two Michaelis-Menten constants $K_{m}^{A,r1}$ and $K_{cat}^{r1}$: see the
+bottom right cell of figure @fig-laplace. This comparison demonstrates that
+the Laplace method is not able to capture the correct relationships between
+parameters’ distributions.
 
@@
 
-![](./figures/laplace.png)
+![](../figures/laplace.png)
 
-CAPTION: FILL THIS IN!
+Graphical comparison of approximate posterior samples generated using Laplace
+sampling (blue-green) with posterior samples generated using MCMC (dark
+grey). (**Top left**) the two sets of samples clearly have different marginal
+distributions for the overall log probability variable, indicating that the
+Laplace samples do not accurately approximate the target distribution. (**Top
+right**) The distribution of marginal posterior predictive log likelihood values
+for training data flux measurements shows that the Laplace method tended to
+yield much worse predictions compared with the true model. (**Bottom left**)
+The Laplace method also tended to produce worse flux predictions for held-back
+test measurements. (**Bottom right**) The marginal joint distribution of two
+parameters: $K_m^{A,rl}$ and $K_{cat}^{r1}$. The Laplace method is not able
+to track the correct joint distribution for this pair of parameters. This is
+unsurprising given that the target distribution has position-dependent scales
+which are difficult for a linear approximation to capture.
 
@@
 equations.
-- Description of the procedure used to generate the artificial case study
-datasets.
+- Details of the procedure used to generate our case study results, including
+generating the artificial case study datasets, specifying priors and carrying
+out computation.
 - Specification of the prior distribution used for the case studies

### 2023-10-30 paper/main.tex
Update paper

@@
 \setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+\setcounter{secnumdepth}{5}
 % Make \paragraph and \subparagraph free-standing
@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, borderline west={3pt}{0pt}{shadecolor}, frame hidden, enhanced, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners, boxrule=0pt, interior hidden, breakable]}{\end{tcolorbox}}\fi
 
@@
 
-\hypertarget{computation}{%
-\subsubsection{Computation}\label{computation}}
-
-We conducted adaptive Hamiltonian Monte Carlo sampling for the full and
-missing- -data datasets. For the full dataset we obtained 1000
-post-warmup samples each from 4 independent Markov chains after 1000
-warm-up samples and ``hot-starting'' with a mass metric output by a
-previous model run.
-
-For the missing-data dataset XXX.
-
-For the Laplace approximation comparison, we used Maud's Laplace mode.
-Full details as well as instructions for reproducing our analysis can be
-found at XXX.
-
 \hypertarget{findings}{%
@@
 
-MCMC sampling for our comparison input yielded XXX samples in XXX time;
-Laplace sampling yielded XXX samples in XXX time. The diagnostics
-indicated that our algorithm was able to find the maximum a posteriori
-parameter configuration, approximate the Hessian and use these
-quantities to generate approximate posterior samples. The results can be
-found at
+MCMC sampling for our comparison input yielded 200 samples in 625
+minutes; Laplace sampling yielded the same number of samples in only one
+minute. The diagnostics indicated that our algorithm was able to find
+the maximum a posteriori parameter configuration, approximate the
+Hessian and use these quantities to generate approximate posterior
+samples. The results can be found at
 \url{https://github.com/biosustain/Methionine_model/tree/main/results}.
@@
 true posterior and the Laplace approximation for the pairwise marginal
-distributions of two Michaelis-Menten constants {[}XXX{]}- and
-{[}XXX{]}: see the bottom right cell of figure Figure~\ref{fig-laplace}.
-This comparison demonstrates that the Laplace method is not able to
-capture the correct relationships between parameters' distributions.
+distributions of two Michaelis-Menten constants \(K_{m}^{A,r1}\) and
+\(K_{cat}^{r1}\): see the bottom right cell of figure
+Figure~\ref{fig-laplace}. This comparison demonstrates that the Laplace
+method is not able to capture the correct relationships between
+parameters' distributions.
 
@@
 
-\includegraphics{./figures/laplace.png}
+\includegraphics{../figures/laplace.png}
 
@@
 
-\caption{\label{fig-laplace}CAPTION: FILL THIS IN!}
+\caption{\label{fig-laplace}Graphical comparison of approximate
+posterior samples generated using Laplace sampling (blue-green) with
+posterior samples generated using MCMC (dark grey). (\textbf{Top left})
+the two sets of samples clearly have different marginal distributions
+for the overall log probability variable, indicating that the Laplace
+samples do not accurately approximate the target distribution.
+(\textbf{Top right}) The distribution of marginal posterior predictive
+log likelihood values for training data flux measurements shows that the
+Laplace method tended to yield much worse predictions compared with the
+true model. (\textbf{Bottom left}) The Laplace method also tended to
+produce worse flux predictions for held-back test measurements.
+(\textbf{Bottom right}) The marginal joint distribution of two
+parameters: \(K_m^{A,rl}\) and \(K_{cat}^{r1}\). The Laplace method is
+not able to track the correct joint distribution for this pair of
+parameters. This is unsurprising given that the target distribution has
+position-dependent scales which are difficult for a linear approximation
+to capture.}
 
@@
 \item
-  Description of the procedure used to generate the artificial case
-  study datasets.
+  Details of the procedure used to generate our case study results,
+  including generating the artificial case study datasets, specifying
+  priors and carrying out computation.
 \item

### 2023-10-30 paper/main.qmd
Update paper and supplementary

@@
       abbreviations: IR,NMR,UV
-#  acs-html:
-#   include-in-header: mathjax.html
-#   embed-resources: true
-#  docx: default
+  acs-html:
+   include-in-header: mathjax.html
+   embed-resources: true
+  docx: default
 ---

### 2023-10-30 paper/main.tex
Update paper and supplementary

@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners, boxrule=0pt, interior hidden, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, frame hidden, boxrule=0pt, enhanced, sharp corners]}{\end{tcolorbox}}\fi
 

### 2023-10-30 plot_figure_3.py
functions for plotting figure 3

@@ -15,9 +15,8 @@
 # along with this program.  If not, see <https://www.gnu.org/licenses/>.
 
 """Code for plotting posterior distribution."""
+import os
 from pathlib import Path
-import tempfile
-
 import arviz as az
 import numpy as np
 import pandas as pd
@@ -29,7 +28,8 @@ import matplotlib.transforms as transforms
 from matplotlib.gridspec import GridSpec
 from matplotlib.lines import Line2D
 
-from maud import io
+from maud.loading_maud_inputs import load_maud_input
+from maud.getting_idatas import get_idata
 
 plt.style.use('ipynb')
 
@@ -41,28 +41,27 @@ predictions. Measurements are also included in these
 plots.
 """
 
-MAUD_OUTPUT_MISSING_AHCYS = Path.cwd() / "results" / "methionine_cycle_6_missing_ahcys_validation"
-MAUD_OUTPUT = Path.cwd() / "results" / "methionine_cycle_6_validation"
-PLOT_OUTPUT = Path.cwd() / "plots" / "figure_4"
+MAUD_OUTPUT_MISSING_AHCYS = Path.cwd() / "results" / "methionine_missing_ahcys"
+MAUD_OUTPUT = Path.cwd() / "results" / "methionine"
+PLOT_OUTPUT = Path.cwd() / "figures" / "figure_3.png"
 
 colour_scheme = {
-    "complete_measurement_set": "#001219",
-    "missing_ahcys_measurement": "#005F73"
+    "complete measurement set": "#001219",
+    "missing ahcys measurement": "#005F73"
 }
 
 OFFSET = 0.05
 cm = 1/2.54  # centimeters in inches
 
 def plot_residual_state_variables(residual_df, ax):
-
     offset = lambda p: transforms.ScaledTranslation(p/72.,0, plt.gcf().dpi_scale_trans)
-    for id, sample in residual_df.loc[residual_df["model"]=="missing ahcys_c measurement"].reset_index().iterrows():
+    for id, sample in residual_df.loc[residual_df["model"]=="missing ahcys measurement"].reset_index().iterrows():
         ax.vlines(
             x=id,
             ymin=sample[0.05],
             ymax=sample[0.95],
             linewidth=2.5,
-            color=colour_scheme["missing_ahcys_measurement"],
+            color=colour_scheme["missing ahcys measurement"],
             transform=ax.transData+offset(+1),
             alpha=0.7,
         )
[…]

### 2023-10-30 plot_figure_2.py
plotting figure 2

@@ -15,10 +15,9 @@ import arviz as az
 from maud.loading_maud_inputs import load_maud_input
 from maud.getting_idatas import get_idata
 
-from plotting_functions import get_measurements_of_type, get_csvs
 
-MAUD_OUTPUT = Path.cwd() / "methionine_model"/ "results" / "methionine"
-PLOT_OUTPUT = Path.cwd() / "methionine_model"/ "plots" / "figure_2"
+MAUD_OUTPUT = Path.cwd() / "results" / "methionine"
+PLOT_OUTPUT = Path.cwd() / "figures" / "figure_2"
 
 plt.style.use('ipynb')
 
@@ -27,17 +26,15 @@ colour_scheme = {
     "sample": "#8F2727"
 }
 
-def plot_flux_forest(mi, infd, axes):
+def plot_flux_forest(mi, idata_methionine, axes, flux_meas_df):
     """Return forest plot axes of fluxes."""
-    flux_measurements = get_measurements_of_type("flux", "yflux", mi.measurements)
-    flux_measurements["id"] = flux_measurements["experiments"] + '-' + flux_measurements["flux"]
-    flux_samples = infd.posterior["flux"].to_dataframe().reset_index()
-    flux_samples["id"] = flux_samples["experiments"] + '-' + flux_samples["reactions"]
-    flux_samples = flux_samples.groupby("id") \
+    flux_samples = idata_methionine.posterior["flux_train"].to_dataframe().reset_index()
+    flux_samples["id"] = flux_samples["experiments"] + flux_samples["reactions"]
+    flux_samples = flux_samples[["id", "flux_train"]].groupby("id") \
                                .quantile([0.05, 0.95]) \
                                .reset_index() \
-                               .pivot(index="id", columns="level_1", values="flux")
-    sorted_fluxes = list(flux_measurements.sort_values("measurement")["id"])
+                               .pivot(index="id", columns="level_1", values="flux_train")
+    sorted_fluxes = list(flux_meas_df.sort_values("flux").index)
     for id, sample in flux_samples.reindex(sorted_fluxes).iterrows():
         axes[0,0].vlines(
             x=id,
@@ -46,32 +43,30 @@ def plot_flux_forest(mi, infd, axes):
             linewidth=2.5,
             color=colour_scheme["sample"],
         )
[…]

### 2023-10-31 plot_figure_5.py
figure 5

@@ -16,38 +16,32 @@
 
 """Code for plotting DOE and answering question about Allostery"""
 import os
-import argparse
 from pathlib import Path
 import tempfile
-import shutil
 
 import arviz as az
 import numpy as np
 import pandas as pd
-import plotnine as p9
-from plotnine import aes, geom_point, ggplot, labs
-import scipy
 import seaborn as sns
 import matplotlib.pyplot as plt
 import matplotlib.patches as mpatches
 import matplotlib.image as mpimg
 from matplotlib.gridspec import GridSpec
 
-from plotting_functions import get_measurements_of_type
-
-from maud import io
+from maud.loading_maud_inputs import load_maud_input
+from maud.getting_idatas import get_idata
 
 plt.style.use('ipynb')
 
-met_13_path = Path.cwd() / "results" / "methionine_cycle_13"
-met_13_true_values = Path.cwd() / "simulate"
-PLOT_OUTPUT = Path.cwd() / "plots" / "figure_5"
+MAUD_OUTPUT = Path.cwd() / "results" / "methionine"
+MAUD_SIM = Path.cwd()  / "results" / "methionine" / "simulate"
+PLOT_OUTPUT = Path.cwd() / "figures" 
 
 def plot_regulation(regulation_df, true_df, ax):
     sns.violinplot(data=regulation_df, x="value", y="regulation", color="#005F73",
-              linewidth=1, orient="h", ax=ax, inner="quart", marker="")
+              linewidth=1, orient="h", ax=ax, inner="quart")
     for line in ax.lines:
-        line.set_marker(None)
+        line.set_marker("")
     ax.set_xlabel(r"ln $\frac{Regulatory\ Components_{2}}{Regulatory\ Components_{1}}$")
     ax.axvline(x=0, color="k")
     for i, value in enumerate(true_df):
@@ -55,37 +49,35 @@ def plot_regulation(regulation_df, true_df, ax):
     return ax
 
 def plot_flux(flux_df, ax):
-    sns.violinplot(data=flux_df, x="flux", y="experiments", color="#005F73",
-              linewidth=1, orient="h", ax=ax, inner="quart", marker="")
+    sns.violinplot(data=flux_df, x="flux_train", y="experiments", color="#005F73",
+              linewidth=1, orient="h", ax=ax, inner="quart")
     for line in ax.lines:
[…]

### 2023-10-31 paper/supplementary.qmd

@@
 ---
+
+The results of all reported Maud runs can be found at <https://github.com/biosustain/Methionine_model/blob/main/results>.
 
@@
 
-The results of all reported Maud runs can be found at <https://github.com/biosustain/Methionine_model/blob/main/results>.
+## Computation
+
+We conducted adaptive Hamiltonian Monte Carlo sampling for the full and missing-
+-data datasets. For the full dataset we obtained 1000 post-warmup samples each
+from 4 independent Markov chains after 1000 warm-up samples and "hot-starting"
+with a mass metric output by a previous model run.
+
+For the missing-data dataset XXX.
+
+# Laplace approximation case study
+
+To compare MCMC sampling with Laplace approximation we used a different model.
+ The full Maud input folders can be found at <https://github.com/biosustain/Methionine_model/tree/main/data/example_ode> 
+and <https://github.com/biosustain/Methionine_model/tree/main/data/example_ode_laplace>
+
+To generate Laplace samples we used Maud's Laplace mode.
 

### 2023-10-31 plot_figure_3.py
updating missing measurement model to within sample

@@ -75,7 +75,7 @@ def plot_residual_state_variables(residual_df, ax):
             transform=ax.transData+offset(-1),
             alpha=0.7
         )
-    ax.axhline(y=0, color="red")
+    ax.axhline(y=0, xmin=0, xmax=len(residual_df)/2, color="red")
     ax.set_ylabel("log residual plot of balanced metabolites")
     ax.get_xaxis().set_visible(False)
     return ax
@@ -113,10 +113,11 @@ def plot_parameter_hist(parameter_df, ax):
 
 def plot_figure():
     mi = load_maud_input(data_path=MAUD_OUTPUT / "user_input")
-    csvs_methionine = [MAUD_OUTPUT/ "validation" / file for file in os.listdir(MAUD_OUTPUT/ "validation") if ".csv" in file]
-    infd_methionine = get_idata(csvs_methionine, mi=mi, mode="test")
-    csvs_methionine_missing = [MAUD_OUTPUT_MISSING_AHCYS/ "validation" / file for file in os.listdir(MAUD_OUTPUT_MISSING_AHCYS/ "validation") if ".csv" in file]
-    infd_methionine_missing = get_idata(csvs_methionine_missing, mi=mi, mode="test")
+    mi_missing = load_maud_input(data_path=MAUD_OUTPUT_MISSING_AHCYS / "user_input")
+    csvs_methionine = [MAUD_OUTPUT/ "samples" / file for file in os.listdir(MAUD_OUTPUT/ "samples") if ".csv" in file]
+    infd_methionine = get_idata(csvs_methionine, mi=mi, mode="train")
+    csvs_methionine_missing = [MAUD_OUTPUT_MISSING_AHCYS/ "samples" / file for file in os.listdir(MAUD_OUTPUT_MISSING_AHCYS/ "samples") if ".csv" in file]
+    infd_methionine_missing = get_idata(csvs_methionine_missing, mi=mi_missing, mode="train")
 
     flux_measurements = [{
         "flux": meas.value, 
@@ -126,7 +127,7 @@ def plot_figure():
     } 
         for exp in mi.experiments 
         for meas in exp.measurements 
-        if (meas.target_type=="flux") & (exp!="validation")
+        if (meas.target_type=="flux")
     ]
     flux_meas_df = pd.DataFrame.from_records(flux_measurements)
     flux_meas_df["id"] = flux_meas_df["experiment"]+flux_meas_df["reaction"]
@@ -141,21 +142,21 @@ def plot_figure():
     } 
[…]

### 2023-10-31 paper/cover_letter.qmd

---
title: Cover letter

---

# Manuscript title

Bayesian regression facilitates quantitative modelling of cell metabolism

# Name of the corresponding author

Teddy Groves (email: <mailto:tedgro@dtu.dk>)

# Names of the other authors

- Nicholas Cowie (email: <mailto:nicow@dtu.dk>)
- Lars Keld Nielsen (email: <mailto:lars.nielsen@uq.edu.au>)

# Why the paper is appropriate for ACS Synthetic Biology

Our paper is intended of the virtual special issue of ACS Synthetic Biology
concentrating on quantitative synthetic biology. It presents Maud, a software
application that we developed for fitting Bayesian statistical models of fluxes
through systems of metabolic reactions at steady state, enabling inference about
these systems from quantitative metabolomics, fluxomics and proteomics data.

Our software is a "quantitative technology and tool for synthetic
biology", as described in the [call for papers](https://axial.acs.org/biology-and-biological-chemistry/cfp-quantitative-synthetic-biology),
because it is intended to be used as a synthetic biology tool. For example, as
we illustrate in our paper, Maud can be used to identify drivers of regulation
through a pathway, thereby helping to identify or rule out knockout targets.

# Short lay summary

This paper presents Maud, a command-line application that implements
Bayesian statistical inference for kinetic models of biochemical metabolic
reaction networks.

Maud takes into account quantitative information from omics experiments
and background knowledge, as well as structural information about kinetic
mechanisms, regulatory interactions and enzyme knockouts.

Our paper reviews alternative options, illustrates Maud's use with two case
studies, and explains Maud's implementation and intended usage.

### 2023-10-31 paper/main.qmd

@@
 ---
-title: "Bayesian regression modelling facilitates quantitative modelling of cell metabolism"
+title: "Bayesian regression facilitates quantitative modelling of cell metabolism"
 keywords: Bayesian inference, kinetic models of cell metabolism 

### 2023-10-31 paper/main.tex

@@
 \hypersetup{
-  pdftitle={Bayesian regression modelling facilitates quantitative modelling of cell metabolism},
+  pdftitle={Bayesian regression facilitates quantitative modelling of cell metabolism},
   pdfauthor={Teddy Groves; Nicholas Luke Cowie; Lars Keld Nielsen},
@@
 
-\title[]{Bayesian regression modelling facilitates quantitative
-modelling of cell metabolism}
+\title[]{Bayesian regression facilitates quantitative modelling of cell
+metabolism}
 \makeatletter
@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, frame hidden, boxrule=0pt, enhanced, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, interior hidden, breakable, enhanced, frame hidden]}{\end{tcolorbox}}\fi
 

### 2023-10-31 plot_figure_3.py
fix figure 3

@@ -31,7 +31,7 @@ from matplotlib.lines import Line2D
 from maud.loading_maud_inputs import load_maud_input
 from maud.getting_idatas import get_idata
 
-plt.style.use('ipynb')
+plt.style.use("ipynb")
 
 
 HELP_MSG = """
@@ -47,52 +47,49 @@ PLOT_OUTPUT = Path.cwd() / "figures" / "figure_3.png"
 
 colour_scheme = {
     "complete measurement set": "#001219",
-    "missing ahcys measurement": "#005F73"
+    "missing ahcys measurement": "#005F73",
 }
 
 OFFSET = 0.05
-cm = 1/2.54  # centimeters in inches
+cm = 1 / 2.54  # centimeters in inches
+
 
 def plot_residual_state_variables(residual_df, ax):
-    offset = lambda p: transforms.ScaledTranslation(p/72.,0, plt.gcf().dpi_scale_trans)
-    for id, sample in residual_df.loc[residual_df["model"]=="missing ahcys measurement"].reset_index().iterrows():
+    # offset = lambda p: transforms.ScaledTranslation(
+    #     p / 72.0, 0, plt.gcf().dpi_scale_trans
+    # )
+    for model, direction in zip(
+        ["missing ahcys measurement", "complete measurement set"], [1, -1]
+    ):
+        df = residual_df.loc[lambda df: df["model"] == model]
+        x = np.arange(len(df)) + 0.1 * direction
         ax.vlines(
-            x=id,
-            ymin=sample[0.05],
-            ymax=sample[0.95],
+            x=x,
+            ymin=df[0.05],
+            ymax=df[0.95],
             linewidth=2.5,
-            color=colour_scheme["missing ahcys measurement"],
-            transform=ax.transData+offset(+1),
+            color=colour_scheme[model],
             alpha=0.7,
         )
-    for id, sample in residual_df.loc[residual_df["model"]=="complete measurement set"].reset_index().iterrows():
-        ax.vlines(
-            x=id,
-            ymin=sample[0.05],
-            ymax=sample[0.95],
-            linewidth=2.5,
-            color=colour_scheme["complete measurement set"],
-            transform=ax.transData+offset(-1),
-            alpha=0.7
-        )
-    ax.axhline(y=0, xmin=0, xmax=len(residual_df)/2, color="red")
[…]

### 2023-10-31 paper/main.qmd
change condition 2 to dataset 12

@@
 
-To demonstrate how Maud's output can be used to yield useful metabolic insights
-we used the results of our case study to explain why the flux of the enzyme
-$GNMT$ is higher in condition 1 than in condition 2. GNMT is an irreversible
-enzyme that is homotropically activated by its substrate, competitively
-inhibited by its product and heterotropically inhibited. The diverse regulation
-makes it the ideal test case to elucidate regulatory changes.
+To demonstrate how Maud's output can be used to yield useful metabolic
+insights we used the results of our case study to explain why the flux of
+the enzyme $GNMT$ is higher in dataset 1 than in dataset 12. GNMT is
+an irreversible enzyme that is homotropically activated by its substrate,
+competitively inhibited by its product and heterotropically inhibited by 5,10-
+methylenetetrahydrofolate (mlthf). The diverse regulation makes it the ideal
+test case to elucidate regulatory changes.
 
@@
 results our main case study analysis. Each curve shows the marginal posterior
-distribution of the ratio of the corresponding regulatory component in condition
-1 compared with condition 2. A positive value indicates that the component
-was increased in condition 1 relative to condition 2, with 0 indicating no
+distribution of the ratio of the corresponding regulatory component in dataset
+1 compared with dataset 12. A positive value indicates that the component
+was increased in dataset 1 relative to dataset 12, with 0 indicating no
 difference. The probability, according to our model, of the component acting in
@@
 Our model correctly inferred that saturation and allosteric effects were the
-main drivers of regulation between the two conditions in this case, with the
+main drivers of regulation between the two datasets in this case, with the
 curves for each component aligning with the ground truth shown in red.
@@
 uncertainty, including uncertainty about the true values of the flux in each
-condition, and propagates this uncertainty into the final conclusion. Our result
+dataset, and propagates this uncertainty into the final conclusion. Our result
 shows that Maud could be used in this realistic case not only to provide a user

### 2023-10-31 plot_figure_2.py
change figure names

@@ -17,7 +17,7 @@ from maud.getting_idatas import get_idata
 
 
 MAUD_OUTPUT = Path.cwd() / "results" / "methionine"
-PLOT_OUTPUT = Path.cwd() / "figures" / "figure_2"
+PLOT_OUTPUT = Path.cwd() / "figures" / "posterior.png"
 
 plt.style.use('ipynb')
 

### 2023-10-31 plot_figure_3.py

@@ -43,7 +43,7 @@ plots.
 
 MAUD_OUTPUT_MISSING_AHCYS = Path.cwd() / "results" / "methionine_missing_ahcys"
 MAUD_OUTPUT = Path.cwd() / "results" / "methionine"
-PLOT_OUTPUT = Path.cwd() / "figures" / "figure_3.png"
+PLOT_OUTPUT = Path.cwd() / "figures" / "missing.png"
 
 colour_scheme = {
     "complete measurement set": "#001219",

### 2023-10-31 plot_figure_5.py

@@ -151,7 +151,7 @@ def plot_figure_5():
 
     plt.figure(dpi = 300)
     fig.tight_layout()
-    fig.savefig(PLOT_OUTPUT / "figure_5.png")
+    fig.savefig(PLOT_OUTPUT / "decomposition.png")
 
 if __name__ == "__main__":
     plot_figure_5()

### 2023-10-31 paper/main.qmd
paper update

--- 2023-10-31 paper/main.qmd
@@
 
-Posterior predictive checking indicated that our model achieved a good fit
-to the simulated reaction and metabolite concentration measurements, as shown
-by the graphs in the top row of figure 3. Note that the fit was good for both
-training and validation measurements.
+Posterior predictive checking indicated that our model achieved a good fit to
+the simulated reaction and metabolite concentration measurements, as shown by
+the graphs in the top row of figure @fig-posterior.
 
@@
 distribution contained enough information to make accurate out of sample
-predictions. In some cases, there were two-dimensional correlations such
-as the one shown in the bottom left of figure 3; in this case the marginal
-distribution of the two parameters is roughly banana shaped. More commonly,
-however, two-dimensional pair plots were insufficient to reveal the underlying
-correlation structure, as seen in the bottom-right plot in figure 3. This does
-not mean that the parameters were uncorrelated, but rather that the correlations
-involve more than two parameters.
+predictions. In some cases, there were two-dimensional correlations such as
+the one shown in the bottom left of figure @fig-posterior; in this case the
+marginal distribution of the two parameters is roughly banana shaped. More
+commonly, however, two-dimensional pair plots were insufficient to reveal the
+underlying correlation structure, as seen in the bottom-right plot in figure
+@fig-posterior. This does not mean that the parameters were uncorrelated, but
+rather that the correlations involve more than two parameters.
 
@@
 approximation is a suitable candidate.
+
+::: {#fig-posterior layout-ncol=1}
+
+![](../figures/posterior.png)
+
+Results of removing concentration measurements for the metabolite $ahcys_c$
+from our case study dataset.  (**Top**) A comparison of metabolite concentration
+residuals between the full measurement dataset (blue-green) and the missing-data
+dataset (grey), displayed on natural logarithmic scale. The missing-data
+model was unable to predict the withheld $ahcys_c$ concentrations.  (**Bottom
+Left**) The marginal posterior distribution for the Michaelis constant
+$K_m^{AHC1,ahcys_c}$ in each model, alongside the true parameter value used to
+generate both datasets. The true value is recovered by the complete-data model
+but not by the missing-data model.  (**Bottom Right**) The distribution of total
+log-likelihood for out-of-sample flux measurements in both models. There is a
+significant overlap between the two distributions, suggesting that removing the
+$ahcys_c$ measurement did not cause catastrophic prediction failure. However,
+overall the complete-data model tended to make better predictions.
+
+:::
 

### 2023-10-31 paper/main.qmd
change decomposition caption

--- 2023-10-31 paper/main.qmd
@@
 Left**) Comparison of marginal posterior distributions for $GNMT$ flux in
-conditions 1 and 2. (**Right**) Log-scale ratios of the regulatory elements
+datasets 1 and 2. (**Right**) Log-scale ratios of the regulatory elements
 defined in equation \eqref{eq-decomposition}. Note that the $Reversibility$
 and $k_cat$ components are excluded: this is because this reaction was modelled
-as irreversible, and $k_{cat}$ was modelled as constant across conditions.
-These plots identify why flux in condition 2 is higher than in condition 1: the
+as irreversible, and $k_{cat}$ was modelled as constant across datasets.
+These plots identify why flux in dataset 12 is higher than in dataset 1: the
 flux increase is due to allostery and saturation with no control from enzyme

### 2023-10-31 plot_figure_2.py
Update posterior plot

@@ -1,4 +1,3 @@
-import builtins
 from pathlib import Path
 import tempfile
 import shutil
@@ -10,7 +9,6 @@ import matplotlib.patches as mpatches
 import matplotlib.image as mpimg
 import numpy as np
 import pandas as pd
-import arviz as az
 
 from maud.loading_maud_inputs import load_maud_input
 from maud.getting_idatas import get_idata
@@ -19,74 +17,78 @@ from maud.getting_idatas import get_idata
 MAUD_OUTPUT = Path.cwd() / "results" / "methionine"
 PLOT_OUTPUT = Path.cwd() / "figures" / "posterior.png"
 
-plt.style.use('ipynb')
+plt.style.use("ipynb")
 
-colour_scheme = {
-    "measurement": "0.6",
-    "sample": "#8F2727"
-}
+colour_scheme = {"measurement": "#8F2727", "sample": "black"}
+
 
 def plot_flux_forest(mi, idata_methionine, axes, flux_meas_df):
     """Return forest plot axes of fluxes."""
     flux_samples = idata_methionine.posterior["flux_train"].to_dataframe().reset_index()
     flux_samples["id"] = flux_samples["experiments"] + flux_samples["reactions"]
-    flux_samples = flux_samples[["id", "flux_train"]].groupby("id") \
-                               .quantile([0.05, 0.95]) \
-                               .reset_index() \
-                               .pivot(index="id", columns="level_1", values="flux_train")
+    flux_samples = (
+        flux_samples[["id", "flux_train"]]
+        .groupby("id")
+        .quantile([0.05, 0.95])
+        .reset_index()
+        .pivot(index="id", columns="level_1", values="flux_train")
+    )
     sorted_fluxes = list(flux_meas_df.sort_values("flux").index)
     for id, sample in flux_samples.reindex(sorted_fluxes).iterrows():
-        axes[0,0].vlines(
+        axes[0, 0].vlines(
             x=id,
             ymin=sample[0.05],
             ymax=sample[0.95],
             linewidth=2.5,
             color=colour_scheme["sample"],
+            zorder=0,
         )
-    for id, meas in flux_meas_df.reindex(sorted_fluxes).iterrows():
-        axes[0,0].vlines(
-            x=id,
-            ymin=meas["flux"]-1.65*meas["uncertainty"],
[…]

### 2023-10-31 paper/main.qmd
Update paper

@@
 
-Results of removing concentration measurements for the metabolite $ahcys_c$
-from our case study dataset.  (**Top**) A comparison of metabolite concentration
-residuals between the full measurement dataset (blue-green) and the missing-data
-dataset (grey), displayed on natural logarithmic scale. The missing-data
-model was unable to predict the withheld $ahcys_c$ concentrations.  (**Bottom
-Left**) The marginal posterior distribution for the Michaelis constant
-$K_m^{AHC1,ahcys_c}$ in each model, alongside the true parameter value used to
-generate both datasets. The true value is recovered by the complete-data model
-but not by the missing-data model.  (**Bottom Right**) The distribution of total
-log-likelihood for out-of-sample flux measurements in both models. There is a
-significant overlap between the two distributions, suggesting that removing the
-$ahcys_c$ measurement did not cause catastrophic prediction failure. However,
-overall the complete-data model tended to make better predictions.
+Marginal posterior distributions from our main model run. (**Top left**)
+Comparison of posterior predictive intervals with simulated flux measurement
+values. All the flux measurements are within the predictive intervals,
+indicating a good fit. (**Top right**) Comparison of posterior predictive
+intervals with simulated concentration measurement values. These also show a
+good fit. (**Bottom left**) Pairwise marginal posterior distribution for two
+uncorrelated parameters, namely $K_m^{AHC1,hcys-L}$ and $K_m^{AHC1_ahcys}$.
+(**Bottom right**) Pairwise marginal posterior distribution for two correlated
+parameters, namely $K_m^{MAT3,atp}$ and $K_{cat}^{MAT3}$.
 

### 2023-10-31 paper/main.tex
Update paper

@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, interior hidden, breakable, enhanced, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, breakable, enhanced, boxrule=0pt, interior hidden, frame hidden, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
 
@@
 fit to the simulated reaction and metabolite concentration measurements,
-as shown by the graphs in the top row of figure 3. Note that the fit was
-good for both training and validation measurements.
+as shown by the graphs in the top row of figure
+Figure~\ref{fig-posterior}.
 
@@
 two-dimensional correlations such as the one shown in the bottom left of
-figure 3; in this case the marginal distribution of the two parameters
-is roughly banana shaped. More commonly, however, two-dimensional pair
-plots were insufficient to reveal the underlying correlation structure,
-as seen in the bottom-right plot in figure 3. This does not mean that
-the parameters were uncorrelated, but rather that the correlations
-involve more than two parameters.
+figure Figure~\ref{fig-posterior}; in this case the marginal
+distribution of the two parameters is roughly banana shaped. More
+commonly, however, two-dimensional pair plots were insufficient to
+reveal the underlying correlation structure, as seen in the bottom-right
+plot in figure Figure~\ref{fig-posterior}. This does not mean that the
+parameters were uncorrelated, but rather that the correlations involve
+more than two parameters.
 
@@
 approximation is a suitable candidate.
+
+\begin{figure}
+
+\begin{minipage}[t]{\linewidth}
+
+{\centering
+
+\raisebox{-\height}{
+
+\includegraphics{../figures/posterior.png}
+
+}
+
+}
+
+\end{minipage}%
+
+\caption{\label{fig-posterior}Marginal posterior distributions from our
+main model run. (\textbf{Top left}) Comparison of posterior predictive
+intervals with simulated flux measurement values. All the flux
+measurements are within the predictive intervals, indicating a good fit.
+(\textbf{Top right}) Comparison of posterior predictive intervals with
+simulated concentration measurement values. These also show a good fit.
+(\textbf{Bottom left}) Pairwise marginal posterior distribution for two
+uncorrelated parameters, namely \(K_m^{AHC1,hcys-L}\) and
+\(K_m^{AHC1_ahcys}\). (\textbf{Bottom right}) Pairwise marginal
+posterior distribution for two correlated parameters, namely
+\(K_m^{MAT3,atp}\) and \(K_{cat}^{MAT3}\).}
+
+\end{figure}
 
@@
 insights we used the results of our case study to explain why the flux
-of the enzyme \(GNMT\) is higher in condition 1 than in condition 2.
-GNMT is an irreversible enzyme that is homotropically activated by its
+of the enzyme \(GNMT\) is higher in dataset 1 than in dataset 12. GNMT
+is an irreversible enzyme that is homotropically activated by its
 substrate, competitively inhibited by its product and heterotropically
-inhibited. The diverse regulation makes it the ideal test case to
-elucidate regulatory changes.
+inhibited by 5,10- methylenetetrahydrofolate (mlthf). The diverse
+regulation makes it the ideal test case to elucidate regulatory changes.
 
@@
 the marginal posterior distribution of the ratio of the corresponding
-regulatory component in condition 1 compared with condition 2. A
-positive value indicates that the component was increased in condition 1
-relative to condition 2, with 0 indicating no difference. The
-probability, according to our model, of the component acting in each
-direction is given by the relative area under the curve on each side of
-the 0 point.
+regulatory component in dataset 1 compared with dataset 12. A positive
+value indicates that the component was increased in dataset 1 relative
+to dataset 12, with 0 indicating no difference. The probability,
+according to our model, of the component acting in each direction is
+given by the relative area under the curve on each side of the 0 point.
 
 Our model correctly inferred that saturation and allosteric effects were
-the main drivers of regulation between the two conditions in this case,
+the main drivers of regulation between the two datasets in this case,
 with the curves for each component aligning with the ground truth shown
@@
 sources of uncertainty, including uncertainty about the true values of
-the flux in each condition, and propagates this uncertainty into the
-final conclusion. Our result shows that Maud could be used in this
-realistic case not only to provide a user with an explanation for an
-observed difference in fluxes, but also a reasonable judgement as to the
+the flux in each dataset, and propagates this uncertainty into the final
+conclusion. Our result shows that Maud could be used in this realistic
+case not only to provide a user with an explanation for an observed
+difference in fluxes, but also a reasonable judgement as to the
 explanation's robustness.
@@
 Left}) Comparison of marginal posterior distributions for \(GNMT\) flux
-in conditions 1 and 2. (\textbf{Right}) Log-scale ratios of the
-regulatory elements defined in equation \eqref{eq-decomposition}. Note
-that the \(Reversibility\) and \(k_cat\) components are excluded: this
-is because this reaction was modelled as irreversible, and \(k_{cat}\)
-was modelled as constant across conditions. These plots identify why
-flux in condition 2 is higher than in condition 1: the flux increase is
-due to allostery and saturation with no control from enzyme
-concentration changes.}
+in datasets 1 and 2. (\textbf{Right}) Log-scale ratios of the regulatory
+elements defined in equation \eqref{eq-decomposition}. Note that the
+\(Reversibility\) and \(k_cat\) components are excluded: this is because
+this reaction was modelled as irreversible, and \(k_{cat}\) was modelled
+as constant across datasets. These plots identify why flux in dataset 12
+is higher than in dataset 1: the flux increase is due to allostery and
+saturation with no control from enzyme concentration changes.}
 

### 2023-10-31 plot_figure_3.py
Update missing measurements figure:

@@ -65,6 +65,11 @@ def plot_residual_state_variables(residual_df, ax):
             color=colour_scheme[model],
             alpha=0.7,
         )
+    rect = mpatches.Rectangle((5.5, 3.3), 6, 3.9, facecolor="none", edgecolor="orange")
+    ax.add_patch(rect)
+    ax.annotate(
+        "ahcys measurements", (12, 5), (15, 6), arrowprops={"facecolor": "black"}
+    )
     ax.axhline(y=0, xmin=0, xmax=len(residual_df) / 2, color="red")
     ax.set_ylabel("log residual plot of balanced metabolites")
     ax.get_xaxis().set_visible(False)

### 2023-10-31 paper/main.qmd
edits

--- 2023-10-31 paper/main.qmd
@@
 
-MCMC sampling for our comparison input yielded 200 samples in 625 minutes;
+MCMC sampling for our comparison input yielded 800 samples in 625 minutes;
 Laplace sampling yielded the same number of samples in only one minute.
@@
 dataset (grey), displayed on natural logarithmic scale. The missing-data
-model was unable to predict the withheld $ahcys_c$ concentrations.  (**Bottom
+model was unable to estimate the withheld $ahcys_c$ concentrations.  (**Bottom
 Left**) The marginal posterior distribution for the Michaelis constant
@@
 Importantly, this form of analysis takes into account all modelled sources of
-uncertainty, including uncertainty about the true values of the flux in each
-dataset, and propagates this uncertainty into the final conclusion. Our result
-shows that Maud could be used in this realistic case not only to provide a user
-with an explanation for an observed difference in fluxes, but also a reasonable
-judgement as to the explanation's robustness.
+uncertainty, including uncertainty about the measured values of the flux in each
+dataset. Our result shows that Maud could be used in this realistic case not
+only to provide a user with an explanation for an observed difference in fluxes,
+but also a reasonable judgement as to the explanation's robustness.
 

### 2023-10-31 paper/main.qmd
Update paper

--- 2023-10-31 paper/main.qmd
@@
 
-![](./figures/methionine-reactions.png)
+![](../figures/methionine-reactions.png)
 
@@
 
-![](./figures/missing.png)
+![](../figures/missing.png)
 
@@
 
-![](./figures/decomposition.png)
+![](../figures/decomposition.png)
 
@@
 
-Maud is a command line application implementing Bayesian inference for
-a wide range of realistic kinetic models. Maud is written in Python
-[@vanrossumPythonReferenceManual2009], designed for use on Windows, macOS
-and Linux, registered on the Python Package Index as `maud-metabolic-models`,
-documented at <https://maud-metabolic-models.readthedocs.io> and actively
-developed and maintained at <https://github.com/biosustain/Maud/>.
+Maud is a command line application implementing Bayesian inference
+for a wide range of realistic kinetic models. Maud is written in
+Python [@vanrossumPythonReferenceManual2009], designed for use
+on Windows, macOS and Linux, is `pip` installable from the Python
+Package Index as `maud-metabolic-models`, documented at <https://maud-metabolic-models.readthedocs.io> 
+and actively developed and maintained at <https://github.com/biosustain/Maud/>.
 
 To use Maud, a user must first collate appropriate input information, represent
-it in files with Maud’s required formats (see section (REFERENCE)below).  Maud’s
-command line interface provides commands for inference, simulation and making
-out-of-sample predictions. Results are stored in files, using a structured,
-interoperable format.
+it in files with Maud’s required formats (see supporting information section on
+input format).  Maud’s command line interface provides commands for inference
+via a range of algorithms including adaptive Hamiltonian Monte Carlo, as well as
+commands for simulation and making out-of-sample predictions. Results are stored
+in files, using a structured, interoperable format.
 
-## Input format
+As well as parameter values, Maud also performs inference for derived quantities
+including the components of the regulatory decomposition described below in
+equation \eqref{eq-decomposition}, log likelihoods, simulated measurement values
+and metabolic control analysis coefficients [@kacser_control_1973].
 
-Maud inputs are structured directories, somewhat inspired by the PEtab
-format [@SchmiesterSch2021]. A Maud input directory must contain a toml
-[@preston-wernertomandgedampradyunTOMLSpecification0rc2020] file called
-`config.toml` which gives the input a name, configures how Maud will be run
-and tells Maud where to find the other files, allowing these to have custom
-names. It must also include a file containing a kinetic model definition,
-a file specifying information about parameters and a file with information
-experiments. The required structure of these files is documented at <https://
-maud-metabolic-models.readthedocs.io/en/latest/inputting.html>. The input is
-validated against a Pydantic [@pydanticdevelopersPydantic2022] data model which can be
-inspected at <https://github.com/biosustain/Maud/tree/main/maud/data_model>.
-
-We chose to implement a custom input format despite the existence of standard
-formats in similar areas, including SBML [@keatingSBMLLevelExtensible2020]
-and PEtab [@SchmiesterSch2021]. This choice was partly motivated by the need
-to ensure flexibility as Maud was developed, but there are also features of
-SBML and PEtab that make them structurally unsuitable in this context. Our
-requirements for an input format included that it be mathematics-free, so that
-all mathematical details are encapsulated in source code, and that it has a
-detailed, verifiable structure. These requirements made toml more attractive
-than SBML: toml is easier for humans to read and edit and can straightforwardly
-be validated using tools like Pydantic. Further, an SBML representation of our
-desired input would not contain differential equations.  It would therefore not
-be interoperable with most SBML targeting software, which typically assumes that
-differential equations are available and does not know about Maud’s structure.
+In the rest of this section we describe Maud's kinetic model at high level and
+discuss Maud's statistical model, implementation and how it solves the crucial
+steady state problem. Further details about Maud's kinetic model can be found in
+the supporting information.
 
@@
 Maud uses the Python library click [@clickdevelopersClickPythonComposable2022]
-to implement a command line interface. The command line interface loads input
-files as Python dictionaries, which are parsed using the Python library `toml`
-[@pearsonTomlPythonLibrary2020] and then validated and converted into structured
-`MaudInput` objects using Pydantic [@pydanticdevelopersPydantic2022]. Maud’s
-statistical model is implemented in the probabilistic programming language Stan
-[@carpenterStanProbabilisticProgramming2017] and accessed using the interface
-cmdstanpy [@standevelopmentteamCmdStanPy2022]. For posterior sampling, Maud
-uses the `MaudInput` to create a Stan input json file and obtain configuration
-information for cmdstanpy, which it uses to (if necessary) generate a
-model executable file and then trigger posterior sampling using adaptive
-Hamiltonian Monte Carlo. When sampling is complete, Maud converts to the
-output into the standard format `InferenceData` using the Python library arviz
-[@kumarArviZUnifiedLibrary2019] and saves it as a json file, along with some
-information for debugging.
+to implement a command line interface. The command line interface loads
+input files as Python dictionaries, which are parsed using the Python library
+`toml` [@pearsonTomlPythonLibrary2020] and then validated and converted into
+structured `MaudInput` objects using Pydantic [@pydanticdevelopersPydantic2022].
+Maud’s statistical model is implemented in the probabilistic programming
+language Stan [@carpenterStanProbabilisticProgramming2017] and accessed using
+the interface cmdstanpy [@standevelopmentteamCmdStanPy2022]. For posterior
+sampling, Maud uses the `MaudInput` to create an input file for Stan and
+then trigger posterior sampling using adaptive Hamiltonian Monte Carlo. See
+@betancourtConceptualIntroductionHamiltonian2018 for more about this algorithm.
 
-Two details of Maud’s implementation are important to highlight: the method of
-posterior sampling and how Maud solves the steady state problem \eqref{eq-3}.
+When sampling is complete, Maud converts to the output into the standard format
+`InferenceData` using the Python library arviz [@kumarArviZUnifiedLibrary2019]
+and saves it as a json file, along with some information for debugging.
+This format allows for easy checking of MCMC diagnostics including divergent
+transitions, effective sample size and the improved $\hat{R}$ statistic proposed
+in @vehtariRankNormalizationFoldingLocalization2021.
 
-### Posterior sampling
+## Solving the steady state problem
 
-Although integrals of the joint probability model for kinetic models
-are typically analytically intractable, they can be approximated
-numerically using Markov Chain Monte Carlo (MCMC) and other methods.
-Maud uses MCMC primarily because there exist many methods for verifying
-that MCMC samples really do approximate the target probability
-distribution: see @vehtariRankNormalizationFoldingLocalization2021 and
-@taltsValidatingBayesianInference2018 for discussion of this point. In
-addition, there are several examples of successful Bayesian kinetic modelling
-projects using MCMC including @st.johnBayesianInferenceMetabolic2018 and
-@xingModelingKineticsLargescale2010.
+Maud's parameters are connected with measurable quantities via the steady state
+equation \eqref{eq-steady}. Posterior sampling using adaptive Hamiltonian Monte
+Carlo requires repeatedly solving equation \eqref{eq-steady} and finding the
+gradients of its solution with respect to all model parameters. This must be
+done numerically as analytic solutions are not available for realistic kinetic
+models.
 
-### Solving the steady state problem
+The speed at which this problem can be solved is tightly coupled with the
+size and complexity of metabolic network that can practically be modelled. See
+@timonenImportanceSamplingApproach2022a for more about considerations involved
+in this kind of modelling.
 
-In order to implement adaptive Hamiltonian Monte Carlo, Maud must
-repeatedly solve the steady state problem \eqref{eq-3} and find its
-gradients with respect to all parameters. To achieve this, Maud uses a
-hybrid method involving two numerical solvers from the SUNDIALS suite
-[@serbanCVODESSensitivityEnabledODE2005]: CVODES and IDAS. These solvers are
-accessed via their interface from Stan. The hybrid method follows that proposed
-by @margossianComputingSteadyStates2018, and involves numerically evolving
-the ODE system for a short period of time, then using the difference between
-the evolved and starting concentrations as the target for a numerical algebra
-solver.
+To solve equation \eqref{eq-steady} and find its gradients, Maud uses
+a hybrid method involving two numerical solvers from the SUNDIALS
+suite [@serbanCVODESSensitivityEnabledODE2005]: CVODES and IDAS via
+their interface from Stan. The hybrid method follows that proposed by
+@margossianComputingSteadyStates2018, and involves numerically evolving the ODE
+system for a short period of time, then using the difference between the evolved
+and starting concentrations as the target for a numerical algebra solver.
 

### 2023-10-31 paper/main.tex
Update paper

@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, breakable, enhanced, boxrule=0pt, interior hidden, frame hidden, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, enhanced, boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden]}{\end{tcolorbox}}\fi
 
@@
 
-\includegraphics{./figures/methionine-reactions.png}
+\includegraphics{../figures/methionine-reactions.png}
 
@@
 
-\includegraphics{./figures/missing.png}
+\includegraphics{../figures/missing.png}
 
@@
 
-\includegraphics{./figures/decomposition.png}
+\includegraphics{../figures/decomposition.png}
 

### 2023-10-31 paper/supplementary.qmd
Update paper

@@
 The results of all reported Maud runs can be found at <https://github.com/biosustain/Methionine_model/blob/main/results>.
+
+# Maud's input format
+
+Maud inputs are structured directories, somewhat inspired by the PEtab
+format [@SchmiesterSch2021]. A Maud input directory must contain a toml
+[@preston-wernertomandgedampradyunTOMLSpecification0rc2020] file called
+`config.toml` which gives the input a name, configures how Maud will be run
+and tells Maud where to find the other files, allowing these to have custom
+names. It must also include a file containing a kinetic model definition,
+a file specifying information about parameters and a file with information
+experiments. The required structure of these files is documented at <https://
+maud-metabolic-models.readthedocs.io/en/latest/inputting.html>. The input is
+validated against a Pydantic [@pydanticdevelopersPydantic2022] data model which can be
+inspected at <https://github.com/biosustain/Maud/tree/main/maud/data_model>.
+
+We chose to implement a custom input format despite the existence of standard
+formats in similar areas, including SBML [@keatingSBMLLevelExtensible2020]
+and PEtab [@SchmiesterSch2021]. This choice was partly motivated by the need
+to ensure flexibility as Maud was developed, but there are also features of
+SBML and PEtab that make them structurally unsuitable in this context. Our
+requirements for an input format included that it be mathematics-free, so that
+all mathematical details are encapsulated in source code, and that it has a
+detailed, verifiable structure. These requirements made toml more attractive
+than SBML: toml is easier for humans to read and edit and can straightforwardly
+be validated using tools like Pydantic. Further, an SBML representation of our
+desired input would not contain differential equations.  It would therefore not
+be interoperable with most SBML targeting software, which typically assumes that
+differential equations are available and does not know about Maud’s structure.
 

### 2023-10-31 paper/main.qmd
Paper updates

@@
 abstract: >
-  This paper presents Maud, a command-line application that implements
-  Bayesian statistical inference for kinetic models of biochemical metabolic
-  reaction networks. Maud takes into account quantitative information from omics
+  This paper presents Maud, a command-line application that implements Bayesian
+  statistical inference for kinetic models of biochemical metabolic reaction
+  networks. Maud takes into account quantitative information from omics
   experiments and background knowledge, as well as structural information about
-  kinetic mechanisms, regulatory interactions and enzyme knockouts. Below, we
-  review the existing options in this area, explain how Maud improves on the state
-  of the art, describe the intended modelling workflow and illustrate its use with
-  an example application.
+  kinetic mechanisms, regulatory interactions and enzyme knockouts. Our paper
+  reviews the existing options in this area, presents a case study illustrating
+  how Maud can be used to analyse a metabolic network and explains the
+  biological, statistical and computational design decisions underpinning Maud.
 number-sections: true
@@
     - id: UQ
-    - name: The University of Queensland
-    - department: Australian Institute for Bioengineering and Nanotechnology (AIBN)
-    - city: St Lucia
-    - country: Australia
+      name: The University of Queensland
+      department: Australian Institute for Bioengineering and Nanotechnology (AIBN)
+      city: St Lucia
+      country: Australia
 bibliography: bibliography.bib
@@
       hide-email: false # if enabled, hides email addresses (and fax/phone)
-      abbreviations: IR,NMR,UV
   acs-html:
@@
 can be found at <https://github.com/biosustain/Methionine_model/tree/main/data/methionine>
-and is described in supplementary information section [REFERENCE].
+and is described in supporting information section 3.
 
@@
 The prior distributions and corresponding true parameter values used in our
-case study are shown in supplementary materials section [REFERENCE]. The first
-two columns show the 1% and 99% quantiles of each marginal prior distribution.
-True parameter value are shown in column three, and the last column shows the
-z-score on log scale of the true parameter value according the marginal prior
-distribution. As can be seen from the table, there are 7 parameters for which
-the true value is outside the 1%-99% range. This is desirable, making the case
-study more realistic, because extreme deviance from the prior distribution is
-likely to occur in practice due to in vivo to in vitro measurement differences.
+case study are shown in supporting information section 3.2. They were chosen to
+reflect a plausible pre-experimental information state. In 7 cases, the marginal
+prior distribution for a parameter disagrees with the true parameter values used
+to generate the data. A similar situation is likely to occur in practice due to
+*in vivo* vs *in vitro* measurement differences.
 
@@
 regulation. Full details of all mathematical aspects of Maud’s kinetic model can
-be found in supplementary material section (REFERENCE).
+be found in supporting information section 2.
 
@@
 other parameters by solving the steady state problem. See table (REFERENCE) in
-this paper's supplementary materials for a description of all these parameters
+this paper's supporting information for a description of all these parameters
 and their dimensions. Note that some quantities in Maud’s kinetic model are
@@
 
-The document `supplementary.pdf` contains the following information:
+The analyses described in this paper, as well as instructions for reproducing
+our results, figures and paper, can be found at <https://biosustain/Methionine_model>.
 
-- Description of Maud's kinetic model including parameter dimensions and rate
-equations.
+The document `supporting_information.pdf` contains the following information:
+
+- Description of Maud's input format
+- Description of Maud's kinetic model including rate equations as well as
+parameters and their dimensions.
 - Details of the procedure used to generate our case study results, including
@@
 out computation.
-- Specification of the prior distribution used for the case studies
+- Specification of the prior distribution used for the case studies.
 

### 2023-10-31 paper/main.tex
Paper updates

@@
 \affiliation{DTU Biosustain, DTU, Kongens Lyngby, Denmark}
-\affiliation{   }
+\affiliation{Australian Institute for Bioengineering and Nanotechnology
+(AIBN), The University of Queensland, St Lucia, Australia}
 
@@
 structural information about kinetic mechanisms, regulatory interactions
-and enzyme knockouts. Below, we review the existing options in this
-area, explain how Maud improves on the state of the art, describe the
-intended modelling workflow and illustrate its use with an example
-application.
+and enzyme knockouts. Our paper reviews the existing options in this
+area, presents a case study illustrating how Maud can be used to analyse
+a metabolic network and explains the biological, statistical and
+computational design decisions underpinning Maud.
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, enhanced, boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, frame hidden, enhanced, boxrule=0pt, breakable]}{\end{tcolorbox}}\fi
 
@@
 \url{https://github.com/biosustain/Methionine_model/tree/main/data/methionine}
-and is described in supplementary information section {[}REFERENCE{]}.
+and is described in supporting information section 3.
 
@@
 The prior distributions and corresponding true parameter values used in
-our case study are shown in supplementary materials section
-{[}REFERENCE{]}. The first two columns show the 1\% and 99\% quantiles
-of each marginal prior distribution. True parameter value are shown in
-column three, and the last column shows the z-score on log scale of the
-true parameter value according the marginal prior distribution. As can
-be seen from the table, there are 7 parameters for which the true value
-is outside the 1\%-99\% range. This is desirable, making the case study
-more realistic, because extreme deviance from the prior distribution is
-likely to occur in practice due to in vivo to in vitro measurement
-differences.
+our case study are shown in supporting information section 3.2. They
+were chosen to reflect a plausible pre-experimental information state.
+In 7 cases, the marginal prior distribution for a parameter disagrees
+with the true parameter values used to generate the data. A similar
+situation is likely to occur in practice due to \emph{in vivo} vs
+\emph{in vitro} measurement differences.
 
@@
 
-MCMC sampling for our comparison input yielded 200 samples in 625
+MCMC sampling for our comparison input yielded 800 samples in 625
 minutes; Laplace sampling yielded the same number of samples in only one
@@
 dataset (grey), displayed on natural logarithmic scale. The missing-data
-model was unable to predict the withheld \(ahcys_c\) concentrations.
+model was unable to estimate the withheld \(ahcys_c\) concentrations.
 (\textbf{Bottom Left}) The marginal posterior distribution for the
@@
 Importantly, this form of analysis takes into account all modelled
-sources of uncertainty, including uncertainty about the true values of
-the flux in each dataset, and propagates this uncertainty into the final
-conclusion. Our result shows that Maud could be used in this realistic
-case not only to provide a user with an explanation for an observed
-difference in fluxes, but also a reasonable judgement as to the
-explanation's robustness.
+sources of uncertainty, including uncertainty about the measured values
+of the flux in each dataset. Our result shows that Maud could be used in
+this realistic case not only to provide a user with an explanation for
+an observed difference in fluxes, but also a reasonable judgement as to
+the explanation's robustness.
 
@@
 \citep{vanrossumPythonReferenceManual2009}, designed for use on Windows,
-macOS and Linux, registered on the Python Package Index as
-\texttt{maud-metabolic-models}, documented at
+macOS and Linux, is \texttt{pip} installable from the Python Package
+Index as \texttt{maud-metabolic-models}, documented at
 \url{https://maud-metabolic-models.readthedocs.io} and actively
@@
 To use Maud, a user must first collate appropriate input information,
-represent it in files with Maud's required formats (see section
-(REFERENCE)below). Maud's command line interface provides commands for
-inference, simulation and making out-of-sample predictions. Results are
-stored in files, using a structured, interoperable format.
-
-\hypertarget{input-format}{%
-\subsection{Input format}\label{input-format}}
-
-Maud inputs are structured directories, somewhat inspired by the PEtab
-format \citep{SchmiesterSch2021}. A Maud input directory must contain a
-toml \citep{preston-wernertomandgedampradyunTOMLSpecification0rc2020}
-file called \texttt{config.toml} which gives the input a name,
-configures how Maud will be run and tells Maud where to find the other
-files, allowing these to have custom names. It must also include a file
-containing a kinetic model definition, a file specifying information
-about parameters and a file with information experiments. The required
-structure of these files is documented at \textless https://
-maud-metabolic-models.readthedocs.io/en/latest/inputting.html\textgreater.
-The input is validated against a Pydantic
-\citep{pydanticdevelopersPydantic2022} data model which can be inspected
-at \url{https://github.com/biosustain/Maud/tree/main/maud/data_model}.
-
-We chose to implement a custom input format despite the existence of
-standard formats in similar areas, including SBML
-\citep{keatingSBMLLevelExtensible2020} and PEtab
-\citep{SchmiesterSch2021}. This choice was partly motivated by the need
-to ensure flexibility as Maud was developed, but there are also features
-of SBML and PEtab that make them structurally unsuitable in this
-context. Our requirements for an input format included that it be
-mathematics-free, so that all mathematical details are encapsulated in
-source code, and that it has a detailed, verifiable structure. These
-requirements made toml more attractive than SBML: toml is easier for
-humans to read and edit and can straightforwardly be validated using
-tools like Pydantic. Further, an SBML representation of our desired
-input would not contain differential equations. It would therefore not
-be interoperable with most SBML targeting software, which typically
-assumes that differential equations are available and does not know
-about Maud's structure.
+represent it in files with Maud's required formats (see supporting
+information section on input format). Maud's command line interface
+provides commands for inference via a range of algorithms including
+adaptive Hamiltonian Monte Carlo, as well as commands for simulation and
+making out-of-sample predictions. Results are stored in files, using a
+structured, interoperable format.
+
+As well as parameter values, Maud also performs inference for derived
+quantities including the components of the regulatory decomposition
+described below in equation \eqref{eq-decomposition}, log likelihoods,
+simulated measurement values and metabolic control analysis coefficients
+\citep{kacser_control_1973}.
+
+In the rest of this section we describe Maud's kinetic model at high
+level and discuss Maud's statistical model, implementation and how it
+solves the crucial steady state problem. Further details about Maud's
+kinetic model can be found in the supporting information.
 
@@
 details of all mathematical aspects of Maud's kinetic model can be found
-in supplementary material section (REFERENCE).
+in supporting information section 2.
 
@@
 derived from the values of other parameters by solving the steady state
-problem. See table (REFERENCE) in this paper's supplementary materials
+problem. See table (REFERENCE) in this paper's supporting information
 for a description of all these parameters and their dimensions. Note
@@
 interface cmdstanpy \citep{standevelopmentteamCmdStanPy2022}. For
-posterior sampling, Maud uses the \texttt{MaudInput} to create a Stan
-input json file and obtain configuration information for cmdstanpy,
-which it uses to (if necessary) generate a model executable file and
-then trigger posterior sampling using adaptive Hamiltonian Monte Carlo.
+posterior sampling, Maud uses the \texttt{MaudInput} to create an input
+file for Stan and then trigger posterior sampling using adaptive
+Hamiltonian Monte Carlo. See
+\citet{betancourtConceptualIntroductionHamiltonian2018} for more about
+this algorithm.
+
 When sampling is complete, Maud converts to the output into the standard
@@
 \citep{kumarArviZUnifiedLibrary2019} and saves it as a json file, along
-with some information for debugging.
-
-Two details of Maud's implementation are important to highlight: the
-method of posterior sampling and how Maud solves the steady state
-problem \eqref{eq-3}.
-
-\hypertarget{posterior-sampling}{%
-\subsubsection{Posterior sampling}\label{posterior-sampling}}
-
-Although integrals of the joint probability model for kinetic models are
-typically analytically intractable, they can be approximated numerically
-using Markov Chain Monte Carlo (MCMC) and other methods. Maud uses MCMC
-primarily because there exist many methods for verifying that MCMC
-samples really do approximate the target probability distribution: see
-\citet{vehtariRankNormalizationFoldingLocalization2021} and
-\citet{taltsValidatingBayesianInference2018} for discussion of this
-point. In addition, there are several examples of successful Bayesian
-kinetic modelling projects using MCMC including
-\citet{st.johnBayesianInferenceMetabolic2018} and
-\citet{xingModelingKineticsLargescale2010}.
+with some information for debugging. This format allows for easy
+checking of MCMC diagnostics including divergent transitions, effective
+sample size and the improved \(\hat{R}\) statistic proposed in
+\citet{vehtariRankNormalizationFoldingLocalization2021}.
 
 \hypertarget{solving-the-steady-state-problem}{%
-\subsubsection{Solving the steady state
+\subsection{Solving the steady state
 problem}\label{solving-the-steady-state-problem}}
 
-In order to implement adaptive Hamiltonian Monte Carlo, Maud must
-repeatedly solve the steady state problem \eqref{eq-3} and find its
-gradients with respect to all parameters. To achieve this, Maud uses a
+Maud's parameters are connected with measurable quantities via the
+steady state equation \eqref{eq-steady}. Posterior sampling using
+adaptive Hamiltonian Monte Carlo requires repeatedly solving equation
+\eqref{eq-steady} and finding the gradients of its solution with respect
+to all model parameters. This must be done numerically as analytic
+solutions are not available for realistic kinetic models.
+
+The speed at which this problem can be solved is tightly coupled with
+the size and complexity of metabolic network that can practically be
+modelled. See \citet{timonenImportanceSamplingApproach2022a} for more
+about considerations involved in this kind of modelling.
+
+To solve equation \eqref{eq-steady} and find its gradients, Maud uses a
 hybrid method involving two numerical solvers from the SUNDIALS suite
-\citep{serbanCVODESSensitivityEnabledODE2005}: CVODES and IDAS. These
-solvers are accessed via their interface from Stan. The hybrid method
-follows that proposed by \citet{margossianComputingSteadyStates2018},
-and involves numerically evolving the ODE system for a short period of
-time, then using the difference between the evolved and starting
-concentrations as the target for a numerical algebra solver.
+\citep{serbanCVODESSensitivityEnabledODE2005}: CVODES and IDAS via their
+interface from Stan. The hybrid method follows that proposed by
+\citet{margossianComputingSteadyStates2018}, and involves numerically
+evolving the ODE system for a short period of time, then using the
+difference between the evolved and starting concentrations as the target
+for a numerical algebra solver.
 
@@
 
-The document \texttt{supplementary.pdf} contains the following
+The analyses described in this paper, as well as instructions for
+reproducing our results, figures and paper, can be found at
+\url{https://biosustain/Methionine_model}.
+
+The document \texttt{supporting\_information.pdf} contains the following
 information:
@@
 \item
-  Description of Maud's kinetic model including parameter dimensions and
-  rate equations.
+  Description of Maud's input format
+\item
+  Description of Maud's kinetic model including rate equations as well
+  as parameters and their dimensions.
 \item
@@
 \item
-  Specification of the prior distribution used for the case studies
+  Specification of the prior distribution used for the case studies.
 \end{itemize}

### 2023-10-31 paper/supporting_information.qmd
Paper updates

--- 2023-10-31 paper/supplementary.qmd
@@
 bibliography: bibliography.bib
+number-sections: true
 format:
@@
 ---
+
+This document provides information in support of our article "Bayesian
+regression facilitates quantitative modelling of cell metabolism".
 
@@
 a file specifying information about parameters and a file with information
-experiments. The required structure of these files is documented at <https://
-maud-metabolic-models.readthedocs.io/en/latest/inputting.html>. The input is
-validated against a Pydantic [@pydanticdevelopersPydantic2022] data model which can be
-inspected at <https://github.com/biosustain/Maud/tree/main/maud/data_model>.
+experiments. The required structure of these files is documented at <https://maud-metabolic-models.readthedocs.io/en/latest/inputting.html>.
+The input is validated against a Pydantic [@pydanticdevelopersPydantic2022] 
+data model.
 
@@
 @tbl-case-study-params shows the prior distributions we used for independent
-parameters.
+parameters. The first two columns show the 1% and 99% quantiles of each
+marginal prior distribution. True parameter value are shown in column three,
+and the last column shows the z-score on log scale of the true parameter value
+according the marginal prior distribution. As can be seen from the table, there
+are 7 parameters for which the true value is outside the 1%-99% range.
 
@@
 
-We conducted adaptive Hamiltonian Monte Carlo sampling for the full and missing-
--data datasets. For the full dataset we obtained 1000 post-warmup samples each
-from 4 independent Markov chains after 1000 warm-up samples and "hot-starting"
-with a mass metric output by a previous model run.
+We conducted adaptive Hamiltonian Monte Carlo sampling for the full and
+missing-data datasets. For the full dataset we obtained 1000 post-warmup
+samples each from 4 independent Markov chains after 1000 warm-up samples and
+“hot-starting” with a mass metric output by a previous model run.
 
@@
 
-To compare MCMC sampling with Laplace approximation we used a different model.
- The full Maud input folders can be found at <https://github.com/biosustain/Methionine_model/tree/main/data/example_ode> 
-and <https://github.com/biosustain/Methionine_model/tree/main/data/example_ode_laplace>
+To compare MCMC sampling with Laplace approximation we used a different model
+with fewer parameters and state variables. This model was chosen because
+we were not able to generate results for our methionine model using Laplace
+approximation. The simpler case still serves to illustrate the general issues
+with approximating the posterior distributions of Bayesian kinetic models using
+the Laplace method, and that the associated numerical instability is another
+reason to prefer other methods where possible.
+
+The full Maud input folders used for our Laplace approximation case study can
+be found at <https://github.com/biosustain/Methionine_model/tree/main/data/example_ode>
+and <https://github.com/biosustain/Methionine_model/tree/main/data/example_ode_laplace>.
 

### 2023-10-31 paper/main.qmd
Update paper

@@
         country: Denmark
+    attributes:
+      corresponding: true
+      equal-contributor: true
+    note: These authors contributed equally to this research.
   - name: "Nicholas Luke Cowie"
     affiliations: 
-    - ref: biosustain
+      - id: biosustain
+        name: DTU
+        department: DTU Biosustain
+        city: Kongens Lyngby
+        country: Denmark
+    attributes:
+      equal-contributor: true
+    note: These authors contributed equally to this research.
   - name: "Lars Keld Nielsen"
     affiliations: 
-    - ref: biosustain
-    - id: UQ
-      name: The University of Queensland
+      - name: The University of Queensland
         department: Australian Institute for Bioengineering and Nanotechnology (AIBN)
@@
         country: Australia
+      - id: biosustain
+        name: DTU
+        department: DTU Biosustain
+        city: Kongens Lyngby
+        country: Denmark
 bibliography: bibliography.bib

### 2023-10-31 paper/main.tex
Update paper

@@
 
+\altaffiliation{These authors contributed equally to this research.}
+
 \email{tedgro@biosustain.dtu.dk}
@@
 
+\altaffiliation{These authors contributed equally to this research.}
+
 \author{Lars Keld Nielsen}
-\affiliation{DTU Biosustain, DTU, Kongens Lyngby, Denmark}
 \affiliation{Australian Institute for Bioengineering and Nanotechnology
 (AIBN), The University of Queensland, St Lucia, Australia}
+\affiliation{DTU Biosustain, DTU, Kongens Lyngby, Denmark}
 
@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, frame hidden, enhanced, boxrule=0pt, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, sharp corners]}{\end{tcolorbox}}\fi
 

### 2023-10-31 paper/main.qmd
Paper update

@@
 The parameters must include sufficient boundary concentrations and fluxes to
-solve \eqref{eq-1}.
+solve equation \eqref{eq-1}.
 
@@
 
-In a fermentation context, \eqref{eq-3} captures the rapid kinetics inside
-the cell, while another set of ODEs would be used to describe the external
-substrate and product concentrations, which could act as boundary parameters
-to \eqref{eq-3}.
+In a fermentation context, equation \eqref{eq-3} captures the rapid kinetics
+inside the cell, while another set of ODEs would be used to describe the
+external substrate and product concentrations, which could act as boundary
+parameters to equation \eqref{eq-3}.
 
@@
 the set of potential models that describe the observed data. As we demonstrate
-in our case study below, capturing this correlation structure is difficult
-outside a Bayesian context.
+in @sec-laplace below, capturing this correlation structure is difficult outside
+a fully Bayesian context.
 
@@
 without the use of MCMC. Examples of alternative inference methods include
-variational inference as in @st.johnBayesianInferenceMetabolic2018, rejection
-sampling and approximate Bayesian computation @saa_construction_2016 and Laplace
-approximation, in which the Fisher information matrix is used to calculate a
-normal approximation around the maximum a posteriori parameter configuration
-[@liebermeister_model_2021; @gopalakrishnan_k-fit_2020; @stapor_pesto_2018;
-@raue_joining_2013]. Non-MCMC-based Bayesian kinetic models have limited
-utility because they lack reliable diagnostic tools for verifying that their
-results approximate the target posterior distribution. This is a problem because
-realistic kinetic models tend to induce highly correlated, non-Gaussian, joint
-probability distributions [@gutenkunst_2007; @stapor_pesto_2018].
+variational inference [@st.johnBayesianInferenceMetabolic2018], rejection
+sampling and approximate Bayesian computation [@saa_construction_2016] and
+Laplace approximation, in which the Fisher information matrix is used to
+calculate a normal approximation around the maximum a posteriori parameter
+configuration [@liebermeister_model_2021; @gopalakrishnan_k-fit_2020;
+@stapor_pesto_2018; @raue_joining_2013]. Non-MCMC-based Bayesian kinetic
+models have limited utility because they lack reliable diagnostic tools for
+verifying that their results approximate the target posterior distribution.
+This is a problem because realistic kinetic models tend to induce highly
+correlated, non-Gaussian, joint probability distributions [@gutenkunst_2007;
+@stapor_pesto_2018].
 
-Our application Maud is the first Bayesian kinetic model to combine biologically
-realistic mechanistic accuracy---including accurate rate laws, post-
-translational modification and thermodynamics---with fast, robust MCMC sampling
-using adaptive Hamiltonian Monte Carlo. Further, Maud is a general-purpose
-application that can be used to fit a wide range of Bayesian kinetic models.
+Our application Maud provides the first Bayesian kinetic model to combine
+biologically realistic mechanistic accuracy---including accurate rate laws,
+post- translational modification and thermodynamics---with fast, robust
+MCMC sampling using adaptive Hamiltonian Monte Carlo. Further, Maud is a
+general-purpose application that can be used to fit a wide range of Bayesian
+kinetic models.
 
@@
 To demonstrate our application’s capabilities, we used Maud to analyse an
-artificial dataset based on the human methionine cycle. We first used Maud
-to generate simulated training and validation measurements based on plausible
-parameter values, then performed posterior sampling. Next, we used Maud to
-predict the validation data.
+artificial dataset based on the human methionine cycle. We generated this
+dataset using Maud, by simulating measurements for a set of training and
+validation conditions based on plausible parameter values. Next, we performed
+posterior inference for the training measurements and prediction for both
+training and validation measurements.
 
-To show that Maud is robust to missing measurements we compared the results
-of fitting the full dataset with an intentionally incomplete dataset. To
+We investigated Maud's sensitivity to missing measurements by comparing the
+results of fitting a full dataset with an intentionally incomplete dataset. To
 demonstrate why a full Bayesian approach is preferable to an approach based on a
-Laplace approximation of the posterior distribution, we also fit our methionine
-cycle model using the latter method and compared the results with MCMC sampling.
+Laplace approximation of the posterior distribution, we compared the results of
+analysing a representative metabolic network using both methods.
 
-Finally, we investigated our results to find out what our model learned about
-the contributions of different regulatory factors to the flux through GNMT, an
-important reaction. This analysis illustrates how Maud can be used to generate
-actionable insights about metabolism without the need for further statistical
-analysis.
+Finally, we dug into our results to find out what our full dataset methionine
+model learned about the contributions of different regulatory factors to the
+flux through GNMT, an important reaction. This analysis illustrates how Maud can
+be used to generate actionable insights about metabolism without the need for
+further statistical analysis.
 

### 2023-10-31 paper/main.qmd
paper update

@@
 
-The training testing split was selected to achieve a large difference between
-the fluxes of the training and testing dataset. The split was determined as
-we are interested in showing how our model can fit to varied conditions, and
+The split between training and validation conditions was selected to achieve a
+large difference in fluxes between the two categories. The split was determined
+as we are interested in showing how our model can fit to varied conditions, and
 conditions closer to the training set are likely to be predicted well without
@@
 prior distribution for a parameter disagrees with the true parameter values used
-to generate the data. A similar situation is likely to occur in practice due to
+to generate the data; a similar situation is likely to occur in practice due to
 *in vivo* vs *in vitro* measurement differences.

### 2023-10-31 paper/main.tex
paper update

@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, boxrule=0pt, sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, interior hidden]}{\end{tcolorbox}}\fi
 
@@
 The parameters must include sufficient boundary concentrations and
-fluxes to solve \eqref{eq-1}.
+fluxes to solve equation \eqref{eq-1}.
 
@@
 
-In a fermentation context, \eqref{eq-3} captures the rapid kinetics
-inside the cell, while another set of ODEs would be used to describe the
-external substrate and product concentrations, which could act as
-boundary parameters to \eqref{eq-3}.
+In a fermentation context, equation \eqref{eq-3} captures the rapid
+kinetics inside the cell, while another set of ODEs would be used to
+describe the external substrate and product concentrations, which could
+act as boundary parameters to equation \eqref{eq-3}.
 
@@
 correlation structure represents the set of potential models that
-describe the observed data. As we demonstrate in our case study below,
-capturing this correlation structure is difficult outside a Bayesian
-context.
+describe the observed data. As we demonstrate in
+Section~\ref{sec-laplace} below, capturing this correlation structure is
+difficult outside a fully Bayesian context.
 
@@
 models without the use of MCMC. Examples of alternative inference
-methods include variational inference as in
-\citet{st.johnBayesianInferenceMetabolic2018}, rejection sampling and
-approximate Bayesian computation \citet{saa_construction_2016} and
+methods include variational inference
+\citep{st.johnBayesianInferenceMetabolic2018}, rejection sampling and
+approximate Bayesian computation \citep{saa_construction_2016} and
 Laplace approximation, in which the Fisher information matrix is used to
@@
 
-Our application Maud is the first Bayesian kinetic model to combine
-biologically realistic mechanistic accuracy---including accurate rate
-laws, post- translational modification and thermodynamics---with fast,
-robust MCMC sampling using adaptive Hamiltonian Monte Carlo. Further,
-Maud is a general-purpose application that can be used to fit a wide
-range of Bayesian kinetic models.
+Our application Maud provides the first Bayesian kinetic model to
+combine biologically realistic mechanistic accuracy---including accurate
+rate laws, post- translational modification and thermodynamics---with
+fast, robust MCMC sampling using adaptive Hamiltonian Monte Carlo.
+Further, Maud is a general-purpose application that can be used to fit a
+wide range of Bayesian kinetic models.
 
@@
 To demonstrate our application's capabilities, we used Maud to analyse
-an artificial dataset based on the human methionine cycle. We first used
-Maud to generate simulated training and validation measurements based on
-plausible parameter values, then performed posterior sampling. Next, we
-used Maud to predict the validation data.
-
-To show that Maud is robust to missing measurements we compared the
-results of fitting the full dataset with an intentionally incomplete
+an artificial dataset based on the human methionine cycle. We generated
+this dataset using Maud, by simulating measurements for a set of
+training and validation conditions based on plausible parameter values.
+Next, we performed posterior inference for the training measurements and
+prediction for both training and validation measurements.
+
+We investigated Maud's sensitivity to missing measurements by comparing
+the results of fitting a full dataset with an intentionally incomplete
 dataset. To demonstrate why a full Bayesian approach is preferable to an
 approach based on a Laplace approximation of the posterior distribution,
-we also fit our methionine cycle model using the latter method and
-compared the results with MCMC sampling.
-
-Finally, we investigated our results to find out what our model learned
-about the contributions of different regulatory factors to the flux
-through GNMT, an important reaction. This analysis illustrates how Maud
-can be used to generate actionable insights about metabolism without the
-need for further statistical analysis.
+we compared the results of analysing a representative metabolic network
+using both methods.
+
+Finally, we dug into our results to find out what our full dataset
+methionine model learned about the contributions of different regulatory
+factors to the flux through GNMT, an important reaction. This analysis
+illustrates how Maud can be used to generate actionable insights about
+metabolism without the need for further statistical analysis.
 
@@
 
-The training testing split was selected to achieve a large difference
-between the fluxes of the training and testing dataset. The split was
-determined as we are interested in showing how our model can fit to
-varied conditions, and conditions closer to the training set are likely
-to be predicted well without necessarily learning the system.
+The split between training and validation conditions was selected to
+achieve a large difference in fluxes between the two categories. The
+split was determined as we are interested in showing how our model can
+fit to varied conditions, and conditions closer to the training set are
+likely to be predicted well without necessarily learning the system.
 
@@
 In 7 cases, the marginal prior distribution for a parameter disagrees
-with the true parameter values used to generate the data. A similar
+with the true parameter values used to generate the data; a similar
 situation is likely to occur in practice due to \emph{in vivo} vs

### 2023-10-31 paper/main.qmd
add methionine references

--- 2023-10-31 paper/main.qmd
@@
 the methionine cycle is highly regulated, with 6 known allosteric effectors
-(REFERENCE). This complex regulation means that quantitative modelling of the
+[@martinov2000; @nijhout2006; @korendyaseva_allosteric_2008]. This complex regulation means that quantitative modelling of the
 methionine cycle requires a detailed kinetic model: this is why we chose it as a

### 2023-10-31 paper/main.tex

@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, boxrule=0pt, sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, interior hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, boxrule=0pt, enhanced, breakable, frame hidden, borderline west={3pt}{0pt}{shadecolor}, interior hidden]}{\end{tcolorbox}}\fi
 
@@
 
-The split between training and validation conditions was selected to
-achieve a large difference in fluxes between the two categories. The
-split was determined as we are interested in showing how our model can
-fit to varied conditions, and conditions closer to the training set are
-likely to be predicted well without necessarily learning the system.
+The training testing split was selected to achieve a large difference
+between the fluxes of the training and testing dataset. The split was
+determined as we are interested in showing how our model can fit to
+varied conditions, and conditions closer to the training set are likely
+to be predicted well without necessarily learning the system.
 
@@
 In 7 cases, the marginal prior distribution for a parameter disagrees
-with the true parameter values used to generate the data; a similar
+with the true parameter values used to generate the data. A similar
 situation is likely to occur in practice due to \emph{in vivo} vs

### 2023-10-31 paper/main.tex

--- 2023-10-31 paper/main.tex
@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, boxrule=0pt, sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, interior hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, breakable]}{\end{tcolorbox}}\fi
 
@@
 simultaneously, the methionine cycle is highly regulated, with 6 known
-allosteric effectors (REFERENCE). This complex regulation means that
-quantitative modelling of the methionine cycle requires a detailed
-kinetic model: this is why we chose it as a case study for Maud.
+allosteric effectors
+\citep{martinov2000, nijhout2006, korendyaseva_allosteric_2008}. This
+complex regulation means that quantitative modelling of the methionine
+cycle requires a detailed kinetic model: this is why we chose it as a
+case study for Maud.
 

### 2023-10-31 plot_figure_4.py
fix laplace plot

@@ -36,7 +36,7 @@ from maud.loading_maud_inputs import load_maud_input
 from maud.getting_idatas import get_idata
 
 
-plt.style.use('ipynb')
+plt.style.use("ipynb")
 
 
 HELP_MSG = """
@@ -52,231 +52,311 @@ MAUD_OUTPUT_LAPLACE_TRAINING = Path.cwd() / "results" / "example_ode_laplace"
 MAUD_OUTPUT_VALIDATION = MAUD_OUTPUT_LAPLACE_TRAINING / "validation"
 PLOT_OUTPUT = Path.cwd() / "figures" / "laplace.png"
 
-colour_scheme = {
-    "measurement": "0.2",
-    "laplace": "#005F73",
-    "posterior": "#001219"
-}
+colour_scheme = {"measurement": "0.2", "laplace": "#005F73", "posterior": "#001219"}
 
 OFFSET = 0.05
-cm = 1/2.54  # centimeters in inches
+cm = 1 / 2.54  # centimeters in inches
+
 
 def plot_lp_testing(axes, lp):
-    color_map = {
-        "posterior": "#001219",
-        "laplace_approximation": "#005F73"
-    }
+    color_map = {"posterior": "#001219", "laplace_approximation": "#005F73"}
     sns.histplot(
-    data=lp, x="lp", hue="sample",
-    log_scale=False, element="step", fill=False,
-    cumulative=True, stat="probability", marker="",
-    ax=axes[0,0],
-    palette=color_map,
+        data=lp,
+        x="lp",
+        hue="sample",
+        log_scale=False,
+        element="step",
+        fill=False,
+        cumulative=True,
+        stat="probability",
+        marker="",
+        ax=axes[0, 0],
+        palette=color_map,
     )
-    axes[0,0].get_legend().remove()
-    axes[0,0] = axes[0,0].twinx()
+    axes[0, 0].get_legend().remove()
+    axes[0, 0] = axes[0, 0].twinx()
     sns.histplot(
-    data=lp, x="lp", hue="sample",
-    cumulative=False,
-    ax=axes[0,0],
-    palette=color_map,
+        data=lp,
+        x="lp",
+        hue="sample",
+        cumulative=False,
+        ax=axes[0, 0],
+        palette=color_map,
     )
-    axes[0,0].set(xlim=(-500, -0))
-    axes[0,0].get_legend().remove()
+    axes[0, 0].set(xlim=(-500, -0))
+    axes[0, 0].get_legend().remove()
[…]

### 2023-10-31 paper/main.qmd
Update paper

@@
 the simulated reaction and metabolite concentration measurements, as shown by
-the graphs in the top row of figure @fig-posterior.
+the graphs in the top row of @fig-posterior.
 
@@
 predictions. In some cases, there were two-dimensional correlations such as
-the one shown in the bottom left of figure @fig-posterior; in this case the
+the one shown in the bottom left of @fig-posterior; in this case the
 marginal distribution of the two parameters is roughly banana shaped. More
@@
 
-Overall, our results show that Maud can fit a realistic pathway-sized dataset.
-This was achieved without fixing the marginal values of kinetic parameters: the
-information required to make good predictions was contained in the correlation
-structure of the joint posterior distribution. This finding is consistent with
-previous analyses of biological systems that found they are “sloppy”, that is,
-sensitive to parameter combinations rather than marginal parameter values, with
-important combinations, scales and regions of sensitivity being difficult to
-ascertain in advance [@gutenkunst_2007; @poirier_revising_1998].
+These results show that Maud can achieve an adequate fit to a realistic
+pathway-sized dataset. This was achieved without fixing the marginal values
+of kinetic parameters: the information required to make good predictions was
+contained in the correlation structure of the joint posterior distribution.
+This finding is consistent with previous analyses of biological systems that
+found they are “sloppy”, that is, sensitive to parameter combinations rather
+than marginal parameter values, with important combinations, scales and regions
+of sensitivity being difficult to ascertain in advance [@gutenkunst_2007;
+@poirier_revising_1998].
 
@@
 distribution manifests not only in the parameter space, but also in the
-measurement space. Figure @fig-laplace frame B compares the 5%-95% interval
-for flux measurement log likelihoods in the true posterior with the Laplace
-approximation; lower log likelihood values indicate that the modelled and
-measured values are further away. The graph shows that the Laplace approximation
-yielded significantly worse predictions than the true posterior, even for the
-training data.
+measurement space. The upper right (training data) and lower left (validation
+data) plots in @fig-laplace show 5%-95% marginal flux measurement log likelihood
+intervals for both methods; lower log likelihood values indicate that the
+modelled and measured values are further away. The graph shows that the Laplace
+approximation yielded significantly worse predictions than the true posterior,
+even for the training data.
 
@@
 of two Michaelis-Menten constants $K_{m}^{A,r1}$ and $K_{cat}^{r1}$: see the
-bottom right cell of figure @fig-laplace. This comparison demonstrates that
+bottom right cell of @fig-laplace. This comparison demonstrates that
 the Laplace method is not able to capture the correct relationships between

### 2023-10-31 paper/main.tex
Update paper

@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, sharp corners, boxrule=0pt, interior hidden, breakable, enhanced, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
 
@@
 fit to the simulated reaction and metabolite concentration measurements,
-as shown by the graphs in the top row of figure
-Figure~\ref{fig-posterior}.
+as shown by the graphs in the top row of Figure~\ref{fig-posterior}.
 
@@
 two-dimensional correlations such as the one shown in the bottom left of
-figure Figure~\ref{fig-posterior}; in this case the marginal
-distribution of the two parameters is roughly banana shaped. More
-commonly, however, two-dimensional pair plots were insufficient to
-reveal the underlying correlation structure, as seen in the bottom-right
-plot in figure Figure~\ref{fig-posterior}. This does not mean that the
-parameters were uncorrelated, but rather that the correlations involve
-more than two parameters.
-
-Overall, our results show that Maud can fit a realistic pathway-sized
-dataset. This was achieved without fixing the marginal values of kinetic
-parameters: the information required to make good predictions was
-contained in the correlation structure of the joint posterior
-distribution. This finding is consistent with previous analyses of
-biological systems that found they are ``sloppy'', that is, sensitive to
-parameter combinations rather than marginal parameter values, with
-important combinations, scales and regions of sensitivity being
-difficult to ascertain in advance
+Figure~\ref{fig-posterior}; in this case the marginal distribution of
+the two parameters is roughly banana shaped. More commonly, however,
+two-dimensional pair plots were insufficient to reveal the underlying
+correlation structure, as seen in the bottom-right plot in figure
+Figure~\ref{fig-posterior}. This does not mean that the parameters were
+uncorrelated, but rather that the correlations involve more than two
+parameters.
+
+These results show that Maud can achieve an adequate fit to a realistic
+pathway-sized dataset. This was achieved without fixing the marginal
+values of kinetic parameters: the information required to make good
+predictions was contained in the correlation structure of the joint
+posterior distribution. This finding is consistent with previous
+analyses of biological systems that found they are ``sloppy'', that is,
+sensitive to parameter combinations rather than marginal parameter
+values, with important combinations, scales and regions of sensitivity
+being difficult to ascertain in advance
 \citep{gutenkunst_2007, poirier_revising_1998}.
@@
 posterior distribution manifests not only in the parameter space, but
-also in the measurement space. Figure Figure~\ref{fig-laplace} frame B
-compares the 5\%-95\% interval for flux measurement log likelihoods in
-the true posterior with the Laplace approximation; lower log likelihood
-values indicate that the modelled and measured values are further away.
-The graph shows that the Laplace approximation yielded significantly
-worse predictions than the true posterior, even for the training data.
+also in the measurement space. The upper right (training data) and lower
+left (validation data) plots in Figure~\ref{fig-laplace} show 5\%-95\%
+marginal flux measurement log likelihood intervals for both methods;
+lower log likelihood values indicate that the modelled and measured
+values are further away. The graph shows that the Laplace approximation
+yielded significantly worse predictions than the true posterior, even
+for the training data.
 
@@
 distributions of two Michaelis-Menten constants \(K_{m}^{A,r1}\) and
-\(K_{cat}^{r1}\): see the bottom right cell of figure
-Figure~\ref{fig-laplace}. This comparison demonstrates that the Laplace
-method is not able to capture the correct relationships between
-parameters' distributions.
+\(K_{cat}^{r1}\): see the bottom right cell of Figure~\ref{fig-laplace}.
+This comparison demonstrates that the Laplace method is not able to
+capture the correct relationships between parameters' distributions.
 

### 2023-10-31 paper/main.qmd
multiomics inaccuracies

--- 2023-10-31 paper/main.qmd
@@
 and in vitro measurements relating to kinetic parameters, both contain vital
-information but are notoriously inaccurate (REFERENCES). Practically useful
+information but are notoriously inaccurate [siskos2017; tabb2010]. Practically useful
 kinetic modelling therefore requires a principled statistical approach that

### 2023-10-31 paper/main.tex
include renders

--- 2023-10-31 paper/main.tex
@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden, enhanced, sharp corners, breakable]}{\end{tcolorbox}}\fi
 
@@
 parameters, both contain vital information but are notoriously
-inaccurate (REFERENCES). Practically useful kinetic modelling therefore
-requires a principled statistical approach that encompasses multiple
-possible model parameterisations.
+inaccurate {[}siskos2017; tabb2010{]}. Practically useful kinetic
+modelling therefore requires a principled statistical approach that
+encompasses multiple possible model parameterisations.
 

### 2023-10-31 paper/main.qmd
Update paper

--- 2023-10-31 paper/main.qmd
@@
 
-To demonstrate how Maud's output can be used to yield useful metabolic
-insights we used the results of our case study to explain why the flux of
-the enzyme $GNMT$ is higher in dataset 1 than in dataset 12. GNMT is
-an irreversible enzyme that is homotropically activated by its substrate,
-competitively inhibited by its product and heterotropically inhibited by 5,10-
-methylenetetrahydrofolate (mlthf). The diverse regulation makes it the ideal
-test case to elucidate regulatory changes.
+To demonstrate how Maud's output can be used to yield useful metabolic insights
+we used the results of our case study to explain why the flux of the enzyme
+GNMT is higher in dataset 1 than in dataset 12. GNMT is an irreversible enzyme
+that is homotropically activated by its substrate, competitively inhibited by
+its product and heterotropically inhibited by 5,10- methylenetetrahydrofolate
+(mlthf). The diverse regulation makes it the ideal test case to elucidate
+regulatory changes.
 
@@
 each direction is given by the relative area under the curve on each side of the
-0 point.
+zero point.
 

### 2023-10-31 paper/main.tex
Update paper

--- 2023-10-31 paper/main.tex
@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, sharp corners, boxrule=0pt, interior hidden, breakable, enhanced, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, frame hidden, interior hidden, enhanced, breakable]}{\end{tcolorbox}}\fi
 
@@
 insights we used the results of our case study to explain why the flux
-of the enzyme \(GNMT\) is higher in dataset 1 than in dataset 12. GNMT
-is an irreversible enzyme that is homotropically activated by its
-substrate, competitively inhibited by its product and heterotropically
-inhibited by 5,10- methylenetetrahydrofolate (mlthf). The diverse
-regulation makes it the ideal test case to elucidate regulatory changes.
+of the enzyme GNMT is higher in dataset 1 than in dataset 12. GNMT is an
+irreversible enzyme that is homotropically activated by its substrate,
+competitively inhibited by its product and heterotropically inhibited by
+5,10- methylenetetrahydrofolate (mlthf). The diverse regulation makes it
+the ideal test case to elucidate regulatory changes.
 
@@
 according to our model, of the component acting in each direction is
-given by the relative area under the curve on each side of the 0 point.
+given by the relative area under the curve on each side of the zero
+point.
 

### 2023-10-31 paper/main.tex

@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, frame hidden, interior hidden, enhanced, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, breakable, enhanced, sharp corners, interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt]}{\end{tcolorbox}}\fi
 

### 2023-10-31 paper/main.qmd
keywords

--- 2023-10-31 paper/main.qmd
@@
 title: "Bayesian regression facilitates quantitative modelling of cell metabolism"
-keywords: Bayesian inference, kinetic models of cell metabolism 
+keywords: Bayesian inference, kinetic models of cell metabolism, 
+          multiomics integration, ordinary differential equations,
+          regulatory anlaysis
 abstract: >

### 2023-10-31 paper/main.tex
paper

@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, breakable, enhanced, sharp corners, interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, frame hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable]}{\end{tcolorbox}}\fi
 

### 2023-10-31 plot_figure_5.py
Tidy figure 5

@@ -73,6 +73,7 @@ def plot_flux(flux_df, ax):
     for line in ax.lines:
         line.set_marker("")
     ax.set_ylabel("")
+    ax.set_xlabel("Flux")
     ax.set_yticklabels(["Experiment 1", "Experiment 12"])
     return ax
 

### 2023-10-31 paper/main.qmd
paper
Update paper

@@
 Importantly, this form of analysis takes into account all modelled sources of
-uncertainty, including uncertainty about the measured values of the flux in each
-dataset. Our result shows that Maud could be used in this realistic case not
-only to provide a user with an explanation for an observed difference in fluxes,
-but also a reasonable judgement as to the explanation's robustness.
+uncertainty, including uncertainty about the measured values of the flux in
+each dataset. Our result shows that Maud could be used in this realistic case
+not only to explain an observed difference in fluxes, but also to provide a
+reasonable judgement as to the explanation's robustness.
 
@@
 metabolite concentrations and fluxes, which are derived from the values of
-other parameters by solving the steady state problem. See table (REFERENCE) in
+other parameters by solving the steady state problem. See table 1 in
 this paper's supporting information for a description of all these parameters
@@
 Maud's parameters are connected with measurable quantities via the steady state
-equation \eqref{eq-steady}. Posterior sampling using adaptive Hamiltonian Monte
-Carlo requires repeatedly solving equation \eqref{eq-steady} and finding the
+equation \eqref{eq-3}. Posterior sampling using adaptive Hamiltonian Monte
+Carlo requires repeatedly solving equation \eqref{eq-3} and finding the
 gradients of its solution with respect to all model parameters. This must be
@@
 
-To solve equation \eqref{eq-steady} and find its gradients, Maud uses
+To solve equation \eqref{eq-3} and find its gradients, Maud uses
 a hybrid method involving two numerical solvers from the SUNDIALS
@@
 
-- Description of Maud's input format
+- Description of Maud's input format.
 - Description of Maud's kinetic model including rate equations as well as

### 2023-10-31 paper/main.tex
Update paper

@@
   pdfauthor={Teddy Groves; Nicholas Luke Cowie; Lars Keld Nielsen},
-  pdfkeywords={Bayesian inference, kinetic models of cell metabolism},
+  pdfkeywords={Bayesian inference, kinetic models of cell metabolism,
+multiomics integration, ordinary differential equations, regulatory
+anlaysis},
   colorlinks=true,
@@
 
-\keywords{Bayesian inference, kinetic models of cell metabolism}
+\keywords{Bayesian inference, kinetic models of cell metabolism,
+multiomics integration, ordinary differential equations, regulatory
+anlaysis}
 
@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, frame hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden, sharp corners, breakable, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 of the flux in each dataset. Our result shows that Maud could be used in
-this realistic case not only to provide a user with an explanation for
-an observed difference in fluxes, but also a reasonable judgement as to
-the explanation's robustness.
+this realistic case not only to explain an observed difference in
+fluxes, but also to provide a reasonable judgement as to the
+explanation's robustness.
 
@@
 derived from the values of other parameters by solving the steady state
-problem. See table (REFERENCE) in this paper's supporting information
-for a description of all these parameters and their dimensions. Note
-that some quantities in Maud's kinetic model are not treated as
-parameters: for example, temperatures, compartment volumes and the
-formation energy of water. Maud treats these quantities as if they were
-known precisely: they can be configured by the user or default values
-can be used. Although in practice there can be considerable uncertainty
-regarding these quantities, we chose to disregard this uncertainty in
-the interest of simplicity.
+problem. See table 1 in this paper's supporting information for a
+description of all these parameters and their dimensions. Note that some
+quantities in Maud's kinetic model are not treated as parameters: for
+example, temperatures, compartment volumes and the formation energy of
+water. Maud treats these quantities as if they were known precisely:
+they can be configured by the user or default values can be used.
+Although in practice there can be considerable uncertainty regarding
+these quantities, we chose to disregard this uncertainty in the interest
+of simplicity.
 
@@
 Maud's parameters are connected with measurable quantities via the
-steady state equation \eqref{eq-steady}. Posterior sampling using
-adaptive Hamiltonian Monte Carlo requires repeatedly solving equation
-\eqref{eq-steady} and finding the gradients of its solution with respect
-to all model parameters. This must be done numerically as analytic
+steady state equation \eqref{eq-3}. Posterior sampling using adaptive
+Hamiltonian Monte Carlo requires repeatedly solving equation
+\eqref{eq-3} and finding the gradients of its solution with respect to
+all model parameters. This must be done numerically as analytic
 solutions are not available for realistic kinetic models.
@@
 
-To solve equation \eqref{eq-steady} and find its gradients, Maud uses a
+To solve equation \eqref{eq-3} and find its gradients, Maud uses a
 hybrid method involving two numerical solvers from the SUNDIALS suite
@@
 \item
-  Description of Maud's input format
+  Description of Maud's input format.
 \item

### 2023-10-31 paper/supporting_information.qmd

@@
 
-For the missing-data dataset XXX.
+For the missing-data dataset 250 post-warmup samples were taken from 4
+indpendent Markov  chains after 100 warmup samples. The sampling was initialised
+using the mass matric from the complete measurement dataset and the warmup 
+consisted of step size adaption for 100 samples. The resulting posterior distribution
+had an $\hat{R} = 1.01$ for the log-probability and did not exhibit post-warmup 
+divergences that were not a result of differential equation errors.
 

### 2023-10-31 paper/main.tex
Move figures into paper folder

@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden, sharp corners, breakable, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, enhanced, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, interior hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
 

### 2023-10-31 plot_figure_2.py

@@ -15,7 +15,7 @@ from maud.getting_idatas import get_idata
 
 
 MAUD_OUTPUT = Path.cwd() / "results" / "methionine"
-PLOT_OUTPUT = Path.cwd() / "figures" / "posterior.png"
+PLOT_OUTPUT = Path.cwd() / "paper" / "figures" / "posterior.png"
 
 plt.style.use("ipynb")
 

### 2023-10-31 plot_figure_3.py

@@ -40,7 +40,7 @@ plots.
 
 MAUD_OUTPUT_MISSING_AHCYS = Path.cwd() / "results" / "methionine_missing_ahcys"
 MAUD_OUTPUT = Path.cwd() / "results" / "methionine"
-PLOT_OUTPUT = Path.cwd() / "figures" / "missing.png"
+PLOT_OUTPUT = Path.cwd() / "paper" / "figures" / "missing.png"
 
 colour_scheme = {
     "complete measurement set": "#001219",

### 2023-10-31 plot_figure_4.py

@@ -50,7 +50,7 @@ MAUD_OUTPUT_TRAINING = Path.cwd() / "results" / "example_ode"
 MAUD_OUTPUT_VALIDATION = MAUD_OUTPUT_TRAINING / "validation"
 MAUD_OUTPUT_LAPLACE_TRAINING = Path.cwd() / "results" / "example_ode_laplace"
 MAUD_OUTPUT_VALIDATION = MAUD_OUTPUT_LAPLACE_TRAINING / "validation"
-PLOT_OUTPUT = Path.cwd() / "figures" / "laplace.png"
+PLOT_OUTPUT = Path.cwd() / "paper" / "figures" / "laplace.png"
 
 colour_scheme = {"measurement": "0.2", "laplace": "#005F73", "posterior": "#001219"}
 

### 2023-10-31 plot_figure_5.py

@@ -35,7 +35,7 @@ plt.style.use("ipynb")
 
 MAUD_OUTPUT = Path.cwd() / "results" / "methionine"
 MAUD_SIM = Path.cwd() / "results" / "methionine" / "simulate"
-PLOT_OUTPUT = Path.cwd() / "figures"
+PLOT_OUTPUT = Path.cwd() / "paper" / "figures"
 
 
 def plot_regulation(regulation_df, true_df, ax):

### 2023-11-01 paper/main.qmd
updating including methionine cycle and updating paths

@@
 and in vitro measurements relating to kinetic parameters, both contain vital
-information but are notoriously inaccurate [siskos2017; tabb2010]. Practically useful
+information but are notoriously inaccurate [@siskos2017; @tabb2010; @lu2017]. Practically useful
 kinetic modelling therefore requires a principled statistical approach that
@@
 
-![](../figures/methionine-reactions.png)
+![](./figures/methionine-reactions.png)
 
@@
 
-![](../figures/posterior.png)
+![](./figures/posterior.png)
 
@@
 
-![](../figures/laplace.png)
+![](./figures/laplace.png)
 
@@
 
-![](../figures/missing.png)
+![](./figures/missing.png)
 
@@
 
-![](../figures/decomposition.png)
+![](./figures/decomposition.png)
 

### 2023-11-01 paper/main.qmd
Implement Lars's paper comments

@@
 ---
-title: "Bayesian regression facilitates quantitative modelling of cell metabolism"
+title: "Maud: Bayesian regression modelling facilitates quantitative modelling of cell metabolism"
 keywords: Bayesian inference, kinetic models of cell metabolism, 
@@
 
-In equation \eqref{eq-1}, C represents a vector of metabolite concentrations,
-$v$ is a vector of reaction rates, and $\mu$ is the growth rate. The second term
-represents the dilution due to cell growth.
+where C represents a vector of metabolite concentrations, $v$ is a vector of
+reaction rates, and $\mu$ is the growth rate. The second term represents the
+dilution due to cell growth.
 
 In a kinetic model, the rates, $v$, are expressed as a function of the enzyme
-concentrations, $E$, the metabolite concentrations, $C$, and a set of parameters,
-$\theta$ as shown in equation \eqref{eq-2}
+concentrations, $E$, the metabolite concentrations, $C$, and a set of parameters
+$\theta$:
 
@@
 The parameters must include sufficient boundary concentrations and fluxes to
-solve equation \eqref{eq-1}.
+solve \eqref{eq-1}.
 
@@
 
-In a fermentation context, equation \eqref{eq-3} captures the rapid kinetics
+In a fermentation context, \eqref{eq-3} captures the rapid kinetics
 inside the cell, while another set of ODEs would be used to describe the
 external substrate and product concentrations, which could act as boundary
-parameters to equation \eqref{eq-3}.
+parameters to \eqref{eq-3}.
 
@@
 represented in Bayesian data generation models with arbitrarily high fidelity.
-See @gelmanBayesianDataAnalysis2020a for more about Bayesian inference and
-@gelmanBayesianWorkflow2020 for a discussion of practical Bayesian workflow.
+See @gelmanBayesianDataAnalysis2020a for more details about Bayesian inference
+and @gelmanBayesianWorkflow2020 for a discussion of practical Bayesian workflow.
 
 Another advantage is that Bayesian inference problems are well-posed even when
-not all parameters are strongly identified. Sloppy models in which measurable
-quantities are sensitive to combinations of parameters but not to individual
-marginal parameter values are ubiquitous in models of biological systems
-[@gutenkunst_2007; @white_2016]. The parameter correlation structure represents
-the set of potential models that describe the observed data. As we demonstrate
-in @sec-laplace below, capturing this correlation structure is difficult outside
-a fully Bayesian context.
+not all parameters are strongly identified. Sloppy parameter models in which
+measurable quantities are sensitive to combinations of parameters but not to
+individual marginal parameter values are ubiquitous in models of biological
+systems [@gutenkunst_2007; @white_2016]. The parameter correlation structure
+represents the set of potential models that describe the observed data. As we
+demonstrate later, capturing this correlation structure is difficult outside a
+fully Bayesian context.
 
@@
 
-The most popular algorithm for fitting Bayesian statistical models is
-Markov Chain Monte Carlo (MCMC). Modern MCMC algorithms allow exploration of
-high- dimensional posterior distributions, have robust failure diagnostics
+The most popular algorithm for fitting Bayesian statistical models is Markov
+Chain Monte Carlo (MCMC). Modern MCMC algorithms enable exploration of
+high-dimensional posterior distributions, have robust failure diagnostics
 [@vehtariRankNormalizationFoldingLocalization2021] and can incorporate fast
@@
 efficient sampling algorithm but uses approximate kinetics, namely lin-log
-kinetics @visser_dynamic_2003, limiting the scope of interpreting parameters and
+kinetics [@visser_dynamic_2003], limiting the scope of interpreting parameters and
 inferring cellular behaviour in experimental conditions outside the reference
@@
 @stapor_pesto_2018; @raue_joining_2013]. Non-MCMC-based Bayesian kinetic
-models have limited utility because they lack reliable diagnostic tools for
-verifying that their results approximate the target posterior distribution.
-This is a problem because realistic kinetic models tend to induce highly
-correlated, non-Gaussian, joint probability distributions [@gutenkunst_2007;
-@stapor_pesto_2018].
+models suffer from a lack of reliable diagnostic tools for verifying that their
+results approximate the target posterior distribution. This is a problem because
+realistic kinetic models tend to induce highly correlated, non-Gaussian, joint
+probability distributions [@gutenkunst_2007; @stapor_pesto_2018].
 
@@
 the description in @korendyaseva_allosteric_2008. The ordinary differential
-equation system describing this model is shown in Equation \eqref{eq-meth-ode}.
+equation system describing this model is
 
@@
 These steady states were then used to generate simulated measurements using the
-measurement model described below in @sec-methionine-measurement-model.
+measurement model.
 
-### Measurement model {#sec-methionine-measurement-model}
-
-Generating an artificial dataset required a specification of the true
-measurement error. For enzyme and metabolite concentration measurements
-we specified a standard deviation of 0.1 on natural logarithmic scale,
-corresponding to approximately 10% measurement error. For each reaction
-measurement a measurement standard deviation of approximately 10% of the
-simulated value.
+For enzyme and metabolite concentration measurements we specified a standard
+deviation of 0.1 on natural logarithmic scale, corresponding to approximately
+10% measurement error. For each reaction measurement a measurement standard
+deviation of approximately 10% of the simulated value.
 
@@
 were measured, and that there was a reaction measurement for each of the
-network’s elementary flux modes.
+network’s free fluxes.
 
-### Training/validation split
-
-The split between training and validation conditions was selected to achieve a
-large difference in fluxes between the two categories. The split was determined
-as we are interested in showing how our model can fit to varied conditions, and
-conditions closer to the training set are likely to be predicted well without
-necessarily learning the system.
-
-### Additional dataset with missing measurements
-
-To gain insight into our model’s robustness to missing measurements, we
-also performed a model run with the same 6 experimental datasets, but with
-measurements of the metabolite S-Adenosyl-L-homocysteine, or “ahcys” removed.
-Since ahcys regulates three enzymes in the methionine cycle, including one
-enzyme which is also thermodynamically regulated, we expected the removal of
-these measurements to yield interesting results.
-
-### Maud input specification
+The simulated experiments and measurements were split into a training and a
+validation dataset in a way that achieved a large difference in flux between the
+two categories. This was done to evaluate whether the fitted model is able to
+extrapolate to conditions well outside the training dataset rather than merely
+interpolating between the training data without necessarily learning the system.
 
@@
 inputs can be found at <https://github.com/biosustain/Methionine_model/tree/main/data>.
+
+## Posterior inference
 
@@
 *in vivo* vs *in vitro* measurement differences.
-
-## Findings
-
-### Posterior inference
 
@@
 
-Analysis of the posterior distributions for kinetic parameters indicated that
-these are highly correlated. The marginal posterior distributions for most
-kinetic parameters did not shrink significantly compared with the corresponding
-marginal prior distributions, even though these parameters’ joint posterior
-distribution contained enough information to make accurate out of sample
-predictions. In some cases, there were two-dimensional correlations such as
-the one shown in the bottom left of @fig-posterior; in this case the
+Analysis of the posterior distributions for kinetic parameters indicated
+that these are highly correlated. The marginal posterior distributions
+for most kinetic parameters did not shrink significantly compared to the
+corresponding marginal prior distributions, even though these parameters’
+joint posterior distribution contained enough information to make accurate out
+of sample predictions. In some cases, there were two-dimensional correlations
+such as the one shown in the bottom left of @fig-posterior; in this case the
 marginal distribution of the two parameters is roughly banana shaped. More
@@
 underlying correlation structure, as seen in the bottom-right plot in figure
-@fig-posterior. This does not mean that the parameters were uncorrelated, but
-rather that the correlations involve more than two parameters.
+@fig-posterior, which shows two marginally independent parameters. This kind of
+marginal independence does not mean that the parameters are independent in the
+full joint posterior distribution.
 
@@
 correlated posterior space would also be easier to summarise. We address this
-question below in @sec-laplace.
-
-This case study illustrates the type of kinetic model and dataset that Maud
-can fit. The model we analysed has 10 reactions, 5 state variables and 212
-parameters. Generalising from our ability to fit this model in a reasonable time
-using Maud, we expect that Maud can be used to fit realistic Bayesian models
-of approximately the same size, but not, for example, genome-scale kinetic
-models. To fit larger models, faster steady state solving methods or alternative
-inference algorithms will be required. @sec-laplace addresses whether Laplace
-approximation is a suitable candidate.
+question in the next section.
 
@@
 indicating a good fit. (**Top right**) Comparison of posterior predictive
-intervals with simulated concentration measurement values. These also show a
-good fit. (**Bottom left**) Pairwise marginal posterior distribution for two
-uncorrelated parameters, namely $K_m^{AHC1,hcys-L}$ and $K_m^{AHC1_ahcys}$.
-(**Bottom right**) Pairwise marginal posterior distribution for two correlated
-parameters, namely $K_m^{MAT3,atp}$ and $K_{cat}^{MAT3}$.
+intervals with simulated concentration measurement values. These also show
+a good fit. (**Bottom left**) Pairwise marginal posterior distribution
+for two pairwise correlated parameters, namely $K_m^{AHC1,hcys-L}$ and
+$K_m^{AHC1_ahcys}$. (**Bottom right**) Pairwise marginal posterior distribution
+for two pairwise-uncorrelated parameters, namely $K_m^{MAT3,atp}$ and $K_{cat}^{MAT3}$.
 
@@
 
-We found that the Laplace method was not able to produce an accurate posterior
-approximation for our model and dataset.
+This current case study illustrates the type of kinetic model and dataset that
+Maud can fit. The model we analysed has 10 reactions, 5 state variables and 212
+parameters. Generalising from our ability to fit this model in a reasonable time
+using Maud, we expect that Maud can be used to fit realistic Bayesian models
+of approximately the same size, but not, for example, genome-scale kinetic
+models. To fit larger models, faster steady state solving methods or alternative
+inference algorithms will be required. The Laplace approximation, in which the
+Fisher information matrix is used to calculate a normal approximation around
+the maximum a posteriori parameter configuration, is a popular strategy for
+generating approximate posterior samples while avoiding a full MCMC approach.
 
-Our overall strategy was to compare samples generated using MCMC, which we
-were confident could be treated as draws from the true posterior distribution,
-with approximate posterior samples generated using the Laplace method. We were
-unable to generate approximate posterior samples for our main methionine cycle
-case study using Laplace approximation, as the algorithm could not recover from
-solver failures caused by unrealistic parameter configurations. We therefore
-made a comparison for a simpler model. This input can be found at <https://github.com/biosustain/Methionine_model/tree/main/data/example_ode>.
+Using the recently implemented Laplace approximation in Stan, we were unable to
+generate approximate posterior samples for our main methionine cycle case study
+using Laplace approximation, as the algorithm could not recover from solver
+failures caused by unrealistic parameter configurations. This is not an inherent
+issue of the Laplace transformation but highlights some of the challenges around
+solving \eqref{eq-3} for real problems.
 
-MCMC sampling for our comparison input yielded 800 samples in 625 minutes;
+In order still to assess the potential use of the Laplace approximation, we made
+a comparison for a simpler model. This input can be found at <https://github.com/biosustain/Methionine_model/tree/main/data/example_ode>.
+MCMC sampling for this simpler model yielded 800 samples in 625 minutes, while
 Laplace sampling yielded the same number of samples in only one minute.
@@
 
-@fig-laplace summarises the results of comparing the samples generated using
-each method. As can be seen from the top left plot, the Laplace method does not
-provide a good approximation to the true posterior distribution in this case,
-as the marginal distribution of the total log probability density is clearly
-different. This impression was confirmed using the Kolmogorov-Smirnov test,
-which is a test to differentiate two empirical univariate distributions. The
-two distributions were significantly different with a p-value indistinguishable
-from zero.
+Comparing the samples generated using each method shows that the Laplace
+approximation does not provide a good approximation to the true posterior
+distribution in this case (@fig-laplace). As can be seen from the top left
+plot, the total log probability density is clearly different and this was
+confirmed by a test of the equality of two empirical univariate distributions
+(Kolmogorov-Smirnov, $p=1.7\times10^{-65}$).
 
@@
 distribution manifests not only in the parameter space, but also in the
-measurement space. The upper right (training data) and lower left (validation
-data) plots in @fig-laplace show 5%-95% marginal flux measurement log likelihood
-intervals for both methods; lower log likelihood values indicate that the
-modelled and measured values are further away. The graph shows that the Laplace
-approximation yielded significantly worse predictions than the true posterior,
-even for the training data.
+measurement space for both training data (upper right) and validation data
+(lower left). The lower log likelihood values indication that the modelled and
+measured values are further away and that the Laplace approximation yielded
+significantly worse predictions than the true posterior, even for the training
+data.
 
@@
 posterior and the Laplace approximation for the pairwise marginal distributions
-of two Michaelis-Menten constants $K_{m}^{A,r1}$ and $K_{cat}^{r1}$: see the
-bottom right cell of @fig-laplace. This comparison demonstrates that
-the Laplace method is not able to capture the correct relationships between
-parameters’ distributions.
+of two Michaelis-Menten constants $K_{m}^{A,r1}$ and $K_{cat}^{r1}$
+(@fig-laplace, bottom right). This comparison demonstrates that the Laplace
+method is not able to capture the correct relationships between parameters’
+distributions.
 
 This result shows that MCMC, while slower than Laplace approximation, is
-unfortunately preferable for posterior inference in this case. We expect that
-the Laplace method will produce worse approximations the more complex the target
-model. Since the approximation is already unacceptable for our simple test
-model, we recommend that Maud users use MCMC sampling in preference to Laplace
-approximation if possible when fitting realistic Bayesian kinetic models.
+required for posterior inference in this case. We expect that the Laplace method
+will produce worse approximations the more complex the target model. Since the
+approximation is already unacceptable for our simple test model, we recommend
+that Maud users use MCMC sampling in preference to Laplace approximation if
+possible when fitting realistic Bayesian kinetic models.
 
@@
 Laplace samples do not accurately approximate the target distribution. (**Top
-right**) The distribution of marginal posterior predictive log likelihood values
-for training data flux measurements shows that the Laplace method tended to
-yield much worse predictions compared with the true model. (**Bottom left**)
-The Laplace method also tended to produce worse flux predictions for held-back
-test measurements. (**Bottom right**) The marginal joint distribution of two
-parameters: $K_m^{A,rl}$ and $K_{cat}^{r1}$. The Laplace method is not able
-to track the correct joint distribution for this pair of parameters. This is
-unsurprising given that the target distribution has position-dependent scales
-which are difficult for a linear approximation to capture.
+right**) The distribution of marginal posterior predictive log likelihood
+values for training data flux measurements shows that the Laplace method
+tended to yield much worse predictions compared with the true model (lower log
+likelihood values indicate that the modelled and measured values are further
+away). (**Bottom left**) The Laplace method also tended to produce worse flux
+predictions for held-back test measurements. (**Bottom right**) The marginal
+joint distribution of two parameters: $K_m^{A,rl}$ and $K_{cat}^{r1}$. The
+Laplace method is not able to track the correct joint distribution for this
+pair of parameters. This is unsurprising given that the target distribution
+has position-dependent scales which are difficult for a linear approximation
+to capture.
 
@@
 
-### Effect of missing metabolite concentration measurements
+## Effect of missing metabolite concentration measurements
+
+To gain insight into our model’s robustness to missing measurements, we
+also performed a model run with the same 6 experimental datasets, but with
+measurements of the metabolite S-Adenosyl-L-homocysteine, or “ahcys” removed.
+Since ahcys regulates three enzymes in the methionine cycle, including one
+enzyme which is also thermodynamically regulated, we expected the removal of
+these measurements to yield interesting results.
 
@@
 
-@fig-missing shows that, as might be expected, the model with missing
-measurements did not correctly infer the missing ahcys concentrations.
-Nonetheless, the remaining measured metabolites were still well predicted,
-suggesting that information about the network is still preserved despite the
-missing measurements. Comparison of flux measurements in both models also
-indicated that removing the ahcys measurement did not result in catastrophic
-model failure.
+As might be expected, the model with missing measurements did not correctly
+infer the missing ahcys concentrations (@fig-missing). Nonetheless, the
+remaining measured metabolites were still well predicted, suggesting
+that information about the network is still preserved despite the missing
+measurements. Comparison of flux measurements in both models also indicated that
+removing the ahcys measurement did not result in catastrophic model failure.
 
 The missing measurements did affect Maud’s ability to infer parameter values
-correctly. The lower left plot of @fig-missing shows that the model with full
-metabolomics learned the true value for the displayed dissociation constant,
-despite this value being far from the mean of the corresponding marginal prior
-distribution. In contrast, the model with missing measurements stayed in the
-neighbourhood of the prior.
+correctly (@fig-missing, lower left). The model with full metabolomics learned
+the true value for the displayed dissociation constant, despite this value being
+far from the mean of the corresponding marginal prior distribution. In contrast,
+the model with missing measurements stayed in the neighbourhood of the prior.
 
@@
 
-### Application to regulatory understanding
+## Application to regulatory understanding
 
@@
 we used the results of our case study to explain why the flux of the enzyme
-GNMT is higher in dataset 1 than in dataset 12. GNMT is an irreversible enzyme
-that is homotropically activated by its substrate, competitively inhibited by
-its product and heterotropically inhibited by 5,10- methylenetetrahydrofolate
-(mlthf). The diverse regulation makes it the ideal test case to elucidate
-regulatory changes.
+GNMT is higher in dataset 1 than in dataset 12 (@fig-decomposition). GNMT
+is an irreversible enzyme that is homotropically activated by its substrate,
+competitively inhibited by its product and heterotropically inhibited by 5,10-
+methylenetetrahydrofolate (mlthf). The complex regulation makes it the ideal
+test case to elucidate regulatory changes.
 
-@fig-decomposition shows the regulatory description of GNMT, according to the
-results our main case study analysis. Each curve shows the marginal posterior
-distribution of the ratio of the corresponding regulatory component in dataset
-1 compared with dataset 12. A positive value indicates that the component
+Regulation can be separated into enzyme abundance, allostery and saturation,
+and we can plot the marginal posterior distribution of the ratio of the
+corresponding regulatory component in dataset 1 compared with dataset 12
+(@fig-decomposition, right panel). A positive value indicates that the component
 was increased in dataset 1 relative to dataset 12, with 0 indicating no
@@
 datasets 1 and 2. (**Right**) Log-scale ratios of the regulatory elements
-defined in equation \eqref{eq-decomposition}. Note that the $Reversibility$
+defined in \eqref{eq-decomposition}. Note that the $Reversibility$
 and $k_cat$ components are excluded: this is because this reaction was modelled
@@
 including the components of the regulatory decomposition described below in
-equation \eqref{eq-decomposition}, log likelihoods, simulated measurement values
+\eqref{eq-decomposition}, log likelihoods, simulated measurement values
 and metabolic control analysis coefficients [@kacser_control_1973].
@@
 
-Maud’s kinetic model decomposes into factors as shown in equation
-\eqref{eq-decomposition}.
+Maud’s kinetic model decomposes into factors:
 
@@
 quantitative metabolomics and proteomics experiments. The likelihood functions
-for such measurements are shown in equations Equation \eqref{eq-yconc} and
-Equation \eqref{eq-yenz}.
+for such measurements are:
 
@@
 link function (the natural logarithm $\ln$) and known standard deviation
-$\sigma_{𝑐𝑜𝑛𝑐}$. The use of this measurement model is motivated by the
+$ \sigma_{𝑐𝑜𝑛𝑐}$. The use of this measurement model is motivated by the
 consideration that concentrations are constrained to be non-negative, so the
@@
 metabolite concentration values. In addition, we expect the precision of most
-metabolomics and proteomics experiments to be roughly proportional to the value
-of the true measured quantity, which supports a measurement model with constant
+metabolomics and proteomics experiments to be roughly proportional to the true
+value of the measured quantity, which supports a measurement model with constant
 coefficient of variation. The measurement standard deviations $\sigma_{𝑐𝑜𝑛𝑐}$
@@
 values can be elicited by considering the likely coefficient of variation of the
-measuring apparatus.
+measuring apparatus. Our measurement model improves on analyses of metabolomics
+and proteomics data that assume a regression model with normally distributed
+errors, whether explicitly using a standard linear model or implicitly using
+ordinary least squares fitting.
 
@@
 are modelled using a likelihood function from a standard linear regression
-model, as shown in equation \eqref{eq-yflux}. Flux measurements can be obtained
-from the results of isotope labelling experiments using metabolic flux analysis,
-for example as described in (Young 2014). When entering flux measurements,
-it is important only to specify measurements for a network’s free fluxes, as
-the values of some steady state fluxes in a metabolic network are constrained
-by others, with the result that dependent fluxes cannot typically be measured
-separately. If measurements of multiple dependent fluxes are entered,
-information will inappropriately be double counted.
+model \eqref{eq-yflux}. Flux measurements can be obtained from isotope labelling
+experiments using metabolic flux analysis, for example as described in (Young
+2014). When entering flux measurements, it is important only to specify
+measurements for a network’s free fluxes, as the values of some steady state
+fluxes in a metabolic network are constrained by others, with the result that
+dependent fluxes cannot typically be measured separately. If measurements of
+multiple dependent fluxes are entered, information will inappropriately be
+double counted.
 
@@
 \end{equation}
-
-Our measurement model improves on analyses of metabolomics and proteomics
-data that assume a regression model with normally distributed errors, whether
-explicitly using a standard linear model or implicitly using ordinary least
-squares fitting. This assumption is undesirable because it implies that the
-measured quantity could in principle be negative, and assumes an additive
-underlying random process, whereas multiplicative processes tend to better
-describe real concentration data.
 
@@
 equation \eqref{eq-3}. Posterior sampling using adaptive Hamiltonian Monte
-Carlo requires repeatedly solving equation \eqref{eq-3} and finding the
-gradients of its solution with respect to all model parameters. This must be
-done numerically as analytic solutions are not available for realistic kinetic
-models.
+Carlo requires repeatedly solving \eqref{eq-3} and finding the gradients of its
+solution with respect to all model parameters. This must be done numerically as
+analytic solutions are not available for realistic kinetic models.
 
@@
 
-To solve equation \eqref{eq-3} and find its gradients, Maud uses
-a hybrid method involving two numerical solvers from the SUNDIALS
-suite [@serbanCVODESSensitivityEnabledODE2005]: CVODES and IDAS via
-their interface from Stan. The hybrid method follows that proposed by
+To solve \eqref{eq-3} and find its gradients, Maud uses a hybrid
+method involving two numerical solvers from the SUNDIALS suite
+[@serbanCVODESSensitivityEnabledODE2005]: CVODES and IDAS via their
+interface from Stan. The hybrid method follows that proposed by
 @margossianComputingSteadyStates2018, and involves numerically evolving the ODE
@@
 
-This research was funded by the Novo Nordisk foundation using grant number
-NNF14OC0009473.
+This research was funded by the Novo Nordisk Foundation (Grant numbers
+NNF20CC0035580 NNF14OC0009473).
 

### 2023-11-01 paper/main.tex
Implement Lars's paper comments

@@
 \hypersetup{
-  pdftitle={Bayesian regression facilitates quantitative modelling of cell metabolism},
+  pdftitle={Maud: Bayesian regression modelling facilitates quantitative modelling of cell metabolism},
   pdfauthor={Teddy Groves; Nicholas Luke Cowie; Lars Keld Nielsen},
@@
 
-\title[]{Bayesian regression facilitates quantitative modelling of cell
-metabolism}
+\title[]{Maud: Bayesian regression modelling facilitates quantitative
+modelling of cell metabolism}
 \makeatletter
@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, enhanced, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, interior hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, enhanced, interior hidden, breakable, sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 parameters, both contain vital information but are notoriously
-inaccurate {[}siskos2017; tabb2010{]}. Practically useful kinetic
-modelling therefore requires a principled statistical approach that
-encompasses multiple possible model parameterisations.
+inaccurate \citep{siskos2017, tabb2010, lu2017}. Practically useful
+kinetic modelling therefore requires a principled statistical approach
+that encompasses multiple possible model parameterisations.
 
@@
 
-In equation \eqref{eq-1}, C represents a vector of metabolite
-concentrations, \(v\) is a vector of reaction rates, and \(\mu\) is the
-growth rate. The second term represents the dilution due to cell growth.
+where C represents a vector of metabolite concentrations, \(v\) is a
+vector of reaction rates, and \(\mu\) is the growth rate. The second
+term represents the dilution due to cell growth.
 
@@
 enzyme concentrations, \(E\), the metabolite concentrations, \(C\), and
-a set of parameters, \(\theta\) as shown in equation \eqref{eq-2}
+a set of parameters \(\theta\):
 
@@
 The parameters must include sufficient boundary concentrations and
-fluxes to solve equation \eqref{eq-1}.
+fluxes to solve \eqref{eq-1}.
 
@@
 
-In a fermentation context, equation \eqref{eq-3} captures the rapid
-kinetics inside the cell, while another set of ODEs would be used to
-describe the external substrate and product concentrations, which could
-act as boundary parameters to equation \eqref{eq-3}.
+In a fermentation context, \eqref{eq-3} captures the rapid kinetics
+inside the cell, while another set of ODEs would be used to describe the
+external substrate and product concentrations, which could act as
+boundary parameters to \eqref{eq-3}.
 
@@
 generation models with arbitrarily high fidelity. See
-\citet{gelmanBayesianDataAnalysis2020a} for more about Bayesian
+\citet{gelmanBayesianDataAnalysis2020a} for more details about Bayesian
 inference and \citet{gelmanBayesianWorkflow2020} for a discussion of
@@
 Another advantage is that Bayesian inference problems are well-posed
-even when not all parameters are strongly identified. Sloppy models in
-which measurable quantities are sensitive to combinations of parameters
-but not to individual marginal parameter values are ubiquitous in models
-of biological systems \citep{gutenkunst_2007, white_2016}. The parameter
-correlation structure represents the set of potential models that
-describe the observed data. As we demonstrate in
-Section~\ref{sec-laplace} below, capturing this correlation structure is
+even when not all parameters are strongly identified. Sloppy parameter
+models in which measurable quantities are sensitive to combinations of
+parameters but not to individual marginal parameter values are
+ubiquitous in models of biological systems
+\citep{gutenkunst_2007, white_2016}. The parameter correlation structure
+represents the set of potential models that describe the observed data.
+As we demonstrate later, capturing this correlation structure is
 difficult outside a fully Bayesian context.
@@
 The most popular algorithm for fitting Bayesian statistical models is
-Markov Chain Monte Carlo (MCMC). Modern MCMC algorithms allow
-exploration of high- dimensional posterior distributions, have robust
+Markov Chain Monte Carlo (MCMC). Modern MCMC algorithms enable
+exploration of high-dimensional posterior distributions, have robust
 failure diagnostics
@@
 utilises an efficient sampling algorithm but uses approximate kinetics,
-namely lin-log kinetics \citet{visser_dynamic_2003}, limiting the scope
+namely lin-log kinetics \citep{visser_dynamic_2003}, limiting the scope
 of interpreting parameters and inferring cellular behaviour in
@@
 \citep{liebermeister_model_2021, gopalakrishnan_k-fit_2020, stapor_pesto_2018, raue_joining_2013}.
-Non-MCMC-based Bayesian kinetic models have limited utility because they
-lack reliable diagnostic tools for verifying that their results
-approximate the target posterior distribution. This is a problem because
-realistic kinetic models tend to induce highly correlated, non-Gaussian,
-joint probability distributions
-\citep{gutenkunst_2007, stapor_pesto_2018}.
+Non-MCMC-based Bayesian kinetic models suffer from a lack of reliable
+diagnostic tools for verifying that their results approximate the target
+posterior distribution. This is a problem because realistic kinetic
+models tend to induce highly correlated, non-Gaussian, joint probability
+distributions \citep{gutenkunst_2007, stapor_pesto_2018}.
 
@@
 
-\includegraphics{../figures/methionine-reactions.png}
+\includegraphics{./figures/methionine-reactions.png}
 
@@
 using the description in \citet{korendyaseva_allosteric_2008}. The
-ordinary differential equation system describing this model is shown in
-Equation\,\eqref{eq-meth-ode}.
+ordinary differential equation system describing this model is
 
@@
 were then used to generate simulated measurements using the measurement
-model described below in Section~\ref{sec-methionine-measurement-model}.
-
-\hypertarget{sec-methionine-measurement-model}{%
-\subsubsection{Measurement
-model}\label{sec-methionine-measurement-model}}
-
-Generating an artificial dataset required a specification of the true
-measurement error. For enzyme and metabolite concentration measurements
-we specified a standard deviation of 0.1 on natural logarithmic scale,
-corresponding to approximately 10\% measurement error. For each reaction
-measurement a measurement standard deviation of approximately 10\% of
-the simulated value.
+model.
+
+For enzyme and metabolite concentration measurements we specified a
+standard deviation of 0.1 on natural logarithmic scale, corresponding to
+approximately 10\% measurement error. For each reaction measurement a
+measurement standard deviation of approximately 10\% of the simulated
+value.
 
@@
 concentrations were measured, and that there was a reaction measurement
-for each of the network's elementary flux modes.
-
-\hypertarget{trainingvalidation-split}{%
-\subsubsection{Training/validation
-split}\label{trainingvalidation-split}}
-
-The split between training and validation conditions was selected to
-achieve a large difference in fluxes between the two categories. The
-split was determined as we are interested in showing how our model can
-fit to varied conditions, and conditions closer to the training set are
-likely to be predicted well without necessarily learning the system.
-
-\hypertarget{additional-dataset-with-missing-measurements}{%
-\subsubsection{Additional dataset with missing
-measurements}\label{additional-dataset-with-missing-measurements}}
-
-To gain insight into our model's robustness to missing measurements, we
-also performed a model run with the same 6 experimental datasets, but
-with measurements of the metabolite S-Adenosyl-L-homocysteine, or
-``ahcys'' removed. Since ahcys regulates three enzymes in the methionine
-cycle, including one enzyme which is also thermodynamically regulated,
-we expected the removal of these measurements to yield interesting
-results.
-
-\hypertarget{maud-input-specification}{%
-\subsubsection{Maud input
-specification}\label{maud-input-specification}}
+for each of the network's free fluxes.
+
+The simulated experiments and measurements were split into a training
+and a validation dataset in a way that achieved a large difference in
+flux between the two categories. This was done to evaluate whether the
+fitted model is able to extrapolate to conditions well outside the
+training dataset rather than merely interpolating between the training
+data without necessarily learning the system.
 
@@
 \url{https://github.com/biosustain/Methionine_model/tree/main/data}.
+
+\hypertarget{posterior-inference}{%
+\subsection{Posterior inference}\label{posterior-inference}}
 
@@
 \emph{in vitro} measurement differences.
-
-\hypertarget{findings}{%
-\subsection{Findings}\label{findings}}
-
-\hypertarget{posterior-inference}{%
-\subsubsection{Posterior inference}\label{posterior-inference}}
 
@@
 that these are highly correlated. The marginal posterior distributions
-for most kinetic parameters did not shrink significantly compared with
-the corresponding marginal prior distributions, even though these
+for most kinetic parameters did not shrink significantly compared to the
+corresponding marginal prior distributions, even though these
 parameters' joint posterior distribution contained enough information to
@@
 correlation structure, as seen in the bottom-right plot in figure
-Figure~\ref{fig-posterior}. This does not mean that the parameters were
-uncorrelated, but rather that the correlations involve more than two
-parameters.
+Figure~\ref{fig-posterior}, which shows two marginally independent
+parameters. This kind of marginal independence does not mean that the
+parameters are independent in the full joint posterior distribution.
 
@@
 correct for. A linearly correlated posterior space would also be easier
-to summarise. We address this question below in
-Section~\ref{sec-laplace}.
-
-This case study illustrates the type of kinetic model and dataset that
-Maud can fit. The model we analysed has 10 reactions, 5 state variables
-and 212 parameters. Generalising from our ability to fit this model in a
-reasonable time using Maud, we expect that Maud can be used to fit
-realistic Bayesian models of approximately the same size, but not, for
-example, genome-scale kinetic models. To fit larger models, faster
-steady state solving methods or alternative inference algorithms will be
-required. Section~\ref{sec-laplace} addresses whether Laplace
-approximation is a suitable candidate.
+to summarise. We address this question in the next section.
 
@@
 
-\includegraphics{../figures/posterior.png}
+\includegraphics{./figures/posterior.png}
 
@@
 (\textbf{Bottom left}) Pairwise marginal posterior distribution for two
-uncorrelated parameters, namely \(K_m^{AHC1,hcys-L}\) and
+pairwise correlated parameters, namely \(K_m^{AHC1,hcys-L}\) and
 \(K_m^{AHC1_ahcys}\). (\textbf{Bottom right}) Pairwise marginal
-posterior distribution for two correlated parameters, namely
+posterior distribution for two pairwise-uncorrelated parameters, namely
 \(K_m^{MAT3,atp}\) and \(K_{cat}^{MAT3}\).}
@@
 
-We found that the Laplace method was not able to produce an accurate
-posterior approximation for our model and dataset.
-
-Our overall strategy was to compare samples generated using MCMC, which
-we were confident could be treated as draws from the true posterior
-distribution, with approximate posterior samples generated using the
-Laplace method. We were unable to generate approximate posterior samples
-for our main methionine cycle case study using Laplace approximation, as
-the algorithm could not recover from solver failures caused by
-unrealistic parameter configurations. We therefore made a comparison for
-a simpler model. This input can be found at
+This current case study illustrates the type of kinetic model and
+dataset that Maud can fit. The model we analysed has 10 reactions, 5
+state variables and 212 parameters. Generalising from our ability to fit
+this model in a reasonable time using Maud, we expect that Maud can be
+used to fit realistic Bayesian models of approximately the same size,
+but not, for example, genome-scale kinetic models. To fit larger models,
+faster steady state solving methods or alternative inference algorithms
+will be required. The Laplace approximation, in which the Fisher
+information matrix is used to calculate a normal approximation around
+the maximum a posteriori parameter configuration, is a popular strategy
+for generating approximate posterior samples while avoiding a full MCMC
+approach.
+
+Using the recently implemented Laplace approximation in Stan, we were
+unable to generate approximate posterior samples for our main methionine
+cycle case study using Laplace approximation, as the algorithm could not
+recover from solver failures caused by unrealistic parameter
+configurations. This is not an inherent issue of the Laplace
+transformation but highlights some of the challenges around solving
+\eqref{eq-3} for real problems.
+
+In order still to assess the potential use of the Laplace approximation,
+we made a comparison for a simpler model. This input can be found at
 \url{https://github.com/biosustain/Methionine_model/tree/main/data/example_ode}.
-
-MCMC sampling for our comparison input yielded 800 samples in 625
-minutes; Laplace sampling yielded the same number of samples in only one
+MCMC sampling for this simpler model yielded 800 samples in 625 minutes,
+while Laplace sampling yielded the same number of samples in only one
 minute. The diagnostics indicated that our algorithm was able to find
@@
 
-Figure~\ref{fig-laplace} summarises the results of comparing the samples
-generated using each method. As can be seen from the top left plot, the
-Laplace method does not provide a good approximation to the true
-posterior distribution in this case, as the marginal distribution of the
-total log probability density is clearly different. This impression was
-confirmed using the Kolmogorov-Smirnov test, which is a test to
-differentiate two empirical univariate distributions. The two
-distributions were significantly different with a p-value
-indistinguishable from zero.
+Comparing the samples generated using each method shows that the Laplace
+approximation does not provide a good approximation to the true
+posterior distribution in this case (Figure~\ref{fig-laplace}). As can
+be seen from the top left plot, the total log probability density is
+clearly different and this was confirmed by a test of the equality of
+two empirical univariate distributions (Kolmogorov-Smirnov,
+\(p=1.7\times10^{-65}\)).
 
@@
 posterior distribution manifests not only in the parameter space, but
-also in the measurement space. The upper right (training data) and lower
-left (validation data) plots in Figure~\ref{fig-laplace} show 5\%-95\%
-marginal flux measurement log likelihood intervals for both methods;
-lower log likelihood values indicate that the modelled and measured
-values are further away. The graph shows that the Laplace approximation
-yielded significantly worse predictions than the true posterior, even
-for the training data.
+also in the measurement space for both training data (upper right) and
+validation data (lower left). The lower log likelihood values indication
+that the modelled and measured values are further away and that the
+Laplace approximation yielded significantly worse predictions than the
+true posterior, even for the training data.
 
@@
 distributions of two Michaelis-Menten constants \(K_{m}^{A,r1}\) and
-\(K_{cat}^{r1}\): see the bottom right cell of Figure~\ref{fig-laplace}.
-This comparison demonstrates that the Laplace method is not able to
-capture the correct relationships between parameters' distributions.
+\(K_{cat}^{r1}\) (Figure~\ref{fig-laplace}, bottom right). This
+comparison demonstrates that the Laplace method is not able to capture
+the correct relationships between parameters' distributions.
 
 This result shows that MCMC, while slower than Laplace approximation, is
-unfortunately preferable for posterior inference in this case. We expect
-that the Laplace method will produce worse approximations the more
-complex the target model. Since the approximation is already
-unacceptable for our simple test model, we recommend that Maud users use
-MCMC sampling in preference to Laplace approximation if possible when
-fitting realistic Bayesian kinetic models.
+required for posterior inference in this case. We expect that the
+Laplace method will produce worse approximations the more complex the
+target model. Since the approximation is already unacceptable for our
+simple test model, we recommend that Maud users use MCMC sampling in
+preference to Laplace approximation if possible when fitting realistic
+Bayesian kinetic models.
 
@@
 
-\includegraphics{../figures/laplace.png}
+\includegraphics{./figures/laplace.png}
 
@@
 Laplace method tended to yield much worse predictions compared with the
-true model. (\textbf{Bottom left}) The Laplace method also tended to
-produce worse flux predictions for held-back test measurements.
-(\textbf{Bottom right}) The marginal joint distribution of two
-parameters: \(K_m^{A,rl}\) and \(K_{cat}^{r1}\). The Laplace method is
-not able to track the correct joint distribution for this pair of
+true model (lower log likelihood values indicate that the modelled and
+measured values are further away). (\textbf{Bottom left}) The Laplace
+method also tended to produce worse flux predictions for held-back test
+measurements. (\textbf{Bottom right}) The marginal joint distribution of
+two parameters: \(K_m^{A,rl}\) and \(K_{cat}^{r1}\). The Laplace method
+is not able to track the correct joint distribution for this pair of
 parameters. This is unsurprising given that the target distribution has
@@
 \hypertarget{effect-of-missing-metabolite-concentration-measurements}{%
-\subsubsection{Effect of missing metabolite concentration
+\subsection{Effect of missing metabolite concentration
 measurements}\label{effect-of-missing-metabolite-concentration-measurements}}
+
+To gain insight into our model's robustness to missing measurements, we
+also performed a model run with the same 6 experimental datasets, but
+with measurements of the metabolite S-Adenosyl-L-homocysteine, or
+``ahcys'' removed. Since ahcys regulates three enzymes in the methionine
+cycle, including one enzyme which is also thermodynamically regulated,
+we expected the removal of these measurements to yield interesting
+results.
 
@@
 
-Figure~\ref{fig-missing} shows that, as might be expected, the model
-with missing measurements did not correctly infer the missing ahcys
-concentrations. Nonetheless, the remaining measured metabolites were
-still well predicted, suggesting that information about the network is
-still preserved despite the missing measurements. Comparison of flux
-measurements in both models also indicated that removing the ahcys
-measurement did not result in catastrophic model failure.
+As might be expected, the model with missing measurements did not
+correctly infer the missing ahcys concentrations
+(Figure~\ref{fig-missing}). Nonetheless, the remaining measured
+metabolites were still well predicted, suggesting that information about
+the network is still preserved despite the missing measurements.
+Comparison of flux measurements in both models also indicated that
+removing the ahcys measurement did not result in catastrophic model
+failure.
 
 The missing measurements did affect Maud's ability to infer parameter
-values correctly. The lower left plot of Figure~\ref{fig-missing} shows
-that the model with full metabolomics learned the true value for the
-displayed dissociation constant, despite this value being far from the
-mean of the corresponding marginal prior distribution. In contrast, the
-model with missing measurements stayed in the neighbourhood of the
-prior.
+values correctly (Figure~\ref{fig-missing}, lower left). The model with
+full metabolomics learned the true value for the displayed dissociation
+constant, despite this value being far from the mean of the
+corresponding marginal prior distribution. In contrast, the model with
+missing measurements stayed in the neighbourhood of the prior.
 
@@
 
-\includegraphics{../figures/missing.png}
+\includegraphics{./figures/missing.png}
 
@@
 \hypertarget{application-to-regulatory-understanding}{%
-\subsubsection{Application to regulatory
+\subsection{Application to regulatory
 understanding}\label{application-to-regulatory-understanding}}
@@
 insights we used the results of our case study to explain why the flux
-of the enzyme GNMT is higher in dataset 1 than in dataset 12. GNMT is an
-irreversible enzyme that is homotropically activated by its substrate,
-competitively inhibited by its product and heterotropically inhibited by
-5,10- methylenetetrahydrofolate (mlthf). The diverse regulation makes it
-the ideal test case to elucidate regulatory changes.
-
-Figure~\ref{fig-decomposition} shows the regulatory description of GNMT,
-according to the results our main case study analysis. Each curve shows
-the marginal posterior distribution of the ratio of the corresponding
-regulatory component in dataset 1 compared with dataset 12. A positive
-value indicates that the component was increased in dataset 1 relative
-to dataset 12, with 0 indicating no difference. The probability,
-according to our model, of the component acting in each direction is
-given by the relative area under the curve on each side of the zero
-point.
+of the enzyme GNMT is higher in dataset 1 than in dataset 12
+(Figure~\ref{fig-decomposition}). GNMT is an irreversible enzyme that is
+homotropically activated by its substrate, competitively inhibited by
+its product and heterotropically inhibited by 5,10-
+methylenetetrahydrofolate (mlthf). The complex regulation makes it the
+ideal test case to elucidate regulatory changes.
+
+Regulation can be separated into enzyme abundance, allostery and
+saturation, and we can plot the marginal posterior distribution of the
+ratio of the corresponding regulatory component in dataset 1 compared
+with dataset 12 (Figure~\ref{fig-decomposition}, right panel). A
+positive value indicates that the component was increased in dataset 1
+relative to dataset 12, with 0 indicating no difference. The
+probability, according to our model, of the component acting in each
+direction is given by the relative area under the curve on each side of
+the zero point.
 
@@
 
-\includegraphics{../figures/decomposition.png}
+\includegraphics{./figures/decomposition.png}
 
@@
 in datasets 1 and 2. (\textbf{Right}) Log-scale ratios of the regulatory
-elements defined in equation \eqref{eq-decomposition}. Note that the
+elements defined in \eqref{eq-decomposition}. Note that the
 \(Reversibility\) and \(k_cat\) components are excluded: this is because
@@
 quantities including the components of the regulatory decomposition
-described below in equation \eqref{eq-decomposition}, log likelihoods,
-simulated measurement values and metabolic control analysis coefficients
+described below in \eqref{eq-decomposition}, log likelihoods, simulated
+measurement values and metabolic control analysis coefficients
 \citep{kacser_control_1973}.
@@
 
-Maud's kinetic model decomposes into factors as shown in equation
-\eqref{eq-decomposition}.
+Maud's kinetic model decomposes into factors:
 
@@
 of quantitative metabolomics and proteomics experiments. The likelihood
-functions for such measurements are shown in equations
-Equation\,\eqref{eq-yconc} and Equation\,\eqref{eq-yenz}.
+functions for such measurements are:
 
@@
 link function (the natural logarithm \(\ln\)) and known standard
-deviation \(\sigma_{𝑐𝑜𝑛𝑐}\). The use of this measurement model is
+deviation \$ \sigma\_\{𝑐𝑜𝑛𝑐\}\$. The use of this measurement model is
 motivated by the consideration that concentrations are constrained to be
@@
 addition, we expect the precision of most metabolomics and proteomics
-experiments to be roughly proportional to the value of the true measured
+experiments to be roughly proportional to the true value of the measured
 quantity, which supports a measurement model with constant coefficient
@@
 plausible values can be elicited by considering the likely coefficient
-of variation of the measuring apparatus.
+of variation of the measuring apparatus. Our measurement model improves
+on analyses of metabolomics and proteomics data that assume a regression
+model with normally distributed errors, whether explicitly using a
+standard linear model or implicitly using ordinary least squares
+fitting.
 
@@
 analyses, are modelled using a likelihood function from a standard
-linear regression model, as shown in equation\,\eqref{eq-yflux}. Flux
-measurements can be obtained from the results of isotope labelling
-experiments using metabolic flux analysis, for example as described in
-(Young 2014). When entering flux measurements, it is important only to
-specify measurements for a network's free fluxes, as the values of some
-steady state fluxes in a metabolic network are constrained by others,
-with the result that dependent fluxes cannot typically be measured
-separately. If measurements of multiple dependent fluxes are entered,
-information will inappropriately be double counted.
+linear regression model \eqref{eq-yflux}. Flux measurements can be
+obtained from isotope labelling experiments using metabolic flux
+analysis, for example as described in (Young 2014). When entering flux
+measurements, it is important only to specify measurements for a
+network's free fluxes, as the values of some steady state fluxes in a
+metabolic network are constrained by others, with the result that
+dependent fluxes cannot typically be measured separately. If
+measurements of multiple dependent fluxes are entered, information will
+inappropriately be double counted.
 
@@
 \end{equation}
-
-Our measurement model improves on analyses of metabolomics and
-proteomics data that assume a regression model with normally distributed
-errors, whether explicitly using a standard linear model or implicitly
-using ordinary least squares fitting. This assumption is undesirable
-because it implies that the measured quantity could in principle be
-negative, and assumes an additive underlying random process, whereas
-multiplicative processes tend to better describe real concentration
-data.
 
@@
 steady state equation \eqref{eq-3}. Posterior sampling using adaptive
-Hamiltonian Monte Carlo requires repeatedly solving equation
-\eqref{eq-3} and finding the gradients of its solution with respect to
-all model parameters. This must be done numerically as analytic
-solutions are not available for realistic kinetic models.
+Hamiltonian Monte Carlo requires repeatedly solving \eqref{eq-3} and
+finding the gradients of its solution with respect to all model
+parameters. This must be done numerically as analytic solutions are not
+available for realistic kinetic models.
 
@@
 
-To solve equation \eqref{eq-3} and find its gradients, Maud uses a
-hybrid method involving two numerical solvers from the SUNDIALS suite
+To solve \eqref{eq-3} and find its gradients, Maud uses a hybrid method
+involving two numerical solvers from the SUNDIALS suite
 \citep{serbanCVODESSensitivityEnabledODE2005}: CVODES and IDAS via their
@@
 
-This research was funded by the Novo Nordisk foundation using grant
-number NNF14OC0009473.
+This research was funded by the Novo Nordisk Foundation (Grant numbers
+NNF20CC0035580 NNF14OC0009473).
 

### 2023-11-01 paper/main.qmd
Paper comments with Lars

@@
 ---
-title: "Maud: Bayesian regression modelling facilitates quantitative modelling of cell metabolism"
+title: "Bayesian regression facilitates quantitative modelling of cell metabolism"
 keywords: Bayesian inference, kinetic models of cell metabolism, 
@@
         name: DTU
-        department: DTU Biosustain
+        department: The Novo Nordisk Foundation Center for Biosustainability
         city: Kongens Lyngby
@@
   - name: "Nicholas Luke Cowie"
+    orcid: 0000-0003-3367-2105
     affiliations: 
-      - id: biosustain
-        name: DTU
-        department: DTU Biosustain
-        city: Kongens Lyngby
-        country: Denmark
+      - ref: biosustain
     attributes:
@@
   - name: "Lars Keld Nielsen"
+    orcid: 0000-0001-8191-3511
     affiliations: 
-      - name: The University of Queensland
+      - id: uq
+        name: The University of Queensland
         department: Australian Institute for Bioengineering and Nanotechnology (AIBN)
@@
         country: Australia
-      - id: biosustain
-        name: DTU
-        department: DTU Biosustain
-        city: Kongens Lyngby
-        country: Denmark
+      - ref: biosustain
 bibliography: bibliography.bib
@@
 
-Our application Maud provides the first Bayesian kinetic model to combine
+Our application Maud provides a Bayesian kinetic model that combines
 biologically realistic mechanistic accuracy---including accurate rate laws,
@@
 underlying correlation structure, as seen in the bottom-right plot in figure
-@fig-posterior, which shows two marginally independent parameters. This kind of
-marginal independence does not mean that the parameters are independent in the
-full joint posterior distribution.
+@fig-posterior, which shows two marginally independent parameters.
 
@@
 a good fit. (**Bottom left**) Pairwise marginal posterior distribution
-for two pairwise correlated parameters, namely $K_m^{AHC1,hcys-L}$ and
+for two correlated parameters, namely $K_m^{AHC1,hcys-L}$ and
 $K_m^{AHC1_ahcys}$. (**Bottom right**) Pairwise marginal posterior distribution
-for two pairwise-uncorrelated parameters, namely $K_m^{MAT3,atp}$ and $K_{cat}^{MAT3}$.
+for two uncorrelated parameters, namely $K_m^{MAT3,atp}$ and $K_{cat}^{MAT3}$.
 
@@
 this paper's supporting information for a description of all these parameters
-and their dimensions. Note that some quantities in Maud’s kinetic model are
-not treated as parameters: for example, temperatures, compartment volumes and
-the formation energy of water. Maud treats these quantities as if they were
-known precisely: they can be configured by the user or default values can be
-used. Although in practice there can be considerable uncertainty regarding
-these quantities, we chose to disregard this uncertainty in the interest of
-simplicity.
+and their dimensions.
 

### 2023-11-01 paper/main.tex
Paper comments with Lars

@@
 \hypersetup{
-  pdftitle={Maud: Bayesian regression modelling facilitates quantitative modelling of cell metabolism},
+  pdftitle={Bayesian regression facilitates quantitative modelling of cell metabolism},
   pdfauthor={Teddy Groves; Nicholas Luke Cowie; Lars Keld Nielsen},
@@
 \author{Teddy Groves}
-\affiliation{DTU Biosustain, DTU, Kongens Lyngby, Denmark}
+\affiliation{The Novo Nordisk Foundation Center for
+Biosustainability, DTU, Kongens Lyngby, Denmark}
 
@@
 \author{Nicholas Luke Cowie}
-\affiliation{DTU Biosustain, DTU, Kongens Lyngby, Denmark}
+\affiliation{The Novo Nordisk Foundation Center for
+Biosustainability, DTU, Kongens Lyngby, Denmark}
 
@@
 \author{Lars Keld Nielsen}
+\affiliation{The Novo Nordisk Foundation Center for
+Biosustainability, DTU, Kongens Lyngby, Denmark}
 \affiliation{Australian Institute for Bioengineering and Nanotechnology
 (AIBN), The University of Queensland, St Lucia, Australia}
-\affiliation{DTU Biosustain, DTU, Kongens Lyngby, Denmark}
 
@@
 
-\title[]{Maud: Bayesian regression modelling facilitates quantitative
-modelling of cell metabolism}
+\title[]{Bayesian regression facilitates quantitative modelling of cell
+metabolism}
 \makeatletter
@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, enhanced, interior hidden, breakable, sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners, breakable]}{\end{tcolorbox}}\fi
 
@@
 
-Our application Maud provides the first Bayesian kinetic model to
-combine biologically realistic mechanistic accuracy---including accurate
-rate laws, post- translational modification and thermodynamics---with
-fast, robust MCMC sampling using adaptive Hamiltonian Monte Carlo.
-Further, Maud is a general-purpose application that can be used to fit a
-wide range of Bayesian kinetic models.
+Our application Maud provides a Bayesian kinetic model that combines
+biologically realistic mechanistic accuracy---including accurate rate
+laws, post- translational modification and thermodynamics---with fast,
+robust MCMC sampling using adaptive Hamiltonian Monte Carlo. Further,
+Maud is a general-purpose application that can be used to fit a wide
+range of Bayesian kinetic models.
 
@@
 Figure~\ref{fig-posterior}, which shows two marginally independent
-parameters. This kind of marginal independence does not mean that the
-parameters are independent in the full joint posterior distribution.
+parameters.
 
@@
 (\textbf{Bottom left}) Pairwise marginal posterior distribution for two
-pairwise correlated parameters, namely \(K_m^{AHC1,hcys-L}\) and
+correlated parameters, namely \(K_m^{AHC1,hcys-L}\) and
 \(K_m^{AHC1_ahcys}\). (\textbf{Bottom right}) Pairwise marginal
-posterior distribution for two pairwise-uncorrelated parameters, namely
+posterior distribution for two uncorrelated parameters, namely
 \(K_m^{MAT3,atp}\) and \(K_{cat}^{MAT3}\).}
@@
 problem. See table 1 in this paper's supporting information for a
-description of all these parameters and their dimensions. Note that some
-quantities in Maud's kinetic model are not treated as parameters: for
-example, temperatures, compartment volumes and the formation energy of
-water. Maud treats these quantities as if they were known precisely:
-they can be configured by the user or default values can be used.
-Although in practice there can be considerable uncertainty regarding
-these quantities, we chose to disregard this uncertainty in the interest
-of simplicity.
+description of all these parameters and their dimensions.
 

### 2023-11-01 paper/supporting_information.qmd
add reference
keep supporting tex

@@
     include-in-header: mathjax.html
-  pdf: default
+  pdf: 
+    keep-tex: true
   docx: default

### 2023-11-01 paper/main.tex
Noor reference in supporting info

@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, breakable, enhanced, boxrule=0pt, interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
 

### 2023-11-01 paper/supporting_information.tex
Noor reference in supporting info

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\KOMAoption{captions}{tableheading}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Supporting information},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Supporting information}
\author{}
\date{}

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, boxrule=0pt, interior hidden, sharp corners, breakable, enhanced]}{\end{tcolorbox}}\fi

This document provides information in support of our article ``Bayesian
regression facilitates quantitative modelling of cell metabolism''.

The results of all reported Maud runs can be found at
\url{https://github.com/biosustain/Methionine_model/blob/main/results}.

\hypertarget{mauds-input-format}{%
\section{Maud's input format}\label{mauds-input-format}}

Maud inputs are structured directories, somewhat inspired by the PEtab
format (Schmiester et al. 2021). A Maud input directory must contain a
toml (Preston-Werner, Tom and Gedam, Pradyun 2020) file called
\texttt{config.toml} which gives the input a name, configures how Maud
will be run and tells Maud where to find the other files, allowing these
to have custom names. It must also include a file containing a kinetic
model definition, a file specifying information about parameters and a
file with information experiments. The required structure of these files
is documented at
\url{https://maud-metabolic-models.readthedocs.io/en/latest/inputting.html}.
The input is validated against a Pydantic (Pydantic developers 2022)
data model.

We chose to implement a custom input format despite the existence of
standard formats in similar areas, including SBML (Keating et al. 2020)
and PEtab (Schmiester et al. 2021). This choice was partly motivated by
the need to ensure flexibility as Maud was developed, but there are also
features of SBML and PEtab that make them structurally unsuitable in
this context. Our requirements for an input format included that it be
mathematics-free, so that all mathematical details are encapsulated in
source code, and that it has a detailed, verifiable structure. These
requirements made toml more attractive than SBML: toml is easier for
humans to read and edit and can straightforwardly be validated using
tools like Pydantic. Further, an SBML representation of our desired
input would not contain differential equations. It would therefore not
be interoperable with most SBML targeting software, which typically
assumes that differential equations are available and does not know
about Maud's structure.

\hypertarget{mauds-kinetic-model}{%
\section{Maud's kinetic model}\label{mauds-kinetic-model}}

\hypertarget{parameters}{%
\subsection{Parameters}\label{parameters}}

Table~\ref{tbl-params} shows all of Maud's unknown parameters along with
their dimensions

Note that Maud's metabolic model includes some quantities that are not
treated as parameters in its statistical model, including temperatures,
compartment volumes and the formation energy of water. Maud treats these
quantities as if they were known precisely: they can be configured by
the user or default values can be used. Although in practice there can
be considerable uncertainty regarding these quantities, we chose to
disregard this uncertainty in the interest of simplicity.

\hypertarget{tbl-params}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2500}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4722}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2778}}@{}}
\caption{\label{tbl-params}Parameters of Maud's statistical
model}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Parameter
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Modelled quantity
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Dimensions
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Parameter
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Modelled quantity
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Dimensions
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\(\Delta_fG\) & Formation energy & metabolites \\
\(k_M\) & Michaelis Menten constants & Substrates of all
enzyme/reactions and products of reversible enzyme/reactions \\
\(k_I\) & Inhibition constants & Inhibiting metabolite/compartments of
enzyme/reactions exhibiting competitive inhibition \\
\(k_{cat}\) & Rate constants & Enzyme/reactions \\
\(L_0\) & Transfer constants & Allosteric interactions \\
\(e_T\) & T dissociation constants & Modifying metabolites of allosteric
inhibitions \\
\(e_R\) & R dissociation constants & Modifying metabolites of allosteric
activations \\
\(k_{cat\, pme}\) & Rate constants of phosphorylation modifying enzymes
& Phosphorylation modifying enzymes \\
\(v_{drain}\) & Drain fluxes & Drains, experiments \\
\(Enzyme\) & Enzyme concentrations & Enzymes, experiments \\
\(C_{unbalanced}\) & Unbalanced metabolite/compartment concentrations &
Unbalanced metabolite/compartments, experiments \\
\(C_{pme}\) & Phosphorylation modifying enzyme concentrations &
Phosphorylation modifying enzymes, experiments \\
\(\psi\) & Membrane potentials & Experiments \\
\end{longtable}

Solving the steady state problem for a given set of parameters in an
experiment yields a vector \(C_{balanced}\) of balanced metabolite
concentrations. These are combined with the balanced metabolite
concentrations \(C_{unbalanced}\) to produce a vector \(C_{mic}\) with a
concentration for each metabolite/compartment combination.

\(\Delta_fG\) parameters can optionally be fixed; this can be useful for
computational purposes, as for example to avoid estimating the formation
energy of a metabolite about which there is no available information due
to it only participating in irreversible reactions.

\hypertarget{rate-equations}{%
\subsection{Rate equations}\label{rate-equations}}

As discussed in the main text, Maud's kinetic model decomposes into
factors contributing to the flux in a metabolic network in an experiment
as shown in equation \eqref{eq-decomposition}. For succinctness, and
since Maud's model assumes that there are no interactions between
experiments, we omit any notation referring to experiments below. We
also omit any reference to the network's drain reactions: these are
modelled as being exactly determined by the values of the parameter
vector \(v_{drain}\).

\begin{equation}
F(C;\theta) = Enzyme\cdot k_{cat}\cdot Reversibility \cdot Saturation \cdot Allostery \label{eq-decomposition}
\end{equation}

The term \(Enzyme\) in equation \eqref{eq-decomposition} is a vector of
non-negative real numbers representing the concentration of the enzyme
catalysing each reaction.

The term \(k_{cat}\) in equation \eqref{eq-decomposition} is a vector of
non-negative real numbers representing the amount of flux carried per
unit of saturated enzyme.

The term \(Reversibility\) in equation \eqref{eq-decomposition} is a
vector of real numbers capturing the impact of thermodynamic effects on
the reaction's flux, as shown in equation \eqref{eq-reversibility}.

\begin{align} 
  Reversibility &= 1 - \exp(\frac{\Delta_{r}G + RT \cdot S^T \ln(C_{mic})}{RT}) \label{eq-reversibility} \\ 
  \Delta_{r}G &= S^{T}\Delta_{f}G + n F \psi \nonumber 
\end{align}

The terms in \eqref{eq-reversibility} have the following meanings:

\begin{itemize}
\tightlist
\item
  \(T\) is the temperature in Kelvin (a number),
\item
  \(R\) is the gas constant (a number),
\item
  \(\Delta_rG\) is a vector representing the Gibbs free energy change of
  each reaction in standard conditions,
\item
  \(\Delta_fG\) is a vector representing the standard condition Gibbs
  free energy change of each metabolite's formation reaction, or in
  other words each metabolite's `formation energy'.
\item
  \(n\) is a vector representing the number of charges transported by
  each reaction.
\item
  \(F\) is the Faraday constant (a number)
\item
  \(\psi\) is a vector representing each reaction's membrane potential
  (these numbers only matter for reactions that transport non-zero
  charge)
\end{itemize}

Note that, for reactions with zero transported charge, the thermodynamic
effect on each reaction is derived from metabolite formation energies.
This formulation is helpful because, provided that all reactions' rates
are calculated from the same formation energies, they are guaranteed to
be thermodynamically consistent.

The term \(n\) accounts for both the charge and the directionality. For
instance, a reaction that exports 2 protons to the extracellular space
in the forward direction would have -2 charge. If a negatively charged
molecule like acetate is exported in the forward direction, \(n\) would
be 1.

Note that this way of modelling the effect of transported charge does
not take into account that the concentration gradient used by the
transport is that of the dissociated molecules. Thus, this expression is
only correct for ions whose concentration can be expressed in the model
only in the charged form; e.g., protons, \(K^+\), \(Na^+\), \(Cl^-\),
etc.

The term \(Saturation\) in equation \eqref{eq-decomposition} is a vector
of non-negative real numbers representing, for each reaction, the
fraction of enzyme that is saturated, i.e.~bound to one of the
reaction's substrates. To describe saturation we use equation
\eqref{eq-saturation}, which is taken from Liebermeister, Uhlendorf, and
Klipp (2010) and Noor et al. (2013). Additionally, this term captures
competitive inhibition: as competitive inhibitor concentration
increases, the saturation denominator increases, effectively decreasing
the saturation of the substrate on the total enzyme pool. Conversely, as
the substrate concentration increases this term approaches 1.

\begin{align} 
Saturation_r &= a \cdot \text{free enzyme ratio}\label{eq-saturation} \\ 
a &= \prod_{\text{s substrate}}\frac{C_{mic}^s}{k_{M}^{rs}} \nonumber \\ 
\text{free enzyme ratio} &= \begin{cases} 
  \prod_{\text{s sustrate}} (1 + \frac{C_{mic}^s}{k_{M}^{rs}})^{S_sr} 
  + \sum_{\text{c inhibitor}}\frac{C_{mic}^c}{k_I^{rc}} & r\text{ irreversible} \\ 
  -1 
  + \prod_{\text{s sustrate}} (1 + \frac{C_{mic}^s}{k_{M}^{rs}})^{S_sr} 
     + \sum_{\text{c inhibitor}}\frac{C_{mic}^c}{k_I^{rc}} 
     + \prod_{\text{p product}} (1 + \frac{C_{mic}^p}{k_{M}^{rp}})^{S_pr}  & r\text{ reversible} 
    \end{cases} \nonumber
\end{align}

The term \(Allostery\) in equation \eqref{eq-decomposition} is a vector
of non-negative numbers describing the effect of allosteric regulation
on each reaction. Allosteric regulation happens when binding to a
certain molecule changes an enzyme's shape in a way that changes its
catalytic behaviour. We use equation \eqref{eq-allostery} to describe
this phenomenon, following the generalised MWC approach described in
Monod, Wyman, and Changeux (1965), Changeux (2013), Popova and Sel'kov
(1975) and Popova and Sel'kov (1979).

\begin{align} 
  Allostery_r &= \frac{1}{1 + L_0^r \cdot (\text{free enzyme ratio}_r \cdot \frac{Qtense}{Qrelaxed})^{subunits}}\label{eq-allostery} \\ 
       Qtense &= 1 + \sum_{\text{i inhibitor}} \frac{C_{mic}^i}{e_T^{ri}}\nonumber \\ 
     Qrelaxed &= 1 + \sum_{\text{a activator}} \frac{C_{mic}^a}{e_R^{ra}}\nonumber
\end{align}

The parameter \(L_0\) in equation \eqref{eq-decomposition} is called the
transfer constant, and the parameter vectors \(e_T\) and \(e_R\) are
called tense and relaxed dissociation constants respectively.

Finally, the term \(Phosphorylation\) in equation
\eqref{eq-decomposition} captures the important effect whereby enzyme
activity is altered due to a coupled process of phosphorylation and
dephosphorylation. This description achieves a similar behaviour to the
MWC formalism for describing allosteric regulation, but using the rates
of phosphorylation and dephosphorylation rather than concentrations of
metabolites.

\begin{align}
Phosphorylation_r &= (\frac{\alpha}{\alpha + \beta})^{subunits} \label{eq-phosphorylation}\\
\alpha &= \sum_{\text{p phosphoylator}} k_{cat\,pme}^{p} \cdot C_{pme}^p \nonumber \\
\beta &= \sum_{\text{d dephosphoylator}} k_{cat\,pme}^{d} \cdot C_{pme}^d \nonumber 
\end{align}

\hypertarget{sec-methionine-case-study}{%
\section{Methionine case study}\label{sec-methionine-case-study}}

\hypertarget{dataset-generation}{%
\subsection{Dataset generation}\label{dataset-generation}}

Starting with the model in Saa and Nielsen (2016), we extracted values
for enzyme concentrations, boundary conditions and fluxes. We used these
values to generate MCMC samples using Maud using the priors specified in
section Section~\ref{sec-methionine-priors}. When this was finished, we
selected one sample with relatively high log probability to use as a
ground truth in our case study. These parameter values are shown below
in table Table~\ref{tbl-case-study-params}. We manually inspected the
parameter values to screen for any obviously implausible values; we did
not find any of these.

\hypertarget{sec-methionine-priors}{%
\subsection{Prior distributions compared with true parameter
values}\label{sec-methionine-priors}}

Table~\ref{tbl-case-study-params} shows the prior distributions we used
for independent parameters. The first two columns show the 1\% and 99\%
quantiles of each marginal prior distribution. True parameter value are
shown in column three, and the last column shows the z-score on log
scale of the true parameter value according the marginal prior
distribution. As can be seen from the table, there are 7 parameters for
which the true value is outside the 1\%-99\% range.

\hypertarget{tbl-case-study-params}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3211}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1560}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1651}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1101}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2477}}@{}}
\caption{\label{tbl-case-study-params}Parameter specification, marginal
prior distributions and true parameter values used in our case
study.}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
parameter name
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
1\% prior quantile
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
99\% prior quantile
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
true value
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
prior Z-score of true value
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
parameter name
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
1\% prior quantile
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
99\% prior quantile
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
true value
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
prior Z-score of true value
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\(e_{𝑅}^{𝐶𝐵𝑆1,𝑎𝑚𝑒𝑡𝑐}\) & 3.430e-06 & 0.002480 & 9.3e-05 & 0.004 \\
\(e_{𝑅}^{𝐺𝑁𝑀𝑇1,𝑎𝑚𝑒𝑡𝑐}\) & 3.000e-05 & 0.002000 & 2.000e-05 & -2.787 \\
\(e_{𝑅}^{𝑀𝐴𝑇3,𝑎𝑚𝑒𝑡𝑐}\) & 1.000e-04 & 0.001000 & 3.170e-04 & 0.003 \\
\(e_{𝑅}^{𝑀𝐴𝑇3,𝑚𝑒𝑡−𝐿𝑐}\) & 4.500e-04 & 0.000800 & 6.000e-04 & 0.000 \\
\(e_{𝑅}^{𝑀𝑇𝐻𝐹𝑅1,𝑎ℎ𝑐𝑦𝑠𝑐}\) & 1.120e-07 & 0.000081 & 2.000e-06 & -0.101 \\
\(e_{𝑇}^{𝐺𝑁𝑀𝑇1,𝑚𝑙𝑡ℎ𝑓𝑐}\) & 1.120e-05 & 0.008050 & 2.290e-04 & -0.136 \\
\(e_{𝑇}^{𝑀𝑇𝐻𝐹𝑅1,𝑎𝑚𝑒𝑡𝑐}\) & 1.120e-07 & 0.000081 & 1.500e-05 &
0.549306 \\
\(k_{𝑐𝑎𝑡}^{𝐴𝐻𝐶1}\) & 1.200e+02 & 400.000000 & 2.340e+02 & 0.179861 \\
\(k_{𝑐𝑎𝑡}^{𝐵𝐻𝑀𝑇1}\) & 6.000e+00 & 35.000000 & 1.380e+01 & -0.135 \\
\(k_{𝑐𝑎𝑡}^{𝐶𝐵𝑆1}\) & 1.000e+01 & 188.000000 & 7.020e+00 & -2.887 \\
\(k_{𝑐𝑎𝑡}^{𝐺𝑁𝑀𝑇1}\) & 7.000e-01 & 60.000000 & 1.050e+01 & 0.352083 \\
\(k_{𝑐𝑎𝑡}^{𝑀𝐴𝑇1}\) & 8.200e-02 & 59.100000 & 7.900e+00 & 0.44375 \\
\(k_{𝑐𝑎𝑡}^{𝑀𝐴𝑇3}\) & 5.890e-01 & 424.000000 & 1.990e+01 & 0.080556 \\
\(k_{𝑐𝑎𝑡}^{𝑀𝐸𝑇𝐻−𝐺𝑒𝑛}\) & 4.840e-01 & 349.000000 & 1.160e+00 & -1.209 \\
\(k_{𝑐𝑎𝑡}^{𝑀𝑆1kcatMS1}\) & 1.000e+00 & 3.300000 & 1.770e+00 & -0.091 \\
\(k_{𝑐𝑎𝑡}^{𝑀𝑇𝐻𝐹𝑅1}\) & 1.300e+00 & 4.200000 & 3.170e+00 & 0.183333 \\
\(k_{𝑐𝑎𝑡}^{𝑃𝑅𝑂𝑇1}\) & 1.590e-01 & 0.222000 & 2.650e-01 & 0.41875 \\
\(k_{𝐼}^{𝐺𝑁𝑀𝑇1,𝑎ℎ𝑐𝑦𝑠𝑐}\) & 2.000e-06 & 0.001400 & 5.300e-05 & 0.010 \\
\(k_{𝐼}^{𝑀𝐴𝑇1,𝑎𝑚𝑒𝑡𝑐}\) & 3.000e-04 & 0.000400 & 3.470e-04 & 0.014 \\
\(k_{𝐼}^{𝑀𝐸𝑇𝐻−𝐺𝑒𝑛,𝑎ℎ𝑐𝑦𝑠𝑐}\) & 1.000e-06 & 0.000030 & 6.000e-06 &
0.021 \\
\(k_{𝑀}^{𝐴𝐻𝐶1,𝑎ℎ𝑐𝑦𝑠𝑐}\) & 5.220e-05 & 0.037600 & 2.320e-05 & -2.050 \\
\(k_{𝑀}^{𝐴𝐻𝐶1,𝑎𝑑𝑛𝑐}\) & 1.670e-07 & 0.000120 & 5.660e-06 & 0.081944 \\
\(k_{𝑀}^{𝐴𝐻𝐶1,ℎ𝑐𝑦𝑠−𝐿𝑐}\) & 1.580e-07 & 0.000114 & 1.060e-05 &
0.318056 \\
\(k_{𝑀}^{𝐵𝐻𝑀𝑇1,ℎ𝑐𝑦𝑠−𝐿𝑐}\) & 1.200e-05 & 0.000032 & 1.980e-05 & 0.049 \\
\(k_{𝑀}^{𝐵𝐻𝑀𝑇1,𝑔𝑙𝑦𝑏𝑐}\) & 4.720e-05 & 0.034000 & 8.460e-03 & 0.659028 \\
\(k_{𝑀}^{𝐶𝐵𝑆1,ℎ𝑐𝑦𝑠−𝐿𝑐}\) & 1.000e-06 & 0.000025 & 4.240e-05 & 3.090 \\
\(k_{𝑀}^{𝐶𝐵𝑆1,𝑠𝑒𝑟−𝐿𝑐}\) & 2.000e-06 & 0.000004 & 2.830e-06 & 0.004 \\
\(k_{𝑀}^{𝐺𝑁𝑀𝑇1,𝑎𝑚𝑒𝑡𝑐}\) & 1.300e-05 & 0.009400 & 5.200e-04 & 0.1375 \\
\(k_{𝑀}^{𝐺𝑁𝑀𝑇1,𝑎ℎ𝑐𝑦𝑠𝑐}\) & 4.100e-07 & 0.000295 & 1.100e-05 & 0.000 \\
\(k_{𝑀}^{𝐺𝑁𝑀𝑇1,𝑔𝑙𝑦𝑐}\) & 5.480e-05 & 0.039500 & 2.540e-03 & 0.189583 \\
\(k_{𝑀}^{𝐺𝑁𝑀𝑇1,𝑠𝑎𝑟𝑐𝑠𝑐}\) & 3.730e-09 & 0.000003 & 1.000e-07 & 0.000 \\
\(k_{𝑀}^{𝑀𝐴𝑇1,𝑚𝑒𝑡−𝐿𝑐}\) & 1.400e-05 & 0.000720 & 1.070e-04 & 0.074 \\
\(k_{𝑀}^{𝑀𝐴𝑇1,𝑎𝑡𝑝𝑐}\) & 5.270e-05 & 0.038000 & 2.030e-03 & 0.125694 \\
\(k_{𝑀}^{𝑀𝐴𝑇3,𝑚𝑒𝑡−𝐿𝑐}\) & 4.470e-05 & 0.032200 & 1.130e-03 & -0.029 \\
\(k_{𝑀}^{𝑀𝐴𝑇3,𝑎𝑡𝑝𝑐}\) & 5.270e-05 & 0.038000 & 2.370e-03 & 0.179167 \\
\(k_{𝑀}^{𝑀𝐸𝑇𝐻−𝐺𝑒𝑛,𝑎𝑚𝑒𝑡𝑐}\) & 7.000e-06 & 0.000013 & 9.370e-06 &
-0.135 \\
\(k_{𝑀}^{𝑀𝑆1,5𝑚𝑡ℎ𝑓𝑐}\) & 3.320e-06 & 0.002390 & 6.940e-05 & -0.124 \\
\(k_{𝑀}^{𝑀𝑆1,ℎ𝑐𝑦𝑠−𝐿𝑐}\) & 1.000e-06 & 0.000003 & 1.710e-06 & -0.054 \\
\(k_{𝑀}^{𝑀𝑇𝐻𝐹𝑅1,𝑚𝑙𝑡ℎ𝑓𝑐}\) & 7.500e-05 & 0.000088 & 8.080e-05 & -0.158 \\
\(k_{𝑀}^{𝑀𝑇𝐻𝐹𝑅1,𝑛𝑎𝑑𝑝ℎ𝑐}\) & 1.600e-05 & 0.000028 & 2.090e-05 & -0.105 \\
\(k_{𝑀}^{𝑃𝑅𝑂𝑇1,𝑚𝑒𝑡−𝐿𝑐}\) & 4.500e-05 & 0.000085 & 4.390e-05 & -2.507 \\
\(𝐿_{0}^{𝐶𝐵𝑆1}\) & 3.730e-02 & 26.800000 & 1.030e+00 & 0.017 \\
\(𝐿_{0}^{𝐺𝑁𝑀𝑇1}\) & 3.730e-02 & 26.800000 & 1.310e+02 & 0.3875 \\
\(𝐿_{0}^{𝑀𝐴𝑇3}\) & 3.730e-03 & 2.680000 & 1.080e-01 & 0.037 \\
\(𝐿_{0}^{𝑀𝑇𝐻𝐹𝑅1}\) & 1.120e-01 & 80.500000 & 3.920e-01 & -1.018 \\
\end{longtable}

\(\Delta_fG\) parameters for most metabolites were fixed; those that
were modelled as unknown had a multivariate normal prior distribution
derived from eQuilibrator (Beber et al. 2021).

The values for \(\Delta_fG\) parameters, as well as all other model
parameters, can be found by inspecting the file \texttt{priors.toml}
which is online at
\url{https://github.com/biosustain/Methionine_model/blob/main/data/methionine/priors.toml}.

\hypertarget{computation}{%
\subsection{Computation}\label{computation}}

We conducted adaptive Hamiltonian Monte Carlo sampling for the full and
missing-data datasets. For the full dataset we obtained 1000 post-warmup
samples each from 4 independent Markov chains after 1000 warm-up samples
and ``hot-starting'' with a mass metric output by a previous model run.

For the missing-data dataset 250 post-warmup samples were taken from 4
indpendent Markov chains after 100 warmup samples. The sampling was
initialised using the mass matric from the complete measurement dataset
and the warmup consisted of step size adaption for 100 samples. The
resulting posterior distribution had an \(\hat{R} = 1.01\) for the
log-probability and did not exhibit post-warmup divergences that were
not a result of differential equation errors.

\hypertarget{laplace-approximation-case-study}{%
\section{Laplace approximation case
study}\label{laplace-approximation-case-study}}

To compare MCMC sampling with Laplace approximation we used a different
model with fewer parameters and state variables. This model was chosen
because we were not able to generate results for our methionine model
using Laplace approximation. The simpler case still serves to illustrate
the general issues with approximating the posterior distributions of
Bayesian kinetic models using the Laplace method, and that the
associated numerical instability is another reason to prefer other
methods where possible.

The full Maud input folders used for our Laplace approximation case
study can be found at
\url{https://github.com/biosustain/Methionine_model/tree/main/data/example_ode}
and
\url{https://github.com/biosustain/Methionine_model/tree/main/data/example_ode_laplace}.

To generate Laplace samples we used Maud's Laplace mode.

\hypertarget{references}{%
\section{References}\label{references}}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-beberEQuilibratorPlatformEstimation2021}{}}%
Beber, Moritz E., Mattia G. Gollub, Dana Mozaffari, Kevin M. Shebek, and
Elad Noor. 2021. {``{eQuilibrator} 3.0 -- a Platform for the Estimation
of Thermodynamic Constants.''} \emph{arXiv:2103.00621 {[}q-Bio{]}},
February. \url{http://arxiv.org/abs/2103.00621}.

\leavevmode\vadjust pre{\hypertarget{ref-changeux_2013}{}}%
Changeux, Jean-Pierre. 2013. {``50 Years of Allosteric Interactions: The
Twists and Turns of the Models.''} \emph{Nature Reviews. Molecular Cell
Biology} 14 (12): 819--29. \url{https://doi.org/10.1038/nrm3695}.

\leavevmode\vadjust pre{\hypertarget{ref-keatingSBMLLevelExtensible2020}{}}%
Keating, Sarah M, Dagmar Waltemath, Matthias König, Fengkai Zhang,
Andreas Dräger, Claudine Chaouiya, Frank T Bergmann, et al. 2020.
{``{\textsc{SBML}} {Level} 3: An Extensible Format for the Exchange and
Reuse of Biological Models.''} \emph{Molecular Systems Biology} 16 (8):
e9110. \url{https://doi.org/10.15252/msb.20199110}.

\leavevmode\vadjust pre{\hypertarget{ref-liebermeister_modular_2010}{}}%
Liebermeister, Wolfram, Jannis Uhlendorf, and Edda Klipp. 2010.
{``Modular Rate Laws for Enzymatic Reactions: Thermodynamics,
Elasticities and Implementation.''} \emph{Bioinformatics} 26 (12):
1528--34. \url{https://doi.org/10.1093/bioinformatics/btq141}.

\leavevmode\vadjust pre{\hypertarget{ref-monod_nature_1965}{}}%
Monod, J, J Wyman, and J P Changeux. 1965. {``On the Nature of
Allosteric Transitions: A Plausible Model.''} \emph{Journal of Molecular
Biology} 12 (May): 88--118.
\url{https://doi.org/10.1016/S0022-2836(65)80285-6}.

\leavevmode\vadjust pre{\hypertarget{ref-noor_note_2013}{}}%
Noor, Elad, Avi Flamholz, Wolfram Liebermeister, Arren Bar-Even, and Ron
Milo. 2013. {``A Note on the Kinetics of Enzyme Action: {A}
Decomposition That Highlights Thermodynamic Effects.''} \emph{FEBS
Letters}, A century of {Michaelis} - {Menten} kinetics, 587 (17):
2772--77. \url{https://doi.org/10.1016/j.febslet.2013.07.028}.

\leavevmode\vadjust pre{\hypertarget{ref-popova_generalization_1975}{}}%
Popova, S V, and E E Sel'kov. 1975. {``Generalization of the Model by
{Monod}, {Wyman} and {Changeux} for the Case of a Reversible
Monosubtrate Reaction {SR},{TP}.''} \emph{FEBS Letters} 53 (3): 269--73.
\url{https://doi.org/10.1016/0014-5793(75)80034-2}.

\leavevmode\vadjust pre{\hypertarget{ref-popova_description_1979}{}}%
---------. 1979. {``{[}{Description} of the Kinetics of the Two
Substrate Reactions {S1}+ {S2} Goes to and Comes from {S3}+{S4} by a
Generalized {Monod}, { Wyman}, {Changeux} Model{]}.''}
\emph{Molekuliarnaia Biologiia} 13 (1): 129--39.
\url{https://www.ncbi.nlm.nih.gov/pubmed/156878}.

\leavevmode\vadjust pre{\hypertarget{ref-preston-wernertomandgedampradyunTOMLSpecification0rc2020}{}}%
Preston-Werner, Tom and Gedam, Pradyun. 2020. {``{TOML Specification}
1.0.0-Rc.1.''} \url{https://toml.io/en/v1.0.0-rc.1/}.

\leavevmode\vadjust pre{\hypertarget{ref-pydanticdevelopersPydantic2022}{}}%
Pydantic developers. 2022. {``Pydantic.''}
\url{https://pypi.org/project/pydantic/}.

\leavevmode\vadjust pre{\hypertarget{ref-saa_construction_2016}{}}%
Saa, Pedro A, and Lars K Nielsen. 2016. {``Construction of Feasible and
Accurate Kinetic Models of Metabolism: {A} {Bayesian} Approach.''}
\emph{Scientific Reports} 6 (July): 29635.
\url{https://doi.org/10.1038/srep29635}.

\leavevmode\vadjust pre{\hypertarget{ref-SchmiesterSch2021}{}}%
Schmiester, Leonard, Yannik Schälte, Frank T. Bergmann, Tacio Camba,
Erika Dudkin, Janine Egert, Fabian Frö hlich, et al. 2021.
{``{PEtab}\textemdash{{Interoperable}} Specification of Parameter
Estimation Problems in Systems Biology.''} \emph{PLOS Computational
Biology} 17 (1): 1--10.
\url{https://doi.org/10.1371/journal.pcbi.1008646}.

\end{CSLReferences}

\end{document}

### 2023-11-01 paper/main.tex
submitted version

@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, breakable, enhanced, boxrule=0pt, interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, sharp corners, breakable, interior hidden, enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
 

### 2023-11-01 paper/supporting_information.tex
submitted version

@@
 \maketitle
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, boxrule=0pt, interior hidden, sharp corners, breakable, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, boxrule=0pt, frame hidden, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable]}{\end{tcolorbox}}\fi
 
@@
 {Monod}, {Wyman} and {Changeux} for the Case of a Reversible
-Monosubtrate Reaction {SR},{TP}.''} \emph{FEBS Letters} 53 (3): 269--73.
+Monosubtrate Reaction {\(S\overleftrightarrow{(R,T)}P\)}.''} \emph{FEBS
+Letters} 53 (3): 269--73.
 \url{https://doi.org/10.1016/0014-5793(75)80034-2}.

### 2024-01-09 paper/main.qmd
editorial changes

@@
         department: The Novo Nordisk Foundation Center for Biosustainability
-        city: Kongens Lyngby
+        city: Kongens Lyngby 2800
         country: Denmark
@@
         department: Australian Institute for Bioengineering and Nanotechnology (AIBN)
-        city: St Lucia
+        city: St Lucia 4067
         country: Australia
@@
 
-# References
+# Author information
+
+## Author contribution
+
+Teddy Groves and Nicholas Cowie: Writing code and documentation, designing
+statistical model, biological model, software architecture and user interface,
+writing and maintaining code, documentation and manuscript.
+
+Lars Keld Nielsen: Supervision, writing manuscript, designing statistical model,
+biological model, software architecture and user interface.   # References
+
+## Conflict of interest
+
+The authors declare no competing financial interest
 

### 2024-01-09 paper/main.tex
editorial changes

@@
 \affiliation{The Novo Nordisk Foundation Center for
-Biosustainability, DTU, Kongens Lyngby, Denmark}
+Biosustainability, DTU, Kongens Lyngby 2800, Denmark}
 
@@
 \affiliation{The Novo Nordisk Foundation Center for
-Biosustainability, DTU, Kongens Lyngby, Denmark}
+Biosustainability, DTU, Kongens Lyngby 2800, Denmark}
 \affiliation{Australian Institute for Bioengineering and Nanotechnology
-(AIBN), The University of Queensland, St Lucia, Australia}
+(AIBN), The University of Queensland, St Lucia 4067, Australia}
 
@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, sharp corners, breakable, interior hidden, enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, boxrule=0pt, breakable, interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 
-\hypertarget{references}{%
-\section{References}\label{references}}
+\hypertarget{author-information}{%
+\section{Author information}\label{author-information}}
+
+\hypertarget{author-contribution}{%
+\subsection{Author contribution}\label{author-contribution}}
+
+Teddy Groves and Nicholas Cowie: Writing code and documentation,
+designing statistical model, biological model, software architecture and
+user interface, writing and maintaining code, documentation and
+manuscript.
+
+Lars Keld Nielsen: Supervision, writing manuscript, designing
+statistical model, biological model, software architecture and user
+interface. \# References
+
+\hypertarget{conflict-of-interest}{%
+\subsection{Conflict of interest}\label{conflict-of-interest}}
+
+The authors declare no competing financial interest
 

### 2024-01-09 paper/supporting_information.qmd
editorial changes

--- 2023-10-31 paper/supporting_information.qmd
@@
 ---
-title: Supporting information
-bibliography: bibliography.bib
-number-sections: true
+title: Supporting information - Bayesian regression facilitates quantitative modelling of cell metabolism
 format:
@@
     include-in-header: mathjax.html
-  pdf: default
+  acs-pdf: 
+    include-in-header: 
+      - text: \renewcommand{\thepage}{S\arabic{page}}
+    keep-tex: true
+    journal:
+      id: asbcd6  #This is an identifier for the target journal: from https://pubs.acs.org/, the identifier is the part of the URL after https://pubs.acs.org/journal/, e.g.
+for J.
+Org.
+Chem.
+it would be joceah.
+      type: article  #article, note, communication, review, letter, or perspectivee
+      layout: traditional #traditional or twocolumn, twocolumn currently broken for longtables
   docx: default
+number-sections: true
+author: 
+  - name: "Teddy Groves"
+    orcid: 0000-0002-7109-3270
+    email: tedgro@biosustain.dtu.dk
+    affiliations: 
+      - id: biosustain
+        name: DTU
+        department: The Novo Nordisk Foundation Center for Biosustainability
+        city: Kongens Lyngby 2800
+        country: Denmark
+    attributes:
+      corresponding: true
+      equal-contributor: true
+    note: These authors contributed equally to this research.
+  - name: "Nicholas Luke Cowie"
+    orcid: 0000-0003-3367-2105
+    affiliations: 
+      - ref: biosustain
+    attributes:
+      equal-contributor: true
+    note: These authors contributed equally to this research.
+  - name: "Lars Keld Nielsen"
+    orcid: 0000-0001-8191-3511
+    affiliations: 
+      - id: uq
+        name: The University of Queensland
+        department: Australian Institute for Bioengineering and Nanotechnology (AIBN)
+        city: St Lucia 4067
+        country: Australia
+      - ref: biosustain
+bibliography: bibliography.bib
 ---
+
+{{< pagebreak >}}
 
@@
 
-: Parameters of Maud's statistical model {#tbl-params}
+: Table S1 – Parameters of Maud's statistical model {#tbl-params}
 
@@
 To describe saturation we use equation \eqref{eq-saturation}, which is taken
-from @liebermeister_modular_2010. Additionally, this term captures competitive
+from @liebermeister_modular_2010 and @noor_note_2013. Additionally, this term captures competitive
 inhibition: as competitive inhibitor concentration increases, the saturation
@@
 
-: Parameter specification, marginal prior distributions and true parameter values used in our case study. {#tbl-case-study-params}
+: Table S2 – Parameter specification, marginal prior distributions and true parameter values used in our case study. {#tbl-case-study-params}
 

### 2024-01-09 paper/supporting_information.tex
editorial changes

--- 2023-11-01 paper/supporting_information.tex
@@
 %
-\documentclass[
-  letterpaper,
-  DIV=11,
-  numbers=noendperiod]{scrartcl}
+\documentclass[journal=asbcd6,manuscript=article,layout=traditional]{achemso}
+\usepackage[version=3]{mhchem}
+\newcommand*\mycommand[1]{\texttt{\emph{#1}}}
 
@@
 \makeatother
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
-  \fi
-  % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
-\KOMAoption{captions}{tableheading}
+\renewcommand{\thepage}{S\arabic{page}}
 \makeatletter
@@
 \hypersetup{
-  pdftitle={Supporting information},
+  pdftitle={Supporting information - Bayesian regression facilitates quantitative modelling of cell metabolism},
+  pdfauthor={Teddy Groves; Nicholas Luke Cowie; Lars Keld Nielsen},
   colorlinks=true,
@@
 
-\title{Supporting information}
-\author{}
-\date{}
+\author{Teddy Groves}
+\affiliation{The Novo Nordisk Foundation Center for
+Biosustainability, DTU, Kongens Lyngby 2800, Denmark}
 
+\altaffiliation{These authors contributed equally to this research.}
+
+\email{tedgro@biosustain.dtu.dk}
+\author{Nicholas Luke Cowie}
+\affiliation{The Novo Nordisk Foundation Center for
+Biosustainability, DTU, Kongens Lyngby 2800, Denmark}
+
+\altaffiliation{These authors contributed equally to this research.}
+
+\author{Lars Keld Nielsen}
+\affiliation{The Novo Nordisk Foundation Center for
+Biosustainability, DTU, Kongens Lyngby 2800, Denmark}
+\affiliation{Australian Institute for Bioengineering and Nanotechnology
+(AIBN), The University of Queensland, St Lucia 4067, Australia}
+
+\title[]{Supporting information - Bayesian regression facilitates
+quantitative modelling of cell metabolism}
+\makeatletter
 \begin{document}
 \maketitle
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, boxrule=0pt, interior hidden, sharp corners, breakable, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, breakable, sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, interior hidden, frame hidden]}{\end{tcolorbox}}\fi
+
+\newpage{}
 
@@
 Maud inputs are structured directories, somewhat inspired by the PEtab
-format (Schmiester et al. 2021). A Maud input directory must contain a
-toml (Preston-Werner, Tom and Gedam, Pradyun 2020) file called
-\texttt{config.toml} which gives the input a name, configures how Maud
-will be run and tells Maud where to find the other files, allowing these
-to have custom names. It must also include a file containing a kinetic
-model definition, a file specifying information about parameters and a
-file with information experiments. The required structure of these files
-is documented at
+format \citep{SchmiesterSch2021}. A Maud input directory must contain a
+toml \citep{preston-wernertomandgedampradyunTOMLSpecification0rc2020}
+file called \texttt{config.toml} which gives the input a name,
+configures how Maud will be run and tells Maud where to find the other
+files, allowing these to have custom names. It must also include a file
+containing a kinetic model definition, a file specifying information
+about parameters and a file with information experiments. The required
+structure of these files is documented at
 \url{https://maud-metabolic-models.readthedocs.io/en/latest/inputting.html}.
-The input is validated against a Pydantic (Pydantic developers 2022)
-data model.
+The input is validated against a Pydantic
+\citep{pydanticdevelopersPydantic2022} data model.
 
 We chose to implement a custom input format despite the existence of
-standard formats in similar areas, including SBML (Keating et al. 2020)
-and PEtab (Schmiester et al. 2021). This choice was partly motivated by
-the need to ensure flexibility as Maud was developed, but there are also
-features of SBML and PEtab that make them structurally unsuitable in
-this context. Our requirements for an input format included that it be
+standard formats in similar areas, including SBML
+\citep{keatingSBMLLevelExtensible2020} and PEtab
+\citep{SchmiesterSch2021}. This choice was partly motivated by the need
+to ensure flexibility as Maud was developed, but there are also features
+of SBML and PEtab that make them structurally unsuitable in this
+context. Our requirements for an input format included that it be
 mathematics-free, so that all mathematical details are encapsulated in
@@
   >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2778}}@{}}
-\caption{\label{tbl-params}Parameters of Maud's statistical
+\caption{\label{tbl-params}Table S1 -- Parameters of Maud's statistical
 model}\tabularnewline
@@
 reaction's substrates. To describe saturation we use equation
-\eqref{eq-saturation}, which is taken from Liebermeister, Uhlendorf, and
-Klipp (2010) and Noor et al. (2013). Additionally, this term captures
-competitive inhibition: as competitive inhibitor concentration
-increases, the saturation denominator increases, effectively decreasing
-the saturation of the substrate on the total enzyme pool. Conversely, as
-the substrate concentration increases this term approaches 1.
+\eqref{eq-saturation}, which is taken from
+\citet{liebermeister_modular_2010} and \citet{noor_note_2013}.
+Additionally, this term captures competitive inhibition: as competitive
+inhibitor concentration increases, the saturation denominator increases,
+effectively decreasing the saturation of the substrate on the total
+enzyme pool. Conversely, as the substrate concentration increases this
+term approaches 1.
 
@@
 this phenomenon, following the generalised MWC approach described in
-Monod, Wyman, and Changeux (1965), Changeux (2013), Popova and Sel'kov
-(1975) and Popova and Sel'kov (1979).
+\citet{monod_nature_1965}, \citet{changeux_2013},
+\citet{popova_generalization_1975} and \citet{popova_description_1979}.
 
@@
 
-Starting with the model in Saa and Nielsen (2016), we extracted values
-for enzyme concentrations, boundary conditions and fluxes. We used these
-values to generate MCMC samples using Maud using the priors specified in
-section Section~\ref{sec-methionine-priors}. When this was finished, we
-selected one sample with relatively high log probability to use as a
-ground truth in our case study. These parameter values are shown below
-in table Table~\ref{tbl-case-study-params}. We manually inspected the
-parameter values to screen for any obviously implausible values; we did
-not find any of these.
+Starting with the model in \citet{saa_construction_2016}, we extracted
+values for enzyme concentrations, boundary conditions and fluxes. We
+used these values to generate MCMC samples using Maud using the priors
+specified in section Section~\ref{sec-methionine-priors}. When this was
+finished, we selected one sample with relatively high log probability to
+use as a ground truth in our case study. These parameter values are
+shown below in table Table~\ref{tbl-case-study-params}. We manually
+inspected the parameter values to screen for any obviously implausible
+values; we did not find any of these.
 
@@
   >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2477}}@{}}
-\caption{\label{tbl-case-study-params}Parameter specification, marginal
-prior distributions and true parameter values used in our case
-study.}\tabularnewline
+\caption{\label{tbl-case-study-params}Table S2 -- Parameter
+specification, marginal prior distributions and true parameter values
+used in our case study.}\tabularnewline
 \toprule\noalign{}
@@
 were modelled as unknown had a multivariate normal prior distribution
-derived from eQuilibrator (Beber et al. 2021).
+derived from eQuilibrator
+\citep{beberEQuilibratorPlatformEstimation2021}.
 
@@
 
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-beberEQuilibratorPlatformEstimation2021}{}}%
-Beber, Moritz E., Mattia G. Gollub, Dana Mozaffari, Kevin M. Shebek, and
-Elad Noor. 2021. {``{eQuilibrator} 3.0 -- a Platform for the Estimation
-of Thermodynamic Constants.''} \emph{arXiv:2103.00621 {[}q-Bio{]}},
-February. \url{http://arxiv.org/abs/2103.00621}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-changeux_2013}{}}%
-Changeux, Jean-Pierre. 2013. {``50 Years of Allosteric Interactions: The
-Twists and Turns of the Models.''} \emph{Nature Reviews. Molecular Cell
-Biology} 14 (12): 819--29. \url{https://doi.org/10.1038/nrm3695}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-keatingSBMLLevelExtensible2020}{}}%
-Keating, Sarah M, Dagmar Waltemath, Matthias König, Fengkai Zhang,
-Andreas Dräger, Claudine Chaouiya, Frank T Bergmann, et al. 2020.
-{``{\textsc{SBML}} {Level} 3: An Extensible Format for the Exchange and
-Reuse of Biological Models.''} \emph{Molecular Systems Biology} 16 (8):
-e9110. \url{https://doi.org/10.15252/msb.20199110}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-liebermeister_modular_2010}{}}%
-Liebermeister, Wolfram, Jannis Uhlendorf, and Edda Klipp. 2010.
-{``Modular Rate Laws for Enzymatic Reactions: Thermodynamics,
-Elasticities and Implementation.''} \emph{Bioinformatics} 26 (12):
-1528--34. \url{https://doi.org/10.1093/bioinformatics/btq141}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-monod_nature_1965}{}}%
-Monod, J, J Wyman, and J P Changeux. 1965. {``On the Nature of
-Allosteric Transitions: A Plausible Model.''} \emph{Journal of Molecular
-Biology} 12 (May): 88--118.
-\url{https://doi.org/10.1016/S0022-2836(65)80285-6}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-noor_note_2013}{}}%
-Noor, Elad, Avi Flamholz, Wolfram Liebermeister, Arren Bar-Even, and Ron
-Milo. 2013. {``A Note on the Kinetics of Enzyme Action: {A}
-Decomposition That Highlights Thermodynamic Effects.''} \emph{FEBS
-Letters}, A century of {Michaelis} - {Menten} kinetics, 587 (17):
-2772--77. \url{https://doi.org/10.1016/j.febslet.2013.07.028}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-popova_generalization_1975}{}}%
-Popova, S V, and E E Sel'kov. 1975. {``Generalization of the Model by
-{Monod}, {Wyman} and {Changeux} for the Case of a Reversible
-Monosubtrate Reaction {SR},{TP}.''} \emph{FEBS Letters} 53 (3): 269--73.
-\url{https://doi.org/10.1016/0014-5793(75)80034-2}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-popova_description_1979}{}}%
----------. 1979. {``{[}{Description} of the Kinetics of the Two
-Substrate Reactions {S1}+ {S2} Goes to and Comes from {S3}+{S4} by a
-Generalized {Monod}, { Wyman}, {Changeux} Model{]}.''}
-\emph{Molekuliarnaia Biologiia} 13 (1): 129--39.
-\url{https://www.ncbi.nlm.nih.gov/pubmed/156878}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-preston-wernertomandgedampradyunTOMLSpecification0rc2020}{}}%
-Preston-Werner, Tom and Gedam, Pradyun. 2020. {``{TOML Specification}
-1.0.0-Rc.1.''} \url{https://toml.io/en/v1.0.0-rc.1/}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-pydanticdevelopersPydantic2022}{}}%
-Pydantic developers. 2022. {``Pydantic.''}
-\url{https://pypi.org/project/pydantic/}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-saa_construction_2016}{}}%
-Saa, Pedro A, and Lars K Nielsen. 2016. {``Construction of Feasible and
-Accurate Kinetic Models of Metabolism: {A} {Bayesian} Approach.''}
-\emph{Scientific Reports} 6 (July): 29635.
-\url{https://doi.org/10.1038/srep29635}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-SchmiesterSch2021}{}}%
-Schmiester, Leonard, Yannik Schälte, Frank T. Bergmann, Tacio Camba,
-Erika Dudkin, Janine Egert, Fabian Frö hlich, et al. 2021.
-{``{PEtab}\textemdash{{Interoperable}} Specification of Parameter
-Estimation Problems in Systems Biology.''} \emph{PLOS Computational
-Biology} 17 (1): 1--10.
-\url{https://doi.org/10.1371/journal.pcbi.1008646}.
-
-\end{CSLReferences}
+\renewcommand{\bibsection}{}
+\bibliography{bibliography.bib}
 

### 2024-01-12 paper/main.qmd
review response changes

@@
 
-Maud represents information from measurements using generalised linear
-regression models that probabilistically connect realised measurements with
-true values of the measurable quantities. Information from other sources
-is represented using a prior distribution over a set of latent parameters.
-The parameters and the measurable quantities are connected by a generative
-model encompassing Maud’s kinetic model as well as the steady state equation
-\eqref{eq-3}. Together, the measurement model, prior model and generative model
-determine a joint probability function that assigns a probability density to any
-possible combination of measurements and parameters.
+Maud uses Markov Chain Monte Carlo to perform posterior inference on a Bayesian
+statistical model. This section introduces these topics and then describes
+exactly what kind of Bayesian statistical model Maud uses.
 
-Below we describe the prior and measurement models, as the generative model has
-already been discussed above.
+### Bayesian inference
+
+Bayesian statistical inference analyses data by constructing a mathematical
+model with the following three components:
+
+- A measurement model or "likelihood" that probabilistically describes
+how likely any possible measurement set is given the true values
+of the measured quantities, i.e. a probability density function $l:
+\mathfrak{Y}\times\hat{\mathfrak{Y}} \rightarrow [0, 1]$ where $\mathfrak{Y}$
+is the set of all possible measurement sets and $\hat{\mathfrak{Y}}$ is the set
+of all possible true measurable values.
+
+- A deterministic process model that describes the measured quantities in
+terms of unknown, possibly unobserved parameters $\theta$, i.e. a function $d:
+\mathfrak{theta} -> \hat{mathfrak{Y}}$.
+
+- A prior model that probabilistically describes how likely any possible set
+of parameter values is, without considering any information included in the
+measurement model, i.e. a probability density function $\pi : \mathfrak{theta}
+\rightarrow [0, 1]$.
+
+Together these components determine a joint probability function
+that encapsulates the Bayesian statistical model, i.e. a function $p:
+\mathfrak{Y}\times\mathfrak{theta}\rightarrow[0,1]$ such that for any $\theta$
+and $y$, $p(\theta, y) = \pi(\theta)l(y\mid\theta)$. Substantive questions about
+the implications of the assumptions implicit in the model components can be
+formulated in terms of this joint density function.
+
+In particular, questions about what the model implies, given a particular
+measurement set $y$ can be formulated in terms of the conditional probability
+density function $p_{\mid y}:\mathfrak{theta}\rightarrow[0,1]$. Bayes's
+theorem guarantees that this conditional density is proportional to the
+product of the prior and likelihood, i.e. $p_{\mid y}(\theta) \propto
+\pi(\theta)l(y\mid\theta)$ See [@gelmanBayesianDataAnalysis2020a, Ch 1] for an
+extended discussion of the theory of Bayesian statistical inference.
+
+The principal computational challenge for Bayesian statistical inference is to
+approximate the values of integrals of the joint density function conditional
+on a known measurement assignment $y$, also known as the posterior distribution.
+For most non-trivial statistical models the posterior distribution cannot be
+integrated analytically, so numerical approximation is required. Markov Chain
+Monte Carlo or "MCMC" is a popular method for addressing this problem that
+uses a Markov chain with known properties to generate samples from a target
+probability distribution that can be used to perform Monte Carlo integration.
+Hamiltonian Monte Carlo makes it possible, given an appropriate choice of
+hyperparameters, to efficiently generate samples from even a high dimensional
+continuous posterior distribution, by calculating the gradients of the log-scale
+joint density function. Adaptive Hamiltonian Monte Carlo, as used by Stan, uses
+a range of well-tested methods to tune these hyperparameters, allowing efficient
+MCMC for high dimensional posterior distributions with minimal user input.
+
+Alternatives to MCMC for numerically approximating integrals over posterior
+distributions include grid sampling [@gelmanBayesianDataAnalysis2020a,
+Ch. 10], rejection sampling [@gelmanBayesianDataAnalysis2020a, Ch. 10],
+importance sampling [@vehtariParetoSmoothedImportance2022] and distributional
+approaches including variational inference and Laplace approximation
+[@gelmanBayesianDataAnalysis2020a, Ch. 13].
+
+The Laplace approximation is of particular interest because, as mentioned above,
+this has been used for approximate Bayesian inference in a similar context
+to Maud. Laplace approximation of a posterior distribution works by first
+finding the mode of the posterior distribution, i.e. the 'maximum a posteriori'
+parameter configuration with the highest posterior density. Next, the second
+derivative of the posterior distribution at this point is found and used to
+generate a normal distribution that either approximates the target distribution,
+or can in turn be used to generate such an approximation.
+
+Maud employs adaptive Hamiltonian Monte Carlo to perform posterior inference
+for a Bayesian statistical model where the process model is the kinetic model
+described above. The next two subsections describe Maud's prior and measurement
+models.
 

### 2024-01-12 paper/main.tex
review response changes

@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, boxrule=0pt, breakable, interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, interior hidden, frame hidden, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 
-Maud represents information from measurements using generalised linear
-regression models that probabilistically connect realised measurements
-with true values of the measurable quantities. Information from other
-sources is represented using a prior distribution over a set of latent
-parameters. The parameters and the measurable quantities are connected
-by a generative model encompassing Maud's kinetic model as well as the
-steady state equation \eqref{eq-3}. Together, the measurement model,
-prior model and generative model determine a joint probability function
-that assigns a probability density to any possible combination of
-measurements and parameters.
-
-Below we describe the prior and measurement models, as the generative
-model has already been discussed above.
+Maud uses Markov Chain Monte Carlo to perform posterior inference on a
+Bayesian statistical model. This section introduces these topics and
+then describes exactly what kind of Bayesian statistical model Maud
+uses.
+
+\hypertarget{bayesian-inference}{%
+\subsubsection{Bayesian inference}\label{bayesian-inference}}
+
+Bayesian statistical inference analyses data by constructing a
+mathematical model with the following three components:
+
+\begin{itemize}
+\item
+  A measurement model or ``likelihood'' that probabilistically describes
+  how likely any possible measurement set is given the true values of
+  the measured quantities, i.e.~a probability density function
+  \(l: \mathfrak{Y}\times\hat{\mathfrak{Y}} \rightarrow [0, 1]\) where
+  \(\mathfrak{Y}\) is the set of all possible measurement sets and
+  \(\hat{\mathfrak{Y}}\) is the set of all possible true measurable
+  values.
+\item
+  A deterministic process model that describes the measured quantities
+  in terms of unknown, possibly unobserved parameters \(\theta\), i.e.~a
+  function \(d: \mathfrak{theta} -> \hat{mathfrak{Y}}\).
+\item
+  A prior model that probabilistically describes how likely any possible
+  set of parameter values is, without considering any information
+  included in the measurement model, i.e.~a probability density function
+  \(\pi : \mathfrak{theta} \rightarrow [0, 1]\).
+\end{itemize}
+
+Together these components determine a joint probability function that
+encapsulates the Bayesian statistical model, i.e.~a function
+\(p: \mathfrak{Y}\times\mathfrak{theta}\rightarrow[0,1]\) such that for
+any \(\theta\) and \(y\), \(p(\theta, y) = \pi(\theta)l(y\mid\theta)\).
+Substantive questions about the implications of the assumptions implicit
+in the model components can be formulated in terms of this joint density
+function.
+
+In particular, questions about what the model implies, given a
+particular measurement set \(y\) can be formulated in terms of the
+conditional probability density function
+\(p_{\mid y}:\mathfrak{theta}\rightarrow[0,1]\). Bayes's theorem
+guarantees that this conditional density is proportional to the product
+of the prior and likelihood,
+i.e.~\(p_{\mid y}(\theta) \propto \pi(\theta)l(y\mid\theta)\) See
+\citep[Ch 1]{gelmanBayesianDataAnalysis2020a} for an extended discussion
+of the theory of Bayesian statistical inference.
+
+The principal computational challenge for Bayesian statistical inference
+is to approximate the values of integrals of the joint density function
+conditional on a known measurement assignment \(y\), also known as the
+posterior distribution. For most non-trivial statistical models the
+posterior distribution cannot be integrated analytically, so numerical
+approximation is required. Markov Chain Monte Carlo or ``MCMC'' is a
+popular method for addressing this problem that uses a Markov chain with
+known properties to generate samples from a target probability
+distribution that can be used to perform Monte Carlo integration.
+Hamiltonian Monte Carlo makes it possible, given an appropriate choice
+of hyperparameters, to efficiently generate samples from even a high
+dimensional continuous posterior distribution, by calculating the
+gradients of the log-scale joint density function. Adaptive Hamiltonian
+Monte Carlo, as used by Stan, uses a range of well-tested methods to
+tune these hyperparameters, allowing efficient MCMC for high dimensional
+posterior distributions with minimal user input.
+
+Alternatives to MCMC for numerically approximating integrals over
+posterior distributions include grid sampling \citep[ Ch.
+10]{gelmanBayesianDataAnalysis2020a}, rejection sampling \citep[Ch.
+10]{gelmanBayesianDataAnalysis2020a}, importance sampling
+\citep{vehtariParetoSmoothedImportance2022} and distributional
+approaches including variational inference and Laplace approximation
+\citep[Ch. 13]{gelmanBayesianDataAnalysis2020a}.
+
+The Laplace approximation is of particular interest because, as
+mentioned above, this has been used for approximate Bayesian inference
+in a similar context to Maud. Laplace approximation of a posterior
+distribution works by first finding the mode of the posterior
+distribution, i.e.~the `maximum a posteriori' parameter configuration
+with the highest posterior density. Next, the second derivative of the
+posterior distribution at this point is found and used to generate a
+normal distribution that either approximates the target distribution, or
+can in turn be used to generate such an approximation.
+
+Maud employs adaptive Hamiltonian Monte Carlo to perform posterior
+inference for a Bayesian statistical model where the process model is
+the kinetic model described above. The next two subsections describe
+Maud's prior and measurement models.
 

### 2024-01-12 paper/supporting_information.qmd
review response changes

@@
 
+# Multimodal posterior distributions
+
+While we have not yet observed this in practice, we expect that Maud is capable
+of accurately sampling mildly multi-modal posterior distributions, i.e. those
+for which regions of parameter space with high probability density are not
+very sharply separated. This is because this ability depends primarily on the
+underlying inference algorithm, and adaptive Hamiltonian Monte Carlo is known
+to be able to sample many such posterior distributions. See, for example,
+@mangiolaSccompRobustDifferential2023, which reports the successful use of
+adaptive Hamiltonian Monte Carlo to sample posterior distributions involving
+mixtures of Gaussian distributions, which are typically multi-modal.
+
+Moreover, in cases where it fails to sample a multi-modal posterior distribution
+adaptive Hamiltonian Monte Carlo typically exhibits poor mixing and divergent
+transitions, which Maud is set up to detect automatically, making this case easy
+for users to diagnose.
+
+Nonetheless, it is also well known that strongly multi-modal posterior
+distributions pose a problem for adaptive Hamiltonian Monte Carlo, as well as
+other MCMC algorithms. See [@betancourtmichaelMarkovChainMonte2020, §3.2] for
+a general discussion of this issue. @mangoubiDoesHamiltonianMonte2018 explores
+the related question of the relative performance of Hamiltonian Monte Carlo vs
+random walk Metropolis Hastings, finding that, for certain cases where the target
+distribution exhibits highly disconnected and similarly shaped modes, the two
+algorithms have similar performance.
+
+It is therefore important to pay careful attention to MCMC diagnostics and
+monitor developments in computational statistics that might expand the range of
+posterior distributions that can practically be sampled.
+
+@zamora-silleroEfficientCharacterizationHighdimensional2011 introduces
+the method HYPERSPACE, which is designed to identify disconnected regions
+of a biochemical parameter space and characterise them by estimating
+their dimensions. HYPERSPACE does not perform parameter inference and is
+therefore not directly comparable with algorithms like adaptive Hamiltonian
+Monte Carlo as used in Maud, but could plausibly be adapted into such an
+algorithm, for example by supplementation with importance sampling (see
+@vehtariParetoSmoothedImportance2022). However, any such algorithm would
+be limited by HYPERSPACE's dependency on a random walk Metropolis Hastings
+algorithm. This kind of algorithm's performance is known to scale poorly
+compared with Hamiltonian Monte Carlo algorithms as the number of parameters
+increases: see @mangoubiDoesHamiltonianMonte2018. Since Maud aims to fit models
+with hundreds of parameters, this means that it is unlikely that HYPERSPACE can
+directly be used to improve Maud's efficiency.
+
 # References

### 2024-01-12 paper/supporting_information.tex
review response changes

@@
 \maketitle
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, breakable, sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, interior hidden, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, breakable, frame hidden, interior hidden, enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
 
@@
 
+\hypertarget{multimodal-posterior-distributions}{%
+\section{Multimodal posterior
+distributions}\label{multimodal-posterior-distributions}}
+
+While we have not yet observed this in practice, we expect that Maud is
+capable of accurately sampling mildly multi-modal posterior
+distributions, i.e.~those for which regions of parameter space with high
+probability density are not very sharply separated. This is because this
+ability depends primarily on the underlying inference algorithm, and
+adaptive Hamiltonian Monte Carlo is known to be able to sample many such
+posterior distributions. See, for example,
+\citet{mangiolaSccompRobustDifferential2023}, which reports the
+successful use of adaptive Hamiltonian Monte Carlo to sample posterior
+distributions involving mixtures of Gaussian distributions, which are
+typically multi-modal.
+
+Moreover, in cases where it fails to sample a multi-modal posterior
+distribution adaptive Hamiltonian Monte Carlo typically exhibits poor
+mixing and divergent transitions, which Maud is set up to detect
+automatically, making this case easy for users to diagnose.
+
+Nonetheless, it is also well known that strongly multi-modal posterior
+distributions pose a problem for adaptive Hamiltonian Monte Carlo, as
+well as other MCMC algorithms. See
+\citep[§3.2]{betancourtmichaelMarkovChainMonte2020} for a general
+discussion of this issue. \citet{mangoubiDoesHamiltonianMonte2018}
+explores the related question of the relative performance of Hamiltonian
+Monte Carlo vs random walk Metropolis Hastings, finding that, for
+certain cases where the target distribution exhibits highly disconnected
+and similarly shaped modes, the two algorithms have similar performance.
+
+It is therefore important to pay careful attention to MCMC diagnostics
+and monitor developments in computational statistics that might expand
+the range of posterior distributions that can practically be sampled.
+
+\citet{zamora-silleroEfficientCharacterizationHighdimensional2011}
+introduces the method HYPERSPACE, which is designed to identify
+disconnected regions of a biochemical parameter space and characterise
+them by estimating their dimensions. HYPERSPACE does not perform
+parameter inference and is therefore not directly comparable with
+algorithms like adaptive Hamiltonian Monte Carlo as used in Maud, but
+could plausibly be adapted into such an algorithm, for example by
+supplementation with importance sampling (see
+\citet{vehtariParetoSmoothedImportance2022}). However, any such
+algorithm would be limited by HYPERSPACE's dependency on a random walk
+Metropolis Hastings algorithm. This kind of algorithm's performance is
+known to scale poorly compared with Hamiltonian Monte Carlo algorithms
+as the number of parameters increases: see
+\citet{mangoubiDoesHamiltonianMonte2018}. Since Maud aims to fit models
+with hundreds of parameters, this means that it is unlikely that
+HYPERSPACE can directly be used to improve Maud's efficiency.
+
 \hypertarget{references}{%

### 2024-01-12 paper/main.qmd
update gh action and fix equations

@@
 of the measured quantities, i.e. a probability density function $l:
-\mathfrak{Y}\times\hat{\mathfrak{Y}} \rightarrow [0, 1]$ where $\mathfrak{Y}$
-is the set of all possible measurement sets and $\hat{\mathfrak{Y}}$ is the set
+\mathcal{Y}\times\hat{\mathcal{Y}} \rightarrow [0, 1]$ where $\mathcal{Y}$
+is the set of all possible measurement sets and $\hat{\mathcal{Y}}$ is the set
 of all possible true measurable values.
@@
 terms of unknown, possibly unobserved parameters $\theta$, i.e. a function $d:
-\mathfrak{theta} -> \hat{mathfrak{Y}}$.
+\mathcal{\theta} -> \hat{\mathcal{Y}}$.
 
@@
 of parameter values is, without considering any information included in the
-measurement model, i.e. a probability density function $\pi : \mathfrak{theta}
+measurement model, i.e. a probability density function $\pi : \mathcal{\theta}
 \rightarrow [0, 1]$.
@@
 that encapsulates the Bayesian statistical model, i.e. a function $p:
-\mathfrak{Y}\times\mathfrak{theta}\rightarrow[0,1]$ such that for any $\theta$
+\mathcal{Y}\times\mathcal{\theta}\rightarrow[0,1]$ such that for any $\theta$
 and $y$, $p(\theta, y) = \pi(\theta)l(y\mid\theta)$. Substantive questions about
@@
 measurement set $y$ can be formulated in terms of the conditional probability
-density function $p_{\mid y}:\mathfrak{theta}\rightarrow[0,1]$. Bayes's
+density function $p_{\mid y}:\mathcal{theta}\rightarrow[0,1]$. Bayes's
 theorem guarantees that this conditional density is proportional to the

### 2024-01-12 paper/main.tex
update gh action and fix equations

@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, interior hidden, frame hidden, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, frame hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, interior hidden, enhanced]}{\end{tcolorbox}}\fi
 
@@
   the measured quantities, i.e.~a probability density function
-  \(l: \mathfrak{Y}\times\hat{\mathfrak{Y}} \rightarrow [0, 1]\) where
-  \(\mathfrak{Y}\) is the set of all possible measurement sets and
-  \(\hat{\mathfrak{Y}}\) is the set of all possible true measurable
+  \(l: \mathcal{Y}\times\hat{\mathcal{Y}} \rightarrow [0, 1]\) where
+  \(\mathcal{Y}\) is the set of all possible measurement sets and
+  \(\hat{\mathcal{Y}}\) is the set of all possible true measurable
   values.
@@
   in terms of unknown, possibly unobserved parameters \(\theta\), i.e.~a
-  function \(d: \mathfrak{theta} -> \hat{mathfrak{Y}}\).
+  function \(d: \mathcal{\theta} -> \hat{\mathcal{Y}}\).
 \item
@@
   included in the measurement model, i.e.~a probability density function
-  \(\pi : \mathfrak{theta} \rightarrow [0, 1]\).
+  \(\pi : \mathcal{\theta} \rightarrow [0, 1]\).
 \end{itemize}
@@
 encapsulates the Bayesian statistical model, i.e.~a function
-\(p: \mathfrak{Y}\times\mathfrak{theta}\rightarrow[0,1]\) such that for
+\(p: \mathcal{Y}\times\mathcal{\theta}\rightarrow[0,1]\) such that for
 any \(\theta\) and \(y\), \(p(\theta, y) = \pi(\theta)l(y\mid\theta)\).
@@
 conditional probability density function
-\(p_{\mid y}:\mathfrak{theta}\rightarrow[0,1]\). Bayes's theorem
+\(p_{\mid y}:\mathcal{theta}\rightarrow[0,1]\). Bayes's theorem
 guarantees that this conditional density is proportional to the product

### 2024-01-12 paper/supporting_information.tex
update gh action and fix equations

@@
 \maketitle
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, breakable, frame hidden, interior hidden, enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, interior hidden, frame hidden, sharp corners, enhanced, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
 

### 2024-02-08 paper/main.tex
Review responses

@@
 \makeatletter
-\makeatother
-\makeatletter
-\makeatother
-\makeatletter
 \@ifpackageloaded{caption}{}{\usepackage{caption}}
@@
 \makeatletter
+\makeatother
+\makeatletter
 \@ifpackageloaded{caption}{}{\usepackage{caption}}
 \@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
-\makeatother
-\makeatletter
-\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
-\makeatother
-\makeatletter
-\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
-\makeatother
-\makeatletter
-\makeatother
-\makeatletter
 \makeatother
@@
 \fi
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\usepackage{bookmark}
+
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, frame hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, interior hidden, enhanced]}{\end{tcolorbox}}\fi
-
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
+
+\section{Introduction}\label{introduction}
 
@@
 
-\hypertarget{results-and-discussion}{%
-\section{Results and Discussion}\label{results-and-discussion}}
+\section{Results and Discussion}\label{results-and-discussion}
 
@@
 
-\begin{minipage}[t]{\linewidth}
-
-{\centering
-
-\raisebox{-\height}{
+\begin{minipage}{\linewidth}
 
 \includegraphics{./figures/methionine-reactions.png}
-
-}
-
-}
 
@@
 
-\end{figure}
-
-\hypertarget{dataset-and-model-specification}{%
+\end{figure}%
+
 \subsection{Dataset and model
-specification}\label{dataset-and-model-specification}}
+specification}\label{dataset-and-model-specification}
 
@@
 
-\hypertarget{posterior-inference}{%
-\subsection{Posterior inference}\label{posterior-inference}}
+\subsection{Posterior inference}\label{posterior-inference}
 
@@
 
-\begin{minipage}[t]{\linewidth}
-
-{\centering
-
-\raisebox{-\height}{
+\begin{minipage}{\linewidth}
 
 \includegraphics{./figures/posterior.png}
-
-}
-
-}
 
@@
 
-\end{figure}
-
-\hypertarget{sec-laplace}{%
-\subsubsection{Comparison with Laplace
-approximation}\label{sec-laplace}}
+\end{figure}%
+
+\subsubsection{Comparison with Laplace approximation}\label{sec-laplace}
 
@@
 
-\begin{minipage}[t]{\linewidth}
-
-{\centering
-
-\raisebox{-\height}{
+\begin{minipage}{\linewidth}
 
 \includegraphics{./figures/laplace.png}
-
-}
-
-}
 
@@
 
-\end{figure}
-
-\hypertarget{effect-of-missing-metabolite-concentration-measurements}{%
+\end{figure}%
+
 \subsection{Effect of missing metabolite concentration
-measurements}\label{effect-of-missing-metabolite-concentration-measurements}}
+measurements}\label{effect-of-missing-metabolite-concentration-measurements}
 
@@
 
-\begin{minipage}[t]{\linewidth}
-
-{\centering
-
-\raisebox{-\height}{
+\begin{minipage}{\linewidth}
 
 \includegraphics{./figures/missing.png}
-
-}
-
-}
 
@@
 
-\end{figure}
-
-\hypertarget{application-to-regulatory-understanding}{%
+\end{figure}%
+
 \subsection{Application to regulatory
-understanding}\label{application-to-regulatory-understanding}}
+understanding}\label{application-to-regulatory-understanding}
 
@@
 
-\begin{minipage}[t]{\linewidth}
-
-{\centering
-
-\raisebox{-\height}{
+\begin{minipage}{\linewidth}
 
 \includegraphics{./figures/decomposition.png}
-
-}
-
-}
 
@@
 
-\end{figure}
-
-\hypertarget{methods}{%
-\section{Methods}\label{methods}}
+\end{figure}%
+
+\section{Methods}\label{methods}
 
@@
 
-\hypertarget{kinetic-model}{%
-\subsection{Kinetic model}\label{kinetic-model}}
+\subsection{Kinetic model}\label{kinetic-model}
 
@@
 
-\hypertarget{statistical-model}{%
-\subsection{Statistical model}\label{statistical-model}}
+\subsection{Statistical model}\label{statistical-model}
 
@@
 
-\hypertarget{bayesian-inference}{%
-\subsubsection{Bayesian inference}\label{bayesian-inference}}
+\subsubsection{Bayesian inference}\label{bayesian-inference}
 
@@
   how likely any possible measurement set is given the true values of
-  the measured quantities, i.e.~a probability density function
-  \(l: \mathcal{Y}\times\hat{\mathcal{Y}} \rightarrow [0, 1]\) where
+  the measured quantities, i.e.~a probability density function \(l:
+  \mathcal{Y}\times\hat{\mathcal{Y}} \rightarrow [0, 1]\) where
   \(\mathcal{Y}\) is the set of all possible measurement sets and
@@
   in terms of unknown, possibly unobserved parameters \(\theta\), i.e.~a
-  function \(d: \mathcal{\theta} -> \hat{\mathcal{Y}}\).
+  function \(d:
+  \mathcal{\theta} -> \hat{\mathcal{Y}}\).
 \item
@@
   included in the measurement model, i.e.~a probability density function
-  \(\pi : \mathcal{\theta} \rightarrow [0, 1]\).
+  \(\pi : \mathcal{\theta}
+  \rightarrow [0, 1]\).
 \end{itemize}
@@
 Together these components determine a joint probability function that
-encapsulates the Bayesian statistical model, i.e.~a function
-\(p: \mathcal{Y}\times\mathcal{\theta}\rightarrow[0,1]\) such that for
-any \(\theta\) and \(y\), \(p(\theta, y) = \pi(\theta)l(y\mid\theta)\).
+encapsulates the Bayesian statistical model, i.e.~a function \(p:
+\mathcal{Y}\times\mathcal{\theta}\rightarrow[0,1]\) such that for any
+\(\theta\) and \(y\), \(p(\theta, y) = \pi(\theta)l(y\mid\theta)\).
 Substantive questions about the implications of the assumptions implicit
@@
 guarantees that this conditional density is proportional to the product
-of the prior and likelihood,
-i.e.~\(p_{\mid y}(\theta) \propto \pi(\theta)l(y\mid\theta)\) See
-\citep[Ch 1]{gelmanBayesianDataAnalysis2020a} for an extended discussion
-of the theory of Bayesian statistical inference.
+of the prior and likelihood, i.e.~\(p_{\mid y}(\theta) \propto
+\pi(\theta)l(y\mid\theta)\) See \citep[Ch
+1]{gelmanBayesianDataAnalysis2020a} for an extended discussion of the
+theory of Bayesian statistical inference.
 
@@
 
-\hypertarget{prior-model}{%
-\subsubsection{Prior model}\label{prior-model}}
+\subsubsection{Prior model}\label{prior-model}
 
@@
 
-\hypertarget{measurement-model}{%
-\subsubsection{Measurement model}\label{measurement-model}}
+\subsubsection{Measurement model}\label{measurement-model}
 
@@
 
-\hypertarget{implementation}{%
-\subsection{Implementation}\label{implementation}}
+\subsection{Implementation}\label{implementation}
 
@@
 
-\hypertarget{solving-the-steady-state-problem}{%
 \subsection{Solving the steady state
-problem}\label{solving-the-steady-state-problem}}
+problem}\label{solving-the-steady-state-problem}
 
@@
 
-\hypertarget{supporting-information}{%
-\section{Supporting information}\label{supporting-information}}
+\section{Supporting information}\label{supporting-information}
 
@@
 
-\hypertarget{acknowledgements}{%
-\section{Acknowledgements}\label{acknowledgements}}
+\section{Acknowledgements}\label{acknowledgements}
 
@@
 
-\hypertarget{author-information}{%
-\section{Author information}\label{author-information}}
-
-\hypertarget{author-contribution}{%
-\subsection{Author contribution}\label{author-contribution}}
+\section{Author information}\label{author-information}
+
+\subsection{Author contribution}\label{author-contribution}
 
@@
 statistical model, biological model, software architecture and user
-interface. \# References
-
-\hypertarget{conflict-of-interest}{%
-\subsection{Conflict of interest}\label{conflict-of-interest}}
+interface.
+
+\subsection{Conflict of interest}\label{conflict-of-interest}
 
 The authors declare no competing financial interest
+
+\section{References}\label{references}
 

### 2024-02-08 paper/supporting_information.qmd
Review responses

@@
 a &= \prod_{\text{s substrate}}\frac{C_{mic}^s}{k_{M}^{rs}} \nonumber \\ 
-\text{free enzyme ratio} &= \begin{cases} 
-  \prod_{\text{s sustrate}} (1 + \frac{C_{mic}^s}{k_{M}^{rs}})^{S_sr} 
-  + \sum_{\text{c inhibitor}}\frac{C_{mic}^c}{k_I^{rc}} & r\text{ irreversible} \\ 
-  -1 
-  + \prod_{\text{s sustrate}} (1 + \frac{C_{mic}^s}{k_{M}^{rs}})^{S_sr} 
-     + \sum_{\text{c inhibitor}}\frac{C_{mic}^c}{k_I^{rc}} 
-     + \prod_{\text{p product}} (1 + \frac{C_{mic}^p}{k_{M}^{rp}})^{S_pr}  & r\text{ reversible} 
+\text{free enzyme ratio} &= \begin{cases} \begin{aligned}
+  &\prod_{\text{s sustrate}} (1 + \frac{C_{mic}^s}{k_{M}^{rs}})^{S_sr} \\
+  + &\sum_{\text{c inhibitor}}\frac{C_{mic}^c}{k_I^{rc}}\end{aligned} & r\text{ irreversible} \\ 
+  \begin{aligned}
+  -1  &+ \prod_{\text{s sustrate}} (1 + \frac{C_{mic}^s}{k_{M}^{rs}})^{S_sr} \\
+      &+ \prod_{\text{p product}} (1 + \frac{C_{mic}^p}{k_{M}^{rp}})^{S_pr} \\ 
+      &+ \sum_{\text{c inhibitor}}\frac{C_{mic}^c}{k_I^{rc}} 
+  \end{aligned}
+    & r\text{ reversible} 
     \end{cases} \nonumber

### 2024-02-08 paper/supporting_information.tex
Review responses

@@
 \maketitle
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, interior hidden, frame hidden, sharp corners, enhanced, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
 
@@
 
-\hypertarget{mauds-input-format}{%
-\section{Maud's input format}\label{mauds-input-format}}
+\section{Maud's input format}\label{mauds-input-format}
 
@@
 
-\hypertarget{mauds-kinetic-model}{%
-\section{Maud's kinetic model}\label{mauds-kinetic-model}}
+\section{Maud's kinetic model}\label{mauds-kinetic-model}
 
-\hypertarget{parameters}{%
-\subsection{Parameters}\label{parameters}}
+\subsection{Parameters}\label{parameters}
 
@@
 
-\hypertarget{tbl-params}{}
 \begin{longtable}[]{@{}
@@
   >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2778}}@{}}
-\caption{\label{tbl-params}Table S1 -- Parameters of Maud's statistical
-model}\tabularnewline
+\caption{Table S1 -- Parameters of Maud's statistical
+model}\label{tbl-params}\tabularnewline
 \toprule\noalign{}
@@
 
-\hypertarget{rate-equations}{%
-\subsection{Rate equations}\label{rate-equations}}
+\subsection{Rate equations}\label{rate-equations}
 
@@
 
-\hypertarget{sec-methionine-case-study}{%
-\section{Methionine case study}\label{sec-methionine-case-study}}
+\section{Methionine case study}\label{sec-methionine-case-study}
 
-\hypertarget{dataset-generation}{%
-\subsection{Dataset generation}\label{dataset-generation}}
+\subsection{Dataset generation}\label{dataset-generation}
 
@@
 
-\hypertarget{sec-methionine-priors}{%
 \subsection{Prior distributions compared with true parameter
-values}\label{sec-methionine-priors}}
+values}\label{sec-methionine-priors}
 
@@
 
-\hypertarget{tbl-case-study-params}{}
 \begin{longtable}[]{@{}
@@
   >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2477}}@{}}
-\caption{\label{tbl-case-study-params}Table S2 -- Parameter
-specification, marginal prior distributions and true parameter values
-used in our case study.}\tabularnewline
+\caption{Table S2 -- Parameter specification, marginal prior
+distributions and true parameter values used in our case
+study.}\label{tbl-case-study-params}\tabularnewline
 \toprule\noalign{}
@@
 
-\hypertarget{computation}{%
-\subsection{Computation}\label{computation}}
+\subsection{Computation}\label{computation}
 
@@
 
-\hypertarget{laplace-approximation-case-study}{%
 \section{Laplace approximation case
-study}\label{laplace-approximation-case-study}}
+study}\label{laplace-approximation-case-study}
 
@@
 
-\hypertarget{multimodal-posterior-distributions}{%
 \section{Multimodal posterior
-distributions}\label{multimodal-posterior-distributions}}
+distributions}\label{multimodal-posterior-distributions}
 
@@
 
-\hypertarget{references}{%
-\section{References}\label{references}}
+\section{References}\label{references}
 

### 2024-02-12 paper/main.qmd
Review responses
Add callout blocks for new material

@@
 
+<!--:::{.callout-important title="Changes"}-->
 Maud uses Markov Chain Monte Carlo to perform posterior inference on a Bayesian
@@
 models.
+<!--:::-->
 
@@
 
+<!--:::{.callout-important title="Changes"}-->
+
 Teddy Groves and Nicholas Cowie: Writing code and documentation, designing
@@
 Lars Keld Nielsen: Supervision, writing manuscript, designing statistical model,
-biological model, software architecture and user interface.   # References
+biological model, software architecture and user interface.
 
@@
 
-The authors declare no competing financial interest
+The authors declare no competing financial interest.
+<!--:::-->
+
+# References
 

### 2024-02-12 paper/supporting_information.qmd
Add callout blocks for new material

@@
 
+<!--  :::{.callout-important title="Changes"} -->
 While we have not yet observed this in practice, we expect that Maud is capable
@@
 directly be used to improve Maud's efficiency.
+<!-- ::: -->
 

### 2024-02-12 paper/main.qmd
Final resubmission changes

@@
 ---
+affiliations:
+  - id: biosustain
+    name: "DTU"
+    department: "The Novo Nordisk Foundation Center for Biosustainability"
+    city: "Kongens Lyngby 2800"
+    country: "Denmark"
+  - id: uq
+    name: "The University of Queensland"
+    department: "Australian Institute for Bioengineering and Nanotechnology (AIBN)"
+    city: "St Lucia 4067"
+    country: "Australia"
+author: 
+  - name: "Teddy Groves"
+    orcid: "0000-0002-7109-3270"
+    email: "tedgro@biosustain.dtu.dk"
+    affiliations:
+      - ref: biosustain
+  - name: "Nicholas Luke Cowie"
+    affiliations:
+      - ref: biosustain
+    orcid: "0000-0003-3367-2105"
+  - name: "Lars Keld Nielsen"
+    affiliations:
+      - ref: uq
+      - ref: biosustain
+    orcid: "0000-0001-8191-3511"
 title: "Bayesian regression facilitates quantitative modelling of cell metabolism"
@@
 number-sections: true
-author: 
-  - name: "Teddy Groves"
-    orcid: 0000-0002-7109-3270
-    email: tedgro@biosustain.dtu.dk
-    affiliations: 
-      - id: biosustain
-        name: DTU
-        department: The Novo Nordisk Foundation Center for Biosustainability
-        city: Kongens Lyngby 2800
-        country: Denmark
-    attributes:
-      corresponding: true
-      equal-contributor: true
-    note: These authors contributed equally to this research.
-  - name: "Nicholas Luke Cowie"
-    orcid: 0000-0003-3367-2105
-    affiliations: 
-      - ref: biosustain
-    attributes:
-      equal-contributor: true
-    note: These authors contributed equally to this research.
-  - name: "Lars Keld Nielsen"
-    orcid: 0000-0001-8191-3511
-    affiliations: 
-      - id: uq
-        name: The University of Queensland
-        department: Australian Institute for Bioengineering and Nanotechnology (AIBN)
-        city: St Lucia 4067
-        country: Australia
-      - ref: biosustain
 bibliography: bibliography.bib
@@
 
-<!--:::{.callout-important title="Changes"}-->
+<!-- :::{.callout-important title="Changes"} -->
 Maud uses Markov Chain Monte Carlo to perform posterior inference on a Bayesian
@@
 models.
-<!--:::-->
+<!-- ::: -->
 

### 2024-02-12 paper/supporting_information.qmd
Final resubmission changes

@@
 ---
+affiliations:
+  - id: biosustain
+    name: "DTU"
+    department: "The Novo Nordisk Foundation Center for Biosustainability"
+    city: "Kongens Lyngby 2800"
+    country: "Denmark"
+  - id: uq
+    name: "The University of Queensland"
+    department: "Australian Institute for Bioengineering and Nanotechnology (AIBN)"
+    city: "St Lucia 4067"
+    country: "Australia"
+author: 
+  - name: "Teddy Groves"
+    orcid: "0000-0002-7109-3270"
+    email: "tedgro@biosustain.dtu.dk"
+    affiliations:
+      - ref: biosustain
+  - name: "Nicholas Luke Cowie"
+    affiliations:
+      - ref: biosustain
+    orcid: "0000-0003-3367-2105"
+  - name: "Lars Keld Nielsen"
+    affiliations:
+      - ref: uq
+      - ref: biosustain
+    orcid: "0000-0001-8191-3511"
 title: Supporting information - Bayesian regression facilitates quantitative modelling of cell metabolism
